---
title: "Hands-on Exercise 4"
subtitle: "Spatial Weights and Applications"
author: "An Loc"
date: "14 September 2026"
date-modified: "last-modified"
date-format: "D MMM YYYY"
format:
  html:
    toc: true
    toc-depth: 4
    toc-expand: true
    code-fold: false
    code-copy: true
    fig-width: 9
    fig-height: 7
execute:
  echo: true
  eval: true
  warning: false
  message: false
  freeze: true
---

# Overview

Every method in the rest of this course needs an answer to one question first: for a given area,
which other areas count as its neighbours, and how much does each one count? A spatial weights
matrix is that answer written down. Lesson 5 feeds it to Moran's I, Lesson 6 to a constrained
clustering algorithm, Lesson 7 to a local regression. Each of those inherits the weights matrix built
here.

This exercise builds spatial weights for the 88 counties of Hunan province, China, and uses them
to compute four spatially lagged versions of GDP per capita. It follows r4gdsa Chapter 8, the
chapter assigned for Lesson 4.

**Part 1** reproduces Chapter 8: contiguity weights by the Queen and Rook rules, distance weights
at a fixed band and at an adaptive k, inverse distance weights, row standardisation, and the four
lag variables. **Part 2** examines the chapter's method and adds the part of the lecture that
Chapter 8 does not cover.

::: {.callout-note appearance="simple"}

## Sources

The hands-on exercise is r4gdsa [Chapter 8: Spatial Weights and
Applications](https://r4gdsa.netlify.app/chap08.html).

The Lesson 4 slides close on three references. Each is used in a specific section, listed below.

| Reading | What it contributes | Used in |
|---|---|---|
| **de Bellefon, Loonis and Le Gleut (2018)**, chapter 2 of INSEE's *Handbook of Spatial Analysis*. The assigned pre-class reading, called the INSEE chapter below | The standardisation schemes beyond `W` and `B`; an explicit warning about the rule chapter 8 uses to set its distance band; why centroid distance loses information that contiguity keeps; the graph-based neighbour definitions | 4.2, 4.3, 4.4, 4.5 |
| **Bavaud (1998)**, "Models for Spatial Weights: A Systematic Look" | The general position that no weights matrix is universally correct, and that a choice has to suit the phenomenon | 5 |
| **Grubesic and Rosso (2014)**, "The Use of Spatially Lagged Explanatory Variables" | What a lagged variable is actually for, and a way to measure how much lagging changes the picture that can be compared against their published result | 4.7 |

The slide dates Bavaud to 2010. The paper itself states that the final version appeared in
*Geographical Analysis* 30, pp. 153-171, in **1998**, and the DOI on the slide resolves to the
1998 article. This page uses 1998.
:::

## Differences from the chapter

| | The chapter | This page | Why | Where |
|---|---|---|---|---|
| **Inverse distance weights** | `style = "B"`, described as row standardised | mismatch identified in 3.5, the two styles weighed in 4.4 | The chapter's text and its code do not describe the same matrix, and 4.4.1 argues its code may be the better of the two | 3.5, 4.4 |
| **Centroids** | `st_centroid` on WGS 84 degrees | same in Part 1, projected comparison in Part 2 | The shortcut leaves the fixed distance band unchanged and moves 17 of 88 counties in the k-nearest list | 4.2 |
| **Cut-off distance** | 62 km, from a figure computed two ways | same band, with the metric mismatch identified | `knearneigh()` finds the nearest neighbour in degrees and `nbdists()` measures it in kilometres, so the chapter's 61.79 km is not the largest first nearest neighbour distance | 4.3 |
| **Queen against Rook** | link totals reported, difference not examined | the eight counties identified | The difference is four mutual pairs touching at a single corner, which is the slide 11 quiz | 4.1 |
| **Geographically weighted statistics** | not covered | added, using `GWmodel::gwss()` | Four slides of the lecture cover it and the chapter stops before it | 4.6 |
| **Standardisation styles** | `W` and `B` only | all five compared | The INSEE chapter describes four schemes citing Tiefelsdorf, and `S` is the one designed to even out neighbour-count heterogeneity | 4.4 |
| **Neighbour definitions** | contiguity, distance band, adaptive k | four graph methods added | The INSEE chapter's argument for them is that they cannot leave a county isolated, which is the failure mode measured in 4.5 | 4.5 |
| **Use of the lagged variable** | four lags computed, never used | the cross-regressive use and a concordance measure | Grubesic and Rosso give both, and a published number to compare against | 4.7 |
| **`left_join()`** | join column inferred | `by = "County"` stated | The inferred join works here. Naming the column means a renamed field fails loudly instead of quietly | 2.3 |
| **Pipe** | `%>%` | <code>&#124;&gt;</code> | Carried over from Exercise 3 | throughout |
| **Figure heights** | default | set per chunk | Hunan is taller than it is wide, aspect 0.889, so the default canvas leaves white bands inside every map | throughout |
| **Map classification** | `qtm()` default, equal interval | quantile, 5 classes, on every map | GDPPC is right skewed, so equal interval puts 79 of the 88 counties in the lowest two classes and flattens the map | 3.1 |
| **`tmap` version** | v3 idiom | v4 idiom | This machine runs tmap 4.4.1 and the v3 calls emit deprecation messages | throughout |

::: {.callout-note collapse="true"}
## Mapping of Chapter 8 sections to this page

Every section of Chapter 8 is covered.

| Chapter 8 | This page |
|---|---|
| 8.1 Overview | 1 Overview |
| 8.2 The study area and data | 2.2 The data |
| 8.2.1 Getting started | 2.1 Packages |
| 8.3.1 Import shapefile | 2.2.1 County boundaries |
| 8.3.2 Import csv | 2.2.2 Development indicators |
| 8.3.3 Performing relational join | 2.3 Joining the two |
| 8.4 Visualising regional development indicator | 3.1 Mapping GDPPC |
| 8.5.1 Computing Queen contiguity | 3.2.1 Queen contiguity |
| 8.5.2 Creating Rook contiguity | 3.2.2 Rook contiguity |
| 8.5.3 Visualising contiguity weights | 3.2.3 Connectivity graphs |
| 8.6.1 Determine the cut-off distance | 3.3.1 The cut-off distance |
| 8.6.2 Computing fixed distance weight matrix | 3.3.2 Fixed distance band |
| 8.6.3 Computing adaptive distance weight matrix | 3.3.3 Adaptive k nearest neighbours |
| 8.7 Weights based on IDW | 3.4 Inverse distance weights |
| 8.8 Row-standardised weights matrix | 3.5 Row standardisation |
| 8.9.1 Spatial lag with row-standardized weights | 3.6.1 Spatial lag |
| 8.9.2 Spatial lag as a sum of neighboring values | 3.6.2 Lag as a sum |
| 8.9.3 Spatial window average | 3.6.3 Window average |
| 8.9.4 Spatial window sum | 3.6.4 Window sum |
| 8.10 References | 6 References |
:::

# Getting started

## Packages

`spdep` provides every weights function used here. `GWmodel` is needed only for the geographically
weighted statistics in Part 2, and `sp` because `gwss()` predates `sf` and still takes a
`Spatial*` object.

```{r}
pacman::p_load(sf, spdep, tmap, tidyverse, knitr, GWmodel, sp)
```

`tmap` is 4.4.1 on this machine while the chapter was written against v3, so this page uses the
v4 spelling of every mapping call.

## The data

Chapter 8 names both files and links neither.

| Layer | Format | Features | Native CRS | Contents |
|---|---|---|---|---|
| `Hunan` | ESRI shapefile | 88 polygons | WGS 84, EPSG:4326 | County boundaries of Hunan province |
| `Hunan_2012` | CSV | 88 rows, 29 fields | not spatial | 2012 development indicators by county |

The shapefile has no projected CRS. It is in geographic degrees, and section 4.2 measures the
effect of that on the neighbour lists.

### County boundaries

```{r}
hunan <- st_read(dsn = "data/geospatial", layer = "Hunan", quiet = TRUE)
hunan
```

88 features, 7 fields, POLYGON geometry, WGS 84.

### Development indicators

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv", show_col_types = FALSE)
dim(hunan2012)
```

29 fields covering wages, deposits, investment, government revenue and expenditure, GDP,
population and employment. This exercise uses one of them, `GDPPC`, GDP per capita in yuan.

## Joining the two

The join key is `County`, present in both. The chapter lets `left_join()` infer it; naming it
means that a renamed field produces an error rather than a silent cross join.

```{r}
hunan <- left_join(hunan, hunan2012, by = "County") |>
  select(1:4, 7, 15)
names(hunan)
```

`select(1:4, 7, 15)` keeps the four name fields, `County` and `GDPPC`. It selects by position, so
it depends on the column order the join happens to produce. It is correct here, verified against
the printed names above.

# Part 1: Following the chapter

## Mapping GDPPC

```{r}
#| fig-height: 5.06
basemap <- tm_shape(hunan) +
  tm_polygons() +
  tm_text("NAME_3", size = 0.35)

gdppc <- tm_shape(hunan) +
  tm_polygons("GDPPC",
              fill.scale = tm_scale_intervals(style = "quantile", n = 5),
              fill.legend = tm_legend(title = "GDPPC (yuan)", position = c("left", "bottom")))

tmap_arrange(basemap, gdppc, ncol = 2)
```

The chapter uses `qtm()`, whose default equal-interval classes put most counties in the lowest
two, because GDPPC is right skewed. Quantile classes are used here and on every later map so the
spatial pattern is visible.

The high values sit in the north east, around Changsha, and along the eastern edge. The low
values run through the west.

```{r}
summary(hunan$GDPPC)
```

## Contiguity weights

`poly2nb()` builds a neighbours list from shared boundaries. Its `queen` argument chooses the
rule: `TRUE` counts any shared point, `FALSE` requires a shared edge.

### Queen contiguity

```{r}
wm_q <- poly2nb(hunan, queen = TRUE)
summary(wm_q)
```

88 regions, 448 links, 5.09 neighbours on average. The summary also reports one county with 11
neighbours, the most, and two with only one.

The neighbours of county 1 and their GDPPC:

```{r}
wm_q[[1]]
hunan$County[wm_q[[1]]]
hunan$GDPPC[wm_q[[1]]]
```

### Rook contiguity

```{r}
wm_r <- poly2nb(hunan, queen = FALSE)
summary(wm_r)
```

440 links against the Queen rule's 448. Eight fewer, which is four pairs of counties counted once
from each side. Section 4.1 identifies them.

### Connectivity graphs

Plotting a neighbours list needs a point per polygon. The chapter takes the centroid of each
geometry with `map_dbl()`.

```{r}
longitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[1]])
latitude  <- map_dbl(hunan$geometry, ~st_centroid(.x)[[2]])
coords <- cbind(longitude, latitude)
head(coords)
```

These are degrees, and `st_centroid` treats them as if they were planar. Section 4.2 measures how
far the centroids move as a result, and which neighbour lists change.

```{r}
#| fig-height: 5.06
par(mfrow = c(1, 2))
plot(hunan$geometry, border = "lightgrey", main = "Queen contiguity")
plot(wm_q, coords, pch = 19, cex = 0.6, add = TRUE, col = "red")
plot(hunan$geometry, border = "lightgrey", main = "Rook contiguity")
plot(wm_r, coords, pch = 19, cex = 0.6, add = TRUE, col = "red")
```

The two graphs are hard to tell apart at this scale: the difference is eight links out of 448,
identified in section 4.1.

## Distance weights

### The cut-off distance

A fixed distance band needs an upper bound. If it is set too low, some counties end up with no
neighbours at all. `nb2listw()` then stops with `Empty neighbour sets found` unless
`zero.policy = TRUE` is set, and with that set the lagged value at an isolated county comes back
as 0 rather than as missing. The chapter sets the bound from the largest first nearest neighbour
distance, so every county keeps at least one.

```{r}
k1 <- knn2nb(knearneigh(coords))
k1dists <- unlist(nbdists(k1, coords, longlat = TRUE))
summary(k1dists)
```

The chapter reads 61.79 km as the largest first nearest neighbour distance and sets the band just
above it, at 62 km, so that every county keeps at least one neighbour. Section 4.3 checks both the
number and the claim.

`longlat = TRUE` tells `nbdists()` the coordinates are degrees and to return great circle
distances in kilometres. Without it the numbers would be degrees, which are not a distance.

### Fixed distance band

```{r}
wm_d62 <- dnearneigh(coords, 0, 62, longlat = TRUE)
wm_d62
```

324 links, 3.68 on average, against the Queen rule's 448 and 5.09.

> **Quiz from the chapter:** what does "Average number of links: 3.681818" mean?

Across the 88 counties, each has 3.68 neighbours within 62 km on average: 324 links divided by
88 counties. It is an average over a range from 1 to 6, not a typical value. Six counties are at
the minimum of 1. `card()` below returns each county's neighbour count.

```{r}
table(card(wm_d62))
```

```{r}
#| fig-height: 5.06
par(mfrow = c(1, 2))
plot(hunan$geometry, border = "lightgrey", main = "1st nearest neighbour")
plot(k1, coords, add = TRUE, col = "red", length = 0.08)
plot(hunan$geometry, border = "lightgrey", main = "Within 62 km")
plot(wm_d62, coords, add = TRUE, pch = 19, cex = 0.6)
```

The right panel shows the characteristic problem with a fixed band: dense areas accumulate
neighbours, sparse areas keep one or two.

### Adaptive k nearest neighbours

Fixing the count instead of the distance gives every county the same number of neighbours,
letting the radius vary.

```{r}
knn6 <- knn2nb(knearneigh(coords, k = 6))
knn6
```

528 links, exactly 6 each. The output reports it as non-symmetric: if B is among A's six
nearest, A need not be among B's.

```{r}
is.symmetric.nb(knn6)
```

```{r}
#| fig-height: 10.13
plot(hunan$geometry, border = "lightgrey", main = "6 nearest neighbours")
plot(knn6, coords, pch = 19, cex = 0.6, add = TRUE, col = "red")
```

## Inverse distance weights

Rather than treating every neighbour equally, weight each by the reciprocal of its distance, so
that near neighbours count more.

```{r}
dist <- nbdists(wm_q, coords, longlat = TRUE)
ids <- lapply(dist, function(x) 1 / (x))
ids[[1]]
```

These are 1/km, between about 0.008 and 0.040 across the whole province.

```{r}
summary(unlist(ids))
```

## Row standardisation

`nb2listw()` turns a neighbours list into a weights list. `style = "W"` divides each row by its
number of neighbours, so the weights in every row sum to 1.

```{r}
rswm_q <- nb2listw(wm_q, style = "W", zero.policy = TRUE)
rswm_q
```

County 10 has eight neighbours, so each gets 0.125:

```{r}
rswm_q$weights[10]
```

The chapter then writes that it derives "a row standardised distance weight matrix" with the code
below.

```{r}
rswm_ids <- nb2listw(wm_q, glist = ids, style = "B", zero.policy = TRUE)
rswm_ids$weights[1]
```

Those are the raw 1/distance values, unchanged from `ids[[1]]` above. `style = "B"` is the binary
style, which uses `glist` as given. Nothing was standardised. Row standardised means the row sums
to 1, and `style = "W"` does that:

```{r}
rswm_ids_W <- nb2listw(wm_q, glist = ids, style = "W", zero.policy = TRUE)
c(B = sum(rswm_ids$weights[[1]]), W = sum(rswm_ids_W$weights[[1]]))
```

Row standardised rows sum to 1, and these do not, so the chapter's sentence and its code do not
describe the same matrix. Section 4.4 measures the gap and takes up which of the two an inverse
distance weighting should use.

## The four lag variables

A spatially lagged variable summarises each county's neighbourhood into one number. Four versions
follow from two binary choices: whether the weights are row standardised, and whether the county
counts as its own neighbour.

|  | Excludes self | Includes self |
|---|---|---|
| **Row standardised** | spatial lag | window average |
| **Binary** | lag as a sum | window sum |

### Spatial lag

```{r}
hunan$lag_GDPPC <- lag.listw(rswm_q, hunan$GDPPC)
head(hunan[, c("County", "GDPPC", "lag_GDPPC")])
```

Anxiang's neighbours have GDPPC 20981, 34592, 24473, 21311 and 22879. Their mean is 24847.2,
which is the lagged value. With row standardised weights the spatial lag is the neighbourhood
mean, and it is on the same scale as the variable itself, so the two can be mapped against each
other.

```{r}
#| fig-height: 5.06
tmap_arrange(
  tm_shape(hunan) + tm_polygons("GDPPC",
    fill.scale = tm_scale_intervals(style = "quantile", n = 5),
    fill.legend = tm_legend(title = "GDPPC", position = c("left", "bottom"))),
  tm_shape(hunan) + tm_polygons("lag_GDPPC",
    fill.scale = tm_scale_intervals(style = "quantile", n = 5),
    fill.legend = tm_legend(title = "lag GDPPC", position = c("left", "bottom"))),
  ncol = 2)
```

The lagged map is smoother. Isolated high counties are pulled down by poorer neighbours and the
broad north-eastern cluster survives. The lag is useful for that reason: a value that survives
averaging with its neighbours belongs to a regional pattern rather than to one county.

### Lag as a sum

Binary weights, no standardisation, so the result is a total rather than a mean.

```{r}
b_weights <- lapply(wm_q, function(x) 0 * x + 1)
b_weights2 <- nb2listw(wm_q, glist = b_weights, style = "B")
hunan$lag_sum_GDPPC <- lag.listw(b_weights2, hunan$GDPPC)
head(hunan[, c("County", "GDPPC", "lag_GDPPC", "lag_sum_GDPPC")])
```

Anxiang's 124236 is the sum of those same five neighbours. This version confounds two things: how
rich the neighbours are, and how many of them there are.

```{r}
#| fig-height: 5.06
tmap_arrange(
  tm_shape(hunan) + tm_polygons("GDPPC",
    fill.scale = tm_scale_intervals(style = "quantile", n = 5),
    fill.legend = tm_legend(title = "GDPPC", position = c("left", "bottom"))),
  tm_shape(hunan) + tm_polygons("lag_sum_GDPPC",
    fill.scale = tm_scale_intervals(style = "quantile", n = 5),
    fill.legend = tm_legend(title = "lag sum", position = c("left", "bottom"))),
  ncol = 2)
```

The right map is partly a map of connectivity. Taoyuan has the most neighbours of any county at
11, so its total is high because it borders many counties rather than because they are
prosperous. Reading the lag sum means reading how rich the neighbours are and how many of them
there are at the same time.

### Window average

`include.self()` adds each county to its own neighbours list, then row standardisation over the
enlarged list gives the mean of the neighbourhood including the county itself.

```{r}
wm_qs <- include.self(wm_q)
wm_qs
```

536 links against 448, which is 448 plus the 88 diagonal entries.

```{r}
wm_qs_lw <- nb2listw(wm_qs)
hunan$lag_window_avg_GDPPC <- lag.listw(wm_qs_lw, hunan$GDPPC)
```

The chapter's text at this point says "Again, we use `nb2listw()` and `glist()` to explicitly
assign weight values." No `glist` is passed. The call relies on the default `style = "W"`, and
that is the right setting for a window average. Only the sentence is wrong.

### Window sum

Binary weights over the same self-inclusive list.

```{r}
b_weights_s <- lapply(wm_qs, function(x) 0 * x + 1)
b_weights_s2 <- nb2listw(wm_qs, glist = b_weights_s, style = "B")
hunan$w_sum_GDPPC <- lag.listw(b_weights_s2, hunan$GDPPC)
```

```{r}
#| fig-height: 5.06
tmap_arrange(
  tm_shape(hunan) + tm_polygons("lag_sum_GDPPC",
    fill.scale = tm_scale_intervals(style = "quantile", n = 5),
    fill.legend = tm_legend(title = "lag sum", position = c("left", "bottom"))),
  tm_shape(hunan) + tm_polygons("w_sum_GDPPC",
    fill.scale = tm_scale_intervals(style = "quantile", n = 5),
    fill.legend = tm_legend(title = "window sum", position = c("left", "bottom"))),
  ncol = 2)
```

These two are hard to tell apart, and the reason is the spread between counties rather than the
size of any one contribution. Totals differ enormously across Hunan because some counties border
one neighbour and some border eleven, so adding one more value does not reorder them. A mean has
no such spread, and including the county pulls the average toward that county's own value.
Section 3.6.5 maps the two side by side.

### The four compared

```{r}
hunan |>
  st_drop_geometry() |>
  select(County, GDPPC, lag_GDPPC, lag_sum_GDPPC, lag_window_avg_GDPPC, w_sum_GDPPC) |>
  head(10) |>
  kable(digits = 0, format.args = list(big.mark = ","))
```

Two identities hold by construction:

```{r}
all.equal(hunan$w_sum_GDPPC, hunan$lag_sum_GDPPC + hunan$GDPPC)
all.equal(hunan$lag_window_avg_GDPPC, hunan$w_sum_GDPPC / (card(wm_q) + 1))
```

```{r}
#| fig-height: 5.06
tmap_arrange(
  tm_shape(hunan) + tm_polygons("lag_GDPPC",
    fill.scale = tm_scale_intervals(style = "quantile", n = 5),
    fill.legend = tm_legend(title = "lag", position = c("left", "bottom"))),
  tm_shape(hunan) + tm_polygons("lag_window_avg_GDPPC",
    fill.scale = tm_scale_intervals(style = "quantile", n = 5),
    fill.legend = tm_legend(title = "window avg", position = c("left", "bottom"))),
  ncol = 2)
```

The window average differs from the lag most where a county's own value is far from its
neighbours'. The table above shows it: where the two columns diverge, the county itself is
pulling against the neighbourhood around it.

# Part 2: Review and extension

Part 1 reproduced the chapter. This part examines its method, and adds what the lecture and the
readings cover that Chapter 8 does not.

The INSEE chapter divides the problem in two: which areas count as neighbours, its section 2.1,
and how much each one counts, its 2.2. Its 2.2.2 reports that opinions in the field diverge on how
much those choices affect results.

Chapter 8 makes a decision at four points in that scheme, and the first four sections below
measure what each one changes. Which areas count as neighbours: the contiguity rule (4.1), the
centroids the distances are measured between (4.2), and the distance band (4.3). How much
each counts: the standardisation scheme (4.4). The sections after that are extensions beyond
the chapter.

## Queen and Rook contiguity

Section 3.2 built both contiguity matrices and printed the totals: 448 links under Queen, 440
under Rook. Eight links is a 1.8% difference in the structure. Which eight, and does dropping
them change any county's lagged value?

```{r}
differs <- which(card(wm_q) != card(wm_r))
tibble(
  county = hunan$County[differs],
  queen = card(wm_q)[differs],
  rook = card(wm_r)[differs],
  `corner-only contact` = sapply(differs, function(i)
    paste(hunan$County[setdiff(wm_q[[i]], wm_r[[i]])], collapse = ", "))
) |> kable()
```

Four pairs, each appearing twice because contiguity is symmetric: Anxiang with Hanshou, Anren
with Zixing, Yongxing with Yanling, Nan with Taoyuan. In each case the two counties meet at a
single point rather than along a boundary. This is the slide 11 quiz in the lecture deck, where
the Queen row for area 1202 has seven entries and the Rook answer has five.

Whether it matters for the analysis:

```{r}
lag_q <- lag.listw(nb2listw(wm_q, style = "W"), hunan$GDPPC)
lag_r <- lag.listw(nb2listw(wm_r, style = "W"), hunan$GDPPC)
moved <- which(abs(lag_q - lag_r) > 1e-9)
cat("counties whose lagged GDPPC changes:", length(moved), "of 88\n")
cat("largest change:", round(max(abs(lag_q - lag_r))), "yuan, at",
    hunan$County[which.max(abs(lag_q - lag_r))], "\n")
cat("as a share of that county's Rook lag:",
    round(100 * max(abs(lag_q - lag_r) / lag_r), 1), "%\n")
```

Eight counties move, and the largest move is 4811 yuan at Anren, 17.7% of its value. Province
wide the choice is close to irrelevant; for those eight counties it is not. The general point is
that a difference of eight links out of 448 is not a difference of 1.8% in the result, because
the links are not spread evenly.

Queen is the right default here. A corner contact between two Chinese counties is a real
administrative adjacency, and the Rook rule discards it on what amounts to an artefact of where
the boundary vertices were digitised.

## Centroids: unprojected versus projected

Every distance in section 3.3 was measured between county centroids, because a distance band
needs a single point to stand for each polygon. Two questions follow: whether a centroid is the
right thing to measure between at all, and whether these ones were computed correctly.

The INSEE chapter takes up the first question, and its answer is no. Its
Example 2.1 sets out three zones where R2 and R3 are separated in space and both touch R1, and
shows that their centroids are nonetheless equidistant from each other:

> Summarising the proximity between zones by the distance between the centroids results in a
> partial loss of the richness of the spatial relationships.

The INSEE chapter is arguing for contiguity over centroid distance on areal data, and that applies to
§3.3 and §3.4 of this page. It adds that contiguity is the usual choice for
demographic and social data, "in which it may be more important to be on either side of an
administrative boundary than to be located at a certain distance from one another." GDP per
capita by county is that kind of data.

Chapter 8 uses centroids regardless, so the rest of this section asks whether it computes them
accurately.

The chapter runs `st_centroid` on WGS 84 longitude and latitude. `st_centroid` computes a planar
centroid, so on geographic coordinates it treats one degree of longitude as equal to one degree
of latitude. At Hunan's latitude of about 27 degrees, a degree of longitude is roughly 89% of a
degree of latitude, so the centroids are pulled slightly. The comparison below reprojects to
EPSG:4547, the CGCS2000 Gauss-Kruger zone covering Hunan, whose units are metres.

```{r}
hunan_proj <- st_transform(hunan, 4547)
cent_proj <- st_centroid(st_geometry(hunan_proj))
cent_chap <- st_as_sf(data.frame(x = coords[, 1], y = coords[, 2]),
                      coords = c("x", "y"), crs = 4326)
shift <- as.numeric(st_distance(cent_chap, st_transform(cent_proj, 4326), by_element = TRUE))
summary(shift)
```

The centroids move by 18 m at the median and 62 m at most. Against a 62 km distance band that is
nothing, and the fixed band is in fact unaffected:

```{r}
coords_p <- st_coordinates(cent_proj)
band_chap <- dnearneigh(coords,   0, 62,    longlat = TRUE)
band_proj <- dnearneigh(coords_p, 0, 62000)
identical(lapply(band_chap, sort), lapply(band_proj, sort))
```

The k nearest neighbour list is a different matter. It has no distance threshold to absorb a
small error: it takes the nearest six whatever the distances are, so a shift of tens of metres
can reorder the sixth and seventh candidates.

```{r}
knn6_chap <- knn2nb(knearneigh(coords,   k = 6))
knn6_proj <- knn2nb(knearneigh(coords_p, k = 6))
changed <- sum(sapply(1:88, function(i)
  !identical(sort(knn6_chap[[i]]), sort(knn6_proj[[i]]))))
cat("counties whose 6 nearest neighbours change:", changed, "of 88\n")
```

17 of 88 counties get a different neighbour set. Following it through to the lagged variable:

```{r}
lag_knn_chap <- lag.listw(nb2listw(knn6_chap, style = "W"), hunan$GDPPC)
lag_knn_proj <- lag.listw(nb2listw(knn6_proj, style = "W"), hunan$GDPPC)
d <- lag_knn_chap - lag_knn_proj
cat("largest change:", round(max(abs(d))), "yuan at", hunan$County[which.max(abs(d))], "\n")
cat("as a share:", round(100 * max(abs(d / lag_knn_proj)), 1), "%\n")
cat("Spearman correlation of the two lag vectors:",
    round(cor(lag_knn_chap, lag_knn_proj, method = "spearman"), 4), "\n")
```

Longhui's neighbourhood mean moves by 8691 yuan, 41.4% of its value, because a swapped sixth
neighbour was far richer than the one it replaced. The rank correlation across all 88 stays at
0.984, so the province-wide picture is intact and one county's number is badly wrong.

So the shortcut holds for the fixed band, which is the only place the chapter uses these
coordinates, and fails for the k nearest list. Projecting first takes one line, and then both
results are the same whichever coordinates are used:

```{r}
#| eval: false
hunan_proj <- st_transform(hunan, 4547)   # CGCS2000 Gauss-Kruger zone, metres
coords <- st_coordinates(st_centroid(st_geometry(hunan_proj)))
```

## The 62 km cut-off distance

The 62 km band in section 3.3.2 came from one number: the largest distance between any county
and its nearest neighbour, 61.79 km. Setting the band there guarantees nobody is left isolated.
It is a different question whether counties 62 km apart influence each other, which is the claim
a distance band makes. The table below varies the band and counts the counties left with no
neighbour at all.

```{r}
bands <- c(45, 50, 55, 58, 59, 60, 62, 65, 70, 75, 80)
map_dfr(bands, function(d) {
  w <- dnearneigh(coords, 0, d, longlat = TRUE)
  tibble(`band (km)` = d, links = sum(card(w)),
         `avg links` = mean(card(w)), isolated = sum(card(w) == 0))
}) |> kable(digits = 2)
```

At 58 km and below some counties have no neighbour at all. From 59 km up every county has at
least one. So the threshold is 59 km, and not the 61.79 km the chapter derives. The gap between
the two has a cause.

`knearneigh(coords)` finds each county's nearest neighbour by planar distance on longitude and
latitude. `nbdists(coords, longlat = TRUE)` then measures those pairs as great circle kilometres.
The two steps use different metrics, so 61.79 km is the distance to the nearest county *in
degrees*, which is not always the nearest county in kilometres.

```{r}
k1_deg <- knn2nb(knearneigh(coords))
k1_km  <- knn2nb(knearneigh(coords, longlat = TRUE))

tibble(
  `nearest neighbour found in` = c("degrees, as the chapter does", "great circle kilometres"),
  `largest such distance (km)` = c(max(unlist(nbdists(k1_deg, coords, longlat = TRUE))),
                                   max(unlist(nbdists(k1_km,  coords, longlat = TRUE))))
) |> kable(digits = 2)

sum(sapply(seq_along(k1_deg), function(i) k1_deg[[i]] != k1_km[[i]]))
```

Measured consistently in kilometres the largest first nearest neighbour distance is 58.39 km, and
6 of the 88 counties have a different nearest neighbour under the two metrics. 62 km is above
58.39 km as well, so every county still keeps at least one neighbour and the chapter's band
works.

This is the same fault as §4.2, planar arithmetic on angular coordinates, and here it affects the
number the chapter uses to justify its choice.

The INSEE chapter describes the same procedure and warns against using it:

> This "minimum distance" method is not adapted to irregularly spaced data, as the minimum
> distance required for a relatively isolated point having at least one neighbour is much higher
> than the distance to the closest neighbour of a point located in a dense zone. There will
> therefore be significant disparities in the number of neighbours.

The disparity is visible in §3.3.2 above: at 62 km the neighbour counts run from 1 to 6, and six
counties have only one neighbour. The band is set by the most isolated county in the province and then
applied everywhere, so guaranteeing Xinhuang a neighbour gives every dense county a
neighbourhood wider than its own spatial scale.

This is the argument for the adaptive k of §3.3.3, and for the graph methods in §4.5.

The lower end of the band is fixed by the isolated counties. The upper end is not fixed by
anything. Going from 62 km to 80 km more than doubles the link count, from 324 to 570, and there
is nothing in the GDPPC data to show which band is closer to the distance over which counties
affect each other.

## The C, U and S standardisation schemes

Sections 3.5 and 3.6 used two styles without comparing them: `W` for the lagged mean and `B` for
the lagged sum. The style decides whether every county's weights add up to the same total, or
whether counties with more neighbours count for more.

`nb2listw()` offers five. The INSEE chapter describes four, citing Tiefelsdorf (1998), and the
fifth is the unstandardised binary case.

```{r}
styles <- c("B", "W", "C", "U", "S")
map_dfr(styles, function(st) {
  lw <- nb2listw(wm_q, style = st)
  rs <- sapply(lw$weights, sum)
  tibble(style = st,
         `min row sum` = min(rs), `max row sum` = max(rs),
         `CV of row sums` = sd(rs) / mean(rs),
         `all weights total` = sum(unlist(lw$weights)))
}) |> kable(digits = 4)
```

The INSEE chapter's account of each:

| Style | What it does | Consequence |
|---|---|---|
| `B` | binary, no standardisation | A county's total weight is its neighbour count, so Taoyuan with 11 neighbours exerts 11 times Xinhuang's pull |
| `W` | row standardisation | Every row sums to 1. Gives more weight to border counties, which have few neighbours |
| `C` | global standardisation, all weights scaled so they total $n$ | Central counties with many neighbours take more external influence than border ones |
| `U` | uniform, all weights total 1 | Same shape as `C`, different scale |
| `S` | variance stabilisation, Tiefelsdorf | Introduced to reduce the heterogeneity that differences in neighbour count create |

The sum of all elements is exactly $n = 88$ for `W`, `C` and `S`. The INSEE chapter gives this as the
reason autocorrelation statistics computed from different matrices can be compared. `U` totals 1 by
definition and `B` totals 448, the link count.

The row sums of `S` vary less than `B`, `C` and `U` and more than `W`: a coefficient of variation
of 0.184 against 0.349 and 0.000. That is the reduction in heterogeneity `S` was designed for.

`B`, `C` and `U` share a coefficient of variation of exactly 0.349 because all three are the same
binary matrix under a single scalar multiplier. They differ in scale and not in shape, so any
statistic invariant to scale cannot tell them apart.

Moran's I in Lesson 5 is computed from the weights list, so `W` and `S` will not give the same
answer there.

### Which style inverse distance weights should use

The obvious fix is to change `style = "B"` to `style = "W"`, matching the chapter's
sentence. The INSEE chapter argues against doing that.

Inside Chapter 8 the question never arises, because `rswm_ids` is built, printed, and then never
used: every lag in its section 8.9 runs on plain Queen contiguity or binary weights, so the
distance decay computed in 8.7 is never applied to anything.

de Bellefon, Loonis and Le Gleut set out what row standardisation means and then add a caveat
directly about this case:

> Each weight $w_{ij}$ can be interpreted as the fraction of spatial influence on observation $i$
> ascribable to $j$. In contrast, such standardisation implies a certain degree of competition
> between neighbours: the fewer neighbours a zone has, the greater their weight. Moreover, when
> weights are inversely proportional to the distance between the zones, row standardisation makes
> them difficult to interpret.

Row standardisation divides by a row sum that is itself a function of distance, so the result
mixes two things: how far a neighbour is, and how far all the other neighbours happen to be. Every
county's weights are forced to total 1 however far away its neighbours are, so the decay is
preserved between one county's neighbours and erased between counties.

The two produce lagged variables on different scales:

```{r}
lag_B <- lag.listw(rswm_ids,   hunan$GDPPC)
lag_W <- lag.listw(rswm_ids_W, hunan$GDPPC)

tibble(quantity = c("GDPPC itself", "lag, style B", "lag, style W"),
       min = c(min(hunan$GDPPC), min(lag_B), min(lag_W)),
       max = c(max(hunan$GDPPC), max(lag_B), max(lag_W))) |>
  kable(digits = 0, format.args = list(big.mark = ","))

cor(lag_B, lag_W)
```

The `W` version stays on the scale of GDPPC. The `B` version is a sum of yuan-per-kilometre
terms, which cannot be read as a neighbourhood income or plotted against one, and the two still
correlate closely enough that a map of either looks reasonable.

So the chapter's code may be the better choice. Which style is right depends on the use: a
neighbourhood **mean** comparable to GDPPC, as in §3.6.1, requires `W`, while preserving the
distance decay at its true magnitude requires `B`. The text names one and the code does the
other, so a reader cannot tell which was intended. That is the error, whichever style is correct.

This page uses `W` wherever a lagged value is compared against GDPPC.

## Graph-based neighbour definitions

Section 4.3 showed that a fixed band leaves counties with no neighbour below 60 km, which is why
the chapter sets the band from the most isolated county in the province. The INSEE chapter covers
a family of neighbour definitions where that cannot happen, built from the geometry of the point
set rather than from a distance threshold: "they have the advantage that it leaves no unit
without neighbours."

All four are in `spdep`, and each applies a different rule. Delaunay triangulation connects
points into triangles that maximise the minimum angle. The sphere of influence graph joins two
counties when the circles centred on each, drawn out to its own nearest neighbour, overlap.
Gabriel's graph joins two counties when no third county falls inside the circle having them as
its diameter. The relative neighbour graph joins two counties only when no third county is closer
to both of them than they are to each other. Each is a subgraph of the one before it.

```{r}
graphs <- list(
  "Delaunay"            = tri2nb(coords),
  "Sphere of influence" = graph2nb(soi.graph(tri2nb(coords), coords)),
  "Gabriel"             = graph2nb(gabrielneigh(coords), sym = TRUE),
  "Relative neighbour"  = graph2nb(relativeneigh(coords), sym = TRUE),
  "Queen contiguity"    = wm_q,
  "Fixed band, 62 km"   = wm_d62,
  "Adaptive, k = 6"     = knn6)

map_dfr(names(graphs), function(nm) {
  n <- graphs[[nm]]
  tibble(method = nm, links = sum(card(n)), `avg links` = mean(card(n)),
         min = min(card(n)), max = max(card(n)), isolated = sum(card(n) == 0))
}) |> kable(digits = 2)
```

None of the four graph methods isolates a county, as the INSEE chapter claims. Neither does
Queen contiguity on this data, so the advantage is over the distance band rather than over
contiguity, and only below 60 km does that distinction start to matter.

The graphs change the spread. Queen contiguity runs from 1 neighbour to 11; Delaunay
runs from 3 to 9 and the sphere of influence from 2 to 9. Taoyuan's 11 neighbours in §3.2.1 come
from it being a large county surrounded by small ones, which the INSEE chapter names as a known
property of contiguity: "a very large zone surrounded by smaller zones will have a far greater
number of neighbours than its neighbouring zones." The graph methods are built on the centroids
and so do not inherit that.

The relative neighbour graph is much the sparsest at 2.70 links on average, which matches
Toussaint's design goal quoted in the INSEE chapter, that it "adapts better to data by requiring the
fewest links."

```{r}
#| fig-height: 5.06
par(mfrow = c(1, 2))
plot(hunan$geometry, border = "lightgrey", main = "Delaunay")
plot(graphs$Delaunay, coords, pch = 19, cex = 0.5, add = TRUE, col = "red")
plot(hunan$geometry, border = "lightgrey", main = "Relative neighbour")
plot(graphs$`Relative neighbour`, coords, pch = 19, cex = 0.5, add = TRUE, col = "red")
```

The left panel shows the artefact these methods are known for. Delaunay spans the convex hull of
the point set, so counties along the provincial border are joined to each other by long edges
that run outside Hunan and across counties they do not touch. The INSEE chapter does not raise
this, and it is the reason the sparser subgraphs exist: the relative neighbour graph on the right
keeps 238 links against Delaunay's 498, and the long border edges are the first to go.

These are Euclidean-only in R, which the INSEE chapter flags as their limitation: great circle distance
is not available, so on a province the size of Hunan they inherit the same planar assumption
§4.2 measured.

## Geographically weighted summary statistics

Section 3.6.3 computed a mean for each county's neighbourhood instead of one mean for the whole
province. Geographically weighted summary statistics work the same way, but the
weighting and the neighbourhood are both adjustable. The window average uses uniform weights over
the contiguity neighbours; `gwss()` lets the weights decay with distance and lets the
neighbourhood be a fixed count or a fixed radius.

The last four content slides of the Lesson 4 deck cover these and their robust versions, and
Chapter 8 stops before them. `GWmodel::gwss()` computes them, and the deck's slide 21 table is
exactly its output.

Bandwidth is set to 6 adaptive, which is the queen contiguity average of 5.09 neighbours plus the
county itself, so these statistics cover about the same neighbourhood as Part 1.

```{r}
hunan_sp <- as_Spatial(st_transform(hunan, 4547))
gw <- gwss(hunan_sp, vars = "GDPPC", bw = 6, kernel = "bisquare",
           adaptive = TRUE, longlat = FALSE, quantile = TRUE)
names(as.data.frame(gw$SDF))
```

`LM` is the local mean, `LSD` the local standard deviation, `LVar` the variance, `LSKe` the skew
and `LCV` the coefficient of variation; `Median`, `IQR` and `QI` are the robust three.

`quantile = TRUE` is required to get those last three. Without it `gwss()` returns only the first
five and silently omits the median, IQR and quantile imbalance. There is no warning; the columns are simply absent, and code that goes on to read
`GDPPC_Median` fails later with a message about the wrong thing.

```{r}
gwr <- as.data.frame(gw$SDF)
hunan$gw_mean   <- gwr$GDPPC_LM
hunan$gw_median <- gwr$GDPPC_Median
hunan$gw_iqr    <- gwr$GDPPC_IQR

tibble(
  statistic = c("GW mean", "GW median"),
  min = c(min(hunan$gw_mean), min(hunan$gw_median)),
  max = c(max(hunan$gw_mean), max(hunan$gw_median))
) |> kable(digits = 0, format.args = list(big.mark = ","))
```

The single global mean of 24405 yuan stands in for local means running from 10312 to 73533. The
province has no representative county. The deck makes that argument on slide 18 for computing
statistics locally.

```{r}
#| fig-height: 3.38
tmap_arrange(
  tm_shape(hunan) + tm_polygons("GDPPC",
    fill.scale = tm_scale_intervals(style = "quantile", n = 5),
    fill.legend = tm_legend(title = "GDPPC", position = c("left", "bottom"))),
  tm_shape(hunan) + tm_polygons("gw_mean",
    fill.scale = tm_scale_intervals(style = "quantile", n = 5),
    fill.legend = tm_legend(title = "GW mean", position = c("left", "bottom"))),
  tm_shape(hunan) + tm_polygons("gw_median",
    fill.scale = tm_scale_intervals(style = "quantile", n = 5),
    fill.legend = tm_legend(title = "GW median", position = c("left", "bottom"))),
  ncol = 3)
```

The mean and the median diverge where one neighbour is an outlier. That is the case the deck's
robust statistics slides address.

```{r}
gap <- abs(hunan$gw_mean - hunan$gw_median)
cat("largest gap between GW mean and GW median:", round(max(gap)), "yuan at",
    hunan$County[which.max(gap)], "\n")
cat("  GW mean:  ", round(hunan$gw_mean[which.max(gap)]), "\n")
cat("  GW median:", round(hunan$gw_median[which.max(gap)]), "\n")
```

At Xinshao the local mean is 28447 and the local median 11286. One rich neighbour is carrying the
mean and the median ignores it. Reading the mean alone would put Xinshao's neighbourhood in the
middle of the province's distribution when most of it is near the bottom.

### Kernel choice and the quantile imbalance

The quantile imbalance records where the local median sits between the local quartiles. Write
that position as $t = (\text{Med} - Q_1)/(Q_3 - Q_1)$, so $t = 0$ puts the median on the lower
quartile and $t = 1$ on the upper. The deck's formula then reduces to $QI = 2t - 1$: negative means the median sits
nearer the lower quartile, positive nearer the upper, and zero means it is centred. It divides by
the weighted IQR, so it is undefined wherever that IQR is zero.

The bisquare kernel reaches exactly zero at the bandwidth, so at `bw = 6` adaptive the outermost
of the six points is given a weight of 0 and only five carry any weight. With that few, the weighted
first and third quartiles can land on the same observation. A Gaussian kernel never assigns a
zero weight, so every neighbour contributes something.

```{r}
gw_g <- gwss(hunan_sp, vars = "GDPPC", bw = 6, kernel = "gaussian",
             adaptive = TRUE, longlat = FALSE, quantile = TRUE)
gwr_g <- as.data.frame(gw_g$SDF)
tibble(kernel = c("bisquare", "gaussian"),
       `IQR of 0` = c(sum(gwr$GDPPC_IQR == 0), sum(gwr_g$GDPPC_IQR == 0)),
       `NaN imbalance` = c(sum(is.nan(gwr$GDPPC_QI)), sum(is.nan(gwr_g$GDPPC_QI)))) |> kable()
```

Ten counties against none. Under the Gaussian kernel the quantile imbalance runs from -0.777 to
0.632 across the province, so no county's median sits at a quartile but several come close.

The bisquare kernel is the usual default and is the wrong default at a bandwidth this small. The
failure surfaces as `NaN` in a column rather than as an error, the same shape as the
`style = "B"` problem above.

## Lagged explanatory variables in regression

Chapter 8 computes four lagged variables and never says what they are used for. Grubesic and
Rosso are a worked case. Their definition, the one the lecture puts on slide 15, excludes the
county itself from its own lag:

$$[Wx]_i = \sum_{j \neq i} w_{ij} x_j$$

Section 3.6.1 follows that definition, and the two window versions in §3.6.3 and §3.6.4 are a
deliberate departure from it.

Their use for it is as an explanatory variable rather than a mapped quantity. They contrast it
with the more familiar model where the **dependent** variable is lagged:

> rather than creating a multiplier effect as with spatially lagged dependent variables, spatial
> cross-regressive terms can be used directly in a standard regression framework. In other words,
> even ordinary least squares regressions would work with these data and not bias $\gamma$.

So a lagged explanatory variable can go straight into an ordinary regression, where a lagged
dependent variable cannot. They used a queen's contiguity matrix to build theirs. Section 4.1
settled on the same default here for a different reason.

They also give a way to say how much lagging alters the picture: cut both the variable and its
lag into tertiles and measure agreement. On Philadelphia census tracts they report 44% of tracts
in the same tertile, with a Cohen's kappa of 0.2.

The same measurement on Hunan:

```{r}
tertile <- function(x) cut(x, breaks = quantile(x, c(0, 1/3, 2/3, 1)),
                           include.lowest = TRUE, labels = c("low", "mid", "high"))
t_own <- tertile(hunan$GDPPC)
t_lag <- tertile(hunan$lag_GDPPC)
tab <- table(own = t_own, lagged = t_lag)
tab
```

```{r}
po <- sum(diag(tab)) / sum(tab)
pe <- sum(rowSums(tab) * colSums(tab)) / sum(tab)^2
tibble(study = c("Hunan GDPPC, this page", "Philadelphia amenity diversity, Grubesic and Rosso"),
       `same tertile` = c(po, 0.44),
       kappa = c((po - pe) / (1 - pe), 0.2)) |>
  kable(digits = 3)
```

A Hunan county aligns more closely with its surrounding neighbours than a Philadelphia tract
does: 56.8% fall in the same tertile as their neighbourhood, against 44%. Cohen's kappa, which
discounts the agreement that would arise by chance alone, is 0.35 against 0.2. Neither is high. Even here, 43% of counties fall in a different tertile from the
neighbourhood around them, and Lesson 5 identifies those as spatial outliers.

The paper connects its own figure to its result. In their models the lagged variable was not
significant, every interaction term came back above p = 0.2, and the fit did not improve. They
give their own concordance of 0.2 as part of the reason, and set one alternative beside it:
"measures at the participant's own census tract are sufficient to capture relevant neighbourhood
characteristics."

## The lag scatter and Moran's I

The spatial lag from section 3.6.1 is one step away from Lesson 5. Plotting each county's GDPPC
against its neighbourhood mean gives the scatter that Moran's I summarises.

```{r}
#| fig-height: 6
ggplot(hunan, aes(x = GDPPC, y = lag_GDPPC)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, colour = "firebrick") +
  geom_vline(xintercept = mean(hunan$GDPPC), linetype = "dashed", colour = "grey50") +
  geom_hline(yintercept = mean(hunan$lag_GDPPC), linetype = "dashed", colour = "grey50") +
  labs(x = "GDPPC (yuan)", y = "Spatially lagged GDPPC (yuan)",
       title = "County GDPPC against its neighbourhood mean",
       subtitle = "Queen contiguity, row standardised weights") +
  theme_minimal()
```

```{r}
cor(hunan$GDPPC, hunan$lag_GDPPC)
```

The upward slope means prosperous counties tend to sit beside prosperous counties, with a Pearson
correlation of 0.485.

The dashed lines mark the two means and split the counties into four groups: prosperous beside
prosperous (top right), poor beside poor (bottom left), and the two mismatched groups off the
diagonal. Lesson 5 names these four and tests which memberships are stronger than chance.

Every value on the vertical axis depends on the weights matrix built in Part 1.

# Reflection

Whether the weights matrix matters is an open question in the readings themselves.

The INSEE chapter quotes both positions. Bhattacharjee and Jensen-Butler found that results vary
considerably depending on how the weights are defined, and Harris, Moffat and Kravtsova argue
that a researcher picks a weights structure and then has no way of checking whether it was the
right one. LeSage and Pace disagree, calling that worry "the biggest myth in spatial
econometrics" and putting it down to misreading the model's coefficients.

My own results support both, depending on whether the claim is about the province or about one
county.

Across the province, LeSage and Pace look right. Queen and Rook contiguity give different lagged
values for only 8 counties out of 88. Computing centroids in degrees rather than metres moves
them 18 m at the median, which the 62 km band absorbs completely. The k nearest neighbour lists
are disturbed by that shift, and still rank the counties almost identically, at a rank
correlation of 0.984. The north east is prosperous and the west is not, under every weights
choice tested here.

County by county, the critics look right. Longhui's neighbourhood mean moves 41.4% on that same
centroid shift, because a k nearest list has no distance threshold to absorb a small error. Anren
moves 17.7% when one corner contact is dropped. And `style = "B"` returns a figure in yuan per
kilometre, which cannot be read as a neighbourhood income at all, yet it still correlates 0.78
with the correct version, so a map of it looks fine. A claim about one county is a claim about
the weights matrix as much as about the data, the same way Exercise 2 found that a claim about
one hotspot was a claim about the bandwidth.

Bavaud puts the general case: "there is no such thing as 'true', 'universal' spatial weights,
optimal in all situations: good candidates must reflect the properties of the particular
phenomenon." So the weights have to be named and the alternatives shown.

One of my own conclusions changed while reading. I had written that the `style = "B"` problem was
a one-character fix. The INSEE chapter argues that row standardising inverse distance weights
makes them hard to interpret, which means the chapter's code may be the better choice and only
its sentence wrong.

The recurring hazard in this exercise was software that fails quietly. Four calls ran with no
error and no warning while doing something other than what the text around them described:
`style = "B"` returning yuan per kilometre where a neighbourhood mean was described, `gwss()`
dropping the median, IQR and quantile imbalance because `quantile = TRUE` was left out, the
bisquare kernel producing ten `NaN` values, and `knearneigh()` finding each county's nearest
neighbour in degrees while `nbdists()` measured it in kilometres, so 61.79 km was not the largest
first nearest neighbour distance.

With more time I would test the bandwidth in section 4.6 instead of choosing it. Setting it to 6
matches the average number of contiguity neighbours, which is a reason but not a test. `GWmodel`
has no automatic selector for `gwss()`, so the alternative is a sensitivity table across several
bandwidths, like the one in section 4.3.

This page does not test whether Hunan's spatial pattern is stronger than chance would produce.
The scatter in section 4.8 slopes upwards, and Moran's I in Lesson 5 decides whether that slope
means anything.

# References

Course materials:

- Kam, T.S. (2026), [Lesson 4: Spatial Weights and
  Applications](https://isss626-ay2026-27aug.netlify.app/lesson/lesson04/lesson04-spatial_weights),
  ISSS626 Geospatial Analytics and Applications, SMU.
- Kam, T.S. (2026), [Chapter 8: Spatial Weights and
  Applications](https://r4gdsa.netlify.app/chap08.html), *R for Geospatial Data Science and
  Analytics*.

Method:

The three named on the Lesson 4 slides, all read in full:

- de Bellefon, M.-P., Loonis, V. and Le Gleut, R. (2018), ["Codifying the neighbourhood
  structure"](https://www.insee.fr/en/statistiques/fichier/3635545/imet131-f-chapitre-2.pdf),
  chapter 2 of *Handbook of Spatial Analysis: Theory and Application with R*, INSEE. The assigned
  pre-class reading.
- Bavaud, F. (1998), ["Models for Spatial Weights: A Systematic
  Look"](https://onlinelibrary.wiley.com/doi/abs/10.1111/j.1538-4632.1998.tb00394.x),
  *Geographical Analysis*, 30(2), 153-171. The slide dates this 2010; the paper and the DOI both
  give 1998.
- Grubesic, T.H. and Rosso, A.L. (2014), ["The Use of Spatially Lagged Explanatory Variables for
  Modeling Neighborhood Amenities and Mobility in Older
  Adults"](https://www.jstor.org/stable/26326897), *Cityscape*, 16(2), 205-214.

Cited within those, and used in sections 4.4, 4.5 and 5:

- Tiefelsdorf, M. (1998), *Modelling spatial processes: the identification and analysis of spatial
  relationships in regression residuals by means of Moran's I*, PhD thesis. The source of the `S`
  variance-stabilising scheme.
- Bhattacharjee, A. and Jensen-Butler, C. (2005), "Estimation of spatial weights matrix in a
  spatial error model", *CRIEFF Discussion Papers*.
- Harris, R., Moffat, J. and Kravtsova, V. (2011), "In search of 'W'", *Spatial Economic
  Analysis*, 6(3), 249-270.
- LeSage, J.P. and Pace, R.K. (2010), "The biggest myth in spatial econometrics", SSRN 1725503.
- Toussaint, G.T. (1980), "The relative neighbourhood graph of a finite planar set", *Pattern
  Recognition*, 12(4), 261-268.
- Anselin, L. (2002), "Under the hood: issues in the specification and interpretation of spatial
  regression models", *Agricultural Economics*, 27(3), 247-267. The source of the
  cross-regressive argument in 4.7.

Alternative reading named on the outline, not used here:

- Bivand, R.S., Pebesma, E. and Gómez-Rubio, V. (2013), *Applied Spatial Data Analysis with R*,
  2nd edition, chapter 9 to section 9.3.1. Springer.
- Brunsdon, C., Fotheringham, A.S. and Charlton, M. (2002), "Geographically weighted summary
  statistics: a framework for localised exploratory data analysis", *Computers, Environment and
  Urban Systems*, 26(6), 501-524. The method behind `gwss()`.

Packages:

- Bivand, R.S. and Wong, D.W.S. (2018), ["Comparing implementations of global and local indicators
  of spatial association"](https://doi.org/10.1007/s11749-018-0599-x), *TEST*, 27(3), 716-748.
  The `spdep` reference.
- Gollini, I., Lu, B., Charlton, M., Brunsdon, C. and Harris, P. (2015), ["GWmodel: An R Package
  for Exploring Spatial Heterogeneity Using Geographically Weighted
  Models"](https://www.jstatsoft.org/article/view/v063i17), *Journal of Statistical Software*,
  63(17).

Data:

- Hunan county boundary layer and `Hunan_2012.csv`, the two files named in Chapter 8.

# About this document

Written and executed in Quarto with R `r getRversion()`. All figures are generated at render time
from the data in `data/`, so nothing on this page is pasted in.

I used AI assistance while working out the approach and the extensions in Part 2. The analysis
was run and checked by me.
