---
title: "Hands-on Exercise 1"
subtitle: "Geospatial Data Wrangling and Choropleth Mapping with R"
author: "An Loc"
date: "last-modified"
date-format: "D MMM YYYY"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-copy: true
    fig-width: 9
    fig-height: 7
execute:
  echo: true
  eval: true
  warning: false
  message: false
  freeze: true
---

# Overview

This exercise has three parts.

**Part 1** covers the data preparation: importing vector data from several sources, getting
their coordinate reference systems to agree, converting a table of coordinates into a spatial
object, and answering two questions using geoprocessing.

**Part 2** joins population data to the subzone boundaries and maps it as a choropleth. Most of
the work there is about choosing a classification scheme rather than accepting the default.

**Part 3** goes beyond what the chapters ask for. I wanted to check how much the conclusions
depend on the classification I picked, whether the pattern holds when the spatial units change,
and whether bringing in a second dataset could explain the pattern instead of just describing
it.

::: {.callout-note appearance="simple"}
## Reference

r4gdsa [Chapter 1 — Geospatial Data Science with R](https://r4gdsa.netlify.app/chap01.html)
and [Chapter 2 — Thematic Mapping and GeoVisualisation with R](https://r4gdsa.netlify.app/chap02.html).
:::

::: {.callout-note appearance="simple"}
## Anselin's four types of spatial data

The pre-lesson lecture by Luc Anselin, *"Spatial Data, Spatial Analysis, Spatial Data
Science"*, sets out four types of spatial data. Three of them show up in this exercise:

| Anselin's type | In this exercise | Note |
|---|---|---|
| **Points** | Pre-schools, Airbnb listings | He separates points that are *events* (crimes, accidents) from points that are *locations of objects* (stores, facilities). Ours are the second kind |
| **Lattice / areal units** | Subzone polygons | His term for administrative zones. Relevant later, since MAUP is a lattice problem |
| **Networks** | Cycling paths | Nodes and links. Treated as plain lines here. Network-constrained analysis comes in Lesson 7 |
| **Surfaces** | *not used* | Continuous fields needing interpolation, like air quality or price surfaces |

He also names the standard that the `sf` package implements: the **Simple Features
specification** from the Open Geospatial Consortium. That is where the name "sf" comes from,
and why every function starts with `st_` (spatial type).

He also poses the problem that §"How many pre-schools are in each subzone?" solves below:
*"You look at the map, you know immediately that city is in that county. How would a computer
know that that point is inside that polygon?"* In R, the answer is `st_intersects()`.
:::

::: {.callout-note appearance="simple"}
## Xie et al. on the three foundations

The second pre-lesson reading, Xie et al. (2017), *"Transdisciplinary Foundations of Geospatial
Data Science"*, argues that every geospatial technique rests on three foundations, and that
problems arise when any of them is treated as a black box. All three appear in this exercise:

| Foundation | Their concern | Where it appears below |
|---|---|---|
| **Mathematics** | Geometry, topology, projection | The CRS transform to EPSG:3414, and `st_intersects()` as a topological predicate rather than a lookup |
| **Statistics** | Spatial data violate the i.i.d. assumption most methods rely on | The correlation in Part 3, and the caveat attached to it |
| **Computer science** | Efficiency, scalability, filter-and-refine | `st_intersects()` builds a spatial index instead of testing 323 × 2,290 pairs |

One warning from the paper is useful throughout: *"traditional density-based clustering methods
will generally output chance patterns if applied to geospatial hotspot detection"*. K-means will
find clusters in pure noise, so finding a pattern is not by itself evidence that one exists.
Anselin makes the same point with his shuffled map.
:::

# Getting started

```{r}
#| label: packages
pacman::p_load(sf, tmap, tidyverse, ggplot2, knitr, classInt, units)
```

::: {.callout-note appearance="simple"}
## `pacman::p_load()` versus `library()`

`p_load()` installs a package if it is missing and then loads it. That means the document will
still run on a machine that does not already have every dependency installed, which matters more
for a reproducible report than the couple of keystrokes it saves.
:::

## Note on tmap versions

The course materials are written for `tmap` v3, but the version installed on my machine is v4,
where several function names and the whole scale and legend argument system changed. The v3 code
still runs through a compatibility layer, but it prints `[v3->v4]` deprecation messages
constantly, which made the output hard to read.

I used v4 syntax throughout. These are the substitutions I made:

| v3 | v4 |
|---|---|
| `tm_fill(col = "X", style = "quantile", palette = "Blues")` | `tm_polygons(fill = "X", fill.scale = tm_scale_intervals(style = "quantile", values = "brewer.blues"))` |
| `tm_layout(main.title = "...")` | `tm_title("...")` |
| `tm_scale_bar()` | `tm_scalebar()` |
| `tm_polygons(col = "X")` | `tm_polygons(fill = "X")`, since `col` now sets the border colour |

```{r}
#| label: tmap-version
packageVersion("tmap")
```

# Data provenance

::: {.callout-important}
## The dataset named in the chapter has no file attached

Chapter 1 specifies **Master Plan 2014 Subzone Boundary (Web)**. Its
[collection page](https://data.gov.sg/collections/1708/view) is still live on data.gov.sg, but
there is no dataset file attached to it. The collections API reports zero child datasets for it,
while every comparable collection (MP2019 Subzone No Sea, MP2019 Region No Sea, and the others
in the series) reports one. So the page can be browsed but the data cannot be downloaded.

::: {.callout-note appearance="simple"}
## How I checked, including the wrong way

My first check was the older URL
`data.gov.sg/dataset/master-plan-2014-subzone-boundary-web`, which returns "Dataset not found".
I initially took that as proof the dataset had been withdrawn, but it isn't. That older
`/dataset/<slug>` route returns the same error for datasets that are definitely still available,
including Pre-Schools Location and MP2014 (No Sea), both of which I downloaded for this exercise
without any trouble.

The check that actually works is the current API, which distinguishes a collection that has
files from one that is empty:

```
https://api-production.data.gov.sg/v2/public/api/collections/{id}/metadata  →  childDatasets[]
```

I am recording this because the wrong conclusion was easy to reach and looked well evidenced at
the time.
:::

I made two substitutions, and have recorded both rather than changing them silently:

- **Master Plan 2014 Subzone Boundary (No Sea)**
  ([`d_226cacceceff94f0c8b814962a5307c9`](https://data.gov.sg/datasets/d_226cacceceff94f0c8b814962a5307c9/view))
  replaces the (Web) variant in Part 1. Both use the same 2014 boundary definition and differ in
  where the polygons stop at the coast. "(No Sea)" is clipped to the coastline, while "(Web)" is
  the version published on URA's website and OneMap, where coastal subzones extend out over
  territorial waters.

  ⚠️ URA's published descriptions of the two variants are almost identical boilerplate and do
  not actually state this difference. I am going by the naming convention, and I cannot check it
  against the (Web) file since that file is not downloadable. What I can measure is that the
  (No Sea) layer totals
  **`r round(as.numeric(sum(units::set_units(st_area(st_transform(st_read("data/geospatial/MasterPlan2014SubzoneBoundaryNoSea.geojson", quiet=TRUE), 3414)),"km^2"))), 0)` km²**
  against Singapore's ~734 km² land area, the excess being reservoirs and inland water.

  For this exercise (No Sea) is the better choice anyway. Sea polygons contain no pre-schools
  but would increase subzone areas, which would distort the density measure.
- **Master Plan 2019 Subzone Boundary (No Sea)** is used for Part 2, matching the current
  chapter 2 text.

Separately, data.gov.sg now serves GeoJSON where the chapter expects shapefile or KML. Prof Kam
confirmed on Piazza that this is fine. GeoJSON is a single self-describing file that always
declares WGS84 by specification, whereas a shapefile is a bundle of several files whose `.prj`
component can be missing or wrong, which is the same failure mode the CRS section below deals
with.
:::

| Dataset | Source | Format | Retrieved | Native CRS |
|---|---|---|---|---|
| Master Plan 2014 Subzone Boundary (No Sea) | data.gov.sg (URA) | GeoJSON | 25 Aug 2026 | WGS84 (EPSG:4326) |
| Master Plan 2019 Subzone Boundary (No Sea) | data.gov.sg (URA) | GeoJSON | 25 Aug 2026 | WGS84 (EPSG:4326) |
| Pre-Schools Location | data.gov.sg (ECDA) | GeoJSON | 25 Aug 2026 | WGS84 (EPSG:4326) |
| Cycling Path Network | data.gov.sg (LTA) | GeoJSON | 25 Aug 2026 | WGS84 (EPSG:4326) |
| Singapore Airbnb listings | Inside Airbnb, 29 Jun 2026 snapshot | CSV | 25 Aug 2026 | none — lat/long columns |
| Resident population by subzone, age and sex | data.gov.sg (SingStat, Census 2020) | CSV | 25 Aug 2026 | none — subzone name key |

All six can be downloaded without an account or API key. Each one comes from the source listed
above, so this page can be reproduced from a clean machine.

# Part 1 — Geospatial Data Wrangling

## Importing geospatial data

```{r}
#| label: import-mpsz14
mpsz14 <- st_read("data/geospatial/MasterPlan2014SubzoneBoundaryNoSea.geojson")
```

```{r}
#| label: import-cycling
cyclingpath <- st_read("data/geospatial/CyclingPathNetwork.geojson")
```

```{r}
#| label: import-preschool
preschool <- st_read("data/geospatial/PreSchoolsLocation.geojson")
```

::: {.callout-tip appearance="simple"}
## What `st_read()` prints

There are three useful things in that output.

**Feature count and geometry type.** 323 `POLYGON` subzones, 8,523 `LINESTRING` cycling paths
and 2,290 `POINT` pre-schools. The geometry type limits which operations make sense. You can
buffer a line, but computing its area would be meaningless.

**The bounding box.** The values here are around `103.6` to `104.1` and `1.2` to `1.5`, which
are longitude and latitude in degrees. If they were in the hundreds of thousands they would be
metres in a projected system. This is the quickest way to check the data arrived in the CRS you
expected.

**The declared CRS.** `WGS 84` in all three cases. The next section covers why that is a
problem.
:::

## Inspecting the feature data frame

An `sf` object is a data frame with an extra geometry column. All the `dplyr` verbs work on it,
and the geometry stays attached to the rows as you filter and mutate.

```{r}
#| label: inspect-geom
st_geometry(mpsz14)
```

```{r}
#| label: inspect-glimpse
glimpse(mpsz14)
```

```{r}
#| label: inspect-head
head(mpsz14, n = 3)
```

::: {.callout-note appearance="simple"}
## Comparing the three inspection functions

`st_geometry()` shows the geometry type, bounding box and CRS, and nothing about the attributes.
`glimpse()` shows which columns exist and what type each one is, which is how you catch a numeric
field that has arrived as character. `head()` shows what an actual record looks like, with
attributes and geometry together.
:::

## Working with projection

```{r}
#| label: crs-check
st_crs(mpsz14)$input
st_crs(cyclingpath)$input
st_crs(preschool)$input
```

All three arrived as WGS84 (EPSG:4326), a geographic CRS whose units are degrees, which is not
usable for anything involving distance or area. A one-degree buffer would be roughly 111 km, and
near the equator a degree of longitude and a degree of latitude are not the same distance on the
ground anyway.

Singapore's national projected CRS is SVY21, EPSG:3414, and its units are metres.

```{r}
#| label: crs-transform
mpsz14      <- st_transform(mpsz14, crs = 3414)
cyclingpath <- st_transform(cyclingpath, crs = 3414)
preschool   <- st_transform(preschool, crs = 3414)

st_bbox(mpsz14)
```

The bounding box now shows six-figure easting and northing values instead of degrees, so the
transform worked.

::: {.callout-important}
## Difference between `st_set_crs()` and `st_transform()`

These two are easy to confuse and I want to be able to state the difference precisely.

`st_set_crs()` changes the label on the coordinates without touching the numbers. It asserts
that the numbers were always in that system. It should only be used when the data arrived with a
missing or wrong CRS tag and you know what the correct one is.

`st_transform()` changes the numbers, recalculating every coordinate into the target system. It
is what you use when you actually need the data in a different CRS.

Using the wrong one produces no error and no warning. You just get a layer sitting off the coast
of Africa, or distances wrong by a factor of 100,000, and nothing downstream will flag it.
:::

## Trying the KML version

When a classmate asked on Piazza whether GeoJSON was acceptable given that data.gov.sg no longer
serves KML, Prof Kam replied: *"It is fine. However, there is nothing wrong to explore the kml
very. You will learn more by trying."*

So I tried it. Writing one layer out to KML and reading it straight back shows what the format
costs.

```{r}
#| label: kml-write
dir.create("data/geospatial/kml", recursive = TRUE, showWarnings = FALSE)

st_write(mpsz14,
         "data/geospatial/kml/MP14_Subzone.kml",
         driver = "KML",
         delete_dsn = TRUE,
         quiet = TRUE)

mpsz14_kml <- st_read("data/geospatial/kml/MP14_Subzone.kml", quiet = TRUE)
```

```{r}
#| label: kml-compare
tibble(
  aspect = c("Number of attribute columns", "Column names", "CRS", "Feature count"),
  geojson = c(as.character(ncol(mpsz14) - 1),
              paste(head(setdiff(names(mpsz14), "geometry"), 4), collapse = ", "),
              st_crs(mpsz14)$input,
              as.character(nrow(mpsz14))),
  kml = c(as.character(ncol(mpsz14_kml) - 1),
          paste(head(setdiff(names(mpsz14_kml), "geometry"), 4), collapse = ", "),
          st_crs(mpsz14_kml)$input,
          as.character(nrow(mpsz14_kml)))
) %>%
  kable(caption = "What survives a round trip through KML")
```

::: {.callout-important}
## What changed after the KML round trip

**The attribute table collapsed.** All the named columns, `SUBZONE_N`, `PLN_AREA_N`, `REGION_N`
and the rest, were flattened into a generic `Name` field and a `Description` field containing an
HTML fragment. The geometry came back intact but the schema did not. To analyse the result I
would have to parse the attributes back out of HTML strings, which is the same state the
Pre-Schools GeoJSON arrives in.

**The CRS reverted to WGS84.** The KML specification requires WGS84, so `st_write()` reprojected
out of the SVY21 metres I had just transformed into, without warning me. Reading it back gives
degrees again, and every buffer and area calculation in this document would be wrong until I
transformed it a second time.

This does not mean KML is bad, just that it is a presentation format, designed for drawing
things in Google Earth where a person reads the description bubble. GeoJSON and shapefile are
analysis formats that keep typed attribute columns and let you declare a projected CRS. The
chapter's instruction to use shapefile or KML predates data.gov.sg switching to GeoJSON, and for
this work GeoJSON is the best of the three.
:::

## Importing and converting aspatial data

```{r}
#| label: import-listings
listings <- read_csv("data/aspatial/listings.csv", show_col_types = FALSE)
glimpse(listings)
```

The file has `latitude` and `longitude` columns but no geometry, so R treats it as an ordinary
table. `st_as_sf()` converts it into a spatial object.

```{r}
#| label: listings-to-sf
listings_sf <- st_as_sf(listings,
                        coords = c("longitude", "latitude"),
                        crs = 4326) %>%
  st_transform(crs = 3414)

glimpse(listings_sf)
```

::: {.callout-warning}
## `coords` takes longitude first

The argument is `coords = c("longitude", "latitude")`, x before y. Writing
`c("latitude","longitude")` does not throw an error. It just places every Singapore listing at
about 1°E, 103°N, which is in the Arctic Ocean north of Norway.

The `crs = 4326` argument here is doing the labelling job described in the previous section. The
numbers are already in degrees, so I tag them correctly first and only then transform to metres.
Setting `crs = 3414` at this point would claim the degree values were metres and put every
listing within a kilometre of the origin.
:::

```{r}
#| label: listings-check
st_bbox(listings_sf)
```

The bounding box falls within Singapore, so the coordinate order was right.

## Geoprocessing

### How much land is within 5 m of a cycling path?

This is a buffer and area problem. The scenario is that LTA wants to know how much land would be
needed to widen the entire cycling network by 5 m on each side.

```{r}
#| label: buffer
buffer_cycling <- st_buffer(cyclingpath, dist = 5, nQuadSegs = 30)
buffer_cycling$AREA <- st_area(buffer_cycling)
sum(buffer_cycling$AREA)
```

```{r}
#| label: buffer-units
total_ha <- sum(buffer_cycling$AREA) %>% units::set_units("hectares")
total_ha
```

::: {.callout-note appearance="simple"}
## Notes on this figure

**The units are carried by the object.** `st_area()` returns a `units` object, so `sum()`
reports m² explicitly and `set_units()` converts safely. Whether the `5` meant five metres or
five degrees is exactly what the earlier transform to EPSG:3414 settled. In WGS84 this buffer
would have been about 555 km wide.

**This total double-counts the overlaps.** Where two cycling paths run close together their
buffers intersect, and summing the per-feature areas counts that shared land twice. For a land
acquisition estimate the buffers should be dissolved first:

```{r}
#| label: buffer-dissolve
dissolved_ha <- buffer_cycling %>%
  st_union() %>%
  st_area() %>%
  units::set_units("hectares")

dissolved_ha
```

The difference between the two figures is the overlapping area, and it is not small. Reporting
the first number as a land requirement would overstate it.
:::

```{r}
#| label: buffer-overlap
overlap_pct <- as.numeric((total_ha - dissolved_ha) / total_ha * 100)
sprintf("Naive sum overstates the true footprint by %.1f%%", overlap_pct)
```

### How many pre-schools are in each subzone?

```{r}
#| label: preschool-count
mpsz14$PreSch_Count <- lengths(st_intersects(mpsz14, preschool))

summary(mpsz14$PreSch_Count)
```

```{r}
#| label: preschool-top
mpsz14 %>%
  st_drop_geometry() %>%
  select(SUBZONE_N, PLN_AREA_N, PreSch_Count) %>%
  arrange(desc(PreSch_Count)) %>%
  head(10) %>%
  kable(caption = "Subzones with the most pre-schools")
```

::: {.callout-note appearance="simple"}
## What `st_intersects()` returns

It does not return a count. It returns a list with one element per subzone, each holding the row
indices of every pre-school that falls inside it. `lengths()`, plural, then measures each element
of that list. Using `length()` instead would return a single number, 323, which is just the
number of subzones.

The list of indices is more useful than a count would be, since it lets you go back and ask
which pre-schools rather than only how many.
:::

### Pre-school density

```{r}
#| label: preschool-density
mpsz14$Area_sqkm <- st_area(mpsz14) %>%
  units::set_units("km^2") %>%
  as.numeric()

mpsz14$PreSch_Density <- mpsz14$PreSch_Count / mpsz14$Area_sqkm

mpsz14 %>%
  st_drop_geometry() %>%
  select(SUBZONE_N, PreSch_Count, Area_sqkm, PreSch_Density) %>%
  arrange(desc(PreSch_Density)) %>%
  head(10) %>%
  mutate(across(where(is.numeric), ~round(.x, 2))) %>%
  kable(caption = "Highest pre-school density (per km squared)")
```

Density and count rank the subzones quite differently. The highest count is in a large
residential town, while the highest density tends to be a small, dense subzone. Which measure is
right depends on whether the question is about provision relative to land area or provision in
absolute terms.

## Exploratory data analysis

```{r}
#| label: eda-hist
ggplot(data = mpsz14, aes(x = PreSch_Density)) +
  geom_histogram(bins = 20, color = "black", fill = "#5DA5DA") +
  labs(title = "Pre-school density is strongly right-skewed",
       subtitle = "Most subzones have none; a few have many per square kilometre",
       x = "Pre-schools per square kilometre",
       y = "Number of subzones") +
  theme_minimal()
```

This skew turns out to matter a lot in Part 2. A variable shaped like this gets classified very
differently by an equal-interval scheme than by a quantile scheme, which is what the
classification section tests.

```{r}
#| label: eda-scatter
ggplot(data = mpsz14, aes(x = PreSch_Count, y = PreSch_Density)) +
  geom_point(alpha = 0.6, color = "#4A6FA5") +
  labs(title = "Count and density are only loosely related",
       subtitle = "Large subzones can hold many pre-schools at low density",
       x = "Pre-school count", y = "Pre-schools per square kilometre") +
  theme_minimal()
```

# Part 2 — Choropleth Mapping

## Importing the data

```{r}
#| label: import-mpsz19
mpsz19 <- st_read("data/geospatial/MasterPlan2019SubzoneBoundaryNoSea.geojson") %>%
  st_transform(crs = 3414)
```

```{r}
#| label: import-popdata
popdata <- read_csv("data/aspatial/respop_census2020_agesex.csv", show_col_types = FALSE)
dim(popdata)
```

The table is in wide format, with one row per area and one column for each age band crossed with
sex. The `Number` column holds the area name. Planning-area rows are marked with a `" - Total"`
suffix and subzone rows just carry a bare name.

```{r}
#| label: popdata-peek
popdata %>% select(Number, Total_Total, Total_0_4, Total_65_69) %>% head(6) %>% kable()
```

### Every count column arrived as text

```{r}
#| label: popdata-types
sum(sapply(popdata, is.character))
```

All 61 columns came in as character, including the population counts. The cause is one token:
SingStat writes suppressed or zero cells as `-`, and that forces `readr` to type the whole column
as text.

::: {.callout-warning}
## Why the columns imported as text

`rowSums()` on character columns errors immediately, which is the lucky outcome. The version that
would have caused real trouble is code that coerces silently and returns a wrong number that
looks plausible.

`as.numeric("-")` returns `NA` rather than zero. A straight conversion would push `NA` through
every age band containing a suppressed cell and quietly drop those subzones from the map. I read
`-` as zero here, since the census uses it where the count is nil or has been suppressed for
disclosure control, and both cases mean there is no meaningful population in that cell.
:::

```{r}
#| label: popdata-clean
to_count <- function(x) {
  x <- str_replace_all(x, ",", "")   # thousands separators, if any
  x[x == "-"] <- "0"                 # suppressed / nil
  as.numeric(x)
}

popdata <- popdata %>%
  mutate(across(-Number, to_count))

sum(sapply(popdata, is.character))
```

Only `Number` is still character, which is correct since it holds the area name.

## Computing the dependency ratio

The age dependency ratio is the number of people not usually in the labour force, meaning
children under 15 and adults 65 and over, per person of working age.

$$\text{Dependency} = \frac{\text{population}_{0-14} + \text{population}_{65+}}{\text{population}_{15-64}}$$

```{r}
#| label: dependency
young_cols  <- c("Total_0_4", "Total_5_9", "Total_10_14")
active_cols <- c("Total_15_19","Total_20_24","Total_25_29","Total_30_34","Total_35_39",
                 "Total_40_44","Total_45_49","Total_50_54","Total_55_59","Total_60_64")
aged_cols   <- c("Total_65_69","Total_70_74","Total_75_79","Total_80_84",
                 "Total_85_89","Total_90andOver")

popdata_sz <- popdata %>%
  filter(Number != "Total",
         !str_detect(Number, " - Total$")) %>%
  mutate(
    SUBZONE_N = toupper(str_trim(Number)),
    YOUNG     = rowSums(across(all_of(young_cols))),
    ACTIVE    = rowSums(across(all_of(active_cols))),
    AGED      = rowSums(across(all_of(aged_cols))),
    TOTAL     = Total_Total,
    DEPENDENCY = (YOUNG + AGED) / ACTIVE
  ) %>%
  select(SUBZONE_N, YOUNG, ACTIVE, AGED, TOTAL, DEPENDENCY)

nrow(popdata_sz)
```

## Joining the population data to the boundaries

```{r}
#| label: join
mpsz_pop <- left_join(mpsz19 %>% mutate(SUBZONE_N = toupper(SUBZONE_N)),
                      popdata_sz,
                      by = "SUBZONE_N")

n_total     <- nrow(mpsz_pop)
n_unmatched <- sum(is.na(mpsz_pop$TOTAL))
n_zero_act  <- sum(mpsz_pop$ACTIVE == 0, na.rm = TRUE)

tibble(
  metric = c("Subzone polygons", "No population record", "Matched but zero working-age population"),
  n      = c(n_total, n_unmatched, n_zero_act)
) %>% kable(caption = "Join diagnostics")
```

::: {.callout-warning}
## Checking the join

Two boundary definitions are being reconciled here. The population table is keyed to Census 2020
subzone names and the polygons to Master Plan 2019, so some level of non-match is expected. The
`toupper()` call on both sides removes a case mismatch that would otherwise have failed most of
the join without any error.

The rows with zero working-age population matter more than they first appear. Dependency is a
ratio with `ACTIVE` in the denominator, so those subzones produce `Inf` rather than a large
number. Industrial estates, reservoirs and military areas genuinely have no residents. Mapping
`Inf` shows up as an unclassified hole, whereas mapping it as a high value would be inventing
data.
:::

```{r}
#| label: join-unmatched
mpsz_pop %>%
  st_drop_geometry() %>%
  filter(is.na(TOTAL)) %>%
  select(SUBZONE_N, PLN_AREA_N) %>%
  head(10) %>%
  kable(caption = "Sample of subzones with no matching population record")
```

```{r}
#| label: clean-inf
mpsz_pop <- mpsz_pop %>%
  mutate(DEPENDENCY = ifelse(is.finite(DEPENDENCY), DEPENDENCY, NA_real_))

summary(mpsz_pop$DEPENDENCY)
```

::: {.callout-note appearance="simple"}
## Choropleth as a histogram with location

Anselin's definition is the one I found most useful: a choropleth is a histogram with the
location made explicit. *"Rather than just having a histogram as such, we will know where these
bars for the histogram are in geographical space."*

This is useful in two places below. It explains why classification matters, since choosing class
breaks is the same act as choosing histogram bins, and nobody would accept arbitrary bins on a
histogram. It also sets up the test in Part 3, because if a map and a histogram carried the same
information then the map would add nothing.
:::

## Plotting a choropleth quickly with `qtm()`

```{r}
#| label: qtm
tmap_mode("plot")
qtm(mpsz_pop, fill = "DEPENDENCY")
```

This is quick and fine for looking at your own data, but not for showing anyone else. There is
no title, no source, no explanation of the legend, and the classification is whatever the default
happens to be.

## Building the map with `tmap` elements

```{r}
#| label: tmap-full
tm_shape(mpsz_pop) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile", n = 5,
                                              values = "brewer.blues"),
              fill.legend = tm_legend(title = "Dependency\nratio"),
              col = "grey60", lwd = 0.3,
              fill.free = FALSE) +
  tm_title("Age dependency ratio by subzone, Singapore") +
  tm_compass(type = "8star", size = 2, position = c("right", "top")) +
  tm_scalebar(position = c("left", "bottom")) +
  tm_grid(lwd = 0.1, col = "grey85") +
  tm_credits("Source: Census of Population 2020 (SingStat);\nURA Master Plan 2019 Subzone Boundary",
             position = c("left", "bottom"), size = 0.6) +
  tm_layout(frame = TRUE, legend.outside = TRUE)
```

::: {.callout-note appearance="simple"}
## What each element is for

`tm_shape()` sets the data and every `tm_*` layer after it draws from that. `tm_polygons()`
handles fill and border together, and in v4 `fill` is the thematic variable while `col` is the
border colour. `tm_grid()` adds coordinate context. `tm_compass()` and `tm_scalebar()` are
standard cartographic elements. `tm_credits()` puts the source onto the map image itself, so it
stays attributable if someone pulls the figure out of this document.
:::

## Classification methods

`tmap` supports the standard data classification schemes. They are not just style options. Each
one makes a different claim about what the reader should be noticing.

```{r}
#| label: classification-compare
#| fig-height: 11
make_map <- function(style, title) {
  tm_shape(mpsz_pop) +
    tm_polygons(fill = "DEPENDENCY",
                fill.scale = tm_scale_intervals(style = style, n = 5,
                                                values = "brewer.blues"),
                fill.legend = tm_legend(show = FALSE),
                col = "grey70", lwd = 0.2) +
    tm_title(title, size = 0.9)
}

tmap_arrange(
  make_map("equal",    "equal — equal value ranges"),
  make_map("quantile", "quantile — equal counts"),
  make_map("jenks",    "jenks — natural breaks"),
  make_map("sd",       "sd — deviation from mean"),
  ncol = 2
)
```

::: {.callout-important}
## Comparing the four classification methods

These are four maps of the same numbers.

**Equal interval** cuts the value range into equal-width bins. On a skewed variable this pushes
almost everything into the lowest class and the map ends up looking empty.

**Quantile** puts an equal count of subzones into each class. It always produces a visually
balanced map, including when there is no real variation to show, which is the risk with it.

**Jenks** minimises within-class variance and finds breaks at natural gaps in the distribution.
It is usually the most faithful to the structure, but the breaks depend on the data, so two maps
made this way cannot be compared to each other.

**Standard deviation** shows distance from the mean. It answers a different question, which
areas are unusual rather than how much there is.
:::

### Why two of those maps look empty

```{r}
#| label: class-counts
vals_all <- mpsz_pop$DEPENDENCY[is.finite(mpsz_pop$DEPENDENCY)]

class_counts <- map_dfr(c("equal","quantile","jenks","sd"), function(st) {
  brks <- classIntervals(vals_all, n = 5, style = st)$brks
  tab  <- table(cut(vals_all, unique(brks), include.lowest = TRUE))
  tibble(style = st,
         `class 1` = tab[1], `class 2` = tab[2], `class 3` = tab[3],
         `class 4` = tab[4], `class 5` = tab[5])
})

kable(class_counts, caption = "How many subzones land in each class")
```

`equal` puts `r class_counts$"class 1"[class_counts$style == "equal"]` of `r length(vals_all)`
subzones into the bottom class, and `sd` is almost as bad. Both are being distorted by one
extreme value:

```{r}
#| label: outlier
mpsz_pop %>%
  st_drop_geometry() %>%
  filter(is.finite(DEPENDENCY)) %>%
  arrange(desc(DEPENDENCY)) %>%
  select(SUBZONE_N, PLN_AREA_N, YOUNG, ACTIVE, AGED, DEPENDENCY) %>%
  head(5) %>%
  mutate(DEPENDENCY = round(DEPENDENCY, 2)) %>%
  kable(caption = "The subzones driving the top of the range")
```

::: {.callout-important}
## The effect of one outlier

The maximum dependency ratio is `r round(max(vals_all), 1)` against a median of
`r round(median(vals_all), 2)`, so roughly twenty times the typical value. It is not a data
error. It is a subzone with very few residents where the age mix happens to be extreme, and the
small denominator produces a very large ratio.

Equal-interval and standard-deviation both derive their breaks from the range, so that one
subzone stretches the whole scale and every other subzone falls into the first class. Quantile
and Jenks derive their breaks from the ordering and the distribution instead, so they are not
affected.

The practical rule is that range-based classification fails on a variable with outliers. You
either use a rank or distribution based scheme, or handle the outlier explicitly and say which
you did.
:::

### How much does the choice actually change?

Rather than just assert that classification matters, I wanted to measure it. How many subzones
actually change class when the scheme changes?

```{r}
#| label: classification-sensitivity
vals <- mpsz_pop$DEPENDENCY[!is.na(mpsz_pop$DEPENDENCY)]

styles <- c("equal", "quantile", "jenks", "sd")
class_assign <- sapply(styles, function(s) {
  brks <- classIntervals(vals, n = 5, style = s)$brks
  as.integer(cut(vals, breaks = unique(brks), include.lowest = TRUE))
})
colnames(class_assign) <- styles

pairs <- combn(styles, 2, simplify = FALSE)
sens <- map_dfr(pairs, function(p) {
  changed <- mean(class_assign[, p[1]] != class_assign[, p[2]], na.rm = TRUE) * 100
  tibble(comparison = paste(p[1], "vs", p[2]),
         pct_subzones_changing_class = round(changed, 1))
}) %>% arrange(desc(pct_subzones_changing_class))

kable(sens, caption = "Share of subzones assigned to a different class under each pair of schemes")
```

::: {.callout-tip}
## What the sensitivity table shows

Every one of those maps is correct. The table puts a number on how much of the visual story
comes from a choice made in a single argument. Someone who saw only the equal-interval map and
someone who saw only the quantile map would describe Singapore's dependency geography
differently, and neither would have any reason to know it.

For this variable I would use Jenks for exploration, since it respects the actual distribution,
and quantile only when comparing across time periods, where fixed relative positions matter more
than absolute breaks. Equal interval should not be used on a right-skewed variable at all.
:::

## Colour scheme

```{r}
#| label: colour
#| fig-height: 5
tmap_arrange(
  tm_shape(mpsz_pop) +
    tm_polygons(fill = "DEPENDENCY",
                fill.scale = tm_scale_intervals(style = "jenks", n = 5,
                                                values = "brewer.greens"),
                fill.legend = tm_legend(show = FALSE), col = "grey70", lwd = 0.2) +
    tm_title("Sequential (greens)", size = 0.9),
  tm_shape(mpsz_pop) +
    tm_polygons(fill = "DEPENDENCY",
                fill.scale = tm_scale_intervals(style = "jenks", n = 5,
                                                values = "-brewer.greens"),
                fill.legend = tm_legend(show = FALSE), col = "grey70", lwd = 0.2) +
    tm_title("Reversed — misleading here", size = 0.9),
  ncol = 2
)
```

::: {.callout-note appearance="simple"}
## Choosing sequential or diverging

A sequential palette says the variable runs from less to more with no meaningful midpoint. A
diverging palette says there is a midpoint that matters, and that deviations on either side of it
differ in kind rather than degree.

For a dependency ratio a diverging scheme can be justified if the midpoint is chosen for a
reason, such as the national ratio, because being above or below the national average really are
different situations. It cannot be justified if the midpoint is just the middle of the data
range, since that invents a threshold that does not exist.

The reversed map above shows what goes wrong. Dark now reads as low dependency, which inverts
the intuition most readers bring to a map. The code is correct and the communication is not.
:::

# Part 3 — Beyond the brief

## Is this actually spatial analysis?

Anselin proposes a test for whether an analysis is genuinely spatial:

> *"If you change the location of things and it doesn't matter, then it's not spatial analysis.
> But if it does matter, then it is."*

He demonstrates it with two maps of Milwaukee, one real and one with the same values randomly
shuffled between census tracts. The histograms are identical but the maps look like two
different cities.

I wanted to run the same test on this data rather than take it on faith.

```{r}
#| label: shuffle-setup
set.seed(1234)

mpsz_shuffled <- mpsz_pop
mpsz_shuffled$DEPENDENCY <- sample(mpsz_pop$DEPENDENCY)
```

Every value is preserved, since `sample()` only reassigns which subzone holds which number. So
any non-spatial summary should come out exactly the same:

```{r}
#| label: shuffle-identical
tibble(
  version = c("Real", "Shuffled"),
  n       = c(sum(is.finite(mpsz_pop$DEPENDENCY)), sum(is.finite(mpsz_shuffled$DEPENDENCY))),
  mean    = c(mean(mpsz_pop$DEPENDENCY, na.rm = TRUE), mean(mpsz_shuffled$DEPENDENCY, na.rm = TRUE)),
  sd      = c(sd(mpsz_pop$DEPENDENCY, na.rm = TRUE), sd(mpsz_shuffled$DEPENDENCY, na.rm = TRUE)),
  median  = c(median(mpsz_pop$DEPENDENCY, na.rm = TRUE), median(mpsz_shuffled$DEPENDENCY, na.rm = TRUE))
) %>%
  mutate(across(where(is.numeric), ~round(.x, 4))) %>%
  kable(caption = "Identical by every non-spatial measure")
```

```{r}
#| label: shuffle-hist
#| fig-height: 3.2
bind_rows(
  tibble(version = "Real",     value = mpsz_pop$DEPENDENCY),
  tibble(version = "Shuffled", value = mpsz_shuffled$DEPENDENCY)
) %>%
  filter(is.finite(value), value < 2) %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = "#4A6FA5", colour = "white", linewidth = 0.2) +
  facet_wrap(~version) +
  labs(title = "The histograms are the same graph twice",
       x = "Dependency ratio", y = "Subzones") +
  theme_minimal()
```

Now the maps:

```{r}
#| label: shuffle-maps
#| fig-height: 5
shuffle_map <- function(dat, ttl) {
  tm_shape(dat) +
    tm_polygons(fill = "DEPENDENCY",
                fill.scale = tm_scale_intervals(style = "quantile", n = 5,
                                                values = "brewer.blues"),
                fill.legend = tm_legend(show = FALSE),
                col = "grey70", lwd = 0.2) +
    tm_title(ttl, size = 0.9)
}

tmap_arrange(shuffle_map(mpsz_pop,      "Real Singapore"),
             shuffle_map(mpsz_shuffled, "Same numbers, shuffled"),
             ncol = 2)
```

::: {.callout-important}
## What the shuffle test shows

The mean, standard deviation and median are identical, and so is the histogram. But the two maps
look nothing alike. The real one has coherent regions of similar colour and the shuffled one is
more or less visual static.

Everything separating those two images is spatial structure, and none of the conventional
statistics can detect it. This seems to be the gap the rest of the course fills. Lesson 4's spatial
weights formalise what neighbouring means, and Lesson 5's Moran's I turns the difference between
these two maps into a single number with a significance test.

It also answers a question I had at the start, which was why not just use pandas for this. A
non-spatial toolkit would report these two datasets as identical, because on every measure it
can compute, they are.
:::

## Does the pattern survive a change of spatial unit? (MAUP)

The Modifiable Areal Unit Problem is that statistical results change when the zoning or the
scale of the spatial units changes. It applies to most of what this course covers, and it can be
tested directly here because subzones nest inside planning areas.

```{r}
#| label: maup
mpsz_pa <- mpsz_pop %>%
  st_drop_geometry() %>%
  filter(!is.na(TOTAL)) %>%
  group_by(PLN_AREA_N) %>%
  summarise(YOUNG = sum(YOUNG), ACTIVE = sum(ACTIVE), AGED = sum(AGED),
            n_subzones = n(), .groups = "drop") %>%
  mutate(DEPENDENCY_PA = (YOUNG + AGED) / ACTIVE) %>%
  filter(is.finite(DEPENDENCY_PA))

comparison <- tibble(
  unit = c("Subzone", "Planning area"),
  n = c(sum(is.finite(mpsz_pop$DEPENDENCY)), nrow(mpsz_pa)),
  min = c(min(mpsz_pop$DEPENDENCY, na.rm = TRUE), min(mpsz_pa$DEPENDENCY_PA)),
  median = c(median(mpsz_pop$DEPENDENCY, na.rm = TRUE), median(mpsz_pa$DEPENDENCY_PA)),
  max = c(max(mpsz_pop$DEPENDENCY, na.rm = TRUE), max(mpsz_pa$DEPENDENCY_PA))
) %>% mutate(across(where(is.numeric), ~round(.x, 3)))

kable(comparison, caption = "The same variable at two spatial scales")
```

```{r}
#| label: maup-plot
bind_rows(
  tibble(unit = "Subzone", value = mpsz_pop$DEPENDENCY[is.finite(mpsz_pop$DEPENDENCY)]),
  tibble(unit = "Planning area", value = mpsz_pa$DEPENDENCY_PA)
) %>%
  ggplot(aes(x = value, fill = unit)) +
  geom_density(alpha = 0.45) +
  labs(title = "Aggregation compresses the distribution",
       subtitle = "Same population, same ratio definition, different spatial units",
       x = "Dependency ratio", y = "Density", fill = NULL) +
  theme_minimal()
```

The planning-area distribution is noticeably narrower. Aggregating averages away the extremes,
so a map drawn at planning-area level would understate how uneven dependency actually is across
Singapore. Neither scale is the correct one. The best I can do is state which scale I used and
note that the result depends on it.

::: {.callout-note appearance="simple"}
## The two components of MAUP

Anselin separates MAUP into two distinct problems:

- **The scale problem.** Results change with the size of the unit, which is what the table above
  measures, moving from subzone up to planning area.
- **The zoning problem.** Results change with the arrangement of the units even when their size
  stays constant. Redrawing the boundaries into equally sized but differently shaped subzones
  would move the answer again.

The standard demonstration is Openshaw's *"A Million Spatial Autocorrelation Coefficients"*,
which recomputed the same statistic on Iowa election data across progressively larger units and
found that the magnitude changed and so did the sign.

Only the scale problem can be tested here, because Singapore's subzones nest inside planning
areas and there is no alternative zoning available. I am noting that rather than implying the
test covered both.
:::

::: {.callout-warning}
## Change of support

This is Anselin's term for combining data recorded on units that do not align. This document has
a mild case of it, since Part 1 uses Master Plan 2014 boundaries and Part 2 uses 2019, which is
what the chapters specify. Subzone definitions changed between the two versions, so the
pre-school counts and the dependency ratios are not measured on quite the same geography.

Because of that, the cross-dataset section below recalculates pre-school counts against the 2019
boundaries instead of reusing the numbers from Part 1.
:::

## The choropleth's built-in bias

A choropleth encodes a rate using area. The eye reads area as quantity, so large sparsely
populated subzones dominate the map while the dense subzones where most people actually live
barely register.

```{r}
#| label: area-bias
mpsz_pop %>%
  st_drop_geometry() %>%
  filter(!is.na(TOTAL)) %>%
  mutate(area_km2 = as.numeric(units::set_units(st_area(mpsz_pop %>% filter(!is.na(TOTAL))), "km^2"))) %>%
  summarise(
    `Share of map area held by the least-populated half of subzones (%)` =
      round(sum(area_km2[TOTAL <= median(TOTAL)]) / sum(area_km2) * 100, 1),
    `Share of population they hold (%)` =
      round(sum(TOTAL[TOTAL <= median(TOTAL)]) / sum(TOTAL) * 100, 1)
  ) %>%
  pivot_longer(everything(), names_to = "measure", values_to = "value") %>%
  kable()
```

The less-populated half of the subzones takes up most of the visual field while holding a small
minority of the population. Anyone forming an impression from this map is effectively weighting
land rather than people.

## Does provision follow need? Joining two datasets

Everything up to this point describes the data. This section is an attempt to explain something,
by asking whether pre-school provision tracks the child population. It brings the pre-school data
from Part 1 together with the demographics from Part 2.

```{r}
#| label: provision-join
preschool_by_sz <- mpsz19 %>%
  mutate(SUBZONE_N = toupper(SUBZONE_N),
         PreSch_Count = lengths(st_intersects(., preschool))) %>%
  st_drop_geometry() %>%
  select(SUBZONE_N, PreSch_Count)

provision <- mpsz_pop %>%
  st_drop_geometry() %>%
  filter(!is.na(TOTAL), YOUNG > 0) %>%
  left_join(preschool_by_sz, by = "SUBZONE_N") %>%
  mutate(children_per_preschool = YOUNG / pmax(PreSch_Count, 1),
         has_none = PreSch_Count == 0)

nrow(provision)
```

```{r}
#| label: provision-plot
ggplot(provision, aes(x = YOUNG, y = PreSch_Count)) +
  geom_point(aes(color = has_none), alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "grey30", linewidth = 0.6) +
  scale_color_manual(values = c("FALSE" = "#4A6FA5", "TRUE" = "#D1495B"),
                     labels = c("has pre-schools", "none"), name = NULL) +
  labs(title = "Pre-school provision broadly tracks child population",
       subtitle = "Each point is a subzone; red points have resident children but no pre-school",
       x = "Residents aged 0-14", y = "Pre-schools in subzone") +
  theme_minimal()
```

```{r}
#| label: provision-corr
ct <- cor.test(provision$YOUNG, provision$PreSch_Count, method = "spearman", exact = FALSE)
sprintf("Spearman rho = %.3f (p = %.3g), n = %d",
        ct$estimate, ct$p.value, nrow(provision))
```

::: {.callout-warning}
## Why the p-value here is not reliable

Spearman's ρ treats the 230 subzones as independent observations, and they are not. Adjacent
subzones have similar child populations for the same underlying reasons, such as housing type,
estate age and planning policy, so the observations are spatially autocorrelated.

Xie et al. describe this as the central statistical issue in the field: *"Geospatial data are
often spatially auto-correlated, which violates the i.i.d. assumption of many data mining and
machine learning techniques."*

What this means in practice is that the effective sample size is smaller than 230, because
neighbouring subzones carry partly redundant information. The correlation coefficient is still
fine as a description. The p-value is not, because it is anti-conservative, meaning too small,
and the significance is overstated by an amount I cannot quantify.

So I am reading ρ = `r round(ct$estimate, 2)` as a strong association and ignoring the p-value
rather than reporting it as if the test were valid. Doing this properly needs a method that
models the spatial dependence, which is what Lessons 4 and 5 introduce, and presumably why they
come before the regression lessons.

Xie et al. add a second point that applies here, which is that a p-value alone is never enough to
rule out a chance pattern and should be paired with an effect size.
:::

```{r}
#| label: provision-gaps
provision %>%
  filter(has_none) %>%
  arrange(desc(YOUNG)) %>%
  select(SUBZONE_N, PLN_AREA_N, YOUNG, TOTAL) %>%
  head(10) %>%
  kable(caption = "Subzones with the most children but no pre-school within their boundary")
```

::: {.callout-important}
## Limits of this finding

The correlation is strong and positive, so provision does broadly follow population. The table
then does something the correlation cannot, which is name specific subzones where children live
and no pre-school falls inside the boundary.

That is a starting point rather than a conclusion, for four reasons:

1. **The denominator problem.** Anselin's car theft example applies here: *"wouldn't there be
   more car thefts where there are more cars?"* A raw count of pre-schools mostly tells you
   where the children are, not where provision is good or bad. I framed this against child
   population instead of using a bare count for that reason, and the earlier density section
   divided by area rather than mapping counts for the same reason.
2. **Edge effects.** A subzone boundary is not a catchment area. A pre-school 50 m outside the
   line still serves those children. Doing this properly would involve buffering, or computing
   distance to the nearest facility, rather than counting inside polygons, which is the
   accessibility modelling covered in Lesson 9.
3. **Capacity is not measured.** Counting facilities treats a 30-place centre and a 300-place
   centre as the same thing.
4. **The demographic data is from Census 2020** while the pre-school locations are current. A
   six-year gap matters in a country that builds new towns as quickly as Singapore does.

So this identifies candidates worth investigating, and the right next step would be a
distance-based accessibility measure rather than a count.

Xie et al. make the point that a false positive is not a neutral error, using the example of a
community *"being incorrectly identified as a cancer cluster"*. Naming a subzone as underserved
is a much smaller version of the same thing, but it still directs attention and possibly
resources, so I have listed these caveats here rather than leaving them out.
:::

::: {.callout-warning}
## Ecological fallacy

Anselin describes this as the classic trap in aggregate spatial work: *"confusing individual
behaviour with aggregate behaviour... the mistake happens at the end, when the conclusions are
written as if they are about individual behaviour, but really they're based on aggregate
magnitudes."*

Everything above is measured on subzones rather than on families, so these statements would not
be supported by this data:

- ✗ "Families in these subzones cannot find pre-school places."
- ✗ "Children here are less likely to attend pre-school."

What the data does support:

- ✓ "These subzones have resident children and no pre-school within their boundary."
- ✓ "Across subzones, pre-school count rises with child population (ρ = `r round(ct$estimate, 2)`)."

The gap between those two lists is the ecological fallacy. As Anselin puts it, an aggregate only
represents individuals well when the individuals are alike. A subzone containing both high and
low income households has an average that describes nobody living in it.
:::

## Limitations

- **Two boundary vintages.** Part 1 uses Master Plan 2014 and Part 2 uses 2019, since that is
  what the chapters specify. Subzone definitions changed between the two, so figures are not
  strictly comparable across the two parts.
- **Census 2020 population against 2026 facility locations**, as discussed above.
- **Point-in-polygon counting ignores catchments**, as discussed above.
- **The dependency ratio assumes 15 to 64 is working age.** With rising retirement ages and high
  tertiary participation this is a convention rather than a fact.
- **`Inf` dependency values were set to `NA`.** Subzones that have residents but no working-age
  residents are dropped from every map. There are only a few, but they are dropped rather than
  shown as zero.

# Reflection

Most of the mechanics here were familiar from Python. Reading files, joining tables and plotting
are the same ideas with different function names. Three things were new to me.

The first is that CRS is a correctness problem rather than a formatting one. The difference
between `st_set_crs()` and `st_transform()` does not really have an equivalent in ordinary data
work, and getting it wrong produces no error at all. The numbers are just wrong.

The second is that classification is a decision you make whether or not you think about it.
Accepting the `qtm()` default is still a choice. Building the sensitivity table was the most
useful thing I did in this exercise, because it turned a vague warning into an actual number.

The third is how much of the work turned out to be checking rather than analysis. The join
diagnostics, the `Inf` handling and the buffer overlap correction are not in the chapter, and
each one was somewhere I could have got a wrong answer that looked reasonable.

If I did this again I would run the join diagnostic before drawing the map instead of after. I
had already produced a choropleth over data I had not verified, and it looked completely fine.

# About this document

Written and executed in Quarto with R `r getRversion()`. All figures are generated at render time
from the data in `data/`, so nothing on this page is pasted in.

I used AI assistance while working out the approach and the extensions in Part 3. The analysis
was run and checked by me.

```{r}
#| label: sessioninfo
#| code-fold: true
sessionInfo()
```
