---
title: "Hands-on Exercise 6"
subtitle: "Geographical Segmentation with Spatially Constrained Clustering Techniques"
author: "An Loc"
date: "28 September 2026"
date-modified: "last-modified"
date-format: "D MMM YYYY"
format:
  html:
    toc: true
    toc-depth: 4
    toc-expand: true
    code-fold: false
    code-copy: true
    fig-width: 9
    fig-height: 7
execute:
  echo: true
  eval: true
  warning: false
  message: false
  freeze: true
---

# Overview

In Exercises 4 and 5, one variable, GDP per capita, was used to test whether counties with similar
values are neighbours. This exercise uses six variables on information and communication technology
(ICT), five of them in the clustering, for a different question: can the 55 townships of Shan State,
Myanmar, be grouped into a small number of regions, each made of townships with similar access to
ICT, and each a single connected piece of territory?

Such a grouping is called regionalisation or geographic segmentation. On slide 3, regions are
defined as internally homogeneous and contiguous, which combines two requirements:

- from cluster analysis, that the townships in a region are similar on the variables;
- from spatial analysis, that a region is one connected piece, so that it can be mapped and managed
  as one area.

**Part 1** is a reproduction of r4gdsa chapter 12: the data, exploratory analysis, conventional
hierarchical clustering, the SKATER and ClustGeo methods of spatially constrained clustering, and
the profiles of the clusters. **Part 2** is a review of the chapter's method, with the methods on
the Lesson 6 slides that the chapter does not cover: REDCAP, the automatic zoning procedure (AZP)
and max-p. Spatially constrained hierarchical clustering (SCHC) is included as well. It is not on
the slides, and it is provided by the `rgeoda` package used for the other three.

::: {.callout-note appearance="simple"}

## Sources

The hands-on exercise is one chapter, assigned for Lesson 6:

- [Chapter 12: Geographical Segmentation with Spatially Constrained Clustering
  Techniques](https://r4gdsa.netlify.app/chap12.html)

The Lesson 6 outline sets two papers as self-reading.

| Reading | What it contributes | Used in |
|--------|--------------|----|
| **Assunção et al. (2006)**, "Efficient Regionalization Techniques for Socio-Economic Geographical Units Using Minimum Spanning Trees" | SKATER: the minimum spanning tree and the conditions for its uniqueness, the pruning of the tree into regions, attributes measured "in standard deviation units", a one-district region in the unrestricted São Paulo result and the minimum-population restriction that removes it, a comparison with AZP, the sum of squared deviations within regions as the quality of a partition, and the debate on combining homogeneity with compactness | Part 2, 3.2, 4.2, 4.5, 4.6, 4.7 |
| **Chavent et al. (2018)**, "ClustGeo: an R package for hierarchical clustering with spatial constraints" | The Ward-like criterion on two dissimilarity matrices, the choice of α, the equivalence of `hclustgeo()` with `hclust(method = "ward.D")` on the transformed matrix, the two kinds of geographic matrix D1, and the proportion of explained pseudo-inertia | Part 2, 3.3, 4.3, 4.6 |

The outline also lists five references. All five are used on this page:

| Reference | What it contributes | Used in |
|--------|--------------|----|
| **Rovan and Sambt (2003)**, "Socio-economic Differences Among Slovenian Municipalities: A Cluster Analysis Approach" | All variables standardised, with the remark that the method makes no distributional assumptions; the number of clusters read from jumps in the Ward merge heights; clusters named from their mean profiles | 4.2, 4.4, 4.8 |
| **Demeter and Brătucu (2013)**, "Statistical Analysis of the EU Countries from a Touristic Point of View" | Cluster membership read at two, three and four clusters from one tree, showing how the hierarchy nests | 4.4 |
| **Brown and Watson (2012)**, "What can a comprehensive plan really tell us about a region?" | The rule of thumb k = √(n/2); cluster means expressed as a percentage of the overall mean; clusters described by variables that were not used to form them | 4.4, 4.8 |
| **Jaya et al. (2019)**, "Clustering with spatial constraints: The case of diarrhea in Bandung city, Indonesia" | ClustGeo with a distance-based and a contiguity-based D1 on the same data, and the argument that the choice of D1 should follow how the phenomenon spreads | 4.6 |
| **de Souza and Taconeli (2022)**, "Spatial and non-spatial clustering algorithms in the analysis of Brazilian educational data" | A Moran's I test on each variable before clustering; one k for every method; ClustGeo with both kinds of D1 and SKATER compared on the same data; clusters cross-tabulated against official regions and compared by F tests; min-max scaling; the NbClust majority vote for the number of clusters | 4.1, 4.2, 4.4, 4.6, 4.8 |

The outline's "All About R" section has links to four web pages, and three of them are used on this
page. The UC Business Analytics tutorial on hierarchical cluster analysis is the source of chapter
sections 12.7.8 to 12.7.10, and is used in sections 3.1.8, 4.2, 4.3 and 4.4. The `spdep` reference
for `skater()` documents the population criterion used in section 4.5, and the ClustGeo introduction
page documents the two kinds of D1 used in section 4.6.

The Lesson 6 slide deck, dated 18 September 2026, lists five methods. The chapter covers two of
them, SKATER and ClustGeo. REDCAP (slides 9 to 12), AZP and max-p (slide 2) are covered in section
4.7.

:::

## Differences from the chapter

| | The chapter | This page | Why | Where |
|-------|---------|----------|--------------------|----|
| **Scale of the clustering variables** | computes min-max and z-score versions in 12.7.3 and 12.7.4, then clusters on the raw rates | Part 1 follows the chapter; Part 2 repeats the analysis on z-scores | On the raw rates, mobile phone and TV account for 81% of the squared distance between townships and computer for 0.6%. Standardising changes the six clusters substantially (adjusted Rand index 0.32) | 4.2 |
| **Ward's method** | picks Ward from `agnes()`, then runs `hclust(method = "ward.D")` | `ward.D2`, which is the method `agnes()` scored and the one `hclustgeo()` runs | `agnes(method = "ward")` and `hclust(method = "ward.D2")` build identical trees. `ward.D` applied to unsquared distances is a different algorithm, and it changes the cluster sizes | 4.3 |
| **Spatial pattern of each variable** | not tested | Moran's I for each of the six rates, as de Souza and Taconeli (2022) do before clustering | Only radio and land line are significantly autocorrelated. TV and mobile phone, which have the most weight in the raw distances, show no significant spatial pattern | 4.1 |
| **Number of clusters** | k = 1 has the largest gap statistic, and 6 is taken as having "the largest gap statistic" after it | 6 kept, with the evidence from five criteria listed | k = 1 has the largest gap. NbClust's majority vote is 2. Ward's merge heights, read as Rovan and Sambt (2003) read them, jump at 2, 4 and 6 | 4.4 |
| **SKATER singletons** | six regions, two of them single townships, not discussed | the singletons identified, and a minimum number of households per region applied, as in Assunção et al. (2006) | Three of six SKATER regions on the standardised variables are single townships: Muse, Tachileik and Taunggyi | 4.5 |
| **Geographic matrix D1 for ClustGeo** | `st_distance()` between polygons | centroid distances and the neighbourhood matrix of Chavent et al. (2018) compared | The distance between two polygons that share a boundary is zero. All 132 neighbouring pairs, and only those pairs, have D1 = 0, so D1 gives every neighbour the same value, zero, whatever the distance between their centres | 4.6 |
| **Contiguity of the result** | reported from the map | counted | ClustGeo applies a soft constraint. The chapter's α = 0.2 solution has six clusters in 18 separate pieces | 4.6, 4.8 |
| **REDCAP, SCHC, AZP, max-p** | not covered | computed with `rgeoda` and compared | Slides 2 and 10 to 12 list REDCAP, AZP and max-p; SCHC is added from `rgeoda` | 4.7 |
| **Minimum spanning tree** | `mstree()` without a seed | `set.seed()` before `mstree()` | `mstree()` starts from a random node. The tree is the same, but the printed order of its edges and the numbering of the SKATER groups change from run to run | 3.2.3 |
| **Colours for six groups** | `rect.hclust(border = 2:5)` and `groups.colors` with five colours | six colours | With fewer colours than groups, the colours are reused and two groups look like one | 3.1.10, 3.2.4 |
| **Cluster profiles** | the chapter profiles the non-spatial ClustGeo clusters (12.10) | the profiles are drawn for the spatially constrained clusters as well | The purpose of the chapter is the regions, and the non-spatial clusters are fragmented | 3.4, 4.8 |

::: {.callout-note collapse="true"}
## Mapping of chapter 12 to this page

| Chapter | This page |
|---|---|
| 12.1 to 12.3 Overview, question, data | 1 Overview, 2.2 The data |
| 12.3.1 Installing and loading R packages | 2.1 Packages |
| 12.4.1 Importing geospatial data | 2.2.1 Township boundaries |
| 12.4.2 Importing aspatial data | 2.2.2 ICT indicators |
| 12.4.3 Deriving new variables | 2.3 Penetration rates |
| 12.5.2.1 Joining geospatial data with aspatial data | 2.4 Joining the ICT penetration rates to the township boundaries |
| 12.5.1 EDA using statistical graphics | 3.1.1 Distributions |
| 12.5.2 EDA using choropleth map | 3.1.2 Choropleth maps |
| 12.6 Correlation analysis | 3.1.3 Correlation between the variables |
| 12.7.1 Extracting clustering variables | 3.1.4 The clustering variables |
| 12.7.2 to 12.7.5 Standardisation | 3.1.5 Standardisation |
| 12.7.6 Computing proximity matrix | 3.1.6 Proximity matrix |
| 12.7.7 Computing hierarchical clustering | 3.1.7 Hierarchical clustering |
| 12.7.8 Selecting the optimal clustering algorithm | 3.1.8 Choice of agglomeration method |
| 12.7.9 Determining optimal clusters | 3.1.9 Number of clusters |
| 12.7.10 Interpreting the dendrograms | 3.1.10 The dendrogram |
| 12.7.11 Visually-driven hierarchical clustering | 3.1.11 Cluster heatmap |
| 12.7.12 Mapping the clusters formed | 3.1.12 Mapping the clusters |
| 12.8.1 Converting into SpatialPolygonsDataFrame | 3.2.1 Neighbour list |
| 12.8.2 Computing neighbour list | 3.2.1 Neighbour list |
| 12.8.3 Calculating edge costs | 3.2.2 Edge costs |
| 12.8.4 Computing minimum spanning tree | 3.2.3 Minimum spanning tree |
| 12.8.5 Computing SKATER clusters | 3.2.4 SKATER regions |
| 12.8.6 Visualising the clusters | 3.2.5 SKATER map |
| 12.9.1 to 12.9.2 ClustGeo, Ward-like clustering | 3.3.1 Ward-like clustering without a constraint |
| 12.9.3 Spatially constrained hierarchical clustering | 3.3.2 Geographic distances and α, 3.3.3 ClustGeo regions |
| 12.10 Visual interpretation of clusters | 3.4 Cluster profiles |

Chapter 12.8.1 converts `shan_sf` to an sp object because older versions of `poly2nb()` required
one. The current `spdep` accepts sf objects directly, so the conversion is not needed and is not
made. The chapter's `write_rds()` and `read_rds()` round trips save intermediate objects to disk.
This page keeps the objects in memory.
:::

# Getting started

## Packages

`rgeoda` provides the four methods in section 4.7, and `mclust` provides the adjusted Rand index
used in Part 2 to compare two sets of clusters.

```{r}
pacman::p_load(spdep, tmap, sf, ClustGeo, rgeoda,
               ggpubr, cluster, factoextra, NbClust,
               heatmaply, corrplot, psych, GGally, mclust,
               knitr, tidyverse)
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses one function:

```r
pacman::p_load(...)                                       # pacman
```

- `p_load()` installs any package that is missing and then attaches all of them, so one call
  replaces a series of `install.packages()` and `library()` calls.

Two of the packages define a function with the same name. `spdep` and `rgeoda` both have a
`skater()`, and whichever package is attached later hides the other. This page therefore always
writes `spdep::skater()` or `rgeoda::skater()` in full.

| Package | Used for |
|-------|-------------|
| `sf`, `spdep` | reading the boundaries, the neighbour list, the minimum spanning tree and SKATER |
| `tidyverse` | reading and reshaping the table, and `ggplot2` charts |
| `tmap` | choropleth maps |
| `corrplot`, `psych`, `ggpubr`, `GGally` | correlation plot, descriptive statistics, arranged charts, parallel coordinates |
| `cluster`, `factoextra`, `NbClust` | agglomerative coefficient, gap statistic, cluster-number indices |
| `heatmaply` | min-max standardisation and the cluster heatmap |
| `ClustGeo` | Ward-like clustering with a geographic constraint |
| `rgeoda` | REDCAP, SCHC, AZP and max-p (Part 2) |
| `mclust`, `knitr` | adjusted Rand index; tables (Part 2) |
:::

## The data

Both files come from the Myanmar Information Management Unit (MIMU), according to section 12.3, and
are distributed with the exercise on eLearn.

| Layer | Format | Features | Native CRS | Contents |
|----|---|---|----|------|
| `myanmar_township_boundaries` | ESRI shapefile | 330 polygons | WGS 84, EPSG:4326 | Township boundaries of Myanmar |
| `Shan-ICT` | CSV | 55 rows, 11 fields | not spatial | Households owning each ICT item, 2014 Population and Housing Census |

### Township boundaries

```{r}
#| results: hide
shan_sf <- st_read(dsn = "data/geospatial",
                   layer = "myanmar_township_boundaries") |>
  filter(ST %in% c("Shan (East)", "Shan (North)", "Shan (South)")) |>
  select(c(2:7))
```

```{r}
shan_sf
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses three functions:

```r
st_read(dsn, layer)                                       # sf
filter(.data, ...)                                        # dplyr
select(.data, ...)                                        # dplyr
```

- `st_read()` reads the shapefile into an sf data frame with one row per township. `dsn` is the
  folder and `layer` the file name without its extension. It prints a summary that begins with the
  full path of the data folder on the computer that renders the page, and the chunk option
  `results: hide` keeps that summary off the page.
- `filter()` keeps the rows for which a condition is true. The condition here is
  `ST %in% c("Shan (East)", "Shan (North)", "Shan (South)")`, dissected below.
- `select(c(2:7))` keeps columns 2 to 7 by position, also dissected below.

**What `ST`, `DT` and `TS` are**

Myanmar is divided into states and regions, each state or region into districts, and each district
into townships. The file has one row per township, 330 rows, and three of its columns record the
state, the district and the township name of each row:

<div style="border: 2px solid #6c757d; border-radius: 6px; padding: 0.6em 0.8em; max-width: 30em; margin: 1em auto;">
<b><code>ST</code></b>: state or region, for example <code>Shan (North)</code>
<div style="border: 2px solid #6c757d; border-radius: 6px; padding: 0.6em 0.8em; margin-top: 0.5em;">
<b><code>DT</code></b>: district within the state, for example <code>Mongmit</code>
<div style="border: 2px solid #6c757d; border-radius: 6px; padding: 0.6em 0.8em; margin-top: 0.5em; background: #f8f9fa;">
<b><code>TS</code></b>: township within the district, for example <code>Mongmit</code>. One township is one row of the file.
</div></div></div>

A district and a township can share a name, as Mongmit does. The file also splits Shan State into
three parts, `Shan (North)`, `Shan (South)` and `Shan (East)`, and records each part in `ST` as if
it were a separate state. That is why the filter has to name all three.

**The `filter()` line, step by step**

The filter is applied below to three of the 330 rows.

<div style="font-size: 1em; margin: 1em 0;">
<p style="margin-bottom: 0.3em;"><b>Step 1. Three rows of the file</b></p>
<table class="table table-sm" style="width: auto; margin: 0 auto;">
<tr><th><code>ST</code></th><th><code>DT</code></th><th><code>TS</code></th></tr>
<tr><td>Shan (North)</td><td>Mongmit</td><td>Mongmit</td></tr>
<tr><td>Shan (South)</td><td>Taunggyi</td><td>Pindaya</td></tr>
<tr><td>Bago (East)</td><td>Bago</td><td>Waw</td></tr>
</table>
<div style="text-align: center; margin: 0.5em 0; color: #343a40;"><span style="font-size: 1.8em;">&darr;</span><br>The condition tests only the <code>ST</code> column.</div>
<p style="margin-bottom: 0.3em;"><b>Step 2. The values tested</b></p>
<table class="table table-sm" style="width: auto; margin: 0 auto;">
<tr><th style="background: rgba(13, 110, 253, 0.18);"><code>ST</code></th><th><code>DT</code></th><th><code>TS</code></th></tr>
<tr><td style="background: rgba(255, 193, 7, 0.45);">Shan (North)</td><td>Mongmit</td><td>Mongmit</td></tr>
<tr><td style="background: rgba(255, 193, 7, 0.45);">Shan (South)</td><td>Taunggyi</td><td>Pindaya</td></tr>
<tr><td style="background: rgba(220, 53, 69, 0.18);">Bago (East)</td><td>Bago</td><td>Waw</td></tr>
</table>
<div style="text-align: center; margin: 0.5em 0; color: #343a40;"><span style="font-size: 1.8em;">&darr;</span><br>Each value is looked up in the list of three names inside <code>c()</code>.</div>
<p style="margin-bottom: 0.3em;"><b>Step 3. The condition in the code</b></p>
<pre style="width: auto; margin: 0 auto; padding: 0.6em;"><code>filter(<span style="border-bottom: 2px solid #0d6efd;">ST %in% c(&quot;Shan (East)&quot;, <span style="background: rgba(255, 193, 7, 0.45);">&quot;Shan (North)&quot;</span>, <span style="background: rgba(255, 193, 7, 0.45);">&quot;Shan (South)&quot;</span>)</span>)</code></pre>
<p style="margin-top: 0.4em;">The underlined part is the condition. <code>Shan (North)</code> and <code>Shan (South)</code> are
in the list, highlighted in yellow as in step 2, so the condition is true for Mongmit and Pindaya.
<code>Bago (East)</code> is not in the list, so the condition is false for Waw.</p>
<div style="text-align: center; margin: 0.5em 0; color: #343a40;"><span style="font-size: 1.8em;">&darr;</span><br><code>filter()</code> keeps the rows where the condition is true.</div>
<p style="margin-bottom: 0.3em;"><b>Step 4. The outcome</b></p>
<table class="table table-sm" style="width: auto; margin: 0 auto;">
<tr><th><code>ST</code></th><th><code>DT</code></th><th><code>TS</code></th><th></th></tr>
<tr style="background: rgba(25, 135, 84, 0.15);"><td>Shan (North)</td><td>Mongmit</td><td>Mongmit</td><td>kept</td></tr>
<tr style="background: rgba(25, 135, 84, 0.15);"><td>Shan (South)</td><td>Taunggyi</td><td>Pindaya</td><td>kept</td></tr>
<tr style="background: rgba(220, 53, 69, 0.12); text-decoration: line-through; color: #6c757d;"><td>Bago (East)</td><td>Bago</td><td>Waw</td><td>dropped</td></tr>
</table>
</div>

Applied to all 330 rows, the same steps keep the 55 townships of Shan State.

```{r}
#| echo: false
#| results: hide
mmr_all <- st_read(dsn = "data/geospatial", layer = "myanmar_township_boundaries")
mmr_names <- names(mmr_all)
shan_parts <- c("Shan (East)", "Shan (North)", "Shan (South)")
```

The map shows the three parts among the 18 values of `ST`:

```{r}
#| echo: false
#| fig-height: 4.6
mmr_all <- mmr_all |> mutate(part = if_else(ST %in% shan_parts, ST, "other states"))
part_cols <- c(`Shan (East)` = "#1b9e77", `Shan (North)` = "#d95f02", `Shan (South)` = "#7570b3",
               `other states` = "grey85")
map_parts <- ggplot(mmr_all) +
  geom_sf(aes(fill = part), colour = "white", linewidth = 0.05) +
  scale_fill_manual(values = part_cols, breaks = names(part_cols)) +
  labs(fill = "value of ST") + theme_void()
bar_parts <- mmr_all |>
  st_drop_geometry() |>
  count(ST, part) |>
  ggplot(aes(n, reorder(ST, n), fill = part)) +
  geom_col() +
  scale_fill_manual(values = part_cols, guide = "none") +
  labs(x = "townships", y = NULL)
patchwork::wrap_plots(map_parts, bar_parts, widths = c(1.1, 1))
```

The same condition can be written without `%in%`, as one comparison per name joined by `|`, which
means "or":

```r
filter(ST == "Shan (East)" | ST == "Shan (North)" | ST == "Shan (South)")
```

`%in%` compares whole values, so each name must match exactly, including the space and the brackets.
`ST %in% "Shan"` matches `r sum(mmr_all$ST %in% "Shan")` townships, because no value of `ST` is the
word "Shan" on its own.

**The `select()` line**

`select(c(2:7))` keeps columns by their position in the file, so the position of each column has to
be known first. RStudio's data viewer numbers the rows but not the columns, so the positions are
easiest to read in the Console.

`names()` prints the column names in order. The number in square brackets at the start of each
printed line is the position of the first name on that line:

```r
st_read(dsn = "data/geospatial", layer = "myanmar_township_boundaries") |> names()
```

```{r}
#| echo: false
mmr_names
```

`ST` is the second name on the line that starts at `[1]`, so it is column 2. `TS_PCODE` is the
second name on the line that starts at `[6]`, so it is column 7.

For a file with many columns, putting the names into a data frame prints one name per line, with its
position as the row number:

```r
col_names <- st_read(dsn = "data/geospatial", layer = "myanmar_township_boundaries") |> names()
data.frame(names = col_names)
```

```{r}
#| echo: false
data.frame(names = mmr_names)
```

The file has 15 columns, and positions 2 to 7 are the shaded ones:

```{r}
#| echo: false
cols <- mmr_names[1:15]
kept <- seq_along(cols) %in% 2:7
cells <- ifelse(kept, sprintf('<td style="background: rgba(255, 193, 7, 0.30)">%s</td>', cols),
                sprintf("<td>%s</td>", cols))
nums <- ifelse(kept, sprintf('<th style="background: rgba(255, 193, 7, 0.30)">%d</th>', seq_along(cols)),
               sprintf("<th>%d</th>", seq_along(cols)))
htmltools::HTML(paste0('<table class="table table-sm" style="font-size: 0.8em;"><tr>',
                       paste(nums, collapse = ""), "</tr><tr>", paste(cells, collapse = ""),
                       "</tr></table>"))
```

The six kept columns are the state (`ST`), district (`DT`) and township (`TS`) names, each with its
code (`_PCODE`). Column 15, `geometry`, is not listed but is kept, because `select()` on an sf
object always keeps the geometry.

The same six columns can be selected by name, which does not depend on the column order and names
the kept columns in the code:

```r
select(ST, ST_PCODE, DT, DT_PCODE, TS, TS_PCODE)
```
:::

The layer is left in longitude and latitude, as in the chapter. Contiguity is defined by shared
boundaries and does not depend on the projection.

### ICT indicators

```{r}
ict <- read_csv("data/aspatial/Shan-ICT.csv")
summary(ict)
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses two functions:

```r
read_csv(file)                                            # readr
summary(object)                                           # base R
```

- `read_csv()` reads the CSV into a tibble, a kind of data frame, and guesses the type of each
  column from its values. It reports those guesses in a message, which the page setting
  `message: false` keeps off the page.
- `summary()` prints the minimum, quartiles, mean and maximum of each numeric column.
:::

::: callout-note
## What the ICT table contains

ICT stands for information and communication technology: the devices and services a household uses
to receive information and to communicate. The chapter also calls the variables "InfoComm", a
shorter name for the same thing.

The table is an extract of the 2014 Myanmar Population and Housing Census for the 55 townships of
Shan State. Each row is one township, and the eleven columns are in three groups. Three rows of the
file:

<table class="table table-sm" style="font-size: 0.85em;">
<tr>
<th colspan="4" style="background: rgba(13, 110, 253, 0.15); text-align: center;">Which township</th>
<th style="background: rgba(108, 117, 125, 0.18); text-align: center;">How many households</th>
<th colspan="6" style="background: rgba(25, 135, 84, 0.15); text-align: center;">How many of those households own each item</th>
</tr>
<tr>
<th>District Pcode</th><th>District Name</th><th>Township Pcode</th><th>Township Name</th>
<th>Total households</th>
<th>Radio</th><th>Television</th><th>Land line phone</th><th>Mobile phone</th><th>Computer</th><th>Internet at home</th>
</tr>
<tr><td>MMR014D001</td><td>Taunggyi</td><td>MMR014001</td><td>Taunggyi</td><td>82,604</td><td>30,176</td><td>62,388</td><td>6,736</td><td>48,461</td><td>6,705</td><td>9,746</td></tr>
<tr><td>MMR014D001</td><td>Taunggyi</td><td>MMR014002</td><td>Nyaungshwe</td><td>42,634</td><td>13,801</td><td>23,640</td><td>810</td><td>14,973</td><td>669</td><td>1,604</td></tr>
<tr><td>MMR015D007</td><td>Matman</td><td>MMR015024</td><td>Matman</td><td>3,318</td><td>810</td><td>728</td><td>28</td><td>150</td><td>20</td><td>15</td></tr>
</table>

The first row reads: Taunggyi township, in Taunggyi district, has 82,604 households, and 30,176 of
them own a radio, 62,388 a television, and so on across the six items. Matman, the smallest
township, has 3,318 households.

The `Township Pcode` column holds the same codes as `TS_PCODE` in the boundary file, and section 2.4
joins the two tables on it. Section 2.3 turns the six counts into rates, because a count of radios
in Taunggyi and one in Matman are counts out of very different numbers of households.
:::

## Penetration rates

A township with more households has more households with a radio, whatever the level of access. In
section 12.4.3, each count is divided by the number of households and multiplied by 1,000, which
gives the number of households per thousand that own the item.

```{r}
ict_derived <- ict |>
  mutate(RADIO_PR = Radio / `Total households` * 1000,
         TV_PR = Television / `Total households` * 1000,
         LLPHONE_PR = `Land line phone` / `Total households` * 1000,
         MPHONE_PR = `Mobile phone` / `Total households` * 1000,
         COMPUTER_PR = Computer / `Total households` * 1000,
         INTERNET_PR = `Internet at home` / `Total households` * 1000) |>
  rename(DT_PCODE = `District Pcode`, DT = `District Name`,
         TS_PCODE = `Township Pcode`, TS = `Township Name`,
         TT_HOUSEHOLDS = `Total households`,
         RADIO = Radio, TV = Television,
         LLPHONE = `Land line phone`, MPHONE = `Mobile phone`,
         COMPUTER = Computer, INTERNET = `Internet at home`)

summary(ict_derived[, 12:17])
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses two functions:

```r
mutate(.data, ...)                                        # dplyr
rename(.data, new_name = old_name, ...)                   # dplyr
```

- `mutate()` adds new columns computed from existing ones. The chapter writes a separate `mutate()`
  call for each of the six rates. A single `mutate()` call with six arguments gives the same result.
- A column name that contains a space has to be written between backticks, as in ``
  `Total households` ``. That is the reason for the `rename()` that follows.
- `rename()` gives the columns short names without spaces, written `new = old`.
- `ict_derived[, 12:17]` uses R's square brackets, which pick part of a table in the form
  `table[rows, columns]`. Nothing before the comma means every row, and `12:17` after the comma
  means columns 12 to 17. `mutate()` adds the six rates after the eleven original columns, so they
  are columns 12 to 17:

```{r}
#| echo: false
cols17 <- names(ict_derived)
kept17 <- seq_along(cols17) %in% 12:17
cells17 <- ifelse(kept17, sprintf('<td style="background: rgba(255, 193, 7, 0.30)">%s</td>', cols17),
                  sprintf("<td>%s</td>", cols17))
nums17 <- ifelse(kept17, sprintf('<th style="background: rgba(255, 193, 7, 0.30)">%d</th>', seq_along(cols17)),
                 sprintf("<th>%d</th>", seq_along(cols17)))
htmltools::HTML(paste0('<div style="overflow-x: auto;"><table class="table table-sm" style="font-size: 0.75em;"><tr>',
                       paste(nums17, collapse = ""), "</tr><tr>", paste(cells17, collapse = ""),
                       "</tr></table></div>"))
```

  The same brackets are used inside `cor()` in section 3.1.3.

The rates are per thousand households, so a value of 500 for `TV_PR` means that half of the
households own a television. The chapter calls them penetration rates.
:::

::: callout-note
## Why the counts are turned into rates

A township with more households has more households owning a radio, whatever the level of access.
Dividing each count by the township's number of households makes the townships comparable, and
multiplying by 1,000 expresses the result per thousand households:

$$\text{penetration rate} = \frac{\text{households owning the item}}{\text{total households}} \times 1000$$
:::

The rest of the exercise is built on these penetration rates. Their distributions, maps and
correlations are examined in Part 1, five of them are used as the clustering variables, and every
set of clusters and regions on this page is formed from those five.

## Joining the ICT penetration rates to the township boundaries

As in section 12.5.2.1, the penetration rates are joined to the township polygons on the township
code.

```{r}
shan_sf <- left_join(shan_sf, ict_derived, by = c("TS_PCODE" = "TS_PCODE"))
names(shan_sf)
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses one function:

```r
left_join(x, y, by = NULL)                                # dplyr
```

- `left_join()` keeps every row of `x`, the township polygons, and adds the columns of `y` from the
  row with the same township code. `by = c("TS_PCODE" = "TS_PCODE")` names the key in each table.
- The two tables have three other column names in common: `DT` (district name), `DT_PCODE` (district
  code) and `TS` (township name). `left_join()` merges only the key, so both copies of each of the
  three are kept, with a suffix for the table each copy came from: `.x` for `x` (`shan_sf`) and `.y`
  for `y` (`ict_derived`).

**The join, step by step**

The join is shown below for two townships, Mongmit and Pindaya.

<div style="font-size: 1em; margin: 1em 0;">
<p style="margin-bottom: 0.3em;"><b>Step 1. The two tables</b></p>
<p style="margin: 0.2em 0;"><code>x</code> = <code>shan_sf</code>, rows 1 and 2 (the geometry column is not shown):</p>
<div style="overflow-x: auto;">
<table class="table table-sm" style="width: auto; margin: 0 auto; font-size: 0.85em;">
<tr><th style="color: #6c757d;">row</th><th><code>ST</code></th><th><code>ST_PCODE</code></th><th style="background: rgba(255, 193, 7, 0.45);"><code>DT</code></th><th style="background: rgba(255, 193, 7, 0.45);"><code>DT_PCODE</code></th><th style="background: rgba(255, 193, 7, 0.45);"><code>TS</code></th><th style="background: rgba(13, 110, 253, 0.18);"><code>TS_PCODE</code></th></tr>
<tr><td style="color: #6c757d;">1</td><td>Shan (North)</td><td>MMR015</td><td>Mongmit</td><td>MMR015D008</td><td>Mongmit</td><td style="background: rgba(13, 110, 253, 0.18);">MMR015017</td></tr>
<tr><td style="color: #6c757d;">2</td><td>Shan (South)</td><td>MMR014</td><td>Taunggyi</td><td>MMR014D001</td><td>Pindaya</td><td style="background: rgba(13, 110, 253, 0.18);">MMR014006</td></tr>
</table>
</div>
<p style="margin: 0.6em 0 0.2em;"><code>y</code> = <code>ict_derived</code>, rows 6 and 34:</p>
<div style="overflow-x: auto;">
<table class="table table-sm" style="width: auto; margin: 0 auto; font-size: 0.85em;">
<tr><th style="color: #6c757d;">row</th><th style="background: rgba(255, 193, 7, 0.45);"><code>DT_PCODE</code></th><th style="background: rgba(255, 193, 7, 0.45);"><code>DT</code></th><th style="background: rgba(13, 110, 253, 0.18);"><code>TS_PCODE</code></th><th style="background: rgba(255, 193, 7, 0.45);"><code>TS</code></th><th><code>TT_HOUSEHOLDS</code></th><th>…</th><th><code>RADIO_PR</code></th><th>…</th></tr>
<tr><td style="color: #6c757d;">6</td><td>MMR014D001</td><td>Taunggyi</td><td style="background: rgba(13, 110, 253, 0.18);">MMR014006</td><td>Pindaya</td><td>17,544</td><td>…</td><td>417.5</td><td>…</td></tr>
<tr><td style="color: #6c757d;">34</td><td>MMR015D003</td><td>Kyaukme</td><td style="background: rgba(13, 110, 253, 0.18);">MMR015017</td><td>Mongmit</td><td>13,652</td><td>…</td><td>286.2</td><td>…</td></tr>
</table>
</div>
<p style="margin-top: 0.4em;">The key, <code>TS_PCODE</code>, is shaded blue. The three names that are in both tables are
shaded yellow.</p>
<div style="text-align: center; margin: 0.5em 0; color: #343a40;"><span style="font-size: 1.8em;">&darr;</span><br><code>by = c(&quot;TS_PCODE&quot; = &quot;TS_PCODE&quot;)</code> pairs each row of <code>x</code> with the row of <code>y</code> that has the same code.</div>
<p style="margin-bottom: 0.3em;"><b>Step 2. The pairs</b></p>
<table class="table table-sm" style="width: auto; margin: 0 auto; font-size: 0.85em;">
<tr><th style="background: rgba(13, 110, 253, 0.18);"><code>TS_PCODE</code></th><th>row of <code>x</code></th><th>row of <code>y</code></th></tr>
<tr><td style="background: rgba(13, 110, 253, 0.18);">MMR015017</td><td>1 (Mongmit)</td><td>34 (Mongmit)</td></tr>
<tr><td style="background: rgba(13, 110, 253, 0.18);">MMR014006</td><td>2 (Pindaya)</td><td>6 (Pindaya)</td></tr>
</table>
<p style="margin-top: 0.4em;">The pairing uses the code only. Mongmit is row 1 of one table and row 34 of the other.</p>
<div style="text-align: center; margin: 0.5em 0; color: #343a40;"><span style="font-size: 1.8em;">&darr;</span><br>Each pair of rows becomes one longer row: the Mongmit row of <code>ict_derived</code> is attached to the right of the Mongmit row of <code>shan_sf</code>.<br><code>TS_PCODE</code> has the same value in both rows, so it is kept once. <code>DT</code>, <code>DT_PCODE</code> and <code>TS</code> are kept twice, with <code>.x</code> or <code>.y</code> added to the name.</div>
<p style="margin-bottom: 0.3em;"><b>Step 3. The joined rows</b></p>
<div style="overflow-x: auto;">
<table class="table table-sm" style="width: auto; margin: 0 auto; font-size: 0.8em;">
<tr><th><code>ST</code></th><th><code>ST_PCODE</code></th><th style="background: rgba(255, 193, 7, 0.45);"><code>DT.x</code></th><th style="background: rgba(255, 193, 7, 0.45);"><code>DT_PCODE.x</code></th><th style="background: rgba(255, 193, 7, 0.45);"><code>TS.x</code></th><th style="background: rgba(13, 110, 253, 0.18);"><code>TS_PCODE</code></th><th style="background: rgba(25, 135, 84, 0.2);"><code>DT_PCODE.y</code></th><th style="background: rgba(25, 135, 84, 0.2);"><code>DT.y</code></th><th style="background: rgba(25, 135, 84, 0.2);"><code>TS.y</code></th><th><code>TT_HOUSEHOLDS</code></th><th>…</th><th><code>RADIO_PR</code></th><th>…</th></tr>
<tr><td>Shan (North)</td><td>MMR015</td><td style="background: rgba(220, 53, 69, 0.18);">Mongmit</td><td style="background: rgba(220, 53, 69, 0.18);">MMR015D008</td><td>Mongmit</td><td style="background: rgba(13, 110, 253, 0.18);">MMR015017</td><td style="background: rgba(220, 53, 69, 0.18);">MMR015D003</td><td style="background: rgba(220, 53, 69, 0.18);">Kyaukme</td><td>Mongmit</td><td>13,652</td><td>…</td><td>286.2</td><td>…</td></tr>
<tr><td>Shan (South)</td><td>MMR014</td><td>Taunggyi</td><td>MMR014D001</td><td>Pindaya</td><td style="background: rgba(13, 110, 253, 0.18);">MMR014006</td><td>MMR014D001</td><td>Taunggyi</td><td>Pindaya</td><td>17,544</td><td>…</td><td>417.5</td><td>…</td></tr>
</table>
</div>
<p style="margin-top: 0.4em;">Headers shaded yellow hold the copy from <code>shan_sf</code>, and green the copy from
<code>ict_derived</code>. The cells shaded red are where the two copies differ.</p>
</div>

`shan_sf` no longer has a column called `TS`, so the code in section 3.1.4 selects `"TS.x"`. The
join adds 16 columns after the 6 columns of `shan_sf`, so the five clustering rates, `RADIO_PR` to
`COMPUTER_PR`, are columns 17 to 21 in the list printed above, the positions used in section 3.4.
:::

The township name is the same in `TS.x` and `TS.y` for all 55 townships. The district is different
for six townships, because Myanmar reorganised some districts of Shan State in 2014 and 2015: the
census table has the districts at the time of the 2014 census, and the boundary file has the
reorganised ones. The township codes are the same in both files, so the join is not affected.

The analysis uses one township name and no district column, so there is no analytical reason to keep
both copies. The chapter's join is kept here so that the chapter's `TS.x` and `17:21` work
unchanged. Dropping the three duplicates first, with
`left_join(shan_sf, select(ict_derived, -DT_PCODE, -DT, -TS), by = "TS_PCODE")`, would leave the
name as `TS` and move the rates to columns 14 to 18.

The joined `shan_sf` is the table used for the rest of the exercise:

| Columns | Used for | Sections |
|------|---------|-----|
| `geometry` | maps; the neighbour list for SKATER, Moran's I and `rgeoda` | throughout; 3.2.1, 4.1, 4.7 |
| `TS.x` | township labels in the clustering table and in printed results | 3.1.4, 4.5, 4.8 |
| the six rates | maps, clustering variables, Moran's I, cluster profiles | 3.1, 3.4, 4.1, 4.8 |
| `TT_HOUSEHOLDS` | a map of households; a minimum number of households per region; region profiles | 3.1.2, 4.5, 4.7, 4.8 |
| `ST` | comparing the regions with the three parts of the state | 4.8 |
| `DT.x`, `DT_PCODE.x`, `DT.y`, `DT_PCODE.y`, `TS.y` | not used | |

# Part 1: Chapter 12

This part follows r4gdsa [Chapter 12](https://r4gdsa.netlify.app/chap12.html), sections 12.5 to
12.10. In section 3.1 the townships are clustered on their ICT rates alone. In sections 3.2 and 3.3,
SKATER and ClustGeo add the location of the townships to the clustering, and section 3.4 has the
profiles of the clusters.

## Hierarchical cluster analysis

This section follows sections 12.5 to 12.7 of the chapter. The clustering in it uses five of the ICT
rates, without the locations of the townships.

### Distributions

The distributions show whether a variable is skewed or has outlying townships. Both affect how far
apart townships appear in the clustering later. In section 12.5.1, the count of households with a
radio is compared with the penetration rate, first with histograms and then with boxplots.

```{r}
#| fig-height: 3
ggplot(data = ict_derived,
       aes(x = `RADIO`)) +
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "light blue")
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses two functions:

```r
ggplot(data, aes(x))                                      # ggplot2
geom_histogram(bins, color, fill)                         # ggplot2
```

- `ggplot()` names the data, and `aes(x = RADIO)` the column drawn along the horizontal axis. The
  `+` adds a layer to the chart.
- `geom_histogram(bins = 20)` divides the range of `RADIO` into 20 equal bins and draws a bar for
  the number of townships in each. `color` sets the outline of the bars and `fill` their inside.
:::

```{r}
#| fig-height: 1.8
ggplot(data = ict_derived,
       aes(x = `RADIO`)) +
  geom_boxplot(color = "black",
               fill = "light blue")
```

::: {.callout-note collapse="false" title="How the code works"}
- `geom_boxplot()` draws the median as the line inside the box, the middle half of the townships as
  the box, and as separate points any township more than 1.5 times the height of the box beyond it.
  Those points are the outliers.
:::

The same two charts for the penetration rate:

```{r}
#| fig-height: 3
ggplot(data = ict_derived,
       aes(x = `RADIO_PR`)) +
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "light blue")
```

```{r}
#| fig-height: 1.8
ggplot(data = ict_derived,
       aes(x = `RADIO_PR`)) +
  geom_boxplot(color = "black",
               fill = "light blue")
```

The count is strongly right-skewed, and the boxplot marks three townships as outliers. The largest
counts belong to the townships with the most households. The rate is much closer to symmetric, with
one township, Ywangan, above 480 radios per thousand households. The chapter's question is what the
two charts reveal: the skew and the outliers in the count come mostly from township size, and
dividing by the number of households removes most of both.

The chapter then draws the six penetration rates side by side, one histogram each:

```{r}
radio <- ggplot(data = ict_derived,
                aes(x = `RADIO_PR`)) +
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "light blue")

tv <- ggplot(data = ict_derived,
             aes(x = `TV_PR`)) +
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "light blue")

llphone <- ggplot(data = ict_derived,
                  aes(x = `LLPHONE_PR`)) +
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "light blue")

mphone <- ggplot(data = ict_derived,
                 aes(x = `MPHONE_PR`)) +
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "light blue")

computer <- ggplot(data = ict_derived,
                   aes(x = `COMPUTER_PR`)) +
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "light blue")

internet <- ggplot(data = ict_derived,
                   aes(x = `INTERNET_PR`)) +
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "light blue")
```

```{r}
#| fig-height: 5
ggarrange(radio, tv, llphone, mphone, computer, internet,
          ncol = 3,
          nrow = 2)
```

::: {.callout-note collapse="false" title="How the code works"}
The blocks use one new function:

```r
ggarrange(..., ncol, nrow)                                # ggpubr
```

- The first block stores each histogram under a name, `radio`, `tv` and so on, without drawing it.
- `ggarrange()` draws the six stored charts together in a grid of three columns and two rows.
:::

Radio, TV and mobile phone are roughly symmetric. Land line, computer and internet are right-skewed:
most townships have few households with them, and a small number have many.

Three features of the charts are used later on this page:

| Feature of the charts | Used in |
|---------|-----------|
| The count of radios is right-skewed and the rate is close to symmetric, because larger townships have more households with every item | 2.3, the reason for using rates |
| The ranges of the rates differ widely: `r round(min(ict_derived$TV_PR))` to `r round(max(ict_derived$TV_PR))` for TV, `r round(min(ict_derived$COMPUTER_PR))` to `r round(max(ict_derived$COMPUTER_PR))` for computer | 3.1.5 and 4.2: on the raw rates, TV and mobile phone make up 81% of the squared distance between townships, and computer 0.6% |
| Ywangan has the highest radio rate, and computer, land line and internet have long right tails | 3.2.4: Ywangan is one of the two single-township SKATER regions on the raw rates. 4.5: Tachileik, Muse and Taunggyi, the three highest computer rates, are the three single-township regions on z-scores |

### Choropleth maps

Maps show where in the state the high and low values of a rate are, which the histograms of section
3.1.1 cannot show. The first map in section 12.5.2.2 is a quick map of the radio penetration rate.

```{r}
#| fig-height: 4.5
qtm(shan_sf, "RADIO_PR")
```

::: {.callout-note collapse="false" title="How the code works"}
```r
qtm(shp, fill)                                            # tmap
```

- `qtm()` (quick thematic map) draws a choropleth map in one call, with default classes, colours and
  legend. The second argument names the column that sets the colour of each township.
:::

It then maps the number of households beside the number of households with a radio.

```{r}
#| fig-height: 4.2
TT_HOUSEHOLDS.map <- tm_shape(shan_sf) +
  tm_fill(fill = "TT_HOUSEHOLDS",
          fill.scale = tm_scale_intervals(n = 5, style = "jenks"),
          fill.legend = tm_legend(title = "Total households")) +
  tm_borders(col_alpha = 0.5)

RADIO.map <- tm_shape(shan_sf) +
  tm_fill(fill = "RADIO",
          fill.scale = tm_scale_intervals(n = 5, style = "jenks"),
          fill.legend = tm_legend(title = "Number Radio")) +
  tm_borders(col_alpha = 0.5)

tmap_arrange(TT_HOUSEHOLDS.map, RADIO.map,
             asp = NA, ncol = 2)
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses six functions:

```r
tm_shape(shp)                                             # tmap
tm_fill(fill, fill.scale, fill.legend)                    # tmap
tm_scale_intervals(n, style)                              # tmap
tm_legend(title)                                          # tmap
tm_borders(col_alpha)                                     # tmap
tmap_arrange(..., asp, ncol)                              # tmap
```

- `tm_shape()` names the sf object that the layers after it draw, and each `+` adds a layer.
- `tm_fill()` colours each township by the column named in `fill`, without drawing its outline.
  `tm_borders()` draws the outlines as a separate layer, at half opacity.
- `tm_scale_intervals(n = 5, style = "jenks")` divides the values into five classes by the Jenks
  natural breaks method, which places the class boundaries where the gaps between the sorted values
  are largest.
- `tmap_arrange()` places the two stored maps side by side. `asp = NA` lets each map keep the shape
  of Shan State.

The chapter is written for tmap version 3, and the installed version is 4, which renames some
arguments. The chapter's code and the version used here:

| Chapter (tmap 3) | This page (tmap 4) |
|---|---|
| `tm_fill(col = "TT_HOUSEHOLDS", n = 5, style = "jenks", title = "Total households")` | `tm_fill(fill = "TT_HOUSEHOLDS", fill.scale = tm_scale_intervals(n = 5, style = "jenks"), fill.legend = tm_legend(title = "Total households"))` |
| `tm_borders(alpha = 0.5)` | `tm_borders(col_alpha = 0.5)` |

tmap 4 still accepts the chapter's version 3 code, but it translates `tm_borders(alpha = 0.5)` into
a half-transparent fill, which pales the whole map. `col_alpha` applies the transparency to the
borders, which is what the chapter intends.
:::

The chapter then draws the number of households beside the radio penetration rate, as one map with
two panels.

```{r}
#| fig-height: 4.6
tm_shape(shan_sf) +
  tm_polygons(fill = c("TT_HOUSEHOLDS", "RADIO_PR"),
              fill.scale = tm_scale_intervals(style = "jenks"),
              fill.free = TRUE) +
  tm_facets(ncol = 2)
```

::: {.callout-note collapse="false" title="How the code works"}
- Given two column names in `fill`, `tm_polygons()` draws one panel for each, and
  `tm_facets(ncol = 2)` places the panels side by side. `fill.free = TRUE` gives each panel its own
  classes and legend, because the two columns are on different scales.
- The chapter's version 3 code is
  `tm_polygons(c("TT_HOUSEHOLDS", "RADIO_PR"), style = "jenks") + tm_facets(sync = TRUE, ncol = 2)`.
:::

The first pair of maps shows that the count of radios follows the number of households: the
townships in the darkest class are largely the same on both maps. The chapter's question is what
differs between the two panels. The highest radio rates are along the western edge of the state,
from the north-west to the south-west, with one more township in the east. Several of the townships
with the most households, such as the darkest township in the north of the left-hand panel, have
only moderate rates. Dividing by the number of households removes the effect of township size.

Three features of the maps are used later on this page:

| Feature of the maps | Used in |
|---------|-----------|
| The map of radio counts has much the same pattern as the map of households | 2.3, the reason for using rates |
| The highest radio rates are along the western edge of the state | 4.1: radio is one of the two rates with significant spatial autocorrelation (Moran's I 0.44). 4.8: the "High radio" region of REDCAP is on the western edge, and seven of the ten highest radio rates are in it |
| Households are concentrated in a few large townships, Taunggyi the largest with 82,604 | 4.5 and 4.7: a minimum number of households per region. Taunggyi meets the minimum of 50,000 in section 4.5 on its own |

### Correlation between the variables

Two variables that are highly correlated measure much the same thing. If both enter the clustering,
that dimension is counted twice. In section 12.6, the correlations are checked before the variables
are chosen.

```{r}
#| fig-height: 6
cluster_vars.cor <- cor(ict_derived[, 12:17])
corrplot.mixed(cluster_vars.cor,
               lower = "ellipse",
               upper = "number",
               tl.pos = "lt",
               diag = "l",
               tl.col = "black")
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses two functions:

```r
cor(x)                                                    # stats
corrplot.mixed(corr, lower, upper, tl.pos, diag, tl.col)  # corrplot
```

- `cor()` returns the matrix of Pearson correlations between every pair of columns.
- `ict_derived[, 12:17]` picks part of the table. Square brackets after a table take the form
  `table[rows, columns]`. Nothing before the comma means every row, all 55 townships. `12:17` after
  the comma is the sequence 12, 13, 14, 15, 16, 17, which means columns 12 to 17: the six
  penetration rates, as the column table in section 2.3 shows. `cor()` accepts numbers only, so the
  name and code columns (1 to 4) and the counts (5 to 11) are left out.
- `corrplot.mixed()` draws the matrix in two halves. `lower = "ellipse"` draws each correlation
  below the diagonal as an ellipse, narrower for a stronger correlation, and tilted right for a
  positive one. `upper = "number"` prints the coefficient above the diagonal. `tl.pos = "lt"` puts
  the variable names on the left and top, `diag = "l"` draws the diagonal in the style of the lower
  half, and `tl.col` sets the colour of the names.
:::

Computer and internet are correlated at 0.87, the highest pair. The chapter keeps computer and drops
internet. TV, mobile phone and computer are also correlated with each other at about 0.7. Radio is
nearly uncorrelated with the others, and is negatively correlated with land line (−0.31).

::: callout-note
## Why one of two highly correlated variables is dropped

Euclidean distance adds one squared difference for each variable. Computer and internet ownership
are high in the same townships, so a difference on one is usually a difference on the other, and the
same difference is counted twice.

```{r}
#| echo: false
nb_tmp <- poly2nb(shan_sf)
skater_groups <- \(X) {
  set.seed(1234)
  tree <- mstree(nb2listw(nb_tmp, nbcosts(nb_tmp, X), style = "B"))
  spdep::skater(tree[, 1:2], X, ncuts = 5)$groups
}
rates6 <- c("RADIO_PR", "TV_PR", "LLPHONE_PR", "MPHONE_PR", "COMPUTER_PR", "INTERNET_PR")
x6 <- st_drop_geometry(shan_sf)[, rates6]
x5 <- x6[, 1:5]
d1_tmp <- as.dist(st_distance(shan_sf, shan_sf))
ari_hc <- adjustedRandIndex(cutree(hclust(dist(x5), method = "ward.D"), 6),
                            cutree(hclust(dist(x6), method = "ward.D"), 6))
ari_sk <- adjustedRandIndex(skater_groups(x5), skater_groups(x6))
ari_cg <- adjustedRandIndex(cutree(hclustgeo(dist(x5), d1_tmp, alpha = 0.2), 6),
                            cutree(hclustgeo(dist(x6), d1_tmp, alpha = 0.2), 6))
dist_cor <- cor(dist(x5), dist(x6))
```

Each method in Part 1 was run twice on the raw rates, once without internet, as in the chapter, and
once with it. An adjusted Rand index of 1 would mean the same six clusters both times (the index is
explained in section 4.2).

| Section | Use of the Euclidean distance | Adjusted Rand index, without and with internet |
|----|-----------|-----|
| 3.1 Hierarchical clustering | the two closest clusters are merged at each step | `r round(ari_hc, 2)` |
| 3.2 SKATER | the minimum spanning tree is built from the edge costs, and each cut reduces the within-region sum of squares | `r round(ari_sk, 2)` |
| 3.3 ClustGeo, α = 0.2 | D0, the variable part of the criterion | `r round(ari_cg, 2)` |

The distances themselves change very little: the two distance matrices are correlated
at `r round(dist_cor, 4)`. The clusters change more, because they are weakly separated (section
4.4). A small change in the distances can change which two clusters are merged first, and every
later merge depends on that one. SKATER changes least, because nearly the same edges are in its
minimum spanning tree both times.

In the chapter, internet is dropped at a correlation of 0.87, and the pairs at about 0.7, such as TV
and mobile phone (0.72), are kept. No threshold is given in the chapter.
:::

### The clustering variables

The clustering functions take a plain table of numbers: one row per township and one column per
variable, with no geometry and no text columns. This step builds that table from `shan_sf`, keeps
the five rates chosen in section 3.1.3, and uses the township names as row names, so that the
dendrogram in section 3.1.10 labels each leaf with its township.

```{r}
cluster_vars <- shan_sf |>
  st_set_geometry(NULL) |>
  select("TS.x", "RADIO_PR", "TV_PR", "LLPHONE_PR", "MPHONE_PR", "COMPUTER_PR")
row.names(cluster_vars) <- cluster_vars$"TS.x"
shan_ict <- select(cluster_vars, c(2:6))
head(shan_ict, 5)
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses three functions:

```r
st_set_geometry(x, NULL)                                  # sf
select(.data, ...)                                        # dplyr
row.names(x) <- value                                     # base R
```

- `st_set_geometry(NULL)` removes the geometry column, so the result is an ordinary data frame. The
  clustering functions accept only numbers.
- `select()` keeps the township name and five of the rates, leaving out `INTERNET_PR`.
- `row.names(cluster_vars) <- cluster_vars$"TS.x"` makes the township names the row labels, as the
  steps below show.
- The second `select()` keeps columns 2 to 6, dropping the name column, which is now held in the row
  names.

Three rows of the table, before and after the `row.names()` line and the second `select()` line:

<div style="margin: 1em 0;">
<p style="margin-bottom: 0.3em;"><b>Step 1. <code>cluster_vars</code>, the table the pipe produces</b></p>
<table class="table table-sm" style="width: auto; margin: 0 auto;">
<tr><th style="color: #6c757d;">row label</th><th style="background: rgba(255, 193, 7, 0.45);"><code>TS.x</code></th><th><code>RADIO_PR</code></th><th>…</th></tr>
<tr><td style="color: #6c757d;">1</td><td style="background: rgba(255, 193, 7, 0.45);">Mongmit</td><td>286.2</td><td>…</td></tr>
<tr><td style="color: #6c757d;">2</td><td style="background: rgba(255, 193, 7, 0.45);">Pindaya</td><td>417.5</td><td>…</td></tr>
<tr><td style="color: #6c757d;">3</td><td style="background: rgba(255, 193, 7, 0.45);">Ywangan</td><td>484.5</td><td>…</td></tr>
</table>
<div style="text-align: center; margin: 0.5em 0; color: #343a40;"><span style="font-size: 1.8em;">&darr;</span><br><code>cluster_vars&#36;&quot;TS.x&quot;</code> takes out the <code>TS.x</code> column. <code>&#36;</code> means &ldquo;the column called&rdquo;.</div>
<p style="margin-bottom: 0.3em;"><b>Step 2. The values taken out</b></p>
<p style="text-align: center;"><code style="background: rgba(255, 193, 7, 0.45); padding: 0.2em 0.4em;">&quot;Mongmit&quot;  &quot;Pindaya&quot;  &quot;Ywangan&quot;  …</code></p>
<div style="text-align: center; margin: 0.5em 0; color: #343a40;"><span style="font-size: 1.8em;">&darr;</span><br><code>row.names(cluster_vars) &lt;-</code> puts them in place of the row labels.</div>
<p style="margin-bottom: 0.3em;"><b>Step 3. <code>cluster_vars</code> with township names as row labels</b></p>
<table class="table table-sm" style="width: auto; margin: 0 auto;">
<tr><th style="color: #6c757d;">row label</th><th><code>TS.x</code></th><th><code>RADIO_PR</code></th><th>…</th></tr>
<tr><td style="background: rgba(255, 193, 7, 0.45);">Mongmit</td><td>Mongmit</td><td>286.2</td><td>…</td></tr>
<tr><td style="background: rgba(255, 193, 7, 0.45);">Pindaya</td><td>Pindaya</td><td>417.5</td><td>…</td></tr>
<tr><td style="background: rgba(255, 193, 7, 0.45);">Ywangan</td><td>Ywangan</td><td>484.5</td><td>…</td></tr>
</table>
<div style="text-align: center; margin: 0.5em 0; color: #343a40;"><span style="font-size: 1.8em;">&darr;</span><br>The next line, <code>select(cluster_vars, c(2:6))</code>, drops column 1, <code>TS.x</code>.</div>
<p style="margin-bottom: 0.3em;"><b>Step 4. <code>shan_ict</code>: numbers only, each row still labelled</b></p>
<table class="table table-sm" style="width: auto; margin: 0 auto;">
<tr><th style="color: #6c757d;">row label</th><th><code>RADIO_PR</code></th><th><code>TV_PR</code></th><th>…</th></tr>
<tr><td style="background: rgba(255, 193, 7, 0.45);">Mongmit</td><td>286.2</td><td>554.1</td><td>…</td></tr>
<tr><td style="background: rgba(255, 193, 7, 0.45);">Pindaya</td><td>417.5</td><td>505.1</td><td>…</td></tr>
<tr><td style="background: rgba(255, 193, 7, 0.45);">Ywangan</td><td>484.5</td><td>260.6</td><td>…</td></tr>
</table>
</div>

The township name is no longer a column, which the clustering functions could not use, but it
remains as a label on each row, and the distance matrix and the dendrogram take their labels from
it.
:::

::: callout-note
## Why the township is the unit of analysis

Each row of `shan_ict` is one township. The township name can label the rows because no two
townships share a name. A district name is shared by several townships, so it cannot label the rows.

The township is also the most detailed level in the data. The maps show computer ownership at three
levels on one colour scale. A district or part of the state has the rate of all its households
together.

```{r}
#| echo: false
#| fig-height: 4.2
level_rates <- \(col) shan_sf |>
  group_by(unit = .data[[col]]) |>
  summarise(rate = sum(COMPUTER) / sum(TT_HOUSEHOLDS) * 1000,
            geometry = st_union(st_buffer(geometry, 50)))
levels_sf <- list(`3 parts of the state (ST)` = level_rates("ST"),
                  `13 districts (DT.x)` = level_rates("DT.x"),
                  `55 townships (TS.x)` = level_rates("TS.x"))
rate_lims <- range(unlist(map(levels_sf, "rate")))
level_plots <- imap(levels_sf, \(g, name) ggplot(g) +
  geom_sf(aes(fill = rate), colour = "white", linewidth = 0.3) +
  scale_fill_viridis_c(limits = rate_lims, option = "magma", direction = -1,
                       name = "computers per\n1,000 households") +
  labs(title = name) + theme_void())
patchwork::wrap_plots(level_plots, ncol = 3, guides = "collect") &
  theme(legend.position = "bottom")
tgy_ts <- levels_sf[[3]]$rate[levels_sf[[3]]$unit == "Taunggyi"]
tgy_dt <- levels_sf[[2]]$rate[levels_sf[[2]]$unit == "Taunggyi"]
```

When townships are combined into districts, the differences between them are lost: Taunggyi
township, at `r round(tgy_ts)` computers per thousand households, is in a district with a rate
of `r round(tgy_dt)`.
:::

### Standardisation

The clustering methods below measure how far apart two townships are with Euclidean distance, the
square root of the sum of squared differences across the variables. A variable with a larger spread
contributes larger squared differences. Sections 12.7.3 and 12.7.4 compute two standardised versions
of the variables.

```{r}
shan_ict.std <- normalize(shan_ict)
summary(shan_ict.std)
```

```{r}
shan_ict.z <- scale(shan_ict)
describe(shan_ict.z)
```

::: {.callout-note collapse="false" title="How the code works"}
The two blocks use three functions:

```r
normalize(x)                                              # heatmaply
scale(x, center = TRUE, scale = TRUE)                     # base R
describe(x)                                               # psych
```

- `normalize()` applies min-max scaling to each column: it subtracts the minimum and divides by the
  range, so every column runs from 0 to 1.
- `scale()` subtracts each column's mean and divides by its standard deviation, which gives z-scores
  with mean 0 and standard deviation 1. It returns a matrix.
- `describe()` prints the mean, standard deviation, median, range, skew and kurtosis of each column.
  The chapter uses it here because `summary()` does not show the standard deviation.

:::

```{r}
#| fig-height: 3.5
r <- ggplot(data = ict_derived,
            aes(x = `RADIO_PR`)) +
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "light blue") +
  ggtitle("Raw values without standardisation")

shan_ict_s_df <- as.data.frame(shan_ict.std)
s <- ggplot(data = shan_ict_s_df,
            aes(x = `RADIO_PR`)) +
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "light blue") +
  ggtitle("Min-Max Standardisation")

shan_ict_z_df <- as.data.frame(shan_ict.z)
z <- ggplot(data = shan_ict_z_df,
            aes(x = `RADIO_PR`)) +
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "light blue") +
  ggtitle("Z-score Standardisation")

ggarrange(r, s, z,
          ncol = 3,
          nrow = 1)
```

::: {.callout-note collapse="false" title="How the code works"}
- The three charts are histograms of the same variable, `RADIO_PR`, on the three scales. `ggtitle()`
  adds a title above each.
- `normalize()` returns a data frame but `scale()` returns a matrix, and `ggplot()` takes a data
  frame, so `as.data.frame()` converts both before plotting.
- `ggarrange()` draws the three stored charts in one row.
:::

The chapter's question is what can be concluded from the three histograms. The shapes are identical.
Both methods are linear rescalings, which move and stretch the axis without changing the shape of
the distribution. The range changes, and so does the weight of the variable in a Euclidean distance.

The chapter then continues with the **raw** rates, `shan_ict`, for every clustering step that
follows, and the standardised versions are not used again. The effect of that decision is measured
in section 4.2.

::: callout-note
## How the scale of a variable affects its weight in the distance

The Euclidean distance between two townships adds up the squared difference on each variable. A
variable measured on a wide range produces large differences, and a variable on a narrow range
produces small ones, whatever their importance to the question.
:::

### Proximity matrix

The clustering in section 3.1.7 groups townships whose ICT rates are close. `hclust()` takes as its
input the distance between every pair of townships, and not the rates themselves, because at each
step it compares only distances, first between townships and then between clusters. Computing the
distances first, with `dist()`, sets the kind of distance separately from the clustering, and the
same distances are used again for ClustGeo in section 3.3.

```{r}
proxmat <- dist(shan_ict, method = "euclidean")
as.matrix(proxmat)[1:4, 1:4]
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses one function:

```r
dist(x, method = "euclidean")                             # stats
```

- `dist()` computes the distance between every pair of rows. With 55 townships there are 55 × 54 / 2
  = 1,485 pairs. The result is stored as the lower triangle only, in an object of class `dist`.
- `as.matrix()` converts it to a full square matrix so that a corner of it can be printed. The
  chapter prints the whole object, which is 1,485 numbers.
:::

::: callout-note
## What the Euclidean distance between two townships measures

The distance is not a distance on the map. Each township is treated as a point whose coordinates are
its rates, and the distance is the length of the straight line between two such points, in
households per thousand. With two rates the points can be drawn, and the distance is the long side
of a right-angled triangle whose other two sides are the differences in the two rates:

$$d = \sqrt{(x_2 - x_1)^2 + (y_2 - y_1)^2}$$

```{r}
#| echo: false
#| fig-height: 4.6
pair2 <- shan_ict[c("Mongmit", "Pindaya"), ]
x1 <- pair2$TV_PR[1]; y1 <- pair2$MPHONE_PR[1]
x2 <- pair2$TV_PR[2]; y2 <- pair2$MPHONE_PR[2]
ggplot() +
  annotate("segment", x = x1, xend = x2, y = y1, yend = y1, linetype = "dashed") +
  annotate("segment", x = x2, xend = x2, y = y1, yend = y2, linetype = "dashed") +
  annotate("segment", x = x1, xend = x2, y = y1, yend = y2, colour = "red", linewidth = 1) +
  annotate("point", x = c(x1, x2), y = c(y1, y2), colour = "red", size = 3) +
  annotate("text", x = c(x1, x2), y = c(y1, y2), label = rownames(pair2),
           hjust = c(-0.25, 1.25), size = 4) +
  annotate("text", x = (x1 + x2) / 2, y = y1 + 7, size = 3.4,
           label = sprintf("difference in TV_PR = %.1f", abs(x1 - x2))) +
  annotate("text", x = x2 - 3, y = (y1 + y2) / 2, hjust = 1, size = 3.4,
           label = sprintf("difference in\nMPHONE_PR = %.1f", abs(y1 - y2))) +
  annotate("text", x = (x1 + x2) / 2 + 6, y = (y1 + y2) / 2 - 6, hjust = 0, size = 3.4,
           colour = "red", label = sprintf("√(%.1f² + %.1f²) = %.1f",
                                           abs(x1 - x2), abs(y1 - y2),
                                           sqrt((x1 - x2)^2 + (y1 - y2)^2))) +
  coord_equal(xlim = c(x2 - 75, x1 + 55), ylim = c(y2 - 12, y1 + 15)) +
  labs(x = "TV_PR, households per thousand", y = "MPHONE_PR, households per thousand")
```

With all five rates the formula has five squared differences under the square root, one for each
rate, and the points can no longer be drawn. For Mongmit and Pindaya the five squared differences
give a distance of `r round(as.matrix(proxmat)["Mongmit", "Pindaya"], 1)`, the value printed above.
:::

### Hierarchical clustering

Agglomerative hierarchical clustering starts with every township in its own cluster and at each step
merges the two closest clusters, until one cluster remains. The methods differ in how they measure
the distance between two clusters that each contain several townships. Ward's method merges the pair
whose merger increases the total within-cluster sum of squares by the least.

```{r}
hclust_ward <- hclust(proxmat, method = "ward.D")
```

```{r}
#| fig-height: 6
plot(hclust_ward, cex = 0.6)
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses two functions:

```r
hclust(d, method = "complete")                            # stats
plot(x, cex)                                              # base R, for hclust objects
```

- `hclust()` runs agglomerative clustering on a `dist` object. `method` selects the rule for the
  distance between clusters. It provides eight, among them `"ward.D"` and `"ward.D2"`, which give
  different trees on the same distances.
- `plot()` on an `hclust` object draws the dendrogram. `cex = 0.6` shrinks the township labels.
:::

::: callout-note
## How agglomerative clustering builds the tree

The procedure is easiest to follow on six points. Each point starts as its own cluster. At each step
the two clusters whose merger adds least to the within-cluster sum of squares are joined, which is
Ward's criterion, and the height of the join in the dendrogram records that increase. After five
merges one cluster remains. Cutting the tree at a chosen height gives the clusters.

```{r}
#| echo: false
#| fig-height: 3.4
toy <- tibble(id = LETTERS[1:6], x = c(1, 1.4, 1.2, 4, 4.5, 4.2), y = c(1, 1.3, 2.1, 1.2, 1.5, 3))
toy_tree <- hclust(dist(toy[, c("x", "y")]), method = "ward.D2")
toy_tree$labels <- toy$id
toy$cluster <- factor(cutree(toy_tree, 2))

p_pts <- ggplot(toy, aes(x, y, colour = cluster)) +
  geom_point(size = 3) + geom_text(aes(label = id), nudge_x = 0.2, colour = "black") +
  coord_equal() + labs(title = "Six points, cut into two clusters") +
  theme(legend.position = "none")

p_tree <- fviz_dend(toy_tree, k = 2, k_colors = scales::hue_pal()(2)[cutree(toy_tree, 2)[toy_tree$order][c(1, 6)]],
                    cex = 0.9, main = "The dendrogram and the cut",
                    ylab = "merge height") +
  geom_hline(yintercept = mean(rev(toy_tree$height)[1:2]), linetype = "dashed", colour = "red")

patchwork::wrap_plots(p_pts, p_tree)
```

The dashed line cuts the tree below its last merge and leaves two clusters, {A, B, C} and {D, E, F}.
The same procedure on the 55 townships gives the dendrogram above, and `cutree()` makes the cut in
section 3.1.12.
:::

### Choice of agglomeration method

Ward's method was used in section 3.1.7 without a comparison. The chapter chooses the method by
building a tree with each of four methods, computing the agglomerative coefficient of each tree, and
keeping the method with the highest coefficient.

```{r}
m <- c("average", "single", "complete", "ward")
names(m) <- c("average", "single", "complete", "ward")

ac <- function(x) {
  agnes(shan_ict, method = x)$ac
}

map_dbl(m, ac)
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses two functions:

```r
agnes(x, method = "average")                              # cluster
map_dbl(.x, .f)                                           # purrr
```

- `m` is a vector of the four method names. The methods differ in how the distance between two
  clusters is measured: single linkage uses the closest pair of townships, complete linkage the
  farthest pair, average linkage the mean over all pairs, and Ward the increase in the
  within-cluster sum of squares. Slide 11 uses the same three linkages for REDCAP.
- `names(m) <- ...` gives each element of `m` a name, here the same text as its value. `map_dbl()`
  copies these names to its result, so each coefficient is printed under the name of its method.
- `ac <- function(x) { ... }` defines a new function called `ac`. `function(x)` means that it takes
  one input, called `x` inside the braces, and the value of the last line in the braces is what it
  returns. `ac("ward")` runs `agnes(shan_ict, method = "ward")`.
- `agnes()` runs the same kind of agglomerative clustering as `hclust()`, and computes the Euclidean
  distances from `shan_ict` itself. It returns a list, and `$ac` takes the element called `ac` from
  it, the agglomerative coefficient, in the same way as `$` takes a column from a table in section
  3.1.4.
- `map_dbl(m, ac)` calls `ac()` once for each element of `m` and collects the four numbers into one
  named vector. `dbl` stands for double, R's type for decimal numbers.

The block gives the same result as the four calls written out:

```r
c(average  = agnes(shan_ict, method = "average")$ac,
  single   = agnes(shan_ict, method = "single")$ac,
  complete = agnes(shan_ict, method = "complete")$ac,
  ward     = agnes(shan_ict, method = "ward")$ac)
```
:::

::: callout-note
## What the agglomerative coefficient measures

For each township $i$, `agnes()` computes the ratio

$$m(i) = \frac{\text{height at which township } i \text{ is first merged}}{\text{height of the final merge}}$$

and the agglomerative coefficient is the average of one minus that ratio over the $n$ townships (the
`agnes()` documentation; Kaufman and Rousseeuw, 1990):

$$\text{AC} = \frac{1}{n} \sum_{i=1}^{n} \bigl(1 - m(i)\bigr)$$

A township merged into its group low in the tree has $m(i)$ close to 0, and $1 - m(i)$ close to 1.

A coefficient close to 1 therefore means that most townships are in tight groups, and the groups are
joined to each other only much higher in the tree. The UC Business Analytics tutorial, the source of
this part of the chapter, calls this a strong clustering structure. A lower coefficient means that
the merges are spread through the tree, with no clear gap between groups. The two examples have six
points each:

```{r}
#| echo: false
#| fig-height: 5.5
ac_example <- function(x, y, title) {
  a <- agnes(data.frame(x, y, row.names = LETTERS[1:6]), method = "ward")
  pts <- ggplot(tibble(x, y, id = LETTERS[1:6]), aes(x, y, label = id)) +
    geom_point(colour = "steelblue", size = 2.5) + geom_text(nudge_y = 0.12) +
    coord_equal(ylim = c(0.5, 2)) + labs(title = title, x = NULL, y = NULL)
  tree <- fviz_dend(as.hclust(a), cex = 0.8, main = sprintf("coefficient %.2f", a$ac),
                    ylab = "merge height") +
    geom_hline(yintercept = max(a$height), linetype = "dashed", colour = "red")
  list(pts, tree)
}
ex_tight <- ac_example(c(1, 1.3, 1.1, 5, 5.2, 4.9), c(1, 1.2, 1.4, 1.1, 1.4, 1.2),
                       "Two tight groups")
ex_spread <- ac_example(c(1, 2, 3, 4, 5, 6), c(1, 1.3, 1, 1.2, 1, 1.3), "Evenly spread")
patchwork::wrap_plots(c(ex_tight, ex_spread), ncol = 2, byrow = FALSE, heights = c(1, 1.4))
```

The dashed red line is the final merge. In the first example every point is merged into its group
far below it. In the second, the merge heights increase steadily up to it.
:::

Ward's method scores highest, 0.943, and the chapter uses Ward's method from here on.

### Number of clusters

The dendrogram can be cut into any number of clusters, k. The chapter uses the gap statistic to
choose k from the data: it compares how tightly the townships are grouped at each k with how tightly
data with no clusters would be grouped. The number chosen, six, is used for every set of clusters on
the rest of the page.

```{r}
set.seed(12345)
gap_stat <- clusGap(shan_ict,
                    FUN = hcut,
                    nstart = 25,
                    K.max = 10,
                    B = 50)
print(gap_stat, method = "firstmax")
```

```{r}
#| fig-height: 3.5
fviz_gap_stat(gap_stat)
```

::: {.callout-note collapse="false" title="How the code works"}
The two blocks use three functions:

```r
clusGap(x, FUNcluster, K.max, B, ...)                     # cluster
hcut(x, k, hc_method = "ward.D2", ...)                    # factoextra
fviz_gap_stat(gap_stat)                                   # factoextra
```

- `clusGap()` clusters the data for k = 1 to `K.max`, and for each k also clusters `B` = 50
  reference data sets generated without structure. The gap is the difference between the reference
  and observed log dispersion. `set.seed()` fixes the random reference sets so that the result
  repeats.
- `FUN = hcut` makes `clusGap()` cluster with `hcut()`, which runs hierarchical clustering and cuts
  the tree into k groups. Its default method is `"ward.D2"`. `nstart = 25` is passed on to `hcut()`,
  which does not use it, because `nstart` is an argument of k-means.
- `print(method = "firstmax")` reports the first local maximum of the gap as the recommended k.
- `fviz_gap_stat()` plots the gap for each k with an error bar, explained in the callout below. Its
  dashed line marks the k chosen by its default rule, `firstSEmax`: the smallest k whose gap is
  within one standard error of the first peak.
:::

The recommended number is 1. The chapter reads the chart as showing that "the 6-cluster gives the
largest gap statistic" and retains six clusters. In the printed table, k = 1 has the largest gap
(0.274), and the gap for k = 6 (0.222) is within 0.02 of every other value from k = 2 to 10, while
each standard error is about 0.04. This part keeps six clusters, so that the results can be compared
with the chapter. The number of clusters on this page is therefore not set by the gap statistic,
which is compared with other criteria in section 4.4.

::: callout-note
## What the gap statistic compares

$W_k$, the within-cluster dispersion, is the sum of the squared distances from each township to the
centre of its cluster when the tree is cut into $k$ clusters. It falls as $k$ rises, because more
clusters always fit the data more closely. `clusGap()` compares its fall on the data with its fall
on 50 reference data sets of the same size, spread evenly over the range of the data, which have no
clusters:

$$\text{gap}(k) = \text{average of } \log W_k \text{ over the reference sets} - \log W_k \text{ of the data}$$

The two terms are the two curves below, drawn for the Shan rates. The gap at each $k$ is the
vertical distance between them.

How to read the gap:

- **A smaller $W_k$ alone means nothing.** Adding a cluster always makes $W_k$ smaller, in the data
  and in the reference sets alike.
- **A large gap means unusually tight clusters.** The townships are much closer to the centres of
  their clusters than points with no clusters would be.
- **A rise in the gap marks real groups.** The gap rises at a $k$ where the data have groups, and
  stays flat where an extra cluster only splits the data as it would split random points.
- **The peak gives the number of clusters.** A flat gap, as for the Shan rates, gives no clear
  number.

Why each $k$ has an error bar:

- **The reference sets are random.** Each of the 50 reference data sets gives a slightly different
  $\log W_k$, so the gap is an estimate.
- **The bar is one standard error each way.** It runs from one standard error below the gap to one
  above, from the spread of the 50 reference values. The short horizontal lines mark its ends.

How to read the chart of `fviz_gap_stat()`:

- **Each point is the gap at one $k$**, and the line joins the points.
- **A long bar** means an uncertain gap.
- **Two bars that overlap** belong to gaps that are not clearly different. On the Shan chart all the
  bars overlap, so no $k$ is clearly better than another.
- **The dashed vertical line** is the $k$ chosen by the `firstSEmax` rule.

```{r}
#| echo: false
#| fig-height: 3.2
as_tibble(gap_stat$Tab) |>
  mutate(k = row_number()) |>
  pivot_longer(c(logW, E.logW), names_to = "curve", values_to = "value") |>
  mutate(curve = recode(curve, logW = "observed, log(W)", E.logW = "reference, E[log(W)]")) |>
  ggplot(aes(k, value, colour = curve)) +
  geom_line() + geom_point() +
  scale_x_continuous(breaks = 1:10) +
  scale_colour_manual(values = c("steelblue", "grey50")) +
  labs(y = "log within-cluster dispersion", colour = NULL) +
  theme(legend.position = "bottom")
```

On the Shan rates the two curves are almost parallel, so the gap hardly changes with $k$.

Two generic examples, of 90 points each, show the gap with and without clusters, with the same bars
as the chart of `fviz_gap_stat()` above:

```{r}
#| echo: false
#| fig-height: 5.5
set.seed(42)
gap_examples <- list(
  `three clear clusters` = tibble(x = c(rnorm(30, 0, 0.4), rnorm(30, 4, 0.4), rnorm(30, 2, 0.4)),
                                 y = c(rnorm(30, 0, 0.4), rnorm(30, 0, 0.4), rnorm(30, 3.5, 0.4))),
  `no clusters` = tibble(x = runif(90, -1, 5), y = runif(90, -1, 4.5)))
gap_plots <- imap(gap_examples, \(d, name) {
  g <- clusGap(as.data.frame(d), FUN = hcut, K.max = 8, B = 50)
  tab <- as_tibble(g$Tab) |> mutate(k = row_number())
  pts <- ggplot(d, aes(x, y)) + geom_point(colour = "steelblue", size = 1) + coord_equal() +
    labs(title = name, x = NULL, y = NULL)
  gp <- ggplot(tab, aes(k, gap)) + geom_line() + geom_point() +
    geom_errorbar(aes(ymin = gap - SE.sim, ymax = gap + SE.sim), width = 0.2) +
    scale_x_continuous(breaks = 1:8) + coord_cartesian(ylim = c(-0.1, 0.95)) + labs(y = "gap")
  list(pts, gp)
})
patchwork::wrap_plots(c(gap_plots[[1]], gap_plots[[2]]), ncol = 2, byrow = FALSE)
```

With three clear clusters, the gap rises sharply to a peak at $k = 3$, above the bars of its
neighbours, and the first maximum is 3. With points spread evenly, the gap is about the same at
every $k$, within the bars, and the first maximum is $k = 1$. The Shan rates in the chart above are
like the second example.
:::

### The dendrogram

The height at which two branches join is the increase in the merging criterion caused by joining
them. Two townships are similar if their branches join low in the tree. Their horizontal position is
not a measure of similarity.

```{r}
#| fig-height: 6
plot(hclust_ward, cex = 0.6)
rect.hclust(hclust_ward, k = 6, border = 1:6)
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses one new function:

```r
rect.hclust(tree, k, border)                              # stats
```

- `rect.hclust()` draws a rectangle around each of the `k` groups obtained by cutting the tree.
  `border` gives the colours. The chapter passes `2:5`, four colours for six groups, so two
  rectangles reuse a colour. `1:6` gives each group its own.
:::

### Cluster heatmap

A cluster heatmap draws the data matrix as coloured cells, with the rows ordered by the dendrogram,
so that the variables on which each group is high or low can be read across.

```{r}
#| fig-height: 9
shan_ict_mat <- data.matrix(shan_ict)

heatmaply(normalize(shan_ict_mat),
          Colv = NA,
          dist_method = "euclidean",
          hclust_method = "ward.D",
          seriate = "OLO",
          colors = Blues,
          k_row = 6,
          margins = c(NA, 200, 60, NA),
          fontsize_row = 4,
          fontsize_col = 5,
          main = "Geographic Segmentation of Shan State by ICT indicators",
          xlab = "ICT Indicators",
          ylab = "Townships of Shan State")
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses two functions:

```r
data.matrix(frame)                                        # base R
heatmaply(x, Colv, dist_method, hclust_method, seriate, colors, k_row, ...)  # heatmaply
```

- `data.matrix()` converts the data frame to a numeric matrix, which `heatmaply()` requires.
- `heatmaply()` draws an interactive heatmap, in which hovering over a cell shows its township,
  variable and value. It clusters the rows itself with `dist_method` and `hclust_method`.
  `Colv = NA` leaves the columns in their original order. `seriate = "OLO"` (optimal leaf ordering)
  reorders the branches so that neighbouring rows are as similar as possible, without changing the
  tree. `k_row = 6` colours six row groups. `margins` and the font sizes set the layout.

The heatmap clusters `normalize(shan_ict_mat)`, the min-max scaled values, while the dendrogram in
section 3.1.10 and the maps in 3.1.12 use the raw values. The six groups on the heatmap are
therefore not the six groups on the map.
:::

::: callout-note
## How to read the cluster heatmap

- Each row is one township and each column one rate.
- The colour of a cell is the township's rate after min-max scaling (section 3.1.5): white for the
  lowest value in the state and dark blue for the highest.
- The rows are in the order of the dendrogram at the left of the heatmap, so rows next to each other
  are similar townships, and the six coloured branches are the six clusters.
- Reading across a row gives one township's profile. Reading down a column within one cluster shows
  whether the cluster is high or low on that rate.

Seven townships, in the same colours:

```{r}
#| echo: false
#| fig-height: 3.2
heat_rows <- c("Taunggyi", "Tachileik", "Muse", "Pindaya", "Ywangan", "Narphan", "Matman")
as_tibble(normalize(shan_ict_mat), rownames = "township") |>
  filter(township %in% heat_rows) |>
  pivot_longer(-township, names_to = "rate", values_to = "value") |>
  mutate(township = factor(township, levels = rev(heat_rows)),
         rate = factor(rate, levels = colnames(shan_ict_mat))) |>
  ggplot(aes(rate, township, fill = value)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = sprintf("%.2f", value), colour = value > 0.6), size = 3) +
  scale_fill_distiller(palette = "Blues", direction = 1, limits = c(0, 1)) +
  scale_colour_manual(values = c(`FALSE` = "black", `TRUE` = "white"), guide = "none") +
  labs(x = NULL, y = NULL, fill = "min-max\nvalue")
```

Taunggyi, Tachileik and Muse are dark on TV, mobile phone and computer. Pindaya and Ywangan are
darkest on radio. Narphan and Matman are light on almost every rate. On the full heatmap, townships
with the same pattern are in adjacent rows and form a block of similar colour.
:::

A dark cell, or a value close to 1, means the township is among the highest in the state on that
rate; a light cell, or a value close to 0, means the township is among the lowest. For example,
Tachileik's computer cell is the darkest in its column, at 1.00, because Tachileik has the highest
computer rate of the 55 townships, `r round(max(shan_ict$COMPUTER_PR))` households per thousand.

### Mapping the clusters

So far the clusters exist only as group numbers in the tree. To see where they are, each township's
group number is attached to its polygon, and the polygons are mapped by group.

```{r}
groups <- as.factor(cutree(hclust_ward, k = 6))
shan_sf_cluster <- cbind(shan_sf, as.matrix(groups)) |>
  rename(CLUSTER = as.matrix.groups.)
table(shan_sf_cluster$CLUSTER)
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses three functions:

```r
cutree(tree, k)                                           # stats
cbind(...)                                                # base R, sf method
rename(.data, ...)                                        # dplyr
```

- `cutree()` cuts the tree at the height that leaves `k` groups and returns the group number of each
  township, in the row order of the data. `as.factor()` makes the numbers categories, so that the
  map draws them as categories and not as quantities.
- `cbind()` binds the group column to `shan_sf` by row position. This is correct only because
  `shan_ict` was built from `shan_sf` in the same row order. The new column is named
  `as.matrix.groups.` after the expression that produced it, and `rename()` changes the name to
  `CLUSTER`.

The same result can be written without the matrix step:
`shan_sf_cluster <- mutate(shan_sf, CLUSTER = groups)`.
:::

```{r}
#| fig-height: 5
qtm(shan_sf_cluster, "CLUSTER")
```

::: {.callout-note collapse="false" title="How the code works"}
- `qtm()` colours each township by its cluster. `CLUSTER` holds text labels, so `qtm()` gives each
  cluster its own colour, with no order between them.
:::

The clusters are fragmented. Cluster 1 appears in the north, the centre and the south, and several
clusters are scattered across the state in small pieces. The townships in one cluster are similar in
their ICT rates, but a cluster is not one connected region. According to the chapter, the
fragmentation is the main limitation of clustering without regard to location. The six clusters are
in 24 separate pieces of territory.

## SKATER

SKATER (Spatial 'K'luster Analysis by Tree Edge Removal) was introduced by Assunção et al. (2006).
Slides 7 and 8 of the lecture and Assunção et al. (2006, sections 3.3 and 3.4) describe it in four
steps:

1. Represent the townships as nodes of a graph, with an edge between every pair of neighbours.
2. Give each edge a cost, the dissimilarity between the two townships it joins.
3. Reduce the graph to a minimum spanning tree: n − 1 edges that connect all n townships at the
   lowest total cost.
4. Remove edges from the tree. Each removal splits one branch into two, and each piece that remains
   is a region. At each step SKATER removes the edge whose removal most reduces the within-region
   sum of squares.

Because every region is a connected piece of the tree, and every edge of the tree joins two
neighbours, every region is contiguous by construction.

### Neighbour list

SKATER can only join townships that share a boundary. The neighbour list records, for each township,
the townships that touch it, and these pairs become the edges of the graph in step 1.

```{r}
shan.nb <- poly2nb(shan_sf)
summary(shan.nb)
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses one function:

```r
poly2nb(pl, queen = TRUE)                                 # spdep
```

- `poly2nb()` finds, for each polygon, the polygons that share at least one boundary point with it
  (Queen contiguity, the default). The chapter first converts `shan_sf` to an sp object with
  `as_Spatial()`. The current version of `spdep` accepts sf objects directly and returns the same
  list.

The 55 townships have 264 neighbour links, 4.8 per township on average. Every township has at least
two neighbours, so no township is isolated and the graph is connected, which SKATER requires.
:::

```{r}
#| fig-height: 6
coords <- st_coordinates(st_centroid(st_geometry(shan_sf)))

plot(st_geometry(shan_sf), border = grey(0.5))
plot(shan.nb, coords, col = "blue", add = TRUE)
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses four functions:

```r
st_geometry(x)                                            # sf
st_centroid(x)                                            # sf
st_coordinates(x)                                         # sf
plot(nb, coords, col, add = TRUE)                         # spdep, for nb objects
```

- `st_geometry(shan_sf)` keeps only the polygons, without the attribute columns. `st_centroid()`
  turns each polygon into one point at its centre, and `st_coordinates()` turns each point into two
  numbers, its longitude (`X`) and latitude (`Y`), one row per township. The figure after this box
  follows Mongmit through the three functions.
- The first `plot()` draws the township boundaries in grey. The second draws a line between the
  centroids of every pair of neighbours, and `add = TRUE` draws it on the same plot. The chapter
  notes that the boundaries go first, because the first call sets the extent of the plot.
:::

The three functions in the first line, applied to one township:

```{r}
#| echo: false
#| fig-height: 7.5
mong <- shan_sf[shan_sf$TS.x == "Mongmit", ]
mong_pt <- st_centroid(st_geometry(mong))
mong_xy <- st_coordinates(mong_pt)
arrow_panel <- function(label) {
  ggplot() +
    annotate("segment", x = 0, xend = 1, y = 0, yend = 0,
             arrow = arrow(length = unit(0.3, "cm")), linewidth = 0.8) +
    annotate("text", x = 0.5, y = 0.25, label = label, family = "mono", size = 3.2) +
    coord_cartesian(xlim = c(-0.1, 1.1), ylim = c(-0.6, 0.6), clip = "off") +
    theme_void()
}
step1 <- ggplot() + geom_sf(data = mong, fill = "grey85", colour = "grey30") +
  labs(title = "1. Mongmit's polygon") + theme_void()
step2 <- ggplot() + geom_sf(data = mong, fill = "grey92", colour = "grey60") +
  geom_sf(data = mong_pt, colour = "red", size = 3) +
  labs(title = "2. One point at its centre") + theme_void()
step3 <- ggplot() +
  annotate("text", x = 0, y = 0, hjust = 0.5, family = "mono", size = 4,
           label = sprintf("     X      Y\n%6.2f %6.2f", mong_xy[1, "X"], mong_xy[1, "Y"])) +
  coord_cartesian(xlim = c(-1, 1), ylim = c(-1, 1)) +
  labs(title = "3. Two numbers:\n   longitude, latitude") + theme_void()
nb_df <- map_dfr(seq_along(shan.nb), \(i) tibble(from = i, to = shan.nb[[i]])) |> filter(from < to)
xy_all <- as_tibble(coords) |> setNames(c("x", "y"))
step4 <- ggplot() +
  geom_sf(data = shan_sf, fill = "grey95", colour = "grey75", linewidth = 0.2) +
  geom_segment(data = tibble(x = xy_all$x[nb_df$from], y = xy_all$y[nb_df$from],
                             xend = xy_all$x[nb_df$to], yend = xy_all$y[nb_df$to]),
               aes(x, y, xend = xend, yend = yend), colour = "steelblue", linewidth = 0.3) +
  geom_point(data = xy_all, aes(x, y), size = 0.8) +
  geom_point(aes(x = mong_xy[1, "X"], y = mong_xy[1, "Y"]), colour = "red", size = 3) +
  labs(title = "4. The same for all 55 townships: one point each, joined to its neighbours",
       subtitle = "plot(shan.nb, coords): Mongmit in red") +
  theme_void()
top <- patchwork::wrap_plots(step1, arrow_panel("st_geometry()\nst_centroid()"), step2,
                             arrow_panel("st_coordinates()"), step3,
                             widths = c(1, 0.7, 1, 0.7, 1))
patchwork::wrap_plots(top, step4, ncol = 1, heights = c(1, 1.8))
```

### Edge costs

Step 2 gives each edge a cost: the Euclidean distance of section 3.1.6 between the ICT rates of the
two townships at its ends, computed only for townships that are neighbours. Neighbours with similar
rates have a low cost. The costs are used in section 3.2.3, where the minimum spanning tree is built
from the cheapest links.

```{r}
lcosts <- nbcosts(shan.nb, shan_ict)
shan.w <- nb2listw(shan.nb, lcosts, style = "B")
summary(shan.w)
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses two functions:

```r
nbcosts(nb, data, method = "euclidean")                   # spdep
nb2listw(neighbours, glist = NULL, style = "W")           # spdep
```

- `nbcosts()` computes, for every neighbour link, the same Euclidean distance as `dist()` in section
  3.1.6, on the raw rates in `shan_ict`. `dist()` computes it for all 1,485 pairs of townships;
  `nbcosts()` only for the pairs that share a boundary.
- `nb2listw()` attaches the costs to the neighbour list as weights, and the result, `shan.w`, is the
  input `mstree()` needs in section 3.2.3. `style = "B"` stores the costs unchanged. The default,
  `"W"`, would divide each township's weights by their sum so that they add up to 1, the row
  standardisation of Exercise 4, and the costs would no longer be distances. This is unrelated to
  the standardisation of the variables in section 3.1.5.

The link between Exercise 4 and this exercise is the weights object. In Exercise 4 the weight of a
link was 1, or 1 divided by the number of neighbours. Here the weight of a link is the dissimilarity
between the two townships.
:::

### Minimum spanning tree

Step 3 reduces the neighbour links to the 54 that connect all 55 townships at the lowest total cost.
The result, the minimum spanning tree, is what SKATER cuts into regions in section 3.2.4.

```{r}
set.seed(1234)
shan.mst <- mstree(shan.w)
dim(shan.mst)
head(shan.mst)
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses one function:

```r
mstree(nbw, ini = NULL)                                   # spdep
```

- `shan.w`, from section 3.2.2, is the neighbour list with the cost of each link. `mstree()` returns
  `shan.mst`, a matrix with one row for each link of the tree: the row numbers of the two townships
  in `shan_sf`, and the cost of the link.
- `mstree()` computes the minimum spanning tree of the weighted graph by Prim's algorithm: start
  from one node, and repeatedly add the cheapest edge that connects a node already in the tree to a
  node not yet in it.
- The tree has 54 rows, one fewer than the number of townships, because a tree that connects n nodes
  has n − 1 edges.
- With `ini = NULL`, the starting node is chosen at random. The tree itself does not depend on the
  start when all costs differ, but the order of its rows does, and so does the numbering of the
  SKATER groups built from it. `set.seed()` fixes the start so the page gives the same printout on
  every render. The chapter does not set a seed, so its printout starts from node 54 and this one
  from a different node.
:::

::: callout-note
## What a minimum spanning tree is

A spanning tree is a set of links that connects every point, directly or through other points,
without any loop, so there is exactly one path between any two points. With $n$ points it always has
$n - 1$ links. A graph has many spanning trees, and the minimum spanning tree is the one whose links
have the lowest total cost. The figure uses five points and seven links, with the cost on each link:

```{r}
#| echo: false
#| fig-height: 2.8
mst_pts <- tibble(id = c("A", "B", "C", "D", "E"), x = c(0, 2, 1, 3, 4.2), y = c(0, 0.3, 2, 2, 0.4))
mst_links <- tibble(from = c("A", "A", "B", "B", "C", "B", "D"),
                    to   = c("B", "C", "C", "D", "D", "E", "E"),
                    cost = c(4, 2, 5, 7, 3, 6, 1))
tree_panel <- function(keep, title) {
  d <- mst_links |> mutate(used = row_number() %in% keep) |>
    left_join(mst_pts, by = c(from = "id")) |>
    left_join(mst_pts, by = c(to = "id"), suffix = c("", "end"))
  ggplot(d) +
    geom_segment(aes(x, y, xend = xend, yend = yend, colour = used, linewidth = used)) +
    geom_label(aes((x + xend) / 2, (y + yend) / 2, label = cost), size = 3, fill = "white",
               label.size = 0, label.padding = unit(0.12, "lines")) +
    geom_point(data = mst_pts, aes(x, y), size = 6, shape = 21, fill = "white") +
    geom_text(data = mst_pts, aes(x, y, label = id), size = 3.2) +
    scale_colour_manual(values = c(`TRUE` = "steelblue", `FALSE` = "grey85"), guide = "none") +
    scale_linewidth_manual(values = c(`TRUE` = 1.3, `FALSE` = 0.5), guide = "none") +
    coord_equal(clip = "off") + theme_void() + labs(title = title) +
    theme(plot.title = element_text(size = 10))
}
patchwork::wrap_plots(
  tree_panel(1:7, "All seven links"),
  tree_panel(c(1, 3, 4, 7), "A spanning tree, cost 4 + 5 + 7 + 1 = 17"),
  tree_panel(c(1, 2, 5, 7), "The minimum spanning tree,\ncost 4 + 2 + 3 + 1 = 10"),
  ncol = 3)
```

SKATER cuts a tree because a tree has no loops: removing any one link splits it into exactly two
connected pieces, so every cut in section 3.2.4 adds one region, and every region is contiguous. The
minimum spanning tree is made of the links between the most similar neighbours, so the links left
for cutting are between townships that are already alike, and a cut separates the most different
parts of a branch.
:::

How Prim's algorithm grows the tree is easiest to see on the same five points. The numbers on the
lines are edge costs. Starting from A, the candidates at each step are the edges that join a point
already in the tree (blue) to a point outside it (dashed), and the cheapest candidate is added
(red):

```{r}
#| echo: false
#| fig-height: 6.5
pts <- tibble(id = c("A", "B", "C", "D", "E"), x = c(0, 2, 1, 3, 4.2), y = c(0, 0.3, 2, 2, 0.4))
edges <- tibble(from = c("A", "A", "B", "B", "C", "B", "D"),
                to   = c("B", "C", "C", "D", "D", "E", "E"),
                cost = c(4, 2, 5, 7, 3, 6, 1))
in_tree <- "A"; tree_edges <- integer(0); frames <- list()
for (k in 1:4) {
  cand <- which(xor(edges$from %in% in_tree, edges$to %in% in_tree))
  pick <- cand[which.min(edges$cost[cand])]
  status <- rep("not yet considered", nrow(edges))
  status[cand] <- "candidate"
  status[tree_edges] <- "in the tree"
  status[pick] <- "added at this step"
  frames[[k]] <- edges |> mutate(status = status,
                                 step = paste0("Step ", k, ": add ", from[pick], "-", to[pick],
                                               " (cost ", cost[pick], ")"))
  tree_edges <- c(tree_edges, pick)
  in_tree <- union(in_tree, c(edges$from[pick], edges$to[pick]))
}
frame_df <- bind_rows(frames) |>
  left_join(pts, by = c(from = "id")) |>
  left_join(pts, by = c(to = "id"), suffix = c("", "end"))
ggplot(frame_df) +
  geom_segment(aes(x, y, xend = xend, yend = yend, colour = status, linetype = status,
                   linewidth = status)) +
  geom_label(aes((x + xend) / 2, (y + yend) / 2, label = cost), size = 3.5, fill = "white",
             label.size = 0, label.padding = unit(0.15, "lines")) +
  geom_point(data = pts, aes(x, y), size = 7, shape = 21, fill = "white") +
  geom_text(data = pts, aes(x, y, label = id)) +
  scale_colour_manual(values = c(`in the tree` = "steelblue", candidate = "orange",
                                 `added at this step` = "red", `not yet considered` = "grey80")) +
  scale_linetype_manual(values = c(`in the tree` = "solid", candidate = "dashed",
                                   `added at this step` = "solid", `not yet considered` = "solid")) +
  scale_linewidth_manual(values = c(`in the tree` = 1.2, candidate = 0.6,
                                    `added at this step` = 1.6, `not yet considered` = 0.4)) +
  facet_wrap(~ step, ncol = 2) +
  scale_y_continuous(expand = expansion(add = 0.35)) +
  scale_x_continuous(expand = expansion(add = 0.35)) +
  coord_equal(clip = "off") + theme_void() +
  labs(colour = NULL, linetype = NULL, linewidth = NULL) +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 10, hjust = 0, margin = margin(b = 6)),
        panel.spacing = unit(1.2, "lines"))
```

After four steps all five points are joined by four edges, with a total cost of 10. B was reached
last, through its cheapest link to the tree, A-B, and the more expensive links B-C, B-D and B-E were
never used. `mstree()` does the same on the 55 townships, and its output lists the 54 edges in the
order they were added, so the first row printed above is the first edge of the tree.

```{r}
#| fig-height: 6
plot(st_geometry(shan_sf), border = gray(0.5))
plot.mst(shan.mst, coords, col = "blue", cex.lab = 0.7, cex.circles = 0.005, add = TRUE)
```

Each township is now connected to the others by exactly one path. The tree keeps, from the 264
neighbour links, the 54 that join the most similar neighbours while keeping every township
connected.

### SKATER regions

Step 4 cuts the tree five times, which leaves six regions.

```{r}
clust6 <- spdep::skater(edges = shan.mst[, 1:2],
                        data = shan_ict,
                        method = "euclidean",
                        ncuts = 5)
ccs6 <- clust6$groups
table(ccs6)
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses one function:

```r
spdep::skater(edges, data, ncuts, crit, vec.crit, method = "euclidean")  # spdep
```

- `edges` is the first two columns of the tree, the pairs of nodes. The costs in column 3 are not
  passed, because `skater()` recomputes the within-region sum of squares from `data` after each cut.
- `ncuts = 5` removes five edges, which leaves six regions. The argument is the number of cuts, one
  fewer than the number of regions.
- `crit` and `vec.crit` set a minimum size for each region. They are not used in the chapter.
- The `spdep::` prefix selects the `spdep` version, because `rgeoda` also has a `skater()`.

The result is a list. `$groups` holds the region of each township, and `$ssto` and `$ssw` hold the
total sum of squares and the within-region sum of squares after each cut.
:::

```{r}
round(clust6$ssw, 0)
```

The six regions have 22, 18, 11, 2, 1 and 1 townships, the sizes printed in the chapter. The first
cut reduces the within-region sum of squares from 12,613 to 10,977, and each later cut reduces it by
less. Two of the six regions are single townships, Ywangan and Narphan.

```{r}
#| fig-height: 6
plot(st_geometry(shan_sf), border = gray(0.5))
plot(clust6, coords, cex.lab = 0.7,
     groups.colors = c("red", "green", "blue", "brown", "pink", "orange"),
     cex.circles = 0.005, add = TRUE)
```

::: {.callout-note collapse="false" title="How the code works"}
- `plot()` on a `skater` object draws the pruned tree, with the edges of each region in the colour
  given for it in `groups.colors`. The chapter gives five colours for six regions, so the sixth
  region reuses the first colour. Six colours are given here.
:::

::: callout-note
## The four steps of SKATER on Shan

The two maps show steps 1 to 3 listed at the start of section 3.2: the neighbour graph with every
Queen link, and the minimum spanning tree with 54 of them.

```{r}
#| echo: false
#| fig-height: 4
xy <- as_tibble(coords) |> setNames(c("x", "y"))
nb_edges <- map_dfr(seq_along(shan.nb), \(i) tibble(from = i, to = shan.nb[[i]])) |>
  filter(from < to)
seg <- \(e) tibble(x = xy$x[e$from], y = xy$y[e$from], xend = xy$x[e$to], yend = xy$y[e$to])
mst_edges <- tibble(from = shan.mst[, 1], to = shan.mst[, 2])

base <- ggplot() + geom_sf(data = shan_sf, fill = "grey95", colour = "grey70", linewidth = 0.2) +
  theme_void()
step_nb <- base + geom_segment(data = seg(nb_edges), aes(x, y, xend = xend, yend = yend),
                               colour = "steelblue", linewidth = 0.3) +
  geom_point(data = xy, aes(x, y), size = 0.7) +
  labs(title = "Steps 1-2: neighbour graph")
step_mst <- base + geom_segment(data = seg(mst_edges), aes(x, y, xend = xend, yend = yend),
                                colour = "steelblue", linewidth = 0.5) +
  geom_point(data = xy, aes(x, y), size = 0.7) +
  labs(title = "Step 3: minimum spanning tree")
patchwork::wrap_plots(step_nb, step_mst, ncol = 2)
```

Each dot is a township and each line an edge.

Step 4 makes the five cuts one at a time. Before each cut, the reduction in the total within-region
sum of squares is computed for the removal of every link in every current region, and the link with
the largest reduction is removed (slide 8). A cut can therefore fall in any region, and each cut
splits one region into two. The six maps show the regions before the first cut and after each cut,
with the region split at that cut in colour:

```{r}
#| echo: false
#| fig-height: 7
cut_groups <- cbind(1, sapply(1:5, \(k) spdep::skater(shan.mst[, 1:2], shan_ict, ncuts = k)$groups))
cut_maps <- lapply(0:5, \(k) {
  if (k == 0) {
    return(ggplot() + geom_sf(data = shan_sf, fill = "grey80", colour = "white", linewidth = 0.2) +
             theme_void() + labs(title = "Before the first cut: 1 region of 55"))
  }
  before <- cut_groups[, k]; after <- cut_groups[, k + 1]
  split <- names(which(tapply(after, before, \(g) length(unique(g))) == 2))
  inside <- before == as.numeric(split)
  part <- ifelse(inside, paste("part", as.integer(factor(after))), "other regions")
  sizes <- sort(table(after[inside]), decreasing = TRUE)
  ggplot() +
    geom_sf(data = mutate(shan_sf, part = part), aes(fill = part), colour = "white",
            linewidth = 0.2) +
    scale_fill_manual(values = setNames(c("grey80", "#1b9e77", "#d95f02"),
                                        c("other regions", sort(unique(part[inside])))),
                      guide = "none") +
    theme_void() +
    labs(title = sprintf("Cut %d: %d townships into %d + %d", k, sum(inside), sizes[1], sizes[2]))
})
patchwork::wrap_plots(cut_maps, ncol = 3)
```

The first cut separates a band of 19 townships in the north and east from the other 36. The second
splits the 36 into 24 and 12. The last three cuts each separate one or two townships, which are the
small regions in the table in section 3.2.4.
:::

### SKATER map

The SKATER regions are attached to the polygons and mapped, beside the hierarchical clusters of
section 3.1.12, to compare the two.

```{r}
#| fig-height: 5
groups_mat <- as.matrix(clust6$groups)
shan_sf_spatialcluster <- cbind(shan_sf_cluster, as.factor(groups_mat)) |>
  rename(SP_CLUSTER = as.factor.groups_mat.)
qtm(shan_sf_spatialcluster, "SP_CLUSTER")
```

The chapter places the two maps side by side.

```{r}
#| fig-height: 5
hclust.map <- qtm(shan_sf_cluster,
                  "CLUSTER") +
  tm_borders(col_alpha = 0.5)

shclust.map <- qtm(shan_sf_spatialcluster,
                   "SP_CLUSTER") +
  tm_borders(col_alpha = 0.5)

tmap_arrange(hclust.map, shclust.map,
             asp = NA, ncol = 2)
```

::: {.callout-note collapse="false" title="How the code works"}
- `cbind()` adds the SKATER region of each township to `shan_sf_cluster` as a new column, and
  `rename()` names it `SP_CLUSTER`, as in section 3.1.12.
- The chapter writes `tm_borders(alpha = 0.5)`. The reason for `col_alpha = 0.5` under tmap 4 is
  given in section 3.1.2.
:::

The two maps show the difference on slide 5: every SKATER region is one connected piece, and the
hierarchical clusters are scattered.

## ClustGeo

ClustGeo (Chavent et al., 2018) adds a second dissimilarity, D1, measured in geographic space, to
the dissimilarity D0 in the space of the variables, and minimises a weighted combination of the two.
The weight on D1 is the mixing parameter α:

- at α = 0, only the variables count, and the result is ordinary Ward clustering;
- at α = 1, only geography counts, and the clusters are compact blocks of territory whatever their
  ICT rates;
- in between, a larger α makes the clusters more compact and less homogeneous.

The paper calls this a soft contiguity constraint. Townships that are not neighbours may still end
up in the same cluster.

### Ward-like clustering without a constraint

With α = 0, ClustGeo is ordinary Ward clustering. Running it first gives the starting point that the
geographic constraint then changes.

```{r}
#| fig-height: 6
nongeo_cluster <- hclustgeo(proxmat)
plot(nongeo_cluster, cex = 0.5)
rect.hclust(nongeo_cluster, k = 6, border = 1:6)
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses one function:

```r
hclustgeo(D0, D1 = NULL, alpha = 0, scale = TRUE, wt = NULL)  # ClustGeo
```

- With only `D0`, `hclustgeo()` runs Ward-like clustering on that matrix. `D0` must be a `dist`
  object.
- According to Chavent et al. (2018, section 2.2), `hclustgeo()` is a wrapper of
  `hclust(method = "ward.D")` applied to the squared dissimilarities divided by 2n. The chapter
  describes it as the same as the `hclust()` call of section 3.1.7. The two produce different trees,
  because `ward.D` is applied to unsquared distances in section 3.1.7.
:::

```{r}
#| fig-height: 5
groups <- as.factor(cutree(nongeo_cluster, k = 6))
shan_sf_ngeo_cluster <- cbind(shan_sf, as.matrix(groups)) |>
  rename(CLUSTER = as.matrix.groups.)
qtm(shan_sf_ngeo_cluster, "CLUSTER")
```

The map is as fragmented as the map in section 3.1.12, and the cluster sizes differ from
it: `r paste(sort(table(shan_sf_ngeo_cluster$CLUSTER), decreasing = TRUE), collapse = ", ")` here
against `r paste(sort(table(shan_sf_cluster$CLUSTER), decreasing = TRUE), collapse = ", ")` there.

### Geographic distances and the choice of α

ClustGeo needs a second matrix, D1, of geographic dissimilarities, and a value of α. `st_distance()`
supplies D1, and `choicealpha()` measures, for each α, how much homogeneity in the ICT rates is
given up for how much geographic homogeneity.

```{r}
dist <- st_distance(shan_sf, shan_sf)
distmat <- as.dist(dist)
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses two functions:

```r
st_distance(x, y)                                         # sf
as.dist(m)                                                # stats
```

- `st_distance()` returns the matrix of distances between every pair of features. For polygons, the
  distance is the shortest distance between their boundaries. For longitude and latitude it is
  computed on the sphere, in metres.
- `as.dist()` converts the square matrix to a `dist` object, as `hclustgeo()` requires.
- The chapter names the object `dist`, the same name as the function `dist()`. R still finds the
  function when it is called, because it looks for a function by that name, but the name is easy to
  misread.

A township and its neighbour touch, so the shortest distance between them is zero.
:::

```{r}
#| fig-height: 4.5
cr <- choicealpha(proxmat, distmat, range.alpha = seq(0, 1, 0.1), K = 6, graph = TRUE)
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses one function:

```r
choicealpha(D0, D1, range.alpha, K, wt = NULL, scale = TRUE, graph = TRUE)  # ClustGeo
```

- `choicealpha()` builds the partition into `K` clusters for each α in `range.alpha`, and measures
  for each partition the proportion of the total pseudo-inertia (a generalised sum of squares)
  explained in D0 space and in D1 space. `graph = TRUE` draws both, first as the raw proportions, Q,
  then as proportions of their own maximum, Qnorm.
:::

```{r}
round(cbind(cr$Q, cr$Qnorm), 3)
```

The chapter uses α = 0.2, read from the charts. Slide 15 of the lecture gives the reasoning with
numbers from this data set: at α = 0 the partition explains 0.76 of the ICT variation, and at α = 1
the geographic partition explains 0.94 of the geographic variation. At α = 0.2 the ICT proportion
falls by 0.11 to 0.65 and the geographic proportion rises by 0.49 to 0.63. The table printed above
gives the same values.

::: callout-note
## How to read the `choicealpha()` charts

Both charts have α on the horizontal axis and one line for each matrix:

- The black line, Q0, is the share of the variation in the ICT rates (D0) that the six clusters
  explain. It is highest at α = 0, where only the rates count, and falls as α rises.
- The red dashed line, Q1, is the share of the geographic variation (D1) that the six clusters
  explain. It is lowest at α = 0 and rises with α.

Each share is one minus the within-cluster sum of squared dissimilarities divided by the total, the
quantity that `choicealpha()` calls the proportion of explained pseudo-inertia:

$$Q = 1 - \frac{\text{within-cluster sum of squared dissimilarities}}{\text{total sum of squared dissimilarities}}$$

The share of variation explained in Part 2 is the same quantity, computed on the z-scores.

The first chart draws Q0 and Q1 themselves. The second, Qnorm, divides each line by its own maximum,
so both lines start from 1 and the two losses can be compared directly. A useful α is one where Q1
has risen a long way while Q0 has fallen only a little. The printed table has the four lines as
columns, Q0, Q1, Q0norm and Q1norm, with one row for each α.
:::

### ClustGeo regions

With α chosen, the tree is built again from both matrices and cut into six clusters.

```{r}
#| fig-height: 5
clustG <- hclustgeo(proxmat, distmat, alpha = 0.2)
groups <- as.factor(cutree(clustG, k = 6))
shan_sf_Gcluster <- cbind(shan_sf, as.matrix(groups)) |>
  rename(CLUSTER = as.matrix.groups.)
qtm(shan_sf_Gcluster, "CLUSTER")
```

The clusters are more compact than at α = 0, and several of them are still split. Cluster 4 in this
printout, for instance, has townships in the north-west and in the north, a strip in the south-west,
and one township in the south-east. The split follows from the soft constraint of Chavent et al.
(2018): geography has a weight in the criterion, and a split is still allowed.

::: callout-note
## What the mixing parameter α does

The four maps cut the ClustGeo tree into six clusters at four values of α, with the chapter's two
matrices. At α = 0 the clusters are the same as those of section 3.3.1. As α rises, geography has
more weight: the clusters become compact, and at α = 1 they are six blocks of territory formed from
distance alone, whatever the ICT rates.

```{r}
#| echo: false
#| fig-height: 7
alpha_maps <- map(c(0, 0.2, 0.5, 1), \(a) {
  g <- cutree(hclustgeo(proxmat, distmat, alpha = a), 6)
  ggplot() + geom_sf(data = mutate(shan_sf, g = factor(g)), aes(fill = g),
                     colour = "white", linewidth = 0.2) +
    scale_fill_brewer(palette = "Set2", guide = "none") + theme_void() +
    labs(title = paste0("alpha = ", a))
})
patchwork::wrap_plots(alpha_maps, ncol = 2)
```
:::

## Cluster profiles

In section 12.10, the clusters are described by their ICT rates. The chapter profiles the
non-spatial `hclustgeo()` clusters of section 3.3.1, and the same clusters are profiled here. The
same profile for a set of contiguous regions is in section 4.8.

```{r}
#| fig-height: 3.5
ggplot(data = shan_sf_ngeo_cluster, aes(x = CLUSTER, y = RADIO_PR)) +
  geom_boxplot()
```

Each box is the spread of the radio rate across the townships of one cluster, and the line inside it
is the median. By median, the clusters
rank `r paste(names(sort(tapply(shan_sf_ngeo_cluster$RADIO_PR, shan_sf_ngeo_cluster$CLUSTER, median), decreasing = TRUE)), collapse = ", ")`
on radio, the order in the chapter. The chapter calls it the order of the mean, which a boxplot does
not show: by mean, the order
is `r paste(names(sort(tapply(shan_sf_ngeo_cluster$RADIO_PR, shan_sf_ngeo_cluster$CLUSTER, mean), decreasing = TRUE)), collapse = ", ")`
(the table of means below).

```{r}
#| fig-height: 5
ggparcoord(data = shan_sf_ngeo_cluster,
           columns = c(17:21),
           scale = "globalminmax",
           alphaLines = 0.2,
           boxplot = TRUE,
           title = "Multiple Parallel Coordinates Plots of ICT Variables by Cluster") +
  facet_grid(~ CLUSTER) +
  theme(axis.text.x = element_text(angle = 30))
```

::: {.callout-note collapse="false" title="How the code works"}
The two blocks use three new functions:

```r
geom_boxplot()                                            # ggplot2
ggparcoord(data, columns, scale, alphaLines, boxplot, title)  # GGally
facet_grid(rows ~ cols)                                   # ggplot2
```


- `columns = c(17:21)` selects the five rates by position. In `shan_sf_ngeo_cluster`, columns 17 to
  21 are `RADIO_PR` to `COMPUTER_PR`, as `names(shan_sf)` printed in section 2.4 shows.
- `scale = "globalminmax"` draws every variable on one common axis without rescaling. Variables with
  larger values, TV and mobile phone, take up most of the height, and computer is compressed near
  zero, which reflects the same difference in scale described in section 3.1.5. `alphaLines = 0.2`
  makes the lines transparent, and `boxplot = TRUE` adds a boxplot on each axis.
- `facet_grid(~ CLUSTER)` draws one panel per cluster.
:::

::: callout-note
## How to read the parallel coordinate plot

A parallel coordinate plot draws each township as one line. The five rates are five vertical axes
side by side, all on the same scale, and a township's line crosses each axis at the height of its
rate. Three townships:

```{r}
#| echo: false
#| fig-height: 3.4
pc_rows <- c("Taunggyi", "Pindaya", "Matman")
as_tibble(shan_ict, rownames = "township") |>
  filter(township %in% pc_rows) |>
  pivot_longer(-township, names_to = "rate", values_to = "value") |>
  mutate(rate = factor(rate, levels = names(shan_ict)),
         township = factor(township, levels = pc_rows)) |>
  ggplot(aes(rate, value, group = township, colour = township)) +
  geom_vline(xintercept = 1:5, colour = "grey60") +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  ggrepel::geom_text_repel(aes(label = round(value)), size = 3, direction = "y", nudge_x = 0.15,
                           hjust = 0, segment.colour = NA, show.legend = FALSE) +
  scale_colour_brewer(palette = "Dark2") +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.1))) +
  labs(x = NULL, y = "households per thousand", colour = NULL)
```

Taunggyi's line is high on TV and mobile phone. Pindaya's line is highest on radio and low on mobile
phone. Matman's line is low on every axis. Townships with similar rates have lines of similar shape.

The plot above draws all 55 townships this way, one panel per cluster, with three additions:

- The lines are transparent, so a dark band is a place where many lines overlap.
- Each axis has a boxplot of the townships in the cluster: the thick line is the median, the box the
  middle half of the townships, and a dot a township far from the rest of its cluster.
- Every panel has the same vertical scale, so the boxes on one axis can be compared across panels by
  their height.

The boxes from left to right give the typical profile of a cluster, and one axis compared across the
panels shows which cluster is high or low on that rate. In panel 4, the TV and mobile phone boxes
are the highest of all six panels, with medians
of `r round(median(shan_sf_ngeo_cluster$TV_PR[shan_sf_ngeo_cluster$CLUSTER == "4"]))`
and `r round(median(shan_sf_ngeo_cluster$MPHONE_PR[shan_sf_ngeo_cluster$CLUSTER == "4"]))`
households per thousand. In panel 5, every box is near the bottom of the scale.
:::

```{r}
shan_sf_ngeo_cluster |>
  st_set_geometry(NULL) |>
  group_by(CLUSTER) |>
  summarise(n = n(),
            mean_RADIO_PR = mean(RADIO_PR),
            mean_TV_PR = mean(TV_PR),
            mean_LLPHONE_PR = mean(LLPHONE_PR),
            mean_MPHONE_PR = mean(MPHONE_PR),
            mean_COMPUTER_PR = mean(COMPUTER_PR)) |>
  kable(digits = 1)
```

::: {.callout-note collapse="false" title="How the code works"}
- `group_by()` splits the table by cluster, and `summarise()` computes one row per cluster: the
  count and the mean of each rate. `kable(digits = 1)` prints the table rounded to one decimal.
:::

Cluster numbers are arbitrary labels, so the chapter's reading of "cluster 4" holds only where the
numbering is the same as in the chapter. In this run it is: cluster 4 is a group of six townships
with the highest TV, land line, mobile phone and computer rates. The chapter describes cluster 5 as
owning "the lowest of all the five ICT". According to the means, that holds for TV, mobile phone and
computer only: cluster 6 has lower radio ownership than cluster 5, and cluster 2 lower land line
ownership.

# Part 2: Review and extension

Part 1 reproduced chapter 12. This part examines the choices in its method, and adds the methods on
the Lesson 6 slides that the chapter does not cover.

| Section | What I do | Source |
|----|--------------------------------------|------------|
| | **Checking the chapter's method** | |
| 4.1 | Test each clustering variable for spatial autocorrelation | de Souza and Taconeli (2022); Lesson 5 |
| 4.2 | Measure how much each variable weighs in the distance, and cluster on standardised values | chapter 12.7.2 to 12.7.5; Assunção et al. (2006); Rovan and Sambt (2003); UC-R tutorial |
| 4.3 | Compare `ward.D` with `ward.D2` | Chavent et al. (2018), section 2.2; UC-R tutorial |
| 4.4 | Examine the evidence for six clusters | chapter 12.7.9; UC-R tutorial; de Souza and Taconeli (2022); Rovan and Sambt (2003); Demeter and Brătucu (2013); Brown and Watson (2012) |
| 4.5 | Explain the single-township SKATER regions and set a minimum number of households | Assunção et al. (2006), section 4 |
| 4.6 | Test the geographic matrix D1 used for ClustGeo | Chavent et al. (2018), sections 4.3 and 4.4; Jaya et al. (2019) |
| | **Adding what the chapter leaves out** | |
| 4.7 | Run REDCAP, AZP and max-p, and SCHC from `rgeoda` | slides 2 and 10 to 12 |
| 4.8 | Compare all the regionalisations on homogeneity and contiguity, and profile the chosen one | slides 4 and 5; de Souza and Taconeli (2022); Brown and Watson (2012) |

Two measures are used to compare the sets of clusters in Part 2:

$$\text{share explained} = 1 - \frac{\text{within-cluster sum of squares}}{\text{total sum of squares}}$$

| Measure | What it is | How to read it | Source |
|----|-----|-----|---------|
| **Share of variation explained** | the formula above, on the standardised variables | 0: the clusters explain nothing. 1: every cluster is internally identical | Assunção et al. (2006, equation 2), as the within-region sum of squares, the `$ssw` of `skater()` in section 3.2.4. Chavent et al. (2018) and de Souza and Taconeli (2022), as the proportion of explained inertia, Q0 on slide 15 (the same number with Euclidean distances and equal weights). `rgeoda`, as "the ratio of between to total sum of squares" |
| **Number of pieces** | the count of separate connected pieces of territory that the clusters are in | 6 for six contiguous regions; more pieces, more fragmented clusters | added on this page, and not used in the readings. It measures the contiguity of slides 3 and 5, with `n.comp.nb()` from `spdep` |

```{r}
cluster_map <- function(data, var = "CLUSTER", title = NULL) {
  tm_shape(data) +
    tm_polygons(fill = var,
                fill.scale = tm_scale_categorical(values = "brewer.set2"),
                fill.legend = tm_legend(title = "Cluster")) +
    tm_title(title)
}

ss_ratio <- function(X, groups) {
  X <- as.matrix(X)
  total <- sum(scale(X, scale = FALSE)^2)
  within <- sum(sapply(split(as.data.frame(X), groups),
                       \(d) sum(scale(as.matrix(d), scale = FALSE)^2)))
  1 - within / total
}

count_pieces <- function(groups, nb = shan.nb) {
  sum(sapply(unique(groups), \(g) {
    if (sum(groups == g) == 1) return(1)
    n.comp.nb(subset(nb, groups == g))$nc
  }))
}

c(explained = ss_ratio(shan_ict.z, shan_sf_cluster$CLUSTER),
  pieces = count_pieces(shan_sf_cluster$CLUSTER))
```

::: {.callout-note collapse="false" title="How the code works"}
The block defines three functions, using these:

```r
tm_scale_categorical(values)                              # tmap
tm_title(text)                                            # tmap
scale(x, center = TRUE, scale = FALSE)                    # base R
split(x, f)                                               # base R
subset(x, subset)                                         # spdep, for nb objects
n.comp.nb(nb.obj)                                         # spdep
```

- `cluster_map()` draws a map of cluster labels with a fixed categorical palette and a title. Part 2
  has many sets of clusters, and the function draws all their maps in one format.
- `ss_ratio()`: `scale(X, scale = FALSE)` subtracts each column mean, so the sum of the squared
  result is the total sum of squares. `split()` divides the rows by cluster, and the same sum
  computed within each cluster gives the within-cluster sum of squares. The ratio is one minus their
  quotient.
- `count_pieces()`: for each cluster, `subset()` keeps only the neighbour links between townships in
  that cluster, and `n.comp.nb()` counts the connected components of what remains. A single township
  is one piece. The sum over clusters is the number of pieces.

For the hierarchical clusters of section 3.1.12, the six clusters explain 61% of the standardised
variation and are in 24 pieces.
:::

## Spatial autocorrelation of the clustering variables

SKATER and ClustGeo in sections 3.2 and 3.3 apply a spatial constraint to all five rates at once,
and the chapter attributes the fragmented map of section 3.1.12 to clustering without regard to
location. Part 1 does not test whether the rates themselves are spatially patterned. If a variable
varies across the state without regard to location, a contiguity constraint can only group townships
that differ on it.

De Souza and Taconeli (2022) apply Moran's I to each of their six educational indicators before
clustering, and report significant spatial autocorrelation for all six, which is their ground for
using spatial algorithms at all. The test is the permutation test of Lesson 5, on Queen contiguity
with row-standardised weights. The question here is which of the six ICT rates are spatially
autocorrelated.

```{r}
set.seed(1234)
lw_shan <- nb2listw(shan.nb, style = "W")
rates <- c("RADIO_PR", "TV_PR", "LLPHONE_PR", "MPHONE_PR", "COMPUTER_PR", "INTERNET_PR")

moran_by_rate <- map_dfr(rates, \(v) {
  m <- moran.mc(shan_sf[[v]], lw_shan, nsim = 999)
  tibble(rate = v, moran_I = unname(m$statistic), p_value = m$p.value)
})
kable(moran_by_rate, digits = 3)
```

::: {.callout-note collapse="false" title="How the code works"}
```r
nb2listw(neighbours, style = "W")                         # spdep
moran.mc(x, listw, nsim)                                  # spdep
map_dfr(.x, .f)                                           # purrr
```

- `nb2listw(style = "W")` turns the Queen neighbour list of section 3.2.1 into row-standardised
  weights, as in Exercise 5.
- `moran.mc()` computes Moran's I and compares it with 999 values obtained by permuting the rates
  across townships. The p-value is the rank of the observed value among the 1,000.
- `map_dfr()` runs the test once for each rate and stacks the six one-row tables.
:::

Only two rates are significantly autocorrelated at the 5% level: radio (I = 0.44) and land line (I =
0.38). Mobile phone is borderline (p = 0.055), and TV, computer and internet show no spatial pattern
that can be distinguished from a random arrangement of the same values.

Unlike de Souza and Taconeli's indicators, then, most of the ICT rates are not spatially structured.
TV and mobile phone, the two variables with the most weight in the raw distances (section 4.2), are
among those without spatial pattern, which is consistent with the fragmented map of section 3.1.12.
A regionalisation of these data can also be expected to give up a substantial share of homogeneity,
because under the constraint, neighbouring townships share a region even where their TV, mobile
phone and computer rates are unrelated.

## Scale of the clustering variables

Two standardised versions of the variables were computed in section 3.1.5, and the raw rates were
used for every clustering step after it. The chapter's own reason for standardising, in section
12.7.2, is to avoid a result "biased to clustering variables with large values". The course's own
sources all include the step. The UC Business Analytics tutorial linked under "All About R", from
which chapter sections 12.7.8 to 12.7.10 are taken, says that "the data must be standardized (i.e.,
scaled) to make variables comparable" and applies `scale()` before `dist()`. Assunção et al. (2006,
section 3.2) define the SKATER edge cost for attributes with "comparable scales, such as when they
are measured in standard deviation units", and both Rovan and Sambt (2003) and de Souza and Taconeli
(2022) standardise every variable before clustering. The bias that the omission introduces is
measured below, and the clustering is repeated on standardised variables.

Euclidean distance adds up squared differences. Summed over all pairs of townships, the squared
differences on one variable are proportional to its variance. The share of each variable in the
total is therefore its share of the summed variance.

```{r}
#| fig-height: 3.5
share <- tibble(variable = names(shan_ict),
                `Raw rates` = sapply(shan_ict, var) / sum(sapply(shan_ict, var)),
                `Min-max` = apply(shan_ict.std, 2, var) / sum(apply(shan_ict.std, 2, var)),
                `z-score` = 1 / 5) |>
  pivot_longer(-variable, names_to = "scale", values_to = "share")

ggplot(share, aes(x = share, y = variable)) +
  geom_col(fill = "steelblue") +
  facet_wrap(~ scale) +
  scale_x_continuous(labels = scales::percent) +
  labs(x = "share of the squared Euclidean distance", y = NULL)
```

::: {.callout-note collapse="false" title="How the code works"}
- `sapply(shan_ict, var)` computes the variance of each column. Dividing by the sum gives each
  variable's share. For z-scores every variance is 1, so every share is one fifth.
- `pivot_longer()` reshapes the three share columns into rows, one row per variable and scale, which
  `facet_wrap()` needs to draw one panel per scale.
:::

```{r}
round(100 * sapply(shan_ict, var) / sum(sapply(shan_ict, var)), 1)
```

On the raw rates, mobile phone and TV together make up 81% of the distance between townships, and
computer 0.6%. The six hierarchical clusters of section 3.1 are therefore mostly a grouping on two
variables.

- Internet was removed in section 12.6 because it is correlated with computer at 0.87. On the raw
  scale, computer itself has almost no weight in the distance, so dropping internet changes the
  distances very little, although the weakly separated clusters still change (section 3.1.3).
- Land line (2.4%) has little weight either, although it is the variable on which three adjacent
  townships in Shan (North) differ most from the rest (section 4.8).
- Radio and land line, the only two rates with significant spatial autocorrelation in section 4.1,
  together account for 18% of the raw distance. On the raw scale the clustering is driven by the
  variables with the least spatial pattern.

After min-max scaling the five shares lie between 18% and 24%, and after z-scores they are equal.
The clustering is repeated on z-scores below. The Ward method used is `ward.D2`, for the reason
given in section 4.3.

```{r}
#| fig-height: 5
h_raw <- cutree(hclust(dist(shan_ict), method = "ward.D2"), k = 6)
h_z <- cutree(hclust(dist(shan_ict.z), method = "ward.D2"), k = 6)
h_mm <- cutree(hclust(dist(shan_ict.std), method = "ward.D2"), k = 6)

tibble(comparison = c("raw against z-score", "raw against min-max", "z-score against min-max"),
       ARI = c(adjustedRandIndex(h_raw, h_z), adjustedRandIndex(h_raw, h_mm),
               adjustedRandIndex(h_z, h_mm))) |>
  kable(digits = 2)
```

::: {.callout-note collapse="false" title="How the code works"}
```r
adjustedRandIndex(x, y)                                   # mclust
```

- `adjustedRandIndex()` takes two vectors of cluster labels for the same townships and returns the
  adjusted Rand index.
:::

::: callout-note
## What the adjusted Rand index measures

The adjusted Rand index compares two clusterings of the same townships. It is computed over every
pair of townships. A pair is a match when the two townships are in the same cluster in both
clusterings, or in different clusters in both. The share of pairs that match is the Rand index. The
adjusted index subtracts the share of matches expected between two random clusterings with the same
cluster sizes and rescales the result, so that it is 1 for identical clusterings and close to 0 for
clusterings that match no better than chance.

Six townships, A to F, in three clusterings:

| | A | B | C | D | E | F |
|---|---|---|---|---|---|---|
| Clustering 1 | 1 | 1 | 1 | 2 | 2 | 2 |
| Clustering 2 | 2 | 2 | 2 | 1 | 1 | 1 |
| Clustering 3 | 1 | 1 | 2 | 2 | 2 | 2 |

- Clustering 2 has the same two groups as clustering 1, with the numbers swapped. All 15 pairs
  match, and the index is `r adjustedRandIndex(c(1, 1, 1, 2, 2, 2), c(2, 2, 2, 1, 1, 1))`. The
  numbers used as labels make no difference.
- In clustering 3, C is in the other group. Ten of the 15 pairs still match, a Rand index of 0.67.
  Two random clusterings with these cluster sizes would also match on many pairs, and the adjusted
  index is `r round(adjustedRandIndex(c(1, 1, 1, 2, 2, 2), c(1, 1, 2, 2, 2, 2)), 2)`.
:::

```{r}
#| fig-height: 5
tmap_arrange(
  cluster_map(mutate(shan_sf, CLUSTER = factor(h_raw)), title = "Raw rates"),
  cluster_map(mutate(shan_sf, CLUSTER = factor(h_z)), title = "z-scores"),
  ncol = 2)
```

The raw and z-score clusterings have an adjusted Rand index
of `r round(adjustedRandIndex(h_raw, h_z), 2)`. Standardisation changes the grouping substantially.
The choice between the two standardisations matters much less: z-score and min-max give almost the
same clusters (index `r round(adjustedRandIndex(h_z, h_mm), 2)`).

In section 12.7.4, the chapter warns that z-score standardisation "should only be used if we would
assume all variables come from some normal distribution". The warning does not hold:

- A z-score is a linear rescaling and assumes nothing about the shape of the distribution. The three
  histograms in section 3.1.5 have the same shape.
- Rovan and Sambt (2003, section 3) standardise all eight of their variables before Ward clustering,
  and note that "the method employed in this study does not make any distributional assumptions, so
  no other transformation of the data has been performed".

The two methods differ for an extreme township:

| | z-score | min-max |
|---|---|---|
| The largest value becomes | its distance from the mean, in standard deviations | 1 |
| An extreme township | moves the mean and the standard deviation | sets the scale of the whole variable on its own |
| Used in | the UC Business Analytics tutorial; Rovan and Sambt (2003) | de Souza and Taconeli (2022) |

Z-scores are used in the rest of Part 2.

## Ward's criterion: ward.D and ward.D2

Ward's method merges, at each step, the two clusters whose union increases the within-cluster sum of
squares by the least. The criterion is defined on squared distances, and `hclust()` has two options
for it. The chapter uses `ward.D` in section 3.1.7. With the ordinary Euclidean distances in
`proxmat`, only `ward.D2` is Ward's method:

| Call | What it does with the Euclidean distances in `proxmat` | Ward's method? |
|-------|--------------------------------------|------|
| `hclust(method = "ward.D")`, section 3.1.7 | applies Ward's merging rule to the distances as given, without squaring them | no |
| `hclust(method = "ward.D2")` | squares the distances first, then applies the rule | yes |
| `agnes(method = "ward")`, section 3.1.8 | builds the same tree as `ward.D2`, with the same merge heights | yes |
| `hclustgeo(proxmat)`, section 3.3.1 | runs `ward.D` on the squared distances divided by 2n (Chavent et al., 2018, section 2.2) | yes |

The six clusters from each call, compared with `ward.D2`:

```{r}
tree_D <- hclust(proxmat, method = "ward.D")
tree_D2 <- hclust(proxmat, method = "ward.D2")
tree_agnes <- as.hclust(agnes(shan_ict, method = "ward"))
tree_geo <- hclustgeo(proxmat)

six <- lapply(list(`ward.D` = tree_D, `ward.D2` = tree_D2, `agnes(ward)` = tree_agnes,
                   hclustgeo = tree_geo), cutree, k = 6)
tibble(call = names(six),
       `cluster sizes` = sapply(six, \(g) paste(sort(table(g), decreasing = TRUE), collapse = ", ")),
       `adjusted Rand index with ward.D2` = sapply(six, adjustedRandIndex, six$ward.D2)) |>
  kable(digits = 2)
```

::: {.callout-note collapse="false" title="How the code works"}
- `as.hclust()` converts the `agnes()` tree to the same class as the other three, so that `cutree()`
  works on all four.
- `lapply()` cuts each tree into six clusters. `sapply()` then gives, for each call, the cluster
  sizes and the adjusted Rand index with the `ward.D2` clusters (section 4.2), which is 1 for the
  same six clusters.
:::

What this means:

- **The chapter's six clusters are not Ward clusters.** The map in section 3.1.12 is of the `ward.D`
  tree, whose cluster sizes differ from those of Ward's method.
- **The agglomerative coefficient belongs to a different tree.** The 0.943 that the chapter uses to
  choose Ward is computed by `agnes()`, so it is the coefficient of the `ward.D2` tree. The map in
  section 3.1.12 is of the `ward.D` tree.
- **`hclustgeo()` is Ward's method.** Its clusters in section 3.3.1 differ from those of section
  3.1.7 because it is `ward.D2` underneath, on the same distances.
- **Part 2 uses `ward.D2`**, as does the UC Business Analytics tutorial, the source of chapter
  sections 12.7.8 to 12.7.10.

## Number of clusters

The gap statistic in section 3.1.9 has its first maximum at one cluster, and the chapter retained
six. A number of clusters can be chosen from the data, by a statistic that compares the fit for each
k, or by the purpose of the analysis, for instance a planning agency that manages six regions. The
question here is whether six clusters can be justified from the standardised variables of section
4.2.

The criteria computed below are:

- the **gap statistic**, with the "one standard error" rule of Tibshirani et al., which picks the
  smallest k whose gap is within one standard error of the best;
- the **average silhouette width**, which for each township compares its average distance to its own
  cluster with its average distance to the nearest other cluster, and ranges from −1 to 1. It is one
  of the three methods listed in chapter section 12.7.9, where only the gap statistic is computed,
  and the UC Business Analytics tutorial computes it with `fviz_nbclust(method = "silhouette")`;
- the **NbClust** majority vote over its indices. The chapter loads `NbClust` and does not run it.
  De Souza and Taconeli (2022, section 3.3) choose their number of clusters this way, from 23
  criteria.

The one-standard-error rule and the silhouette threshold of 0.25 used below come from the
cluster-analysis literature (Tibshirani et al., 2001; Kaufman and Rousseeuw, 1990) and are not part
of the course material.

```{r}
set.seed(12345)
gap_z <- clusGap(shan_ict.z, FUN = hcut, K.max = 10, B = 50)
tab <- gap_z$Tab
c(Tibs2001SEmax = maxSE(tab[, "gap"], tab[, "SE.sim"], method = "Tibs2001SEmax"),
  firstmax = maxSE(tab[, "gap"], tab[, "SE.sim"], method = "firstmax"),
  globalmax = maxSE(tab[, "gap"], tab[, "SE.sim"], method = "globalmax"))
```

```{r}
#| fig-height: 3.2
tree_z <- hclust(dist(shan_ict.z), method = "ward.D2")
sil <- sapply(2:10, \(k) mean(silhouette(cutree(tree_z, k), dist(shan_ict.z))[, "sil_width"]))

tibble(k = 2:10, silhouette = sil) |>
  ggplot(aes(k, silhouette)) +
  geom_line() + geom_point() +
  scale_x_continuous(breaks = 2:10) +
  labs(y = "average silhouette width")
```

```{r}
#| results: hide
#| fig-show: hide
nb_z <- NbClust(shan_ict.z, distance = "euclidean", min.nc = 2, max.nc = 10,
                method = "ward.D2", index = "all")
```

```{r}
table(nb_z$Best.nc["Number_clusters", ])
```

::: {.callout-note collapse="false" title="How the code works"}
The blocks use three functions:

```r
maxSE(f, SE.f, method)                                    # cluster
silhouette(x, dist)                                       # cluster
NbClust(data, distance, min.nc, max.nc, method, index)    # NbClust
```

- `maxSE()` applies a rule to the column of gaps and their standard errors and returns the chosen k.
  The three rules differ in how values close to the maximum are counted.
- `silhouette()` takes the cluster labels and the distance matrix and returns the silhouette width
  of each township. The mean over townships is computed for each k from 2 to 10.
- `NbClust()` computes up to 30 indices for each k and reports the k that scores best on each index.
  `index = "all"` runs them all. It prints a long commentary and draws charts, which the chunk
  options `results: hide` and `fig-show: hide` suppress. `Best.nc` holds the k chosen by each index,
  and `table()` counts the votes.
:::

::: callout-note
## What the silhouette width measures

For one township, a is its average distance to the other townships in its own cluster, and b is its
average distance to the townships of the nearest other cluster. The silhouette width is (b − a) /
max(a, b). It is close to 1 when the township is well inside its cluster, close to 0 when it is on
the border between two clusters, and negative when it is closer to another cluster than to its own.
The average over all townships summarises the whole partition.

The figure shows the widths of the 55 townships for six clusters on the z-scores, grouped by
cluster. Many are close to zero, and some are negative.

```{r}
#| echo: false
#| fig-height: 3.4
fviz_silhouette(silhouette(cutree(tree_z, 6), dist(shan_ict.z)), print.summary = FALSE) +
  labs(title = NULL) + theme(axis.text.x = element_blank())
```
:::

On the z-scores:

- the one-standard-error and first-maximum rules of the gap statistic give **k = 1**, and **10**,
  the largest value tried, under the global maximum. The gap at k = 1 (`r round(tab[1, "gap"], 3)`)
  is not exceeded until k = 10 (`r round(tab[10, "gap"], 3)`), by much less than one standard error
  (about 0.04);
- the silhouette width is highest at **k = 2**, `r round(sil[1], 2)`, and stays between 0.22 and
  0.27 for every k. Kaufman and Rousseeuw's rule of thumb reads a width below 0.25 as no substantial
  structure;
- NbClust's majority vote is **2** (`r max(table(nb_z$Best.nc["Number_clusters", ]))` indices), with
  six chosen by `r sum(nb_z$Best.nc["Number_clusters", ] == 6)`.

The authors of the three references that fix a number of clusters use simpler methods. Rovan and
Sambt (2003, section 3.1) read the Ward dendrogram for "big jumps of the values of the between-group
sum of squares", and find two large jumps, at two and at four groups. Demeter and Brătucu (2013)
read cluster membership at two, three and four clusters from one tree and choose three for
interpretation. Brown and Watson (2012) apply the rule of thumb k = √(n/2), which for 55 townships
gives `r round(sqrt(55 / 2), 1)`. The first two readings are applied below to the Ward tree on
z-scores.

```{r}
#| fig-height: 3.2
merge_heights <- tibble(k = 2:10, height = rev(tree_z$height)[1:9]) |>
  mutate(jump = height - lead(height))

ggplot(merge_heights, aes(k, height)) +
  geom_line() + geom_point() +
  geom_text(aes(label = round(jump, 2)), vjust = -0.8, size = 3, na.rm = TRUE) +
  scale_x_continuous(breaks = 2:10) +
  labs(y = "height of the merge that joins k clusters into k - 1")

nested <- sapply(2:6, \(k) cutree(tree_z, k))
colnames(nested) <- paste0("k", 2:6)
as_tibble(nested) |> count(k2, k3, k4, k5, k6)
```

::: {.callout-note collapse="false" title="How the code works"}
- `tree_z$height` holds the height of each merge in the order the merges were made, so its last
  value is the merge that joins the final two clusters. `rev()` reverses it, so element k − 1 is the
  height of the merge that joins k clusters into k − 1. `lead()` takes the next value, so `jump` is
  how much higher that merge is than the one before it, which joins k + 1 clusters into k. A large
  jump means that stopping at k clusters avoids an expensive merge.
- `cutree()` for k = 2 to 6 gives five columns of labels from the same tree, and `count()` lists
  each combination that occurs. Each row follows one group of townships down the tree, which is the
  reading of cluster membership that Demeter and Brătucu (2013) make at two, three and four
  clusters.
:::

The merge heights have secondary jumps at four and at six clusters. The largest jump is at two
clusters (`r round(merge_heights$jump[1], 2)`), and the next largest are at four
(`r round(merge_heights$jump[3], 2)`) and at six (`r round(merge_heights$jump[5], 2)`). From seven
clusters on, no jump exceeds `r round(max(merge_heights$jump[6:8]), 2)`. The nested table shows how
six arises: each of the two top-level groups divides into three, and no later split undoes an
earlier one.

The gap statistic gives one cluster; the silhouette width, NbClust and the largest merge-height jump
give two; the rule of thumb gives five; and the secondary merge-height jumps give four or six. The
ICT profiles of Shan's townships vary continuously, with weak separation between groups at any k, so
the number of regions cannot be settled by the data alone. De Souza and Taconeli (2022, section 3.3)
fix one k from the unconstrained Ward tree and use it "for all clustering algorithms henceforward",
so that the methods can be compared on equal terms. k = 6 is used in the same way in the rest of
Part 2: it is one of the break points of the tree, and it keeps the comparison with the chapter.

## Singleton regions in SKATER

Two of the six SKATER regions of section 3.2.4 are single townships. SKATER cuts the edge whose
removal most reduces the within-region sum of squares. A township whose values are far from all of
its neighbours' is expensive to keep attached, so cutting it off as a region of one is often the
best single cut. Assunção et al. (2006, section 4) met the same outcome in their own case study.
Their unrestricted regionalisation of the 96 districts of São Paulo into eight regions left "the
north-eastern region" with "one district only", 0.12% of the city's population, beside a region of
36 districts, and they note that "for many applications, this unbalance is not convenient". Their
remedy was a restriction on the regions: a minimum population of 500,000 inhabitants, under which
region populations ranged from 651,513 to 2,543,743. `spdep::skater()` implements the restriction
through its `crit` and `vec.crit` arguments, which its reference page, linked under "All About R",
describes as "limits of group size or limits of population size".

The question here is which townships become singletons on the standardised variables, and what a
minimum size, first in townships and then in households, does to the regions.

```{r}
skater_z <- function(crit = NULL, vec.crit = NULL, seed = 1234) {
  set.seed(seed)
  w <- nb2listw(shan.nb, nbcosts(shan.nb, as.data.frame(shan_ict.z)), style = "B")
  tree <- mstree(w)
  if (is.null(crit)) {
    spdep::skater(tree[, 1:2], as.data.frame(shan_ict.z), ncuts = 5)$groups
  } else if (is.null(vec.crit)) {
    spdep::skater(tree[, 1:2], as.data.frame(shan_ict.z), ncuts = 5, crit = crit)$groups
  } else {
    spdep::skater(tree[, 1:2], as.data.frame(shan_ict.z), ncuts = 5,
                  crit = crit, vec.crit = vec.crit)$groups
  }
}

sk_z <- skater_z()
sort(table(sk_z), decreasing = TRUE)
shan_sf$TS.x[sk_z %in% names(which(table(sk_z) == 1))]
```

::: {.callout-note collapse="false" title="How the code works"}
- `skater_z()` repeats the three steps of section 3.2 on the z-scores: edge costs, minimum spanning
  tree, and SKATER. It returns the region of each township. `crit`, if given, is passed to
  `spdep::skater()` as a minimum per region. Without `vec.crit` the minimum counts townships; with
  `vec.crit`, a value for each township, the minimum applies to the sum of those values over the
  region.
- `names(which(table(sk_z) == 1))` finds the regions with one township, and the last line prints the
  names of those townships.
:::

On the standardised variables, three of the six regions are single townships: Muse, Tachileik and
Taunggyi. The singletons are the extremes on the variables with almost no weight before
standardising (section 4.2). Their z-scores on computer ownership are the three highest in the state
(`r paste(round(sort(shan_ict.z[, "COMPUTER_PR"], decreasing = TRUE)[1:3], 1), collapse = ", ")`).
Taunggyi is the state capital, and Muse and Tachileik are the main border crossings to China and
Thailand. The remaining 52 townships are divided into three large regions.

The singletons on both scales, in red:

```{r}
#| echo: false
#| fig-height: 5
singleton_map <- function(groups, title) {
  d <- mutate(shan_sf, region = factor(groups),
              single = groups %in% as.numeric(names(which(table(groups) == 1))))
  ggplot(d) +
    geom_sf(aes(fill = region), colour = "white", linewidth = 0.2, alpha = 0.35) +
    geom_sf(data = d[d$single, ], fill = "red", colour = "black", linewidth = 0.4) +
    ggrepel::geom_label_repel(data = d[d$single, ], aes(label = TS.x, geometry = geometry),
                              stat = "sf_coordinates", size = 3, min.segment.length = 0,
                              nudge_x = 0.6) +
    scale_fill_brewer(palette = "Set2", guide = "none") +
    theme_void() + labs(title = title)
}
patchwork::wrap_plots(singleton_map(ccs6, "SKATER on the raw rates (section 3.2.4)"),
                      singleton_map(sk_z, "SKATER on z-scores"), ncol = 2)
```

Each of these singleton regions is a township that differs from every neighbour. A region of one
township, however, cannot be planned for as a region. The minimum size is set below to three and to
five townships:

```{r}
#| fig-height: 5
sk_z3 <- skater_z(crit = 3)
sk_z5 <- skater_z(crit = 5)

rbind(`no minimum` = unname(sort(table(sk_z), decreasing = TRUE)),
      `at least 3` = unname(sort(table(sk_z3), decreasing = TRUE)),
      `at least 5` = unname(sort(table(sk_z5), decreasing = TRUE)))

tmap_arrange(
  cluster_map(mutate(shan_sf, CLUSTER = factor(sk_z)), title = "SKATER, no minimum"),
  cluster_map(mutate(shan_sf, CLUSTER = factor(sk_z5)), title = "SKATER, at least 5"),
  ncol = 2)
```

```{r}
c(`no minimum` = ss_ratio(shan_ict.z, sk_z),
  `at least 3` = ss_ratio(shan_ict.z, sk_z3),
  `at least 5` = ss_ratio(shan_ict.z, sk_z5)) |> round(3)
```

With a minimum of five townships, the regions range from 5 to 14 townships, and the share of
variation explained falls from `r round(ss_ratio(shan_ict.z, sk_z), 2)`
to `r round(ss_ratio(shan_ict.z, sk_z5), 2)`. That loss is the cost of absorbing Muse, Tachileik and
Taunggyi into regions with neighbours that differ from them on computer ownership.

A minimum counted in townships gives a township of 3,318 households the same weight as Taunggyi with
82,604. Assunção et al. set their minimum in population, the quantity their regions were balanced
on. The equivalent here is a minimum number of households. Shan
has `r format(sum(shan_sf$TT_HOUSEHOLDS), big.mark = ",")` households, and minima of 50,000 and
100,000 are tried, roughly one twentieth and one tenth of the total.

```{r}
#| fig-height: 5
sk_hh50 <- skater_z(crit = 50000, vec.crit = shan_sf$TT_HOUSEHOLDS)
sk_hh100 <- skater_z(crit = 100000, vec.crit = shan_sf$TT_HOUSEHOLDS)

imap_dfr(list(`at least 50,000 households` = sk_hh50,
              `at least 100,000 households` = sk_hh100),
         \(g, name) tibble(minimum = name,
                           regions = length(unique(g)),
                           townships = paste(sort(table(g), decreasing = TRUE), collapse = ", "),
                           smallest_households = min(tapply(shan_sf$TT_HOUSEHOLDS, g, sum)),
                           explained = ss_ratio(shan_ict.z, g))) |>
  kable(digits = 3, format.args = list(big.mark = ","))

cluster_map(mutate(shan_sf, CLUSTER = factor(sk_hh100)),
            title = "SKATER, at least 100,000 households per region")
```

With a minimum of 50,000 households, Taunggyi remains a region of its own, because its 82,604
households meet the minimum by themselves. Muse and Tachileik are absorbed. With a minimum of
100,000, every region has between 7 and 12 townships, and the share explained
is `r round(ss_ratio(shan_ict.z, sk_hh100), 2)`, lower than with a minimum of five townships. As
Assunção et al. describe, the restriction balances the regions, here in households.

## Geographic dissimilarity in ClustGeo

The geographic matrix D1 of section 3.3.2 was built with `st_distance()` between polygons, and the
distance between two townships that share a boundary is zero. Chavent et al. (2018) illustrate two
kinds of D1: geographic distances between the municipalities (section 4.3), and a neighbourhood
matrix D1 = 1 − A, where A is the adjacency matrix with 1 on its diagonal (section 4.4). The
ClustGeo introduction page, linked under "All About R", specifies the first as "the distances
between the town halls of the 303 municipalities", that is, between points. The chapter's matrix
measures the distance between polygon boundaries, a third kind.

Two of the references use both kinds of D1 on one data set. Jaya et al. (2019) cluster diarrhoea
risk in the 30 districts of Bandung with a distance-based and a contiguity-based D1, and prefer the
distance-based one because diarrhoea may spread "not limited to adjacent locations but also possible
in locations that are far apart". De Souza and Taconeli (2022) run both versions on 399 Paraná
municipalities, choose α separately for each (0.25 for distances and 0.15 for neighbourhoods), and
find that the neighbourhood version separates the clusters more on the educational indicators. In
Jaya et al. and in de Souza and Taconeli, the choice of D1 is a decision about what "near" should
mean for the phenomenon, made before α is chosen.

The question here is what the chapter's D1 measures, and whether the two matrices from the paper
give more contiguous clusters at a smaller loss of homogeneity.

```{r}
A <- nb2mat(shan.nb, style = "B")
D1_poly <- as.matrix(distmat)
off_diag <- row(A) != col(A)

tibble(pairs_at_zero = sum(D1_poly[off_diag] == 0) / 2,
       neighbour_pairs = sum(A) / 2,
       zero_iff_neighbour = all((D1_poly == 0 & off_diag) == (A == 1)))
```

::: {.callout-note collapse="false" title="How the code works"}
- `nb2mat(style = "B")` converts the neighbour list to a 55 × 55 matrix with 1 for neighbours and 0
  otherwise.
- `off_diag` is true everywhere except the diagonal. Pairs are counted in both directions in a full
  matrix, so each count is divided by 2.
- The last value tests whether a pair has zero distance exactly when the two townships are
  neighbours.
:::

The pairs with D1 = 0 are exactly the 132 neighbouring pairs. Under this D1, a township is at the
same distance, zero, from all of its neighbours, however large they are. Between townships that do
not touch, the distance is the gap between their nearest edges, so a long, thin township counts as
close to many others. The two alternatives from the paper are built below. Centroid distances need a
projected CRS, and Shan State is within UTM zone 47N (96°E to 102°E), EPSG:32647.

```{r}
shan_utm <- st_transform(shan_sf, 32647)
D1_centroid <- as.dist(st_distance(st_centroid(st_geometry(shan_utm))))

A1 <- A
diag(A1) <- 1
D1_neighbour <- as.dist(1 - A1)

D0_z <- dist(shan_ict.z)
```

::: {.callout-note collapse="false" title="How the code works"}
- `st_transform(shan_sf, 32647)` projects the layer to UTM zone 47N, in metres. `st_centroid()` then
  finds the centre of each township on the flat plane, and `st_distance()` between the centroids
  gives straight-line distances in metres.
- `D1_neighbour` follows section 4.4 of Chavent et al. (2018): the adjacency matrix with 1 on the
  diagonal, subtracted from 1. Neighbours are at dissimilarity 0 and all other pairs at 1.
- `D0_z` is the distance matrix on z-scores, used for all three versions of D1 so that only D1
  changes.
:::

::: callout-note
## What each D1 measures, for one township

The three maps show the dissimilarity between Taunggyi and every other township under each version
of D1:

- **Polygon distance (chapter):** every neighbour is at 0, however far its centre lies.
- **Centroid distance:** dissimilarity grows smoothly with distance.
- **Neighbourhood matrix, 1 − A:** neighbours are at 0, and every other township is at 1.

```{r}
#| echo: false
#| fig-height: 3.6
ti <- which(shan_sf$TS.x == "Taunggyi")
d1_maps <- list(
  `polygon distance (km)` = as.matrix(distmat)[ti, ] / 1000,
  `centroid distance (km)` = as.matrix(D1_centroid)[ti, ] / 1000,
  `neighbourhood, 1 - A` = as.matrix(D1_neighbour)[ti, ])
d1_plots <- imap(d1_maps, \(v, name) {
  ggplot() + geom_sf(data = mutate(shan_sf, d = as.numeric(v)), aes(fill = d),
                     colour = "white", linewidth = 0.1) +
    geom_sf(data = shan_sf[ti, ], fill = NA, colour = "red", linewidth = 0.6) +
    scale_fill_viridis_c(direction = -1) + theme_void() +
    labs(title = name, fill = NULL) + theme(legend.position = "bottom")
})
patchwork::wrap_plots(d1_plots, ncol = 3)
```

Taunggyi is outlined in red. On the first map, its neighbours have the same colour as Taunggyi
itself: the polygon distance gives every neighbour the same value, zero.
:::

For each D1, `choicealpha()` gives the two proportions for α from 0 to 1, and each partition's
pieces and share of variation explained are computed below.

```{r}
#| fig-height: 4
D1_list <- list(`polygon distance (chapter)` = distmat,
                `centroid distance` = D1_centroid,
                `neighbourhood 1 - A` = D1_neighbour)
alphas <- seq(0, 1, 0.1)

alpha_grid <- imap_dfr(D1_list, \(D1_mat, name) {
  cr <- choicealpha(D0_z, D1_mat, range.alpha = alphas, K = 6, graph = FALSE)
  parts <- lapply(alphas, \(a) cutree(hclustgeo(D0_z, D1_mat, alpha = a), 6))
  tibble(D1 = name, alpha = alphas,
         Q0norm = cr$Qnorm[, 1], Q1norm = cr$Qnorm[, 2],
         pieces = sapply(parts, count_pieces),
         explained = sapply(parts, \(g) ss_ratio(shan_ict.z, g)))
})

alpha_grid |>
  pivot_longer(c(Q0norm, Q1norm), names_to = "criterion", values_to = "value") |>
  ggplot(aes(alpha, value, linetype = criterion)) +
  geom_line() + geom_point(size = 1) +
  facet_wrap(~ D1) +
  scale_linetype_manual(values = c(Q0norm = "solid", Q1norm = "dashed"),
                        labels = c("ICT variables (D0)", "geography (D1)")) +
  labs(y = "normalised proportion explained", linetype = NULL) +
  theme(legend.position = "bottom")
```

::: {.callout-note collapse="false" title="How the code works"}
- `imap_dfr()` calls the function once for each element of `D1_list`, passing the element as
  `D1_mat` and its name as `name`, and stacks the returned tables into one.
- For each D1 it runs `choicealpha()` without a chart, then for each α builds the six-cluster
  partition with `hclustgeo()` and computes its pieces and share of variation explained.
- The matrix is called `D1_mat` and not `D1` because `tibble()` builds its columns in order, and a
  later column can refer to an earlier one by name. A column named `D1` would then hide the matrix
  from the `pieces` and `explained` columns, and `hclustgeo()` would receive the text of the name.
- The chart is the lower panel of `choicealpha()`'s own output, read as in section 3.3.2, and drawn
  for the three versions of D1 together.
:::

```{r}
alpha_grid |>
  filter(alpha <= 0.6) |>
  select(D1, alpha, pieces, explained) |>
  pivot_wider(names_from = D1, values_from = c(pieces, explained), names_vary = "slowest") |>
  kable(digits = 2)
```

- **Polygon distance.** The partition does not become contiguous in the range where α is useful. At
  α = 0.2 the six clusters are still
  in `r alpha_grid |> filter(D1 == "polygon distance (chapter)", alpha == 0.2) |> pull(pieces)`
  pieces. The chapter's own solution in section 3.3.3, on the raw rates, is
  in `r count_pieces(shan_sf_Gcluster$CLUSTER)`.
- **Centroid distance.** The partition has about as many pieces as with the polygon distance up to α
  = 0.3, and fewer above α = 0.3.
- **Neighbourhood matrix.** At α = 0.3 the partition has six pieces, which means six contiguous
  regions, and it
  explains `r alpha_grid |> filter(D1 == "neighbourhood 1 - A", alpha == 0.3) |> pull(explained) |> round(2)`
  of the variation.

With the neighbourhood matrix, contiguity is reached at the smallest α, which is the behaviour
Chavent et al. (2018) report for their data in section 4.4. Jaya et al. chose the distance-based D1
because diarrhoea spreads between districts that are far apart. By the same reasoning, the
neighbourhood matrix is the better choice for ICT access: ownership of a television or a telephone
does not spread between townships as an infection does, and a region is wanted here as an area that
can be administered as one.

Jaya et al. compare their two versions by their Q values at the chosen α. Q1 is computed on a
different matrix in each case, so a higher Q1 under one D1 does not mean more geographic cohesion
than a lower Q1 under another. The table above has no such mismatch: every version of D1 is scored
on the same two measures, the share of variation explained on the z-scores and the number of pieces.

Assunção et al. (2006, section 2) place ClustGeo's approach in a longer debate. They describe
methods that combine nearness in attribute space with nearness in geographic space as a weighted
mean, and report the criticism of Openshaw et al. (1998) that homogeneity and compactness "are not
strictly comparable" and that compactness "should not be combined with homogeneity measures", and
should be handled as a constraint. The mixing parameter α is such a weight, and in the table above
contiguity is not reached until α is large enough. In SKATER, REDCAP and AZP, contiguity is a
constraint.

## Methods from the lecture beyond the chapter

Slide 2 lists five spatially constrained methods, in two families. The hierarchical methods build
the regions by merging or by splitting a tree: SKATER, REDCAP and ClustGeo. The partitioning methods
start from a set of regions and improve it by moving townships between them: AZP and max-p. The
chapter covers SKATER and ClustGeo, run in sections 3.2 and 3.3. The `rgeoda` package, an R
interface to the GeoDa library, provides the others. The question here is how much of the variation
the remaining methods explain on Shan, and how balanced their regions are, against SKATER.

| Method | Family | How the regions are formed | Source |
|-------|------|--------------------------|-----------|
| **REDCAP** | hierarchical | builds a spanning tree by one of four linkages and cuts it as SKATER does; the full-order variants measure the dissimilarity between two clusters on all pairs of their townships, which addresses the SKATER limitation on slide 9; first-order single linkage is identical to SKATER | slides 9 to 12; Guo (2008) |
| **SCHC** (spatially constrained hierarchical clustering) | hierarchical | ordinary agglomerative clustering in which only neighbouring clusters may merge | not on the slides; provided by `rgeoda` beside REDCAP |
| **AZP** (automatic zoning procedure) | partitioning | starts from a random partition into k contiguous regions, and moves townships between neighbouring regions when a move reduces the within-region sum of squares and keeps every region contiguous; the greedy version stops at the first local optimum, and simulated annealing and tabu search accept some worse moves to escape one | slide 2; compared with SKATER in Assunção et al. (2006, section 5) |
| **max-p** | partitioning | takes a minimum per region, such as a number of households, in place of k, and finds the largest number of contiguous regions that all meet it | slide 2; Duque, Anselin and Rey (2012), not in the readings |

::: callout-note
## First-order and full-order linkage in REDCAP

Slide 9 describes two clusters, C1 with five members and C2 with three. There are 15 pairs of
members between them, and only three of the pairs are neighbours. First-order linkage measures the
distance between the clusters on the three neighbouring pairs only. Full-order linkage uses all 15.
SKATER is first-order, and slide 11 lists the six REDCAP variants, which differ in which pairs they
use and how they summarise them.

```{r}
#| echo: false
#| fig-height: 3.4
angle <- c(A = 160, B = 115, C = 200, D = 240, E = 280, F = 310, G = 0, H = 55) * pi / 180
nodes <- tibble(id = names(angle), x = cos(angle), y = sin(angle),
                cluster = c(rep("C1", 5), rep("C2", 3)))
all_pairs <- expand_grid(a = nodes$id[1:5], b = nodes$id[6:8]) |>
  left_join(nodes, by = c(a = "id")) |> left_join(nodes, by = c(b = "id"), suffix = c("", "end"))
first_order <- all_pairs |> filter(paste0(a, b) %in% c("BH", "EH", "EF"))
linkage_panel <- function(pairs, title) {
  ggplot() +
    geom_segment(data = pairs, aes(x, y, xend = xend, yend = yend), colour = "red",
                 linewidth = 0.9) +
    geom_point(data = nodes, aes(x, y, fill = cluster), shape = 21, size = 8) +
    geom_text(data = nodes, aes(x, y, label = id)) +
    scale_fill_manual(values = c(C1 = "lightblue", C2 = "moccasin")) +
    coord_equal(xlim = c(-1.25, 1.25), ylim = c(-1.25, 1.25)) + theme_void() +
    labs(title = title, fill = NULL) + theme(plot.margin = margin(5, 20, 5, 20))
}
patchwork::wrap_plots(linkage_panel(first_order, "First-order: 3 pairs (BH, EH, EF)"),
                      linkage_panel(all_pairs, "Full-order: all 15 pairs"),
                      ncol = 2, guides = "collect")
```

The eight members are placed on a circle so that no line passes through a third member. Their
positions are for illustration and are not those of the slide.
:::

### Contiguity in rgeoda

`rgeoda` builds its own weights from the polygons. Two townships count as Queen neighbours when
their boundaries share a point, and `queen_weights()` tests whether two vertices are the same point
by a precision threshold, zero by default. When adjacent polygons in a shapefile were digitised
separately, their shared boundary may have vertices that differ in the last decimal places, and an
exact comparison then misses the link. `rgeoda`'s neighbours are compared below with those of
`poly2nb()` in section 3.2.1, with and without a threshold.

```{r}
queen_default <- queen_weights(shan_sf)
queen_matched <- queen_weights(shan_sf, precision_threshold = 1e-7)

tibble(weights = c("spdep::poly2nb()", "rgeoda default", "rgeoda, precision_threshold = 1e-7"),
       links = c(sum(card(shan.nb)),
                 sum(sapply(1:55, \(i) length(get_neighbors(queen_default, i)))),
                 sum(sapply(1:55, \(i) length(get_neighbors(queen_matched, i))))),
       min_neighbours_reported = c(min(card(shan.nb)), min_neighbors(queen_default),
                                   min_neighbors(queen_matched))) |>
  kable()

all(sapply(1:55, \(i) setequal(get_neighbors(queen_matched, i), shan.nb[[i]])))
```

::: {.callout-note collapse="false" title="How the code works"}
```r
queen_weights(sf_obj, order = 1, precision_threshold = 0)  # rgeoda
get_neighbors(gda_w, idx)                                  # rgeoda
min_neighbors(gda_w)                                       # rgeoda
```

- `queen_weights()` finds Queen neighbours. With `precision_threshold = 0` two boundary points must
  be exactly equal to count as shared. A positive threshold, here 1e-7 degrees (about 1 cm), counts
  points closer than that as shared.
- `get_neighbors()` returns the neighbours of township `idx`, and `length()` counts them.
- `min_neighbors()` returns the smallest number of neighbours of any township, as `rgeoda` records
  it in the weights object.
- The last line tests whether the matched weights give every township exactly the same neighbours as
  `spdep`.
:::

With exact matching, `rgeoda` misses nine links, including four of Pangsang's five neighbours,
because the boundaries of adjacent townships in this shapefile do not share exactly identical
vertices. The weights object records a minimum of zero neighbours, which `rgeoda` records as an
isolated township. In testing, `rgeoda::skater()` on the default weights crashed the R session. With
a threshold of 1e-7 degrees, the neighbours are the same as `spdep`'s for every township, and those
weights are used for all the methods below.

### Running the methods

`rgeoda` standardises the variables to z-scores by default (`scale_method = "standardize"`), so the
raw rates are passed and the results are comparable with the z-score results above.

```{r}
rs <- 123456789
rg <- list(
  `SKATER (rgeoda)` = rgeoda::skater(6, queen_matched, shan_ict, random_seed = rs, cpu_threads = 1),
  `REDCAP first-order single` = redcap(6, queen_matched, shan_ict, "firstorder-singlelinkage",
                                       random_seed = rs, cpu_threads = 1),
  `REDCAP full-order single` = redcap(6, queen_matched, shan_ict, "fullorder-singlelinkage",
                                      random_seed = rs, cpu_threads = 1),
  `REDCAP full-order average` = redcap(6, queen_matched, shan_ict, "fullorder-averagelinkage",
                                       random_seed = rs, cpu_threads = 1),
  `REDCAP full-order complete` = redcap(6, queen_matched, shan_ict, "fullorder-completelinkage",
                                        random_seed = rs, cpu_threads = 1),
  `REDCAP full-order Ward` = redcap(6, queen_matched, shan_ict, "fullorder-wardlinkage",
                                    random_seed = rs, cpu_threads = 1),
  `SCHC Ward` = schc(6, queen_matched, shan_ict, "ward"),
  `AZP greedy` = azp_greedy(6, queen_matched, shan_ict, random_seed = rs),
  `AZP simulated annealing` = azp_sa(6, queen_matched, shan_ict, cooling_rate = 0.85,
                                     random_seed = rs),
  `AZP tabu` = azp_tabu(6, queen_matched, shan_ict, tabu_length = 10, conv_tabu = 10,
                        random_seed = rs))

rg_table <- imap_dfr(rg, \(r, name) tibble(
  method = name,
  explained = r$`The ratio of between to total sum of squares`,
  pieces = count_pieces(r$Clusters),
  sizes = paste(sort(table(r$Clusters), decreasing = TRUE), collapse = ", ")))
kable(rg_table, digits = 3)
```

::: {.callout-note collapse="false" title="How the code works"}
```r
rgeoda::skater(k, w, df, random_seed, cpu_threads)        # rgeoda
redcap(k, w, df, method, random_seed, cpu_threads)        # rgeoda
schc(k, w, df, method)                                    # rgeoda
azp_greedy(p, w, df, random_seed)                         # rgeoda
azp_sa(p, w, df, cooling_rate, random_seed)               # rgeoda
azp_tabu(p, w, df, tabu_length, conv_tabu, random_seed)   # rgeoda
```

- Every function takes the number of regions, the weights and the data, and returns a list with
  `$Clusters`, the region of each township, and the ratio of between to total sum of squares.
- `method` in `redcap()` names the linkage and the order of constraint. In `schc()` it names the
  linkage.
- `cooling_rate` sets how quickly simulated annealing stops accepting worse moves. `tabu_length`
  sets how many recent moves tabu search forbids reversing, and `conv_tabu` how many moves without
  improvement end the search.
- `random_seed` fixes the random start of the tree or the partition, and `cpu_threads = 1` makes the
  result independent of the number of processor cores.
- `rgeoda::` is written in full for `skater()`, because `spdep` has a function of the same name.
:::

The regions of every method are in six pieces, so all of them are strictly contiguous. They differ
in how much of the variation their regions explain and in how balanced the regions are:

| Methods | Share explained | Singletons | Note |
|------|--|--|----------|
| SKATER (`rgeoda`); REDCAP first-order single | `r sprintf("%.2f", rg_table$explained[1])` | 3 | identical regions, as on slide 10; different regions from `spdep`'s SKATER on the same z-scores (adjusted Rand index `r round(adjustedRandIndex(rg[[1]]$Clusters, sk_z), 2)`), which also has three singletons |
| REDCAP full-order single | `r sprintf("%.2f", rg_table$explained[3])` | 2 | the lowest of the REDCAP variants |
| REDCAP full-order average, complete, Ward | 0.51 to 0.52 | 1 or 2 | the highest of the tree-based methods |
| SCHC Ward | `r sprintf("%.2f", rg_table$explained[7])` | 5 | one region of 50 townships |
| AZP greedy, simulated annealing, tabu | 0.50 to 0.51 | 1 | from a random start; greedy and tabu search give the same share and region sizes |

Assunção et al. (2006, section 5) compare SKATER with AZP on three Brazilian data sets and find
SKATER "superior to AZP in all tests", with AZP's partition quality between 0.678 and 0.988 of
SKATER's. On Shan the order is reversed: AZP explains more than either SKATER implementation. The
authors themselves caution that "there is no decisive way to compare two classification methods,
because performance is dependent on the characteristics of the data set", and their AZP was run five
times from different starts with the best kept, where the runs here use one seed. The single
comparison here is consistent with that caution. On 55 units with weak spatial structure (section
4.1), SKATER's first cuts separate the extreme townships, and AZP's moves of townships between
neighbouring regions give a higher share explained.

### Max-p with a household threshold

Max-p fixes a minimum number of households per region and finds the largest number of contiguous
regions that meet it. It is listed on slide 2 and is not used in the readings; the method is from
Duque, Anselin and Rey (2012). Shan has `r format(sum(shan_sf$TT_HOUSEHOLDS), big.mark = ",")`
households. The threshold is set to 100,000 and 150,000 households, roughly one tenth and one
seventh of the total.

```{r}
#| fig-height: 5
hh <- shan_sf["TT_HOUSEHOLDS"]
maxp_100 <- maxp_greedy(queen_matched, shan_ict, bound_variable = hh, min_bound = 100000,
                        iterations = 99, random_seed = rs, cpu_threads = 1)
maxp_150 <- maxp_greedy(queen_matched, shan_ict, bound_variable = hh, min_bound = 150000,
                        iterations = 99, random_seed = rs, cpu_threads = 1)

imap_dfr(list(`at least 100,000` = maxp_100, `at least 150,000` = maxp_150),
         \(r, name) tibble(threshold = name,
                           regions = length(unique(r$Clusters)),
                           explained = r$`The ratio of between to total sum of squares`,
                           sizes = paste(sort(table(r$Clusters), decreasing = TRUE),
                                         collapse = ", "))) |>
  kable(digits = 3)

tmap_arrange(
  cluster_map(mutate(shan_sf, CLUSTER = factor(maxp_100$Clusters)),
              title = "max-p, at least 100,000 households"),
  cluster_map(mutate(shan_sf, CLUSTER = factor(maxp_150$Clusters)),
              title = "max-p, at least 150,000 households"),
  ncol = 2)
```

::: {.callout-note collapse="false" title="How the code works"}
```r
maxp_greedy(w, df, bound_variable, min_bound, iterations, random_seed, cpu_threads)  # rgeoda
```

- `bound_variable` is a one-column sf data frame holding the quantity that each region must reach,
  here the number of households, and `min_bound` is the threshold.
- `iterations = 99` runs the construction phase 99 times from different random starts and keeps the
  best result.
- Max-p determines the number of regions from the threshold.
:::

A threshold of 150,000 households gives six regions, the same number as the other methods, with no
singletons: every region has at least `r min(table(maxp_150$Clusters))` townships. The share
explained is lower, `r round(maxp_150[["The ratio of between to total sum of squares"]], 2)`,
because under the threshold, townships as different as Taunggyi and its rural neighbours have to
share a region.

## Comparison of the regionalisations

Part 1 has three sets of six clusters, from Ward, SKATER and ClustGeo, and sections 4.2 to 4.7 have
seven more. Slides 4 and 5 show scattered and contiguous clusters side by side. Conventional
clustering finds the most homogeneous groups without regard to contiguity. The constrained methods
guarantee or encourage contiguity, and their groups are less homogeneous. Every set of six clusters
on this page is scored below on the two measures, computed on the same standardised variables.

```{r}
comparison <- bind_rows(
  tibble(method = "Hierarchical, raw rates, ward.D (chapter, 3.1.12)",
         groups = list(as.integer(shan_sf_cluster$CLUSTER))),
  tibble(method = "Hierarchical, z-scores, ward.D2 (4.2)", groups = list(h_z)),
  tibble(method = "SKATER, raw rates (chapter, 3.2)", groups = list(ccs6)),
  tibble(method = "SKATER, z-scores (4.5)", groups = list(sk_z)),
  tibble(method = "SKATER, z-scores, at least 5 (4.5)", groups = list(sk_z5)),
  tibble(method = "ClustGeo, raw, polygon D1, alpha 0.2 (chapter, 3.3)",
         groups = list(as.integer(shan_sf_Gcluster$CLUSTER))),
  tibble(method = "ClustGeo, z, neighbourhood D1, alpha 0.3 (4.6)",
         groups = list(cutree(hclustgeo(D0_z, D1_neighbour, alpha = 0.3), 6))),
  tibble(method = "REDCAP full-order complete (4.7)",
         groups = list(rg$`REDCAP full-order complete`$Clusters)),
  tibble(method = "AZP tabu (4.7)", groups = list(rg$`AZP tabu`$Clusters)),
  tibble(method = "max-p, 150,000 households (4.7)", groups = list(maxp_150$Clusters))) |>
  mutate(explained = map_dbl(groups, \(g) ss_ratio(shan_ict.z, g)),
         pieces = map_dbl(groups, count_pieces),
         smallest = map_dbl(groups, \(g) min(table(g))))

comparison |> select(-groups) |> kable(digits = 2)
```

```{r}
#| fig-height: 5
ggplot(comparison, aes(pieces, explained, label = str_remove(method, " \\(.*"))) +
  geom_point(size = 2.5, colour = "steelblue") +
  ggrepel::geom_text_repel(size = 3, max.overlaps = 20) +
  geom_vline(xintercept = 6, linetype = "dashed", colour = "grey50") +
  annotate("text", x = 6.3, y = 0.3, label = "six contiguous regions", hjust = 0, size = 3,
           colour = "grey40") +
  labs(x = "number of separate pieces in the six clusters",
       y = "share of variation explained (z-scores)")
```

::: {.callout-note collapse="false" title="How the code works"}
- Each row of `comparison` holds a method name and, in a list column, its vector of cluster labels.
  `map_dbl()` applies each measure to every vector. `smallest` is the size of the smallest region.
- `geom_text_repel()` from `ggrepel` places the labels so they do not overlap each other or the
  points. The dashed line marks six pieces, the value for six contiguous regions.
:::

Each point is one set of six clusters. Points further right are in more pieces, and points higher up
explain more of the variation. The dashed line marks six pieces, which means six contiguous regions.

- **Contiguity costs about a quarter of the explained variation.** Unconstrained Ward clustering on
  z-scores explains `r round(comparison$explained[2], 2)` and its clusters are
  in `r comparison$pieces[2]` pieces. The best strictly contiguous solutions, REDCAP full-order and
  AZP, explain about 0.51.
- **The chapter's two constrained solutions are not the strongest.** SKATER on raw rates
  explains `r round(comparison$explained[3], 2)` on the standardised measure, with two singletons,
  and the chapter's ClustGeo solution explains `r round(comparison$explained[6], 2)` while its
  clusters are still in `r comparison$pieces[6]` pieces.
- **Balanced regions cost more.** The solutions with no singletons, SKATER with a minimum of five,
  ClustGeo with the neighbourhood matrix, and max-p, explain between 0.36 and 0.41.

REDCAP full-order complete linkage gives the most homogeneous contiguous regions. Its regions,
profiled by their mean rates:

```{r}
#| fig-height: 5
redcap_sf <- shan_sf |> mutate(REGION = factor(rg$`REDCAP full-order complete`$Clusters))
cluster_map(redcap_sf, "REGION", "REDCAP full-order complete linkage, six regions")
```

```{r}
redcap_sf |>
  st_drop_geometry() |>
  group_by(REGION) |>
  summarise(townships = n(),
            households = sum(TT_HOUSEHOLDS),
            across(RADIO_PR:COMPUTER_PR, mean),
            members = if (n() <= 3) paste(TS.x, collapse = ", ") else "") |>
  kable(digits = 0, format.args = list(big.mark = ","))
```

```{r}
#| fig-height: 4
redcap_sf |>
  st_drop_geometry() |>
  select(REGION, RADIO_PR:COMPUTER_PR) |>
  bind_cols(as_tibble(shan_ict.z) |> rename_with(\(x) paste0(x, "_z"))) |>
  select(REGION, ends_with("_z")) |>
  pivot_longer(-REGION, names_to = "variable", values_to = "z") |>
  mutate(variable = str_remove(variable, "_PR_z")) |>
  ggplot(aes(variable, z)) +
  geom_hline(yintercept = 0, colour = "grey60") +
  geom_boxplot(fill = "light blue") +
  facet_grid(~ REGION) +
  labs(x = NULL, y = "z-score") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The profile is drawn on z-scores, so that the five variables share one scale and a box above the
zero line means a region above the state average.

Brown and Watson (2012) express each cluster mean as a percentage of the mean across all units, and
describe the clusters by variables that "were not used in the cluster analysis". De Souza and
Taconeli (2022, section 4) cross-tabulate their clusters against the official intermediate regions
of Paraná, and compare the clusters on each indicator by an F test. Rovan and Sambt (2003) name
their clusters from the mean profiles. Two variables were left out of the clustering for Shan,
internet at home (dropped in section 3.1.4) and the number of households, and the state has an
official division into three parts.

```{r}
state_mean <- redcap_sf |>
  st_drop_geometry() |>
  summarise(across(c(RADIO_PR:INTERNET_PR, TT_HOUSEHOLDS), mean))

redcap_sf |>
  st_drop_geometry() |>
  group_by(REGION) |>
  summarise(townships = n(), across(c(RADIO_PR:INTERNET_PR, TT_HOUSEHOLDS), mean)) |>
  mutate(across(c(RADIO_PR:INTERNET_PR, TT_HOUSEHOLDS), \(x) 100 * x / state_mean[[cur_column()]])) |>
  rename_with(\(x) str_remove(x, "_PR"), RADIO_PR:INTERNET_PR) |>
  kable(digits = 0, caption = "Region means as a percentage of the state mean. INTERNET and TT_HOUSEHOLDS were not used to form the regions")
```

::: {.callout-note collapse="false" title="How the code works"}
- `state_mean` holds the mean of each rate over all 55 townships.
- In the second block of code, `summarise()` computes the region means, and `mutate(across(...))`
  divides each by the state mean of the same column. `cur_column()` returns the name of the column
  being processed, so each column is divided by its own state mean. A value of 100 is the state
  average.
:::

```{r}
table(part = redcap_sf$ST, region = redcap_sf$REGION)
chisq.test(table(redcap_sf$ST, redcap_sf$REGION), simulate.p.value = TRUE, B = 9999)$p.value

map_dfr(c("RADIO_PR", "TV_PR", "LLPHONE_PR", "MPHONE_PR", "COMPUTER_PR"), \(v) {
  f <- summary(aov(redcap_sf[[v]] ~ redcap_sf$REGION))[[1]]
  tibble(rate = v, F = f$`F value`[1], p_value = format.pval(f$`Pr(>F)`[1], digits = 2))
}) |>
  kable(digits = 1)
```

::: {.callout-note collapse="false" title="How the code works"}
- `table()` counts the townships of each part of the state in each region.
  `chisq.test(simulate.p.value = TRUE)` tests whether regions and parts are independent. Several
  cells are small, so the p-value is taken from 9,999 simulated tables and not from the chi-squared
  approximation.
- `aov()` fits a one-way analysis of variance of each rate on the region, and `summary()` returns
  the F statistic and its p-value. The five rates are tested one at a time, as in de Souza and
  Taconeli (2022).
:::

The six regions differ significantly on every one of the five rates, most strongly on land line.
These F tests are descriptive only: the regions were formed to differ on the same five variables, so
significant differences are expected by construction. Internet at home and the official division of
the state were not used to form the regions.

::: callout-note
## What the F test compares

For one rate, the F test compares how far apart the six region means are with how spread out the
townships are inside the regions:

$$F = \frac{\text{difference between the region means}}{\text{spread of the townships inside the regions}}$$

- **A large F:** the region means are far apart, and the townships in each region are close to their
  region mean. The regions differ on that rate.
- **A small F:** the differences between the region means are small compared with the spread inside
  the regions.
- **The p-value** is the probability of an F at least this large if all six regions had the same
  mean.

Each part of the formula is a sum of squares divided by its degrees of freedom: 6 − 1 = 5 for the
region means and 55 − 6 = 49 for the townships inside the regions.
:::

Internet at home, which was dropped in section 3.1.4 for its correlation with computer ownership, is
close to computer ownership in five of the six regions: both are 165% of the state mean in region 2,
and the Taunggyi region is highest on both. The exception is region 4, where computer ownership is
close to average (92%) and internet at home is half of it (49%). Apart from Laukkaing, Konkyan and
Kunlong (region 4), little information was lost by dropping internet. The regions are also
associated with the official parts of the state (simulated p < 0.01): region 2 is in Shan (East) and
Shan (North), region 3 in Shan (North) and Shan (South), and the three small regions in Shan (North)
or Shan (South). Only the largest region spans all three parts.

Named from their profiles, in the manner of Rovan and Sambt (2003), the six regions are:

| Region | Townships | Name | Profile, as a percentage of the state mean |
|--|--|-----|------------|
| 1 | 24 | Low access | below the state mean on every item, from 55% (land line) to 88% (radio) |
| 2 | 14 | Connected east and north | 154% to 165% on land line, mobile phone, computer and internet |
| 3 | 11 | High radio | radio 158%, TV and mobile phone close to average, land line and computer about 70%; the second-largest townships on average (136% of mean households) |
| 4 | 3 | Land line, low radio | Laukkaing, Konkyan and Kunlong: land line 317%, radio 30% |
| 5 | 2 | Low TV and computer | Narphan and Pangwaun: TV 57%, computer 19%, internet 11% |
| 6 | 1 | Taunggyi | the state capital, above average on everything and 450% of mean households |

# Reflection

Chapter 12 asks, in section 12.2.1, whether Shan State can be delineated into homogeneous regions
using six ICT measures. The answer on this page is yes, with qualifications that the chapter does
not raise.

**The regions depend on the scale of the variables.** On the raw rates, mobile phone and TV make up
81% of the distance, so the chapter's clusters group the townships mainly on those two. After
standardising, the SKATER singletons change from Ywangan and Narphan to Taunggyi, Muse and
Tachileik.

**The number of regions is not settled by the data.** The gap statistic, the silhouette width and
NbClust all give fewer than six clusters, and the silhouette widths show weak structure at every k.
The merge heights of the Ward tree jump at two, four and six clusters. I would report the six
regions as one level of a nested partition, with the two- and four-region levels beside it.

**Most of the rates have no spatial pattern.** De Souza and Taconeli (2022) found every one of their
indicators spatially autocorrelated; here only radio and land line are. TV and mobile phone, which
carry most of the weight in the chapter's clusters, have no significant pattern. That explains the
fragmented map of section 3.1.12 better than the omission of location from the method, and it is why
contiguity costs about a quarter of the explained variation in section 4.8.

**SKATER is not the strongest method on these data.** Before section 4.8, I expected SKATER to be
the natural choice, because it is contiguous by construction and taught first. REDCAP full-order and
AZP explain more, consistent with the SKATER limitation on slide 9, and without a minimum size three
of SKATER's six regions are single townships. Assunção et al. (2006) met the same problem in São
Paulo and set a minimum population; here a minimum of 100,000 households gives six regions of 7 to
12 townships. They also found SKATER better than AZP on every data set they tried. On Shan the order
is reversed, which fits their own remark that performance "is dependent on the characteristics of
the data set".

Three calls in the chapter ran without error and without the effect that the text around them
describes:

- the standardised versions of the data were computed and not used for clustering;
- `hclust(method = "ward.D")` was run where the agglomerative coefficient that justified it belongs
  to `ward.D2`;
- `st_distance()` between polygons gave a geographic matrix in which all neighbours are at distance
  zero.

As in chapters 9 and 10, one object is computed and a different object is passed to the next step.
All three faults are explained in the course's own sources. The UC Business Analytics tutorial, the
source of chapter sections 12.7.8 to 12.7.10, standardises with `scale()` and uses `ward.D2`, and
Chavent et al. (2018) describe what `hclustgeo()` computes and offer the neighbourhood matrix as D1.

**Limits.** Population is not weighted in most of the methods: apart from max-p and the household
minimum in section 4.5, a township of 3,318 households counts as much as Taunggyi with 82,604. The
weights argument of `hclustgeo()` (Chavent et al., 2018, section 2.2) could give households a weight
in ClustGeo. The data are from the 2014 census, so the regions are based on access as it was in
2014.

# References

Course materials:

- Kam, T.S. (2026). *Lesson 6: Geographic Segmentation with Spatial Clustering*. ISSS626 Geospatial
  Analytics and Applications, SMU. Slide deck dated 18 September 2026.
- Kam, T.S. (2026). [*R for Geospatial Data Science and Analytics*, Chapter 12: Geographical
  Segmentation with Spatially Constrained Clustering
  Techniques](https://r4gdsa.netlify.app/chap12.html).

Assigned self-reading:

- Assunção, R.M., Neves, M.C., Câmara, G. and da Costa Freitas, C. (2006). "Efficient
  Regionalization Techniques for Socio-Economic Geographical Units Using Minimum Spanning Trees".
  *International Journal of Geographical Information Science*, 20(7), 797-811.
- Chavent, M., Kuentz-Simonet, V., Labenne, A. and Saracco, J. (2018). "ClustGeo: an R package for
  hierarchical clustering with spatial constraints". *Computational Statistics*, 33(4), 1799-1822.

References listed in the Lesson 6 outline:

- Brown, N.S. and Watson, P. (2012). "What can a comprehensive plan really tell us about a region?:
  A cluster analysis of county comprehensive plans in Idaho". *Western Economics Forum*, 11(2),
  22-37.
- de Souza, D.C. and Taconeli, C.A. (2022). "Spatial and non-spatial clustering algorithms in the
  analysis of Brazilian educational data". *Communications in Statistics: Case Studies, Data
  Analysis and Applications*, 8(4), 588-606.
- Demeter, T. and Brătucu, G. (2013). "Statistical Analysis of the EU Countries from a Touristic
  Point of View". *Bulletin of the Transilvania University of Braşov, Series V: Economic Sciences*,
  6(55), No. 2, 121-130.
- Jaya, I.G.N.M., Ruchjana, B.N., Andriyana, Y. and Agata, R. (2019). "Clustering with spatial
  constraints: The case of diarrhea in Bandung city, Indonesia". *Journal of Physics: Conference
  Series*, 1397, 012068.
- Rovan, J. and Sambt, J. (2003). "Socio-economic Differences Among Slovenian Municipalities: A
  Cluster Analysis Approach". In Ferligoj, A. and Mrvar, A. (eds), *Developments in Applied
  Statistics*, Metodološki zvezki 19, Ljubljana: FDV, 265-278.

All About R, linked from the Lesson 6 outline:

- UC Business Analytics R Programming Guide. [*Hierarchical Cluster
  Analysis*](http://uc-r.github.io/hc_clustering).
- [`skater()` reference page](https://r-spatial.github.io/spdep/reference/skater.html), `spdep`.
- Chavent, M. et al. [*Introduction to
  ClustGeo*](https://cran.r-project.org/web/packages/ClustGeo/vignettes/intro_ClustGeo.html),
  introduction page of the package.

Methods referred to:

- Openshaw, S., Alvanides, S. and Whalley, S. (1998). *Some Further Experiments with Designing
  Output Areas for the 2001 UK Census*. Report, University of Leeds. Cited in Assunção et al.
  (2006).
- Duque, J.C., Anselin, L. and Rey, S.J. (2012). "The max-p-regions problem". *Journal of Regional
  Science*, 52(3), 397-419.
- Guo, D. (2008). "Regionalization with Dynamically Constrained Agglomerative Clustering and
  Partitioning (REDCAP)". *International Journal of Geographical Information Science*, 22(7),
  801-823.
- Kaufman, L. and Rousseeuw, P.J. (1990). *Finding Groups in Data: An Introduction to Cluster
  Analysis*. Wiley.
- Tibshirani, R., Walther, G. and Hastie, T. (2001). "Estimating the Number of Clusters in a Data
  Set via the Gap Statistic". *Journal of the Royal Statistical Society, Series B*, 63(2), 411-423.

Software:

- Charrad, M., Ghazzali, N., Boiteau, V. and Niknafs, A. (2014). "NbClust: An R Package for
  Determining the Relevant Number of Clusters in a Data Set". *Journal of Statistical Software*,
  61(6), 1-36.
- Li, X. and Anselin, L. (2024). *rgeoda: R Library for Spatial Data Analysis*. R package.
- Maechler, M., Rousseeuw, P., Struyf, A., Hubert, M. and Hornik, K. *cluster: Cluster Analysis
  Basics and Extensions*. R package; help page `agnes.object` for the agglomerative coefficient.

# About this document

Written and executed in Quarto with R `r getRversion()`, `spdep` `r packageVersion("spdep")`,
`ClustGeo` `r packageVersion("ClustGeo")`, `rgeoda` `r packageVersion("rgeoda")` and
`tmap` `r packageVersion("tmap")`. All figures and tables are generated at render time from the data
in `data/`, so nothing on this page is pasted in.

I used AI assistance while working out the approach and the extensions in Part 2. The analysis was
run and checked by me.
