---
title: "Hands-on Exercise 2"
subtitle: "Spatial Point Patterns Analysis with R"
author: "An Loc"
date: "last-modified"
date-format: "D MMM YYYY"
format:
  html:
    toc: true
    toc-depth: 4
    toc-expand: true
    code-fold: false
    code-copy: true
    fig-width: 9
    fig-height: 7
execute:
  echo: true
  eval: true
  warning: false
  message: false
  freeze: true
---

# Overview

This exercise applies **Spatial Point Patterns Analysis (SPPA)** to the locations of childcare
centres in Singapore. It follows the two chapters assigned for Lesson 2: r4gdsa Chapter 4 and
Chapter 5. Where I write "the chapters" below, I mean those two.

**Part 1** reproduces the **first-order** methods from r4gdsa Chapter 4: methods that measure
the *intensity* of points: the Clark–Evans nearest-neighbour index and kernel density
estimation, including bandwidth selection, kernel choice, and fixed versus adaptive bandwidth.

**Part 2** reproduces the **second-order** methods from Chapter 5: methods that measure the
*spatial dependence* between points: the G, F, K and L functions, each tested against Complete
Spatial Randomness (CSR) using Monte Carlo envelopes, for two contrasting planning areas.

**Part 3** is where most of my time went. It is a working diary: the choices the chapters
leave to the software, how far the conclusions move when those choices change, what the
week's readings add when applied to this data, and where the analysis runs out.

::: {.callout-note appearance="simple"}

## Reference

r4gdsa [Chapter 4: First-order Spatial Point Patterns Analysis
Methods](https://r4gdsa.netlify.app/chap04.html)
and [Chapter 5: 2nd Order Spatial Point Patterns Analysis
Methods](https://r4gdsa.netlify.app/chap05.html).
:::

# Getting started

## Packages

```{r}
pacman::p_load(sf, spatstat, terra, tmap, tidyverse)
```

## The data

Two datasets are used. Both come from Data.gov.sg, with one caveat to state up front:

- **Childcare centres**: the published table *Listing of Centre Services* is an address list
  with **no coordinates**, so it cannot be mapped as it stands. The layer used here is
  `ChildCareServices.kml`, the geocoded version distributed with the chapters, produced by the
  OneMap workflow Kam demonstrated in Lesson 1.
- **Planning subzones**: URA 2019 Master Plan subzone boundaries (no sea). This is the same
  file used in Hands-on Exercise 1, so the wrangling carries straight over.

The KML stores points as XYZ with a zero Z, which spatstat will not accept, so `st_zm()` drops
the third dimension. Both layers are then transformed to SVY21 (EPSG:3414) so that every
distance, area and bandwidth below is in metres rather than degrees.

```{r}
childcare <- st_read("data/geospatial/ChildCareServices.kml", quiet = TRUE) |>
  st_zm(drop = TRUE, what = "ZM") |>
  st_transform(3414)

mpsz <- st_read("data/geospatial/MasterPlan2019SubzoneBoundaryNoSea.geojson",
                quiet = TRUE) |>
  st_transform(3414)
```

```{r}
st_crs(childcare)$epsg
```

```{r}
nrow(childcare)
```

# Part 1: First-order methods

First-order methods ask **how dense** the points are and **whether that density varies**
across space. Two families are covered: the nearest-neighbour approach (Clark–Evans) and
density estimation (KDE).

## From sf to a point pattern

spatstat works with its own object classes: `ppp` for the point pattern and `owin` for the
observation window. An `sf` point layer becomes a `ppp` in one step.

```{r}
childcare_ppp <- as.ppp(childcare)
```

```{r}
summary(childcare_ppp)
```

The pattern has `r childcare_ppp$n` points. The observation window is the rectangle enclosing
the points,
**not yet** the study area. That matters, because every first-order measure is computed
relative to the window.

## Defining the study area

The window should be the area we actually want to make claims about. For a national-level
analysis, that is the Singapore mainland boundary, which I build by dissolving the subzone
boundaries.

```{r}
sg_boundary <- st_union(mpsz)
sg_owin <- as.owin(sg_boundary)
```

```{r}
plot(sg_owin)
```

Now the points are clipped to the window. Any point outside (there should be none in this
case) would be dropped here.

```{r}
childcareSG_ppp <- childcare_ppp[sg_owin]
```

```{r}
childcareSG_ppp$n
```

## Clark–Evans nearest-neighbour test

Clark and Evans (1954) compare the mean observed nearest-neighbour distance against what would
be expected under **Complete Spatial Randomness (CSR)**. The index *R* is the ratio of the two:

- *R* < 1: points are closer than random, so **clustered**
- *R* ≈ 1: consistent with CSR
- *R* > 1: points are further apart than random, so **regular/dispersed**

```{r}
clarkevans.test(childcareSG_ppp,
                correction = "none",
                clipregion = sg_owin,
                alternative = "clustered")
```

The test can also be run as a Monte Carlo simulation rather than the analytic normal
approximation, which is slower but does not rely on distributional assumptions.

```{r}
clarkevans.test(childcareSG_ppp,
                correction = "none",
                clipregion = sg_owin,
                alternative = "clustered",
                method = "MonteCarlo",
                nsim = 999)
```

## Kernel density estimation

The Clark–Evans index collapses the whole pattern into one number. KDE keeps the geography:
it spreads a kernel over each point and sums the surfaces, producing a continuous density
field that shows *where* intensity is high and low.

### Automatic bandwidth selection

The bandwidth (the `sigma` of the kernel) controls how far each point's influence extends,
and it drives the appearance of the result more than anything else in KDE. The package offers
several automatic selectors; the default is `bw.diggle`.

```{r}
kde_SG_diggle <- density(childcareSG_ppp,
                         sigma = bw.diggle,
                         edge = TRUE,
                         kernel = "gaussian")
```

```{r}
plot(kde_SG_diggle)
```

```{r}
summary(kde_SG_diggle)
```

The densities here are in **per square metre** units, which are awkward to read (the integral
of the surface is the number of points). Rescaling the pattern to kilometres makes the values
interpretable as centres per square kilometre.

```{r}
childcareSG_ppp_km <- rescale.ppp(childcareSG_ppp, 1000, "km")
kde_childcareSG_km <- density(childcareSG_ppp_km,
                              sigma = bw.diggle,
                              edge = TRUE,
                              kernel = "gaussian")
```

```{r}
summary(kde_childcareSG_km)
```

### Comparing the selectors

The four common selectors optimise different criteria, so they return different bandwidths.
Rather than accept the default silently, I compare them.

```{r}
bw.CvL(childcareSG_ppp_km)
```

```{r}
bw.scott(childcareSG_ppp_km)
```

```{r}
bw.ppl(childcareSG_ppp_km)
```

```{r}
bw.diggle(childcareSG_ppp_km)
```

There is an order-of-magnitude spread here, from about 0.39 km (`bw.ppl`) to 4.54 km
(`bw.CvL`). Since all four are defensible choices, the map itself is only as settled as the
selector behind it. Part 3 works through what that does to the surface.

`bw.scott` stands out here, because it is the rule Donthu and Rust use. They pick the
bandwidth that would be optimal if the data were bivariate normal, while admitting "we have no
reason to actually expect the underlying distribution to be bivariate normal". On this data
that rule returns about 2.2 km against `bw.diggle`'s 0.30 km, roughly seven times wider. Since
their paper uses the density surface to choose a store location, the two rules would suggest
visibly different catchments from identical data. The 1989 advice is not wrong, but it is much
smoother than a modern
cross-validation selector, and a siting decision taken from one surface rather than the other
would land in a different place.

### The kernel function

The kernel is the shape of the influence region. spatstat's default is Gaussian; quartic and
Epanechnikov kernels are common alternatives. At the same bandwidth they produce broadly
similar surfaces, differing mostly in how sharply the tails fall off.

```{r}
par(mfrow = c(1, 2))
plot(density(childcareSG_ppp_km, sigma = 0.296, kernel = "gaussian"), main = "Gaussian")
plot(density(childcareSG_ppp_km, sigma = 0.296, kernel = "quartic"), main = "Quartic")
par(mfrow = c(1, 1))
```

### Fixed versus adaptive bandwidth

A fixed bandwidth applies the same smoothing radius everywhere. An **adaptive** bandwidth
shrinks the radius where points are dense and widens it where they are sparse, which is
usually the better assumption for a service like childcare whose supply follows population
density closely.

```{r}
kde_fb <- density(childcareSG_ppp_km, sigma = 0.6, edge = TRUE, kernel = "gaussian")
```

```{r}
kde_ab <- adaptive.density(childcareSG_ppp_km, method = "kernel")
```

```{r}
par(mfrow = c(1, 2))
plot(kde_fb, main = "Fixed bandwidth (0.6 km)")
plot(kde_ab, main = "Adaptive bandwidth")
par(mfrow = c(1, 1))
```

## A cartographic KDE map

The spatstat density object is a pixel image. To map it properly I convert it to a raster,
assign the correct CRS, and map with tmap.

A unit note, because it is easy to get wrong: the density computed on the *metre*-based
pattern is in **centres per square metre**, which is awkward to read. I compute it in metres
(so the raster extent is in metres and the CRS is honest) and multiply the values by 10⁶ to
display them as centres per km².

```{r}
kde_terra <- rast(density(childcareSG_ppp, sigma = bw.diggle, edge = TRUE)) * 1e6
crs(kde_terra) <- "EPSG:3414"
```

```{r}
kde_terra
```

```{r}
tm_shape(mpsz) +
  tm_borders(col = "grey60", lwd = 0.3) +
  tm_shape(kde_terra) +
  tm_raster(col.scale = tm_scale_continuous(values = "viridis"),
            col.legend = tm_legend(title = "Centres per km²",
                                   title.size = 0.8,
                                   text.size = 0.7,
                                   position = tm_pos_in("right", "bottom"),
                                   bg.color = "white",
                                   bg.alpha = 0.7,
                                   frame = TRUE)) +
  tm_graticules(lines = FALSE) +
  tm_layout(legend.frame = TRUE) +
  tm_compass(position = tm_pos_in("left", "top")) +
  tm_scalebar(position = tm_pos_in("left", "bottom"))
```

## Analysis by planning area

National-level statistics hide a lot of local variation. The chapters repeat the Clark–Evans
test and KDE within individual planning areas. I use the four areas the book uses (Punggol,
Tampines, Choa Chu Kang and Jurong West) because they are demographically and physically
different from each other.

```{r}
pg <- mpsz |> filter(PLN_AREA_N == "PUNGGOL")
tm <- mpsz |> filter(PLN_AREA_N == "TAMPINES")
ck <- mpsz |> filter(PLN_AREA_N == "CHOA CHU KANG")
jw <- mpsz |> filter(PLN_AREA_N == "JURONG WEST")
```

```{r}
childcare_pg_ppp <- childcare_ppp[as.owin(pg)]
childcare_tm_ppp <- childcare_ppp[as.owin(tm)]
childcare_ck_ppp <- childcare_ppp[as.owin(ck)]
childcare_jw_ppp <- childcare_ppp[as.owin(jw)]
```

```{r}
c(punggol = childcare_pg_ppp$n, tampines = childcare_tm_ppp$n,
  choa_chu_kang = childcare_ck_ppp$n, jurong_west = childcare_jw_ppp$n)
```

```{r}
clarkevans.test(childcare_ck_ppp, correction = "none", alternative = "two.sided", nsim = 999)
```

```{r}
clarkevans.test(childcare_tm_ppp, correction = "none", alternative = "two.sided", nsim = 999)
```

```{r}
par(mfrow = c(2, 2))
plot(density(rescale.ppp(childcare_pg_ppp, 1000, "km"), sigma = bw.diggle), main = "Punggol")
plot(density(rescale.ppp(childcare_tm_ppp, 1000, "km"), sigma = bw.diggle), main = "Tampines")
plot(density(rescale.ppp(childcare_ck_ppp, 1000, "km"), sigma = bw.diggle), main = "Choa Chu Kang")
plot(density(rescale.ppp(childcare_jw_ppp, 1000, "km"), sigma = bw.diggle), main = "Jurong West")
par(mfrow = c(1, 1))
```

# Part 2: Second-order methods

Second-order methods ask **how the points are arranged relative to each other**: clustered,
dispersed or random, and at what distance. Where Part 1 says "intensity is high here",
Part 2 says "points attract or repel each other at these scales".

For each function I compare **Choa Chu Kang** (a mature HDB town with a distinct centre) and
**Tampines** (a large, densely built town), which lets the same statistic run against two
quite different urban layouts.

## The G function

The G function is the cumulative distribution of nearest-neighbour distances: *G(r)* is the
probability that a random point has a neighbour within distance *r*. If points cluster, *G(r)*
rises quickly for small *r*.

```{r}
G_ck <- Gest(childcare_ck_ppp, correction = "border")
plot(G_ck, xlim = c(0, 500))
```

Testing against CSR requires a Monte Carlo **envelope**: the function is recomputed on many
simulated CSR patterns, and the observed curve is compared against the simulated band.

```{r}
G_ck.csr <- envelope(childcare_ck_ppp, Gest, nsim = 999)
plot(G_ck.csr)
```

```{r}
G_tm <- Gest(childcare_tm_ppp, correction = "best")
G_tm.csr <- envelope(childcare_tm_ppp, Gest, correction = "all", nsim = 999)
plot(G_tm.csr)
```

## The F function

The F function is the complement view: the distance from **arbitrary locations** to the
nearest point, not from points to points. Because it measures empty space rather than
neighbour distances, it responds to gaps in the pattern where G responds to close pairs.

```{r}
F_ck.csr <- envelope(childcare_ck_ppp, Fest, nsim = 999)
plot(F_ck.csr)
```

```{r}
F_tm.csr <- envelope(childcare_tm_ppp, Fest, correction = "all", nsim = 999)
plot(F_tm.csr)
```

## The K function

Ripley's K is the expected number of other points within distance *r*, standardised by
intensity. It summarises dependence across *all* distances, not just nearest neighbours. Under
CSR, *K(r) = πr²*. The plot below uses the variance-stabilised form *K(r) − πr²* so the CSR
baseline is the horizontal line at zero.

```{r}
K_ck <- Kest(childcare_ck_ppp, correction = "Ripley")
plot(K_ck, . - r ~ r, ylab = "K(d) - r", xlab = "d (m)")
```

```{r}
K_ck.csr <- envelope(childcare_ck_ppp, Kest, nsim = 99, rank = 1, glocal = TRUE)
plot(K_ck.csr, . - r ~ r, xlab = "d", ylab = "K(d) - r")
```

```{r}
K_tm <- Kest(childcare_tm_ppp, correction = "Ripley")
K_tm.csr <- envelope(childcare_tm_ppp, Kest, nsim = 99, rank = 1, glocal = TRUE)
plot(K_tm.csr, . - r ~ r, xlab = "d", ylab = "K(d) - r", xlim = c(0, 500))
```

## The L function

The L function is the variance-stabilised transform of K: *L(r) = √(K(r)/π) − r*. Under CSR,
*L(r) = 0* at all distances. Positive *L* indicates clustering; negative *L* indicates
regularity. The transform makes deviations from CSR easier to read at all scales.

```{r}
L_ck.csr <- envelope(childcare_ck_ppp, Lest, nsim = 99, rank = 1, glocal = TRUE)
plot(L_ck.csr, . - r ~ r, xlab = "d", ylab = "L(d) - r")
```

```{r}
L_tm.csr <- envelope(childcare_tm_ppp, Lest, nsim = 99, rank = 1, glocal = TRUE)
plot(L_tm.csr, . - r ~ r, xlab = "d", ylab = "L(d) - r", xlim = c(0, 500))
```

# Part 3: Beyond the chapters

The reproduction above is the assignment. This part records what I noticed while working
through it: which choices the chapters let the software make silently, whether the conclusions
survive when those choices are changed, and what changes when the week's readings and slides
are actually applied rather than read. The sections follow that order.

## Reviewing the chapters' choices

### Checking for coincident points

This is a prerequisite rather than a criticism, and it belongs before any of the distance
methods run. Baddeley, Rubak and Turner (2015), listed in the Lesson 2 references and written
by the authors of spatstat, set out why in section 3.4.4:

> Duplication of points is important, because statistical methodology for spatial point
> processes (as used in this book) is based largely on assumption that processes are simple,
> i.e. that points of the process can never be coincident. When the data have coincident
> points, some statistical procedures designed for simple point processes will be severely
> affected. For example, the pair correlation function will have an infinite value at distance
> zero. It is strongly advisable to check for duplicated points and to decide on a strategy for
> dealing with them if they are present.

Coincident points are plausible for this data, since several childcare centres can operate in
the same mall or HDB block and geocoding would return one coordinate for all of them. So the
check is worth running rather than assuming.

```{r}
c(points = npoints(childcareSG_ppp),
  unique_locations = npoints(unique(childcareSG_ppp)),
  duplicated = sum(duplicated(childcareSG_ppp)),
  nn_distances_of_zero = sum(nndist(childcareSG_ppp) == 0))
```

This layer is clean: every centre has its own coordinate, no nearest-neighbour distance is
zero, and the pair correlation function used later in Part 3 is safe to run. Nothing needs
fixing here, but that is only knowable after running the check.

It is worth knowing what a failure here would look like, because I met one. An earlier draft of
this exercise used a different geocoded childcare file, `CHILDCARE.shp`, distributed with a
later chapter of the same book. That version had 174 duplicated coordinates, including one
location carrying 25 centres, and it changed results: the Clark-Evans index read 0.495 instead
of 0.494, and the adaptive-bandwidth surface in Part 1 peaked at 314 centres per square
kilometre instead of about 47, because an adaptive kernel shrinks its bandwidth wherever
intensity is high and a 25-point stack drives that bandwidth towards zero. The resulting map
was almost entirely dark, since the colour ramp stretched to a maximum reached by a dozen
pixels. Fixed-bandwidth KDE was unaffected, because it spreads every point over the same
radius regardless of how many neighbours it has.


### A clipregion argument that does nothing

A smaller one. The chapter calls the Clark-Evans test like this:

```r
clarkevans.test(childcareSG_ppp, correction = "none",
                clipregion = "sg_owin", alternative = c("clustered"))
```

`clipregion` is passed the **string** `"sg_owin"` rather than the object `sg_owin`. It reads as
though the test is being clipped to the Singapore boundary. It is not: the argument is only
used by the guard-area correction, and with `correction = "none"` it is ignored entirely.
Passing the object, passing the string, or omitting the argument all return R = 0.49444. The
line is harmless but misleading, and I dropped the quotes in Part 1.

### The envelopes are pointwise, and one argument does not exist

Chapter 5 builds the K and L envelopes like this:

```r
K_ck.csr <- envelope(childcare_ck_ppp, Kest, nsim = 99, rank = 1, glocal = TRUE)
```

`glocal` is not an argument of `envelope()`. The argument that controls this behaviour is
called `global`, so `glocal = TRUE` is absorbed by `...`, passed down to `Kest`, and ignored.
The envelopes come out **pointwise**, which is the default, and identical to what you get by
leaving the argument out.

```{r}
set.seed(1)
e_default <- envelope(childcare_ck_ppp, Lest, nsim = 99, verbose = FALSE)
set.seed(1)
e_glocal  <- envelope(childcare_ck_ppp, Lest, nsim = 99, glocal = TRUE, verbose = FALSE)
set.seed(1)
e_global  <- envelope(childcare_ck_ppp, Lest, nsim = 99, global = TRUE, verbose = FALSE)

c(glocal_same_as_default = identical(e_glocal$hi, e_default$hi),
  global_same_as_default = identical(e_global$hi, e_default$hi))
```

The distinction matters, and it is the difference between two questions.

A **pointwise** envelope is built separately at each distance *r*. At any single *r* it has the
stated error rate, but the plot shows hundreds of distances at once, so the chance that the
observed curve strays outside *somewhere* is far higher than the nominal level. Reading a
pointwise band as though a single excursion proves clustering is the multiple-comparison
problem in graphical form.

A **global** envelope is built from the largest deviation each simulated curve reaches anywhere
in the range, so the band is wider and one excursion anywhere is enough to reject at the stated
level.

```{r}
i <- round(length(e_default$r) / 2)
data.frame(
  envelope = c("pointwise", "global"),
  lo = round(c(e_default$lo[i], e_global$lo[i]), 1),
  hi = round(c(e_default$hi[i], e_global$hi[i]), 1),
  frac_r_outside = round(c(mean(e_default$obs > e_default$hi | e_default$obs < e_default$lo),
                           mean(e_global$obs  > e_global$hi  | e_global$obs  < e_global$lo)), 3)
)
```

For Choa Chu Kang's L function the pointwise band puts about 43% of the distance range outside
the envelope, and the global band about 1%. Same data, same simulations, different question.

Because of this I report the second-order results below as a yes-or-no answer under a global
envelope, rather than as a percentage of the range under a pointwise one. The percentage is
useful for describing *where* along the distance axis a departure sits, but it is not a
significance level and should not be read as one.

One practical detail: the pair correlation function is undefined at *r* = 0 and returns `Inf`
there, which a global envelope will always flag. Every test below is evaluated over *r* > 0
with non-finite values dropped.

### Choosing the study window

Every first-order and second-order statistic is computed inside a window, so the window you
choose changes the numbers you get. Three things I noticed:

1. **The national window follows the coastline.** `as.owin(sg_boundary)` produces
   an irregular polygon. The nearest-neighbour distances near the coast are truncated by the
   boundary, which biases *R* upward (points appear more dispersed than they are) unless an
   edge correction is applied. `clarkevans.test` with `correction = "none"` makes no
   allowance for this.

2. **The national Clark–Evans index answers a question I did not mean to ask.** *R* = 0.49
   confirms clustering, but childcare supply follows population, and population in Singapore
   is itself clustered. A nearest-neighbour test against *uniform* CSR asks the wrong
   question when the points track an underlying covariate. The same weakness affects the
   second-order tests in Part 2.

3. **The window changes how strong the result looks.** Clark-Evans compares the observed
   mean nearest-neighbour distance with the distance
   expected under CSR, and that expected distance depends on the area of the window through
   the intensity *n/A*. Enlarge the window without adding points and CSR expects the points to
   be further apart, so the same points look more clustered by comparison.

   The test below holds the 74 Choa Chu Kang centres fixed and only changes the window.

```{r}
ck_pts <- childcare_ck_ppp
windows <- list(
  `planning-area polygon` = as.owin(st_union(ck)),
  `polygon + 500 m buffer` = as.owin(st_buffer(st_union(ck), 500)),
  `convex hull of points`  = convexhull.xy(ck_pts$x, ck_pts$y),
  `bounding box`           = owin(range(ck_pts$x), range(ck_pts$y))
)
do.call(rbind, lapply(names(windows), function(nm) {
  w <- windows[[nm]]
  Y <- suppressWarnings(ppp(ck_pts$x, ck_pts$y, window = w))
  t <- suppressWarnings(clarkevans.test(Y, correction = "none",
                                        alternative = "two.sided", nsim = 999))
  data.frame(window = nm, area_km2 = round(spatstat.geom::area(w) / 1e6, 2),
             R = round(unname(t$statistic), 3), p = round(t$p.value, 3))
}))
```

   All four windows agree that the pattern is clustered, which is reassuring, but they do not
   agree on how strongly. *R* ranges from 0.87 under the convex hull to 0.59 under the buffered
   polygon, and the p-value moves from 0.026 to below 0.001. Nothing changed but the outline I
   drew around the same 74 centres. A result reported as "R = 0.87, p = 0.03" and one reported
   as "R = 0.59, p < 0.001" would be written up quite differently, even though they describe
   the same centres. Wartman and Malasavage's insistence that "the study area should be
   determined objectively, rather than arbitrarily" is doing real work here, and neither
   chapter mentions the issue.

   O'Sullivan and Unwin explain the mechanism behind the buffered case: edge effects arise
   because "sites in the center of the study area can have nearby observations in all
   directions, whereas sites at the edges only have neighbors toward the center". Padding the
   window adds empty space full of edge, which is why *R* falls furthest there.

   Two of the applied papers sit usefully side by side here. Borruso and Porceddu write
   that the bandwidth "represents the only arbitrary variable which defines the extent of the
   searching function", while Wartman and Malasavage insist that "the study area should be
   determined objectively, rather than arbitrarily". Borruso's sentence is about the KDE
   kernel, so the table above is not a counter-example to it. But his paper never treats the
   study area as a choice at all, and on this data the study area is what moved the verdict,
   with no bandwidth involved. Following Borruso alone, an analyst would carefully document
   the one parameter he names and say nothing about the one that mattered.

### How much the bandwidth choice matters

Yin (2020), one of the week's core readings, frames KDE as a bias-variance trade-off: too small
a bandwidth produces a noisy surface that overfits the sample, too large a bandwidth smooths
away the structure we are looking for. The four automatic selectors sit at different points on
that trade-off, and on this data they are far apart.

| Selector | Bandwidth (km) | What it optimises |
|---|---|---|
| `bw.diggle` | 0.30 | Edge-corrected minimisation of mean-square error (the book's default) |
| `bw.ppl` | 0.39 | Likelihood cross-validation, driven by the data rather than a formula |
| `bw.scott` | 2.16 / 1.40 (x / y) | A rule of thumb based on the pattern's own spread |
| `bw.CvL` | 4.54 | Coarser smoothing, low variance, high bias |

The spread is more than tenfold. Since the chapters pick one without saying why, the question
is how much of the map depends on that choice, so I re-ran the national surface under the two
extremes alongside the default.

```{r}
kde_ppl <- density(childcareSG_ppp_km, sigma = bw.ppl, edge = TRUE)
kde_cvl <- density(childcareSG_ppp_km, sigma = bw.CvL, edge = TRUE)
```

```{r}
par(mfrow = c(1, 3))
plot(kde_childcareSG_km, main = "bw.diggle")
plot(kde_ppl, main = "bw.ppl")
plot(kde_cvl, main = "bw.CvL")
par(mfrow = c(1, 1))
```

The broad pattern is stable across all three: the same dense central-east corridor, the same
relative voids in the north. What changes is the granularity. The two small bandwidths break
the surface into many separate peaks, while `bw.CvL` merges them into a few large blobs. So the
statement "childcare is concentrated in the central and eastern regions" is robust, but any
claim about a *particular* hotspot is a claim about the bandwidth as much as about the data.

The trade-off can be measured rather than eyeballed. The coefficient of variation of the
density values says how uneven each surface is.

```{r}
cv_im <- function(x) { v <- as.vector(x$v); v <- v[!is.na(v)]; sd(v) / mean(v) }
c(`bw.ppl` = cv_im(kde_ppl), `bw.diggle` = cv_im(kde_childcareSG_km), `bw.CvL` = cv_im(kde_cvl))
```

The ordering follows the bandwidths rather than the reputations of the methods. `bw.diggle`
returns the smallest bandwidth here at 0.30 km and gives the most uneven surface, `bw.ppl` at
0.39 km is close behind, and `bw.CvL` at 4.54 km is far smoother than either. The book's
default therefore sits at the variance-heavy end of the range on this data rather than in the
middle, which is the opposite of what I assumed before measuring it.


### A unit mismatch in the KDE map

The chapters build the KDE map by rescaling the point pattern to **kilometres**, converting
the result to a raster, and then assigning the raster the CRS `EPSG:3414`, which is a
**metre** projection. The raster's extent (about 0–55 in km units) is then interpreted by
tmap as 0–55 *metres*. The rendered map is a sliver of the coastline about fifty metres wide,
not the country, even though the density surface is correct.

The fix used for the KDE map in Part 1 is to compute the density on the metre-based pattern, so
the raster extent
is genuinely in metres and the CRS assignment is honest, then rescale *only the values* by
10⁶ for a per-km² legend. The change is small, and the map then shows Singapore properly.

What makes this one instructive is that nothing failed. The code ran without error and
produced a map that looked reasonable at a glance. I only caught it by reading the graticule
labels, which spanned about a hundredth of a degree.

### Why adaptive suits this data

The default `density()` uses a fixed bandwidth: every point spreads its influence to the same
radius. Adaptive bandwidth (`adaptive.density`) lets the radius shrink where points are dense.
For childcare, whose supply tracks population density, adaptive is the more defensible
assumption: a centre in a dense town competes with neighbours close by, while one in a
sparse area stands alone.

In the comparison plotted in Part 1, the fixed surface smooths the dense corridor into one
continuous mass while the adaptive surface keeps more local structure. Which one is preferable
depends on what you are asking, so the choice should be stated rather than left at
the default.

### Do the functions agree?

All four functions describe the same underlying dependence from different angles, so reading
them against each other is a way of checking that a result is not an artefact of one
particular statistic:

- **G and K/L both detect clustering** in both planning areas, with the observed curves sitting
  above the CSR band at short distances.
- **F rarely leaves its band.** F is computed from arbitrary locations, so it reflects the
  *average* empty-space structure. With only 74 and 117 points in these two areas, F is blunt; it
  contributes little beyond confirming there is no large-scale regularity.
- **K and L give the cleaner contrast between the two towns.** G looks only at each point's
  nearest neighbour and is sensitive to the border correction, while K and L average over all
  pairs. That stability comes from accumulating counts across distances, which is an advantage
  here and a liability later: the section on nulls below shows the same accumulation producing
  a signal that turns out not to be there.

### Why the two towns differ

Both towns show clustering at short distances, but Tampines is more strongly clustered, with
its K/L curves rising higher above the band.

The likely explanation is the housing. Tampines is one of the earliest and densest new towns,
built around a tight commercial centre with residential blocks close by, while Choa Chu Kang
is younger and more spread out. Childcare centres locate near homes, so the arrangement of the
centres largely follows the arrangement of the flats. That explanation comes from knowing the
towns, not from the curves.

## Applying the week's material

I have already used two of the readings above: Yin (2020) for the bandwidth comparison, and
Yuan et al. (2020) for the argument that CSR is the wrong null when points follow population.
The two methods below come from material outside r4gdsa Chapters 4 and 5. Quadrat analysis is
covered in the Lesson 2 slides and in Anselin's *Point Pattern Analysis: Quadrat Counts*. The
inhomogeneous null then tests Yuan's warning, which is that CSR treats every location as
equally likely to hold a centre, so any service that follows the population will be reported as
clustered whether or not the centres have anything to do with each other.

### Quadrat analysis

Neither r4gdsa Chapter 4 nor Chapter 5 covers quadrat analysis, but the Lesson 2 slides and
Anselin's *Point Pattern Analysis: Quadrat Counts* both do, so I have added it here as a third
first-order method. It is the oldest of them. You divide the study area into a grid, count the
points in each cell, and compare those counts with what CSR would produce. Quadrat analysis
adds a different kind of evidence from Clark-Evans, which measures distances between
centres. Quadrat analysis ignores distance entirely and looks only at how counts vary between
cells, so if the two agree, the finding does not depend on either way of measuring.

```{r}
qc <- quadratcount(childcareSG_ppp_km, nx = 6, ny = 6)
```

```{r}
qc
```

The slides give the reading rule through the variance-mean ratio. Under CSR the counts follow
a Poisson distribution, whose variance equals its mean, so VMR ≈ 1 is random, VMR > 1 is
clustered and VMR < 1 is regular.

```{r}
counts <- as.vector(qc)
c(variance = var(counts), mean = mean(counts), VMR = var(counts) / mean(counts))
```

```{r}
quadrat.test(childcareSG_ppp_km, nx = 6, ny = 6)
```

The chi-square test is unambiguous: X² = 1389 on 30 degrees of freedom, p < 2.2e-16, the same
strong clustering Clark-Evans found. The raw VMR of 143 is also far above 1, but it is
inflated and should not be reported on its own. The count grid inherits the coastline, so
several edge tiles straddle land and sea and sit near zero, while the busiest central tile
holds 277 centres.

This is the one method in the exercise where the **modifiable areal unit problem** applies in
its strict sense, because it is the only one that aggregates points into areal units before
computing anything. O'Sullivan and Unwin describe MAUP as the problem that "the aggregation
units used are arbitrary with respect to the phenomena under investigation, yet the units used
will affect statistics determined on the basis of data reported in this way".

A 6 by 6 grid is exactly such an arbitrary unit. Change `nx` and `ny`, or shift the grid
origin, and the counts, the VMR and the chi-square statistic all move. That sensitivity, which
the Lesson 2 slides list among the weaknesses of quadrat analysis, is why the field moved to
distance-based methods.

So quadrat analysis agrees with Clark-Evans that the pattern is clustered, which is
reassuring, but its headline number depends on a grid I chose arbitrarily. That dependence is
the reason later methods measure distances between points instead of counts inside cells.

### Two functions the chapters leave out

r4gdsa Chapters 4 and 5 stop at G, F, K and L. Two of the other references go further, and both
additions
turned out to matter.

Anselin's *Point Pattern Analysis: F and J Functions* introduces the **J function**, which
combines the two
curves already computed: `J(r) = (1 - G(r)) / (1 - F(r))`. Under CSR it is flat at 1, below 1
means clustering and above 1 means inhibition. It is a tidier answer to the question I asked
earlier about whether G and F agree, since it combines them arithmetically instead of leaving
me to compare two plots by eye.

*Chapter 11 Point Pattern Analysis* of **Intro to GIS and Spatial Analysis**, listed in the
references, makes a second point. K and L are **cumulative**: they count every point within
distance *r*, so clustering at short range keeps inflating the curve at longer ranges. The
**pair correlation function** counts only points in a narrow band at *r*, so each distance is
judged on its own.

::: {.callout-warning appearance="simple"}
## Two different functions, one letter

The pair correlation function is written **lowercase g(r)**, and it is *not* the **G-function**
from the lesson slides and Chapter 5. They are unrelated:

- **G(r)**, `Gest()`, is the cumulative distribution of nearest-neighbour distances, so it uses
  one distance per point and says nothing about what lies further away.
- **g(r)**, `pcf()`, is derived from Ripley's K by differentiating it, so it uses all pairs of
  points and reports the density of pairs at each separation.

Gimond introduces g in section 11.3.3 as a modification of K, not of G, which is the clearest
way to keep them apart. To avoid the collision in code I have named the objects `pcf_*` rather
than `g_*`, since Part 2 already uses `G_ck` and `G_tm` for the G-function.
:::

```{r}
childcare_tm_ppp_km <- rescale.ppp(childcare_tm_ppp, 1000, "km")
childcare_ck_ppp_km <- rescale.ppp(childcare_ck_ppp, 1000, "km")

set.seed(42)
J_tm <- envelope(childcare_tm_ppp_km, Jest, nsim = 99, verbose = FALSE)
pcf_tm <- envelope(childcare_tm_ppp_km, pcf,  nsim = 99, verbose = FALSE)
J_ck <- envelope(childcare_ck_ppp_km, Jest, nsim = 99, verbose = FALSE)
pcf_ck <- envelope(childcare_ck_ppp_km, pcf,  nsim = 99, verbose = FALSE)
```

```{r}
par(mfrow = c(1, 2))
plot(J_tm, main = "J function: Tampines", legend = FALSE)
plot(pcf_tm, main = "Pair correlation g: Tampines", legend = FALSE)
par(mfrow = c(1, 1))
```

```{r}
outside <- function(env, side = c("above", "below")) {
  side <- match.arg(side)
  if (side == "above") mean(env$obs > env$hi, na.rm = TRUE)
  else mean(env$obs < env$lo, na.rm = TRUE)
}
data.frame(
  town        = c("Tampines", "Choa Chu Kang"),
  J_clustered = c(outside(J_tm, "below"), outside(J_ck, "below")),
  g_clustered = c(outside(pcf_tm, "above"), outside(pcf_ck, "above"))
)
```

Those percentages are pointwise, so they describe where along the distance axis each curve
leaves its band, not whether the departure is significant overall. The section on envelopes
above explains why. Repeating both tests under a global envelope, over *r* > 0, gives the
answer that can actually be reported:

```{r}
exits <- function(e) {
  k <- is.finite(e$obs) & e$r > 0
  any(e$obs[k] > e$hi[k] | e$obs[k] < e$lo[k], na.rm = TRUE)
}
glob <- function(X, fun, ...) { set.seed(42)
  envelope(X, fun, nsim = 99, global = TRUE, verbose = FALSE, ...) }

data.frame(
  town = c("Tampines", "Choa Chu Kang"),
  J_global   = c(exits(glob(childcare_tm_ppp_km, Jest)),
                 exits(glob(childcare_ck_ppp_km, Jest))),
  pcf_global = c(exits(glob(childcare_tm_ppp_km, pcf)),
                 exits(glob(childcare_ck_ppp_km, pcf)))
)
```

The two functions part company. The pair correlation function rejects CSR for both towns under
the stricter test, so the clustering it reports is not an artefact of scanning many distances.
The J function does not reject for either town, even though its pointwise band was left over
77% of the range in Tampines. J is built from G and F, both of which are cumulative
nearest-neighbour constructions, and its deviations are correspondingly small and spread out
rather than concentrated. Under a band sized for the whole curve, they no longer stand out.

So on the CSR comparison I would report the pair correlation result and treat the J result as
suggestive only. Under CSR, Tampines is clustered on both pointwise measures, with J outside
its band over about 77% of the range and g over 98%, while Choa Chu Kang is much weaker at
roughly 20% and 7%.

Clark-Evans reports both towns as significantly clustered, at R = 0.67 for Tampines and
R = 0.84 for Choa Chu Kang, and stops there. What the distance-resolved functions add is that
the two are clustered in different ways. Tampines departs from randomness at nearly every
distance, while Choa Chu Kang does so only over a narrow band. A single nearest-neighbour index
cannot express that difference, because it collapses the whole pattern into one number before
distance enters the picture.

The non-cumulative view matters again in the next section, where it changes a conclusion I had
already written down.

### Choosing a null hypothesis

Everything in Part 2 was tested against Complete Spatial Randomness, which assumes a centre
was equally likely to open anywhere in the town. For a service that follows families, that
assumption is close to guaranteed to report clustering. This section builds two better nulls
and then sets out what neither of them can settle.

#### A null estimated from the pattern itself

Yuan et al. (2020) warn that CSR is a weak null for events that track covariates. The warning is
testable directly. The homogeneous L-function in Part 2 compares the observed pattern with a
completely random one, which is loaded here: if childcare follows population density, the
intensity gradient is counted as clustering. The inhomogeneous L-function removes the trend
first. It estimates the intensity (here, the KDE surface) and tests the pattern against an
inhomogeneous Poisson process with that intensity, so any clustering that survives is
interaction between points that the trend cannot explain.

```{r}
set.seed(42)   # envelope simulations are random; fix the seed so the results are reproducible
lambda_tm <- density(childcare_tm_ppp_km, sigma = bw.diggle, edge = TRUE)
lambda_tm$v <- pmax(lambda_tm$v, 0.01)   # Linhom needs strictly positive intensity
L_tm_inhom <- envelope(childcare_tm_ppp_km, Linhom, lambda = lambda_tm,
                       nsim = 99, correction = "Ripley")
```

```{r}
plot(L_tm_inhom, . - r ~ r, main = "Inhomogeneous L: Tampines")
```

```{r}
lambda_ck <- density(childcare_ck_ppp_km, sigma = bw.diggle, edge = TRUE)
lambda_ck$v <- pmax(lambda_ck$v, 0.01)
L_ck_inhom <- envelope(childcare_ck_ppp_km, Linhom, lambda = lambda_ck,
                       nsim = 99, correction = "Ripley")
```

```{r}
plot(L_ck_inhom, . - r ~ r, main = "Inhomogeneous L: Choa Chu Kang")
```

A coarse summary of where the observed curve sits relative to the envelope makes the difference
easy to read.

```{r}
frac_band <- function(env) c(above = mean(env$obs > env$hi, na.rm = TRUE),
                            below = mean(env$obs < env$lo, na.rm = TRUE))
list(Tampines = frac_band(L_tm_inhom),
     `Choa Chu Kang` = frac_band(L_ck_inhom))
```

Under the homogeneous test, both towns look strongly clustered. Under the inhomogeneous null
the picture changes, and the two towns differ. For Tampines the observed curve sits below the
envelope for most of the plotted range (about 92% of it), which is the signature of
regularity, not clustering: once the density of centres is accounted for, centres are further
apart than the trend predicts. For Choa Chu Kang the curve hovers at or below the envelope
(about 49% of the range below it, essentially none above), so the clustering the homogeneous
test reported for this town was almost entirely the intensity trend, with no clear residual
signal left.

Both results line up with Yuan's warning: a homogeneous null will overstate clustering for
any service that follows population.

The Tampines curve sitting below the envelope looks like regularity, and my first reading of it
was that centres compete and space themselves out. That reading was wrong, and the pair
correlation function from the previous section is what shows it. L is cumulative, so a small
deviation repeated across short distances accumulates into a curve that drifts below the band
without any single distance being unusual. Running the same null through the non-cumulative
`pcfinhom` removes that accumulation.

```{r}
set.seed(42)
pcfi_tm <- envelope(childcare_tm_ppp_km, pcfinhom, lambda = lambda_tm,
                  nsim = 99, verbose = FALSE)
pcfi_ck <- envelope(childcare_ck_ppp_km, pcfinhom, lambda = lambda_ck,
                  nsim = 99, verbose = FALSE)
```

```{r}
par(mfrow = c(1, 2))
plot(pcfi_tm, main = "Inhomogeneous g: Tampines", legend = FALSE)
plot(pcfi_ck, main = "Inhomogeneous g: Choa Chu Kang", legend = FALSE)
par(mfrow = c(1, 1))
```

```{r}
data.frame(
  town = c("Tampines", "Choa Chu Kang"),
  g_inhom_outside_envelope = c(outside(pcfi_tm, "above") + outside(pcfi_tm, "below"),
                               outside(pcfi_ck, "above") + outside(pcfi_ck, "below"))
)
```

The inhomogeneous g never leaves its envelope for either town at any distance. So the
regularity I thought I had found was an artefact of L accumulating, not a property of the
centres. Once the intensity trend is accounted for, there is no detectable interaction between
childcare centres at any distance.

The direction of the correction is worth being clear about. A global band is wider, which
makes it harder to reject the null and easier to stay inside. So for a result that rejects,
like the pair correlation test above, the global version is the stricter test. For a result
that finds nothing, like this one, the narrower pointwise band was already the harder test to
pass, and the global check below is a formality. I ran it so both conclusions are reported on
the same footing.

```{r}
data.frame(
  town = c("Tampines", "Choa Chu Kang"),
  pcfinhom_global_exits = c(exits(glob(childcare_tm_ppp_km, pcfinhom, lambda = lambda_tm)),
                            exits(glob(childcare_ck_ppp_km, pcfinhom, lambda = lambda_ck)))
)
```

Neither town's curve leaves the wider band either, as expected. The absence of residual
interaction therefore does not depend on which envelope is used, unlike the J result above.

Neither r4gdsa chapter uses a non-cumulative function, so this correction is not available from
them.

#### A null built from resident population

The inhomogeneous tests above have a circularity in them, which I set out in the section below:
the intensity was estimated from a kernel density of the very points being tested, so genuine
clustering could be absorbed into the "trend" before the test runs. Floch et al. say what the
fix is, namely "a model that links a covariable to the intensity". The covariable for childcare
is obvious and I already have it, since Hands-on Exercise 1 used the same census table:
**resident children aged 0 to 4 by subzone**.

```{r}
pop_sz <- read_csv("../Hands-on_Ex01/data/aspatial/respop_census2020_agesex.csv",
                   show_col_types = FALSE) |>
  filter(!grepl("Total", Number)) |>
  transmute(SUBZONE_N = toupper(Number),
            kids = suppressWarnings(as.numeric(Total_0_4)))

mpsz_pop <- mpsz |>
  left_join(pop_sz, by = "SUBZONE_N") |>
  mutate(kids = replace_na(kids, 0),
         kids_km2 = kids / as.numeric(st_area(mpsz) / 1e6))

c(subzone_polygons = nrow(mpsz_pop),
  name_found_in_census = sum(mpsz_pop$SUBZONE_N %in% pop_sz$SUBZONE_N),
  with_resident_count = sum(mpsz_pop$kids > 0),
  children_0_4 = sum(mpsz_pop$kids))
```

Every one of the 332 subzone names matches the census table, but only 226 carry a resident
count. The other 106 are places like Raffles Place, Boat Quay, Jurong Island and the southern
islands, where the census reports no resident population because nobody lives there. Treating
those as zero is correct rather than a gap in the join, and the total of 183,100 children
agrees with the published national figure.

The next step turns this into a surface: rasterise the subzone densities, crop to the town, and
scale the values so the surface integrates to the number of centres actually observed. That
last step is what makes it a candidate intensity for the point process rather than just a
population map.

```{r}
kids_rast <- terra::rasterize(terra::vect(mpsz_pop),
                              terra::rast(terra::vect(mpsz_pop), resolution = 200),
                              field = "kids_km2")

pop_intensity <- function(town_sf, X) {
  rr  <- terra::crop(kids_rast, terra::vect(town_sf), mask = TRUE)
  mat <- terra::as.matrix(rr, wide = TRUE); mat[is.na(mat)] <- 0
  e   <- as.vector(terra::ext(rr))
  lam <- im(mat[nrow(mat):1, ], xrange = e[1:2] / 1000, yrange = e[3:4] / 1000)
  lam <- lam[Window(X), drop = FALSE]
  lam$v[is.na(lam$v) | lam$v <= 0] <- 1e-6
  lam$v <- lam$v * X$n / (sum(lam$v, na.rm = TRUE) * lam$xstep * lam$ystep)
  lam
}

lam_pop_tm <- pop_intensity(tm, childcare_tm_ppp_km)
lam_pop_ck <- pop_intensity(ck, childcare_ck_ppp_km)
```

```{r}
par(mfrow = c(1, 2))
plot(lam_pop_tm, main = "Children 0-4 intensity: Tampines")
plot(lam_pop_ck, main = "Children 0-4 intensity: Choa Chu Kang")
par(mfrow = c(1, 1))
```

```{r}
set.seed(42)
pcfpop_tm <- envelope(childcare_tm_ppp_km, pcfinhom, lambda = lam_pop_tm,
                    nsim = 99, verbose = FALSE)
pcfpop_ck <- envelope(childcare_ck_ppp_km, pcfinhom, lambda = lam_pop_ck,
                    nsim = 99, verbose = FALSE)
data.frame(
  town = c("Tampines", "Choa Chu Kang"),
  clustered = c(outside(pcfpop_tm, "above"), outside(pcfpop_ck, "above")),
  regular   = c(outside(pcfpop_tm, "below"), outside(pcfpop_ck, "below"))
)
```

The answer barely moves. Against a null built from where the children actually live, rather
than from a smoothed version of the centres themselves, Choa Chu Kang shows nothing outside
its envelope at any distance and Tampines shows a small excess over about 2% of the range.

The agreement is worth more than the size of the residual. The two nulls are built from
completely different information, one from the childcare points and one from the census, and
they point the same way.
The circularity I was worried about does not appear to be driving the earlier result, and the
conclusion now rests on an external covariate of the kind Floch et al. ask for rather than on
the pattern explaining itself.

The small Tampines residual is the one thing I would look at next. It is weak enough to be
noise across 99 simulations, but it is in the direction of clustering, and Tampines is the town
where the housing is densest.

#### What none of this can settle

Section 4.3.2 of Floch et al., one of the core readings, is a warning that applies to
everything above, and it is worth quoting because it states the problem more precisely than I
could:

> [T]he same aggregated or dispersed structures can be obtained with an inhomogeneous Poisson
> process in which the intensity of the process varies in space but the points are independent
> of each other. **A single observation of a spatial distribution does not allow for any
> distinction between first- and second-order properties of a process** in the absence of
> additional information such as that provided by a model that links a covariable to the
> intensity.

That is the identification problem this exercise ran into empirically. A clustered-looking map
can mean centres attract each other (second-order), or that centres are simply more numerous
where families live (first-order). One snapshot cannot separate them. Floch et al. cite Ellison
and co-authors on exactly this: natural advantages, which raise intensity, have an effect "that is
indistinguishable from that of positive externalities" causing aggregation. For childcare, "many
young families here" and "operators cluster near each other" produce the same picture.

It also names the two standard escapes, and they are the two methods that appear in this
exercise: the inhomogeneous K function of Baddeley et al., which I used through `Linhom` and
`pcfinhom`, and the Kd of Duranton and Overman, which I did not.

This forces an honest qualification of my own result. The escape only works if the intensity
comes from **outside** the pattern, from a covariate such as resident population. Mine did not.
I estimated it with a kernel density of the very points being tested, so any genuine
short-range clustering was partly absorbed into the "trend" before the test ran. That biases
the inhomogeneous result towards finding nothing, which is what it found.

My conclusion is therefore narrower than it first appears. The pattern is consistent with
centres being located wherever demand is, and this data cannot demonstrate interaction on top
of that. That is not the same as proving no interaction exists.

Doing better needs the covariate, which is the natural next step and the reason the limitation
below is first on the list.

## Limitations

Where this exercise falls short:

1. **The population null is coarse.** I did test against resident children aged 0 to 4 rather
   than only against CSR, which was the main gap. But the census gives one figure per subzone,
   so the intensity surface is a step function that jumps at subzone boundaries rather than a
   smooth demand surface. Any residual clustering measured at distances shorter than a subzone
   is partly a product of that blockiness.
2. **Edge effects at the coast.** All functions were used with their best available edge
   corrections, but the coastline makes the window irregular and some bias remains at short
   distances.
3. **Description, not explanation.** Even with the population covariate, this is still a
   description of where centres are. It says nothing about capacity, waiting lists or whether
   supply meets demand, and nothing about change over time. The spatio-temporal extension in
   Lesson 3 is where the time dimension starts.
4. **Attribute data is thin.** The current childcare table carries almost no useful attributes
   (capacity, hours, fees), so I could not stratify the analysis, e.g. cluster analysis for
   centres with and without infant-care places, which is the policy-relevant question.

None of these change the results above, but they do limit how far the results can be pushed.

## Summary

At the national scale childcare supply is heavily concentrated, and the concentration sits in
a central-eastern corridor that survives every bandwidth I tried. Clark-Evans, the quadrat
chi-square test and the G, K and L functions all agree on that.

The second-order results need more care than the chapters suggest. Both towns look strongly
clustered against a homogeneous CSR null, but that null assumes centres could have gone
anywhere with equal probability, which is not true of a service that follows housing. Testing
against an intensity-adjusted null instead, and using the non-cumulative pair correlation
function rather than K or L, almost all of it goes. Choa Chu Kang shows no departure from its
envelope at any distance, and Tampines only a small one, over about 2% of the range.

So the reading I would defend is that childcare centres are concentrated because people are
concentrated, and that this data shows little sign of the centres responding to each other on
top of that. That conclusion holds under both a pointwise and a global envelope, which is more
than the J function manages.

Getting there took three corrections to my own first answer: one for the null, one for the
cumulative functions, and one for reading pointwise bands as though they carried a significance
level. None of the three would have surfaced from the chapters alone.

# Reflection

Lesson 1 was mostly about getting data into the right shape. This exercise was less about
mechanics and more about what the tests were comparing against.

The null hypothesis took up most of my time. Every test in Part 2 ran without complaint and
reported strong clustering, and none of them were wrong. But CSR treats a reservoir and a
housing estate as equally likely places for a childcare centre, so a service that follows
families gets reported as clustered more or less by construction. When I rebuilt the null from
census counts of children aged 0 to 4, most of the clustering went away. The code was fine
throughout. I had not thought about what the comparison assumed.

The mistake I spent longest on came from using the wrong file. The chapters read
`ChildCareServices.kml`, which is not in the book's GitHub repository, so I substituted a
different geocoded childcare layer that is, `CHILDCARE.shp` from a later chapter. Both look
official and both are childcare centres in Singapore. But the substitute had 174 coordinates
shared by more than one centre, including 25 at a single point, and that made the
adaptive-bandwidth map in Part 1 look nothing like the book's. I assumed at first that the
chapters had glossed over something. The real explanation was that I was not using their data.
Once I tracked down the actual KML, Choa Chu Kang returned R = 0.84097 and p = 0.008866,
matching the chapter exactly.

Two smaller errors followed the same pattern. I read the inhomogeneous L curve as evidence that
centres space themselves out, when it was an artefact of L accumulating small deviations across
distances; the pair correlation function showed nothing at any distance. And before finding the
right file, I assumed a difference from the chapter came from the newer planning boundaries and
wrote that up, when both boundary editions return the same result. In all three cases I had an
explanation that sounded reasonable and did not test it.

The terminology caught me out once. The pair correlation function is written g(r), and the
lesson slides use G(r) for the nearest-neighbour function. They are different functions, and I
had to go back to Gimond to be sure which one I was reading about.

I only checked the Monte Carlo simulation envelopes properly after a classmate asked on Piazza
whether they were pointwise or global. Until then I had been quoting the share of the distance
range where a curve sits outside its band as though that were a result, and it is not. Those
bands are pointwise, so with hundreds of distances on the axis a curve straying outside
somewhere is close to expected. The `glocal = TRUE` in the chapters does not guard against
this, because no argument by that name exists. Rerunning everything with global envelopes left
the main conclusion standing but removed the J function result completely.

If I did this again I would confirm the exact data file before writing anything, and check for
coincident points and settle the study window as a first step rather than partway through. The
file I read mattered more than any method I picked, which is not where I expected the
sensitivity to be.

# References

Grouped as the Lesson 2 page groups them, with a note on where each one is used here. Sources
listed on that page but not drawn on are left out rather than padded in.

## Hands-on chapters

- r4gdsa [Chapter 4: First-order Spatial Point Patterns Analysis
  Methods](https://r4gdsa.netlify.app/chap04.html) and [Chapter 5: 2nd Order Spatial Point
  Patterns Analysis Methods](https://r4gdsa.netlify.app/chap05.html). Reproduced in Parts 1
  and 2, and reviewed in Part 3.

## Core readings

- Yuan, Y., Qiang, Y., Bin Asad, K., and Chow, T. E. (2020). *Point Pattern Analysis*. The
  Geographic Information Science & Technology Body of Knowledge, 1st Quarter 2020 Edition,
  John P. Wilson (ed.). DOI: 10.22224/gistbok/2020.1.13.
  Source of the warning that CSR is a weak null for events that track a covariate, which drives
  the section on choosing a null hypothesis.
- Yin, P. (2020). *Kernels and Density Estimation*. The Geographic Information Science &
  Technology Body of Knowledge, 1st Quarter 2020 Edition, John P. Wilson (ed.). DOI:
  10.22224/gistbok/2020.1.12.
  The bias-variance framing behind the bandwidth comparison, which I measure rather than
  restate.
- Chapter 4 *Spatial distribution of points* of Floch, J.M., Marcon, E. and Puech, F. (2018)
  *Handbook of Spatial Analysis: Theory and Application with R*.
  Section 4.3.2 states the identification problem between first- and second-order effects that
  limits the whole exercise; section 4.5.1 sets out the Kd function I describe but do not run.

## References

- Baddeley, A., Rubak, E. and Turner, R. (2015) *Spatial Point Patterns: Methodology and
  Applications with R*, Chapman and Hall/CRC.
  Section 3.4.4 is why the exercise checks for coincident points before running anything.
- O'Sullivan, D. and Unwin, D. (2010) *Geographic Information Analysis*, Second Edition. John
  Wiley & Sons, New Jersey.
  Definitions of the modifiable areal unit problem, used in the quadrat section, and of edge
  effects, used when the study window is buffered.
- *Chapter 11 Point Pattern Analysis* of **Intro to GIS and Spatial Analysis**, sections 11.2,
  11.3, 11.3.1 and 11.4.
  Section 11.3.3 of the same chapter introduces the pair correlation function as a
  non-cumulative alternative to K, which is what corrects the reading of the inhomogeneous L
  curve. Rendered version consulted at
  <https://rstudio-pubs-static.s3.amazonaws.com/481933_61ed3eff92e74c04a52d44a48b1b4548.html>.

## Applications

- Donthu, N. and Rust, R. T. (1989) "Estimating Geographic Customer Densities Using Kernel
  Density Estimation", *Marketing Science*, 8(2), pp. 191-203.
  Their normal-reference bandwidth rule is `bw.scott`, tested here against the chapters'
  default.
- Wartman, J. and Malasavage, N. E. (2010) "Spatial Analysis for Identifying Concentrations of
  Urban Damage", in *Methods and Techniques in Urban Engineering*, InTech.
  The argument that the study area must be set objectively, which the window test supports.
- Borruso, G. and Porceddu, A. (2009) "A Tale of Two Cities: Density Analysis of CBD on Two
  Midsize Urban Areas in Northeastern Italy", in *Studies in Computational Intelligence,
  Geocomputation and Urban Planning*, pp. 37-56.
  The claim that bandwidth is the only arbitrary variable, discussed alongside Wartman.

## Enrichment

- Prof. Luc Anselin on point pattern analysis (YouTube). Of the six parts, this exercise uses
  *Point Pattern Analysis: Quadrat Counts* for the quadrat method and the variance-mean ratio,
  and *Point Pattern Analysis: F and J Functions* for the J function. *Point Pattern Analysis:
  K, L and Kd Functions* covers the Kd indicator discussed in Part 3.

## Cited within the above

- Clark, P. J. and Evans, F. C. (1954), for the nearest-neighbour index used throughout Part 1.
- Duranton, G. and Overman, H. G. (2005), for the Kd function, via Floch et al. and Anselin.

---

*Data: childcare centres, Early Childhood Development Agency via Data.gov.sg (geocoded
edition); planning subzones, URA 2019 Master Plan, Data.gov.sg; resident population, Singapore
Census 2020 via Data.gov.sg. Analysis in R with `spatstat`, `sf`, `terra` and `tmap`.*
