---
title: "Hands-on Exercise 5"
subtitle: "Global and Local Measures of Spatial Autocorrelation"
author: "An Loc"
date: "21 September 2026"
date-modified: "last-modified"
date-format: "D MMM YYYY"
format:
  html:
    toc: true
    toc-depth: 4
    toc-expand: true
    code-fold: false
    code-copy: true
    fig-width: 9
    fig-height: 7
execute:
  echo: true
  eval: true
  warning: false
  message: false
  freeze: true
---

# Overview

Exercise 4 built a spatial weights matrix and used it to compute a spatially lagged GDP per capita
(GDPPC) for each of Hunan's 88 counties. At the end of that exercise, the scatterplot of GDPPC
against its spatial lag had a positive slope. This suggested that counties with high GDPPC are
adjacent to other counties with high GDPPC. No test was made of whether that slope is larger than
spatial randomness would produce.

This exercise tests for spatial autocorrelation in two ways. A **global** measure is a single
statistic for the whole province, tested against the null hypothesis of spatial randomness. A
**local** measure is computed for each county, and identifies the counties that belong to a cluster
and the counties that differ from their neighbours.

**Part 1** reproduces r4gdsa chapter 9, Global Measures of Spatial Autocorrelation: Moran's I,
Geary's C, their Monte Carlo versions and the spatial correlogram. **Part 2** reproduces chapter 10,
Local Measures of Spatial Autocorrelation: local Moran's I with the LISA cluster map, and the
Getis-Ord Gi\* hot spot map. **Part 3** examines the method of the two chapters and adds the parts
of the lecture that neither chapter covers.

::: {.callout-note appearance="simple"}

## Sources

The hands-on exercise is two chapters, both assigned for Lesson 5:

- [Chapter 9: Global Measures of Spatial Autocorrelation](https://r4gdsa.netlify.app/chap09.html)
- [Chapter 10: Local Measures of Spatial Autocorrelation](https://r4gdsa.netlify.app/chap10.html)

The Lesson 5 outline sets five papers as self-reading. Each is used in a specific section.

| Reading | What it contributes | Used in |
|---|---|---|
| **Moran (1950)**, "Notes on Continuous Stochastic Phenomena" | The statistic, derived for a regular lattice; the expectation −1/(N−1); and the statement that the variance needs a normality assumption while the expectation does not | 3.3, 5.5 |
| **Geary (1954)**, "The Contiguity Ratio and Statistical Mapping" | The squared-difference form, and the clearest statement of what each of the two null models assumes | 3.4, 5.5, 5.6 |
| **Getis and Ord (1992)**, "The Analysis of Spatial Association by Use of Distance Statistics" | The G statistic in both forms, the binary symmetric weights it is defined on, the requirement of a natural origin, and the demonstration that Moran's I cannot tell high-value clustering from low-value clustering | 4.2, 5.4 |
| **Anselin (1995)**, "Local Indicators of Spatial Association - LISA" | The two conditions a statistic must satisfy to be a LISA; a Bonferroni-corrected worked example run on 9,999 permutations; and the claim that permutation inference is identical for the local Moran and both of Gi and Gi\* | 4.1, 5.3, 5.8 |
| **Ord and Getis (1995)**, "Local Spatial Autocorrelation Statistics: Distributional Issues and an Application" | Simulation evidence for when a normal approximation to a local statistic fails, the extension of Gi to nonbinary weights, and the Bonferroni recommendation | 5.2, 5.3 |

The outline also lists eight references. Three of them are used on this page:

| Reference | What it contributes | Used in |
|---|---|---|
| **Anselin (1996)**, "The Moran Scatterplot as an ESDA Tool" | Moran's I as the slope of the regression of the spatial lag on the variable, and the four-quadrant reading of the scatterplot | 4.1.4 |
| **Getis (2010)**, "Spatial Autocorrelation", chapter B.3 of the *Handbook of Applied Spatial Analysis* | The problem of dependent multiple tests in local statistics, and the false discovery rate as one response to it | 5.3.1 |
| **Mack and Kam (2018)**, "Is There Space for Violence?" | A tabulation of neighbour counts across weights definitions before one is chosen | 5.7.1 |

The other five references on the outline are not used on this page.

:::

## Differences from the chapters

| | The chapter | This page | Why | Where |
|---|---|---|---|---|
| **Projected CRS** | chapter 10 uses EPSG:32650, chapter 9 does not project at all | EPSG:32649 throughout | Hunan spans 108.78°E to 114.25°E, so it lies in UTM zone 49N. Zone 50N starts at 114°E. The wrong zone narrows the fixed bandwidth from 62,955 m to 60,800 m, a 3.4% change on the correct figure | 5.1 |
| **Local p-values** | `nsim = 99` is passed, then `p_ii` is used | `p_ii_sim`, with `nsim = 999` | Both come from the same permutations, but `p_ii` is a normal approximation to a distribution the same call reports as skewed. The choice changes the count of significant counties from 13 to 16 and changes which ones | 5.2 |
| **Number of simulations** | chapter 10 uses 99 for the local statistics | 999 | A rank-based p-value cannot fall below `2 / (nsim + 1)`, so at 99 its minimum is 0.02, above the Bonferroni bound for 88 counties. Chapter 9 already uses 999 | 5.2, 5.3 |
| **Multiple comparisons** | not mentioned | measured, with the positions of both assigned papers | Ord and Getis (1995) recommend Bonferroni and Anselin (1995) applies it. 88 tests at α = 0.05 expect about four false positives, and at `nsim = 999` no county remains significant under either correction | 5.3 |
| **Natural origin of the G variable** | standardises `GDPPC` for the scatterplot and uses the raw values for Gi\*, without saying why | tested by shifting GDPPC | Getis and Ord (1992) state that G is scale-invariant but not location-invariant. Adding 100,000 to GDPPC changes the p-value of global G from 0.0012 to 0.19 and leaves Moran's I unchanged | 5.4.1 |
| **Gi against the local Moran** | presented as separate analyses | the permutation inference compared directly | Anselin (1995) claims it is identical for the local Moran, Gi and Gi\*. On these weights it holds for all three, up to simulation error, so chapter 10's two maps differ in their weights and in how they label counties | 5.8 |
| **Standardised Moran scatterplot** | `scale(lag_GDPPC)` on the vertical axis | the lag of the standardised variable, as in Anselin (1996) | Standardising the lag separately makes the slope of the fitted line 0.485, the Pearson correlation of GDPPC with its lag, where Moran's I is 0.301 | 4.1.4 |
| **Choice of `k` for Gi\*** | `k = 6`, not examined | `k` = 4, 6 and 8 compared | Significant counties rise from 17 to 25 with `k`; twelve keep the same classification under every value | 5.7.1 |
| **Getis-Ord global G** | not covered | added, with `globalG.test()` | Slides 16 and 17 cover it and neither chapter reaches it | 5.4 |
| **Local Geary's c** | not covered | added, with `local_c_perm()` | Slide 26 names it as the other LISA. It finds 21 significant counties where local Moran's I finds 16 | 5.6 |
| **Adaptive distance weights** | computed as `hunan_adw`, never used | used, and compared against the fixed band | The chapter builds the object and then runs Gi\* on the fixed-band weights only. The two disagree on 15 of the 88 counties | 5.7 |
| **Moran's I null model** | `randomisation = TRUE` default, not discussed | the three null models explained from the assigned readings | The default z-score and the permutation test in Part 1 rest on different assumptions, and Moran's own variance assumes normality | 5.5 |

::: {.callout-note collapse="true"}
## Mapping of chapters 9 and 10 to this page

Every substantive section of both chapters is covered. Chapter 10 repeats chapter 9's data import
in its sections 10.2 and 10.3, so those rows are not duplicated below.

| Chapter | This page |
|---|---|
| 9.1 Overview | 1 Overview |
| 9.2 Getting started | 2.1 Packages |
| 9.3.1 Import shapefile | 2.2.1 County boundaries |
| 9.3.2 Import csv | 2.2.2 Development indicators |
| 9.3.3 Performing relational join | 2.3 Joining the two |
| 9.3.4 Visualising regional development indicator | 3.1 Mapping GDPPC |
| 9.4.1 Computing contiguity spatial weights | 3.2 Contiguity weights |
| 9.4.2 Row-standardised weights matrix | 3.2 Contiguity weights |
| 9.5.1 Moran's I test | 3.3.1 The test |
| 9.5.2 Computing Monte Carlo Moran's I | 3.3.2 The permutation test |
| 9.5.3 Visualising Monte Carlo Moran's I | 3.3.3 The simulated distribution |
| 9.6.1 Geary's C test | 3.4.1 The test |
| 9.6.2 Computing Monte Carlo Geary's C | 3.4.2 The permutation test |
| 9.6.3 Visualising the Monte Carlo Geary's C | 3.4.3 The simulated distribution |
| 9.7.1 Compute Moran's I correlogram | 3.5 Spatial correlogram |
| 9.7.2 Compute Geary's C correlogram | 3.5 Spatial correlogram |
| 10.4.1 Computing contiguity spatial weights | 4.1.1 Weights the sfdep way |
| 10.4.2 Row-standardised weights matrix | 4.1.1 Weights the sfdep way |
| 10.4.3 Computing local Moran's I | 4.1.2 Local Moran's I |
| 10.4.3.1 to 10.4.3.4 Mapping | 4.1.3 The two maps |
| 10.4.4.1 Plotting Moran scatterplot | 4.1.4 Moran scatterplot |
| 10.4.4.2 Standardised variable | 4.1.4 Moran scatterplot |
| 10.4.4.3 Preparing LISA map classes | 4.1.5 LISA cluster map |
| 10.4.4.4 Plotting and visualising LISA map | 4.1.5 LISA cluster map |
| 10.5.1 Deriving distance weight matrix | 4.2.1 Distance weights |
| 10.5.2 Computing local Gi\* | 4.2.2 Local Gi\* |
| 10.5.3 Preparing and visualising HCSA map | 4.2.3 Hot and cold spot map |
:::

# Getting started

## Packages

`spdep` provides the global statistics and the correlogram. `sfdep` provides the local statistics,
returning them as columns of a data frame. The two give identical local Moran values. `sfdep` avoids
the matrix that `spdep` returns, which has to be bound to the sf object by row position.

```{r}
pacman::p_load(sf, spdep, sfdep, tmap, tidyverse, knitr)
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses one function:

```r
pacman::p_load(..., install = TRUE)                       # pacman
```

- `p_load()` loads each package named in `...`, and first installs any that are missing, because
  `install = TRUE` is the default. It replaces a separate `install.packages()` and `library()` call
  for each package.
- The `pacman::` prefix calls the function from `pacman` without attaching the package itself.
:::

## The data

Both chapters use the same two files as Chapter 8, described in Exercise 4.

| Layer | Format | Features | Native CRS | Contents |
|---|---|---|---|---|
| `Hunan` | ESRI shapefile | 88 polygons | WGS 84, EPSG:4326 | County boundaries of Hunan province |
| `Hunan_2012` | CSV | 88 rows, 29 fields | not spatial | 2012 development indicators by county |

### County boundaries

The shapefile is in geographic coordinates (degrees). Contiguity is defined by shared boundaries and
is unaffected by the projection, but every distance-based weight depends on the projection. Section
5.1 identifies the UTM zone for Hunan as zone 49N (EPSG:32649), and the layer is projected to that
zone here so that one CRS is used throughout the page.

```{r}
hunan <- st_read(dsn = "data/geospatial", layer = "Hunan", quiet = TRUE) |>
  st_transform(crs = 32649)
st_crs(hunan)$input
nrow(hunan)
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses three functions and the pipe:

```r
st_read(dsn, layer, quiet = FALSE)                        # sf
st_transform(x, crs)                                      # sf
st_crs(x)                                                 # sf
```

- `st_read()` reads a vector data file into an sf data frame: one row per feature, with attribute
  columns and a geometry column. For a shapefile, `dsn` is the folder and `layer` is the file name
  without its extension. `quiet = TRUE` suppresses the summary it otherwise prints, which begins
  with the full path of the data folder on the computer that renders the page.
- `|>` is R's native pipe. It passes the result on its left as the first argument of the function on
  its right, so `st_transform()` receives the layer from `st_read()` as its `x`.
- `st_transform()` recalculates every coordinate in the CRS given in `crs`, here EPSG:32649, UTM
  zone 49N, in metres. `st_set_crs()` would change only the label, and leave the coordinates in
  degrees.
- `st_crs()` returns the CRS of an object, and `$input` extracts the code it was set from. `nrow()`
  counts the counties.

The figure shows what `st_transform()` does to the layer.

```{r}
#| echo: false
#| fig-height: 4
ortho <- "+proj=ortho +lat_0=27.5 +lon_0=111.5 +datum=WGS84"
s2_was <- sf_use_s2(FALSE)
globe_land <- spData::world |> st_crop(xmin = 50, xmax = 170, ymin = -30, ymax = 80)
globe_grid <- st_graticule(st_bbox(c(xmin = 50, ymin = -30, xmax = 170, ymax = 80), crs = 4326),
                           lon = seq(60, 170, 20), lat = seq(-20, 80, 20))
sf_use_s2(s2_was)
globe_disc <- st_sfc(st_buffer(st_point(c(0, 0)), 6371000), crs = ortho)
hunan_deg <- st_transform(hunan, 4326)

before <- ggplot() +
  geom_sf(data = globe_disc, fill = "aliceblue", colour = "steelblue") +
  geom_sf(data = globe_grid, colour = "grey75", linewidth = 0.2) +
  geom_sf(data = globe_land, fill = "grey88", colour = "grey60", linewidth = 0.2) +
  geom_sf(data = hunan_deg, fill = "darkorange", colour = NA) +
  coord_sf(crs = ortho) +
  labs(title = "Before: degrees on the globe", subtitle = "EPSG:4326, latitude and longitude") +
  theme_void()

step <- ggplot() +
  annotate("segment", x = 0, xend = 1, y = 0, yend = 0,
           arrow = arrow(length = unit(0.3, "cm")), linewidth = 0.8) +
  annotate("text", x = 0.5, y = 0.12, label = "st_transform(crs = 32649)",
           family = "mono", size = 2.8) +
  coord_cartesian(xlim = c(-0.1, 1.1), ylim = c(-0.5, 0.5), clip = "off") +
  theme_void()

after <- ggplot() +
  geom_sf(data = hunan, fill = "darkorange", colour = "white", linewidth = 0.1) +
  coord_sf(datum = 32649) +
  scale_x_continuous(labels = scales::comma) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "After: metres on a flat grid", subtitle = "EPSG:32649, UTM zone 49N",
       x = "easting (m)", y = "northing (m)") +
  theme_minimal() +
  theme(axis.text = element_text(size = 7))

patchwork::wrap_plots(before, step, after, widths = c(1, 0.7, 1))
```
:::

### Development indicators

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv", show_col_types = FALSE)
dim(hunan2012)
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses one function:

```r
read_csv(file, show_col_types = should_show_types())      # readr
```

- `read_csv()` reads a comma-separated file into a tibble and guesses the type of each column from
  its values. `show_col_types = FALSE` suppresses the message that lists those guesses.
- `dim()` returns the number of rows and columns.

The two files are read by different functions, which return different kinds of object:

| | `read_csv()` | `st_read()` |
|---|---|---|
| Package | `readr`, part of the tidyverse | `sf` |
| Reads | a table of values, here a CSV file | a spatial file, here a shapefile |
| Returns | a tibble (`tbl_df`), a kind of data frame | an sf data frame: a data frame with a geometry column |
| Geometry and CRS | none | one polygon per row, with its CRS |
| On this page | `Hunan_2012`: 88 rows, 29 columns of indicators | `Hunan`: 88 county polygons with 7 attribute fields |
| Role on this page | supplies GDPPC, the variable every statistic tests | supplies the county boundaries, from which the maps are drawn and the neighbours found |

`left_join()` in the next section attaches the CSV's columns to the county polygons, so that GDPPC
and the geometry are in one sf object.
:::

## Joining the two

```{r}
hunan <- left_join(hunan, hunan2012, by = "County") |>
  select(1:4, 7, 15)
names(hunan)
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses two functions:

```r
left_join(x, y, by = NULL)                                # dplyr
select(.data, ...)                                        # dplyr
```

- `left_join()` keeps every row of `x`, here `hunan` with its geometry, and adds the columns of `y`
  from the matching row. `by` names the column to match on. With the default, `NULL`, it matches on
  every column name the two tables share and prints a message naming them. Chapter 9 relies on that
  default. This chunk names `by = "County"`, as in Exercise 4, which fixes the match to one column.
- `select()` keeps the columns listed in `...`, given as names or as positions. `1:4, 7, 15` are
  positions in the joined table. On an sf object the geometry column is kept without being listed.

The joined table has 36 columns. Columns 8 to 35 come from the CSV. The columns that `select()`
keeps are shaded:

<table class="table table-sm">
<tr><th style="background: rgba(255, 193, 7, 0.30)">1</th><th style="background: rgba(255, 193, 7, 0.30)">2</th><th style="background: rgba(255, 193, 7, 0.30)">3</th><th style="background: rgba(255, 193, 7, 0.30)">4</th><th>5</th><th>6</th><th style="background: rgba(255, 193, 7, 0.30)">7</th><th>8</th><th>…</th><th style="background: rgba(255, 193, 7, 0.30)">15</th><th>…</th><th style="background: rgba(255, 193, 7, 0.30)">36</th></tr>
<tr><td style="background: rgba(255, 193, 7, 0.30)">NAME_2</td><td style="background: rgba(255, 193, 7, 0.30)">ID_3</td><td style="background: rgba(255, 193, 7, 0.30)">NAME_3</td><td style="background: rgba(255, 193, 7, 0.30)">ENGTYPE_3</td><td>Shape_Leng</td><td>Shape_Area</td><td style="background: rgba(255, 193, 7, 0.30)">County</td><td>City</td><td>…</td><td style="background: rgba(255, 193, 7, 0.30)">GDPPC</td><td>…</td><td style="background: rgba(255, 193, 7, 0.30)">geometry</td></tr>
</table>

Selection by position depends on this column order. The printed names confirm that the six
attributes kept are the ones intended.

The same columns can be selected by name, which does not depend on the column order and states what
is kept:

```r
hunan <- left_join(hunan, hunan2012, by = "County") |>
  select(NAME_2, ID_3, NAME_3, ENGTYPE_3, County, GDPPC)
```

Both versions return the same object. The page keeps the chapter's positions so that its code
matches chapter 9.
:::

# Part 1: Global measures of spatial autocorrelation

This part follows r4gdsa [Chapter 9](https://r4gdsa.netlify.app/chap09.html). Each statistic in it
is a single value for the whole province, tested against the null hypothesis of spatial randomness.

## Mapping GDPPC

```{r}
#| fig-height: 4.1
equal <- tm_shape(hunan) +
  tm_polygons(fill = "GDPPC",
              fill.scale = tm_scale_intervals(style = "equal", n = 5,
                                              values = "brewer.blues"),
              fill.legend = tm_legend(title = "GDPPC",
                                      position = tm_pos_out("right", "center"))) +
  tm_borders(col_alpha = 0.5) +
  tm_title("Equal interval classification")

quantile <- tm_shape(hunan) +
  tm_polygons(fill = "GDPPC",
              fill.scale = tm_scale_intervals(style = "quantile", n = 5,
                                              values = "brewer.blues"),
              fill.legend = tm_legend(title = "GDPPC",
                                      position = tm_pos_out("right", "center"))) +
  tm_borders(col_alpha = 0.5) +
  tm_title("Quantile classification")

tmap_arrange(equal, quantile, asp = 1, ncol = 2)
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses eight functions:

```r
tm_shape(shp)                                             # tmap
tm_polygons(fill, fill.scale, fill.legend)                # tmap
tm_scale_intervals(n = 5, style = "pretty", values = NA)  # tmap
tm_legend(title, position)                                # tmap
tm_pos_out(cell.h, cell.v)                                # tmap
tm_borders(col, col_alpha)                                # tmap
tm_title(text)                                            # tmap
tmap_arrange(..., ncol = NA, asp = 0)                     # tmap
```

- `tm_shape()` names the sf object that the layers after it draw. Each `+` adds a layer or a
  setting.
- `tm_polygons()` draws the polygons. Each visual property in tmap v4 has three arguments: `fill`
  names the variable that sets the colour, `fill.scale` sets how its values map to colours, and
  `fill.legend` sets the legend.
- `tm_scale_intervals()` divides a numeric variable into `n` classes by the method in `style`, and
  colours them from the palette in `values`, here the ColorBrewer Blues palette. The default style
  is `"pretty"`. Only `style` differs between the two maps.
- `tm_legend()` sets the legend title and position. `tm_pos_out("right", "center")` places the
  legend outside the map frame, to the right and centred vertically, so that it covers no county.
- `tm_borders()` adds a layer that draws only the county boundaries, here at half opacity
  (`col_alpha = 0.5`). `tm_polygons()` already draws the boundaries, in dark grey at full opacity,
  so on these maps the extra layer changes little. Chapter 9 includes the same layer.
- `tm_title()` adds the title above the map.
- `tmap_arrange()` places stored maps side by side in `ncol` columns. `asp = 1` gives each map an
  aspect ratio of 1, where the default, `0`, fits each map to its cell.
:::

In the chapters, the two maps illustrate the effect of the classification method on the apparent
pattern. With a right-skewed variable such as GDPPC, equal-interval breaks place most of the
province in the lowest class and Changsha alone in the highest. Quantile breaks assign an equal
number of counties to each class and show higher GDPPC in the north east than in the west.

Chapter 10 asks two questions of these maps:

- **Do they reveal outliers or clusters?** Changsha stands out as an outlier, the only county in the
  highest equal-interval class. The quantile map suggests a cluster of higher GDPPC in the north
  east and of lower GDPPC in the west.
- **Do they indicate hot spots or cold spots?** They suggest where these may lie, around Changsha
  and in the west, but a choropleth cannot show whether the pattern is significant. Section 3.3
  tests it for the province, and sections 4.1 and 4.2 for each county.

```{r}
summary(hunan$GDPPC)
```

## Contiguity weights

Chapter 9 builds the neighbour list with Queen contiguity and divides each county's weights by its
neighbour count. Exercise 4 built the same matrix.

```{r}
wm_q <- poly2nb(hunan, queen = TRUE)
summary(wm_q)
```

```{r}
rswm_q <- nb2listw(wm_q, style = "W", zero.policy = TRUE)
rswm_q
```

::: {.callout-note collapse="false" title="How the code works"}
The two blocks above use two functions:

```r
poly2nb(pl, queen = TRUE)                                 # spdep
nb2listw(neighbours, style = "W", zero.policy = NULL)     # spdep
```

- `poly2nb()` builds a neighbour list from polygons: for each county, the row numbers of the
  counties whose boundaries touch it. With `queen = TRUE`, the default, one shared boundary point is
  enough, and `queen = FALSE` requires a shared edge. The call states the default, as chapter 9
  does.
- `nb2listw()` turns the neighbour list into a weights object, which the spdep tests take as their
  `listw` argument. `style = "W"`, the default, row-standardises the weights.
- `zero.policy = TRUE` allows a county with no neighbours. With the default, `NULL`, spdep uses its
  global option, under which a county with no neighbours stops the function with an error.

The figure follows one county, Anxiang in row 1, through the two functions. `poly2nb()` records
the row numbers of the five counties that touch it, and `nb2listw()` gives each of them a weight of
1/5.

```{r}
#| echo: false
#| fig-height: 4.2
nb_ids <- c(1, wm_q[[1]])
nb_focus <- hunan |>
  mutate(row = row_number(),
         role = case_when(row == 1 ~ "Anxiang (row 1)",
                          row %in% wm_q[[1]] ~ "its neighbours",
                          TRUE ~ "other"),
         role = factor(role, levels = c("Anxiang (row 1)", "its neighbours", "other")))
nb_box <- st_bbox(st_buffer(st_union(nb_focus[nb_ids, ]), 5000))
nb_labels <- nb_focus[nb_ids, ] |>
  mutate(txt = paste0(row, "\n", NAME_3)) |>
  st_point_on_surface()

nb_map <- ggplot() +
  geom_sf(data = nb_focus, aes(fill = role), colour = "white", linewidth = 0.3) +
  geom_sf_text(data = nb_labels, aes(label = txt), size = 2.8, lineheight = 0.9) +
  scale_fill_manual(values = c("Anxiang (row 1)" = "darkorange",
                               "its neighbours" = "#fdd49e", "other" = "grey90")) +
  coord_sf(xlim = nb_box[c("xmin", "xmax")], ylim = nb_box[c("ymin", "ymax")], datum = NA) +
  labs(fill = NULL, x = NULL, y = NULL) +
  theme_void() +
  theme(legend.position = "bottom")

nb_steps <- ggplot() +
  annotate("text", x = 0, y = 0.95, label = "poly2nb()", family = "mono", size = 3.4, hjust = 0) +
  annotate("label", x = 0, y = 0.78, hjust = 0, family = "mono", size = 3.4,
           label = paste0("wm_q[[1]]\n", paste(wm_q[[1]], collapse = "    "))) +
  annotate("segment", x = 0.25, xend = 0.25, y = 0.64, yend = 0.46,
           arrow = arrow(length = unit(0.25, "cm"))) +
  annotate("text", x = 0.3, y = 0.55, label = 'nb2listw(style = "W")',
           family = "mono", size = 3.4, hjust = 0) +
  annotate("label", x = 0, y = 0.3, hjust = 0, family = "mono", size = 3.4,
           label = paste0("rswm_q$weights[[1]]\n",
                          paste(rswm_q$weights[[1]], collapse = "  "))) +
  coord_cartesian(xlim = c(0, 1), ylim = c(0.1, 1), clip = "off") +
  theme_void()

patchwork::wrap_plots(nb_map, nb_steps, widths = c(1.2, 1))
```
:::

Row standardisation makes each county's weights sum to 1. A lagged value is thus a weighted average
of the neighbouring values, on the same scale as the variable itself. Every global statistic below
is computed on this weights object.

## Moran's I

### The test

```{r}
moran.test(hunan$GDPPC,
           listw = rswm_q,
           zero.policy = TRUE,
           na.action = na.omit)
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses one function:

```r
moran.test(x, listw, randomisation = TRUE, zero.policy = attr(listw, "zero.policy"),
           alternative = "greater", na.action = na.fail)                  # spdep
```

- `moran.test()` computes Moran's I for `x` with the weights in `listw`, and tests it against its
  expectation under spatial randomness. It reports I, its expectation and variance, and the
  z-score, labelled "standard deviate", with its p-value.
- `randomisation = TRUE`, the default, sets the null model for the variance of I. Section 5.5
  explains the choice.
- `alternative` sets the direction of the test. The three options are:

  | `alternative` | Tests whether I is | Pattern it looks for |
  |---|---|---|
  | `"greater"`, the default | above its expectation | clustering: neighbouring values tend to be similar |
  | `"less"` | below its expectation | dispersion: neighbouring values tend to be dissimilar |
  | `"two.sided"` | different from its expectation in either direction | either pattern |

  This page keeps the default, because the question is whether GDPPC clusters. `moran.mc()` in
  section 3.3.2 has the same default. The local statistics of Part 2 default to `"two.sided"`.
- `zero.policy = TRUE` permits counties with no neighbours. Hunan has no such county, so the
  argument has no effect here.
- `na.action = na.omit` drops counties with missing values, where the default, `na.fail`, stops
  with an error. GDPPC has no missing values, so this argument also has no effect.

The figure places the z-score of this test on the standard normal curve. The shaded bands mark how
far a z-score lies from zero, and the red line marks the result for GDPPC.

```{r}
#| echo: false
#| fig-height: 3.6
z_obs <- unname(moran.test(hunan$GDPPC, listw = rswm_q, zero.policy = TRUE)$statistic)
cuts <- c(-Inf, -2.58, -1.96, -1.65, 1.65, 1.96, 2.58, Inf)
tibble(z = seq(-5.5, 5.5, 0.01), d = dnorm(z)) |>
  mutate(band = cut(z, cuts, labels = c("z < -2.58", "-2.58 to -1.96", "-1.96 to -1.65",
                                        "-1.65 to 1.65", "1.65 to 1.96", "1.96 to 2.58",
                                        "z > 2.58"))) |>
  ggplot(aes(z, d)) +
  geom_area(aes(fill = band, group = band)) +
  geom_line() +
  geom_vline(xintercept = z_obs, colour = "darkred", linewidth = 0.8) +
  annotate("label", x = z_obs, y = 0.2, hjust = 1.05, size = 3.2,
           label = paste0("GDPPC\nz = ", round(z_obs, 2))) +
  annotate("text", x = c(-3.8, 0, 3.8), y = 0.43, size = 3.5,
           label = c("Dispersed", "Random", "Clustered")) +
  scale_fill_manual(values = c("#2166ac", "#67a9cf", "#d1e5f0", "#fff7bc",
                               "#fddbc7", "#ef8a62", "#b2182b")) +
  scale_x_continuous(breaks = c(-2.58, -1.96, 0, 1.96, 2.58)) +
  labs(x = "z-score", y = NULL, fill = "z-score band") +
  theme_minimal() +
  theme(axis.text.y = element_blank(), panel.grid.minor = element_blank())
```
:::

Chapter 9 asks what statistical conclusion the output supports:

- **Moran's I:** 0.301, against an expectation of −0.0115 under spatial randomness. The
  expectation, −1/(n−1), is slightly negative, because a county is not its own neighbour.
- **Test:** z = 4.74 and p = 1.1 × 10⁻⁶, so the null hypothesis of spatial randomness is rejected.

Slide 10 of the Lesson 5 lecture gives the reading of I, shown below on a scale from negative to
positive with the value for GDPPC marked.

```{r}
#| code-fold: true
#| fig-height: 2.4
i_obs <- round(unname(moran.test(hunan$GDPPC, listw = rswm_q)$estimate[1]), 3)
ggplot() +
  annotate("rect", xmin = -1, xmax = -0.1, ymin = -0.25, ymax = 0.25, fill = "#67a9cf", alpha = 0.5) +
  annotate("rect", xmin = -0.1, xmax = 0.1, ymin = -0.25, ymax = 0.25, fill = "#fff7bc") +
  annotate("rect", xmin = 0.1, xmax = 1, ymin = -0.25, ymax = 0.25, fill = "#ef8a62", alpha = 0.5) +
  annotate("segment", x = -1.05, xend = 1.08, y = 0, yend = 0, linewidth = 0.8,
           arrow = arrow(ends = "both", length = unit(0.3, "cm"))) +
  annotate("text", x = c(-0.55, 0, 0.55), y = 0.55, fontface = "bold", size = 4,
           label = c("Negative: dispersed", "About zero: random", "Positive: clustered")) +
  annotate("text", x = c(-0.55, 0, 0.55), y = -0.5, size = 3.2,
           label = c("neighbours tend to be dissimilar", "no spatial pattern",
                     "neighbours tend to be similar")) +
  annotate("point", x = i_obs, y = 0, size = 3.5, colour = "darkred") +
  annotate("label", x = i_obs, y = 0.25, vjust = -0.2, size = 3.2,
           label = paste0("GDPPC, I = ", i_obs)) +
  annotate("text", x = c(-1, 0, 1), y = -0.18, label = c("-1", "0", "+1"), size = 3) +
  coord_cartesian(xlim = c(-1.15, 1.15), ylim = c(-0.7, 0.75)) +
  theme_void()
```

The width of the middle band is schematic: whether a value of I is close enough to zero is decided
by the test above. GDPPC is in the positive case: counties have values more similar to those of
their neighbours than a random allocation of the same 88 values would produce. The test confirms
that the higher GDPPC in the north east and lower GDPPC in the west, shown by the quantile map in
section 3.1, is not an artefact of the visual interpretation of a choropleth.

### The permutation test

Moran's I has an analytical null distribution. It assumes either that the values are drawn from a
normal distribution or that they are a random permutation of the observed values. For a right-skewed
variable such as GDPPC, a test that does not depend on this assumption is preferable. A Monte Carlo
test constructs the null distribution empirically: the 88 values are randomly reassigned to the 88
counties many times, and I is computed for each permutation.

```{r}
set.seed(1234)
bperm <- moran.mc(hunan$GDPPC,
                  listw = rswm_q,
                  nsim = 999,
                  zero.policy = TRUE,
                  na.action = na.omit)
bperm
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses two functions:

```r
set.seed(seed)                                            # base R
moran.mc(x, listw, nsim, zero.policy = attr(listw, "zero.policy"),
         alternative = "greater", na.action = na.fail)    # spdep
```

- `set.seed()` fixes the starting point of R's random number generator. The permutations are then
  the same on every render, and the reported p-value does not change between renders.
- `moran.mc()` runs the Monte Carlo test: it reassigns the values of `x` to the counties at random
  `nsim` times and computes I for each assignment. `nsim` has no default. With `nsim = 999` the
  observed I is ranked among 1,000 values, so the smallest attainable pseudo p-value is 1/1000.
- `alternative = "greater"`, the default, tests for positive spatial autocorrelation. `na.action =
  na.omit` drops counties with missing values, where the default, `na.fail`, stops with an error.
  GDPPC has no missing values, so the argument has no effect here. `zero.policy` defaults to the
  setting stored in the weights object, and `rswm_q` was built with `zero.policy = TRUE`, so setting
  it again changes nothing.
- `bperm$res` holds the 999 simulated values followed by the observed I as its 1,000th element. The
  summaries and the histogram below use `bperm$res[1:999]`, which excludes the observed value.
  Chapter 9 plots the whole of `bperm$res`.
:::

Chapter 9 asks what statistical conclusion the output supports:

- **Rank:** the observed I ranks 1000th of 1000 values, above all 999 permuted values.
- **Pseudo p-value:** 0.001, the smallest attainable with 999 simulations.
- **Conclusion:** the null hypothesis of spatial randomness is rejected, as in section 3.3.1.

### The simulated distribution

```{r}
mean(bperm$res[1:999])
var(bperm$res[1:999])
summary(bperm$res[1:999])
```

```{r}
#| fig-height: 5
ggplot(tibble(I = bperm$res[1:999]), aes(x = I)) +
  geom_histogram(bins = 20, fill = "grey70", colour = "black") +
  geom_vline(xintercept = -1 / (nrow(hunan) - 1), colour = "red") +
  labs(x = "Simulated Moran's I", y = "Count",
       title = "Monte Carlo null distribution, 999 permutations") +
  theme_minimal()
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses these functions:

```r
tibble(...)                                               # tibble
ggplot(data, aes(x))                                      # ggplot2
geom_histogram(bins, fill, colour)                        # ggplot2
geom_vline(xintercept, colour)                            # ggplot2
labs(x, y, title)                                         # ggplot2
theme_minimal()                                           # ggplot2
```

- `tibble(I = bperm$res[1:999])` builds a data frame with one column, holding the 999 simulated
  values of Moran's I. The name `I` is chosen here and is only a column name.
- `ggplot()` starts a plot from that data frame. `aes()` sets the aesthetic mapping: which column
  controls which visual property of the plot, such as position, colour or size. `aes(x = I)` places
  the values of column `I` along the horizontal axis. The observed I is not among them, and the text
  below the plot compares it with the simulated values.
- Each `+` adds a layer or a setting. `geom_histogram()` sorts the values into `bins = 20` intervals
  and draws a bar for each, with `fill` setting the colour of the bars and `colour` their outline.
- `geom_vline()` draws a vertical line at `xintercept`, here the expectation −1/(n−1).
- `labs()` sets the axis labels and the title, and `theme_minimal()` removes the grey background.
:::

Chapter 9 asks what statistical observation the histogram supports:

- **Centre:** the simulated values are distributed approximately symmetrically around the
  expectation −1/(n−1), marked by the red line. Chapter 9 draws the line at zero, which at n = 88 is
  0.0115 away and indistinguishable on this axis.
- **Observed value:** 0.301, beyond the largest of the 999 simulated values, 0.276.

Chapter 9 also sets a challenge: to plot the values with `ggplot2` instead of base graphics. The
histogram above is drawn with `ggplot2`.

## Geary's C

### The test

Moran's I is based on the deviation of each county from the provincial mean. Geary's C is based on
the squared differences between each county and its neighbours, so it is more sensitive to local
differences between neighbouring values.

```{r}
geary.test(hunan$GDPPC, listw = rswm_q)
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses one function:

```r
geary.test(x, listw, randomisation = TRUE, zero.policy = attr(listw, "zero.policy"),
           alternative = "greater")                                        # spdep
```

- `geary.test()` computes Geary's C for `x` with the weights in `listw`, and tests it against its
  expectation of 1 under spatial randomness. The call passes only `x` and `listw`, so every other
  argument keeps its default.
- The "standard deviate" in the output is (1 − C) divided by the standard deviation of C, so it is
  positive when C is below 1. `alternative = "greater"` therefore tests for clustering, as it does
  for Moran's I, although clustering makes C smaller than its expectation.
- `randomisation = TRUE`, the default, sets the null model for the variance of C, as for
  `moran.test()` in section 3.3.1.

The two maps compare the statistics on one county, Anxiang, and its five neighbours from section
3.2. Moran's I compares every county with the provincial mean. Geary's C compares Anxiang directly
with each neighbour and squares the difference, so the mean plays no part.

```{r}
#| echo: false
#| fig-height: 4.6
mc_mu <- mean(hunan$GDPPC)
mc_pts <- st_point_on_surface(nb_focus[nb_ids, ]) |>
  mutate(dev = GDPPC - mc_mu, diff = GDPPC[1] - GDPPC)
mc_xy <- st_coordinates(mc_pts)
mc_links <- tibble(x = mc_xy[1, 1], y = mc_xy[1, 2], xend = mc_xy[-1, 1], yend = mc_xy[-1, 2],
                   diff = mc_pts$diff[-1], name = mc_pts$NAME_3[-1])
fmt <- function(v) paste0(ifelse(v > 0, "+", ""), format(round(v), big.mark = ",", trim = TRUE))
zoom <- coord_sf(xlim = nb_box[c("xmin", "xmax")], ylim = nb_box[c("ymin", "ymax")], datum = NA)

moran_view <- ggplot() +
  geom_sf(data = nb_focus, fill = "grey92", colour = "white", linewidth = 0.3) +
  geom_sf(data = mutate(nb_focus[nb_ids, ], dev = GDPPC - mc_mu), aes(fill = dev),
          colour = "white", linewidth = 0.3) +
  geom_sf_text(data = mc_pts, aes(label = paste0(NAME_3, "\n", fmt(dev))), size = 2.8,
               lineheight = 0.9) +
  scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#b2182b", midpoint = 0,
                       guide = "none") +
  zoom +
  labs(title = "Moran's I: each county against the mean",
       subtitle = paste0("deviation from the provincial mean of ",
                         format(round(mc_mu), big.mark = ","))) +
  theme_void()

geary_view <- ggplot() +
  geom_sf(data = nb_focus, aes(fill = role), colour = "white", linewidth = 0.3) +
  geom_segment(data = mc_links, aes(x, y, xend = xend, yend = yend), colour = "grey30") +
  geom_label(data = mc_links, aes(x = x + 0.7 * (xend - x), y = y + 0.7 * (yend - y),
                 label = paste0(name, ": ", format(abs(diff), big.mark = ",", trim = TRUE))),
             size = 2.6, label.size = 0, fill = "white") +
  scale_fill_manual(values = c("Anxiang (row 1)" = "darkorange",
                               "its neighbours" = "#fdd49e", "other" = "grey92"),
                    guide = "none") +
  zoom +
  labs(title = "Geary's C: Anxiang against each neighbour",
       subtitle = "size of the difference in GDPPC, which C squares") +
  theme_void()

patchwork::wrap_plots(moran_view, geary_view, ncol = 2)
```
:::

Chapter 9 asks what statistical conclusion the output supports. Geary's C for GDPPC is 0.691,
against an expectation of exactly 1 (z = 3.61, p = 1.5 × 10⁻⁴), so the null hypothesis of spatial
randomness is rejected.

Its scale is inverted relative to Moran's I. Slide 12 of the Lesson 5 lecture states that C
approaches 0 and I approaches 1 when similar values cluster:

| Pattern | Moran's I | Geary's C |
|---|---|---|
| clustered: neighbouring values similar | positive, towards 1 | below 1, towards 0 |
| random | about its expectation, −0.0115 | about 1 |
| dispersed: neighbouring values dissimilar | negative | above 1 |
| **GDPPC** | **0.301** | **0.691** |

Both values are on the clustered side, so like Moran's I, C indicates significant positive spatial
autocorrelation.

### The permutation test

```{r}
set.seed(1234)
bperm <- geary.mc(hunan$GDPPC,
                  listw = rswm_q,
                  nsim = 999)
bperm
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses one function:

```r
geary.mc(x, listw, nsim, zero.policy = attr(listw, "zero.policy"),
         alternative = "greater")                                          # spdep
```

- `geary.mc()` runs the Monte Carlo test for Geary's C in the same way as `moran.mc()` in section
  3.3.2: it reassigns the values of `x` to the counties at random `nsim` times and computes C for
  each assignment. `set.seed(1234)` again makes the permutations the same on every render.
- With `alternative = "greater"`, the pseudo p-value counts the permuted values of C that are at or
  below the observed one, since a small C indicates clustering.
- The result is stored in `bperm`, which replaces the Moran's I result of section 3.3.2. As before,
  `bperm$res` holds the 999 simulated values followed by the observed C as its 1,000th element.
:::

Chapter 9 asks what statistical conclusion the output supports:

- **Rank:** the observed C, 0.69072, ranks first of the 1000 values, below all 999 permuted values.
- **Pseudo p-value:** 0.001, the smallest attainable with 999 simulations.
- **Conclusion:** the null hypothesis of spatial randomness is rejected, as in section 3.4.1.

### The simulated distribution

```{r}
mean(bperm$res[1:999])
var(bperm$res[1:999])
summary(bperm$res[1:999])
```

```{r}
#| fig-height: 5
ggplot(tibble(C = bperm$res[1:999]), aes(x = C)) +
  geom_histogram(bins = 20, fill = "grey70", colour = "black") +
  geom_vline(xintercept = 1, colour = "red") +
  labs(x = "Simulated Geary's C", y = "Count",
       title = "Monte Carlo null distribution, 999 permutations") +
  theme_minimal()
```

Chapter 9 asks what statistical observation the histogram supports:

- **Centre:** the reference line at 1 is the expectation under spatial randomness, and the simulated
  mean is approximately equal to it.
- **Observed value:** 0.691 lies beyond the left tail, which for Geary's C is the side that
  indicates clustering.

## Spatial correlogram

Both statistics so far were computed on first-order neighbours only. A spatial correlogram computes
the statistic at successive lag orders, where lag 2 comprises the neighbours of the neighbours, and
shows the range of lags over which spatial autocorrelation persists.

```{r}
#| fig-height: 5
MI_corr <- sp.correlogram(wm_q,
                          hunan$GDPPC,
                          order = 6,
                          method = "I",
                          style = "W")
plot(MI_corr)
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses one function:

```r
sp.correlogram(neighbours, var, order = 1, method = "corr", style = "W")   # spdep
```

- `sp.correlogram()` computes a spatial autocorrelation statistic at each lag from 1 to `order`. It
  takes the neighbour list `wm_q` because it builds the higher-order neighbours itself, so it needs
  no weights object. The lag-2 neighbours of a county are the neighbours of its neighbours,
  excluding the county itself and its lag-1 neighbours, and so on up to `order = 6`.
- `method` selects the statistic. The default, `"corr"`, is a correlation coefficient, so
  `method = "I"` is needed for Moran's I. The Geary correlogram below uses `method = "C"`.
- `style = "W"` row-standardises the weights at each lag. `plot()` draws the value at each lag with
  a bar two standard deviations either side, and a horizontal line at the expected value under
  spatial randomness.

The map shows the lags that `sp.correlogram()` builds, counted from one county, Changsha. Each
county is coloured by the number of steps across shared boundaries that separate it from Changsha.
At lag 1 Changsha is compared with its immediate neighbours, and at lags 5 and 6 with counties in the
west and south of the province.

```{r}
#| echo: false
#| fig-height: 5.5
cs <- which(hunan$NAME_3 == "Changsha")
lags <- nblag(wm_q, 6)
lag_order <- rep(NA_integer_, nrow(hunan))
for (k in 1:6) lag_order[lags[[k]][[cs]]] <- k
lag_order[cs] <- 0L

hunan |>
  mutate(lag = factor(ifelse(is.na(lag_order), "beyond 6", as.character(lag_order)),
                      levels = c(as.character(0:6), "beyond 6"),
                      labels = c("Changsha", paste("lag", 1:6), "beyond lag 6"))) |>
  ggplot() +
  geom_sf(aes(fill = lag), colour = "white", linewidth = 0.3) +
  scale_fill_manual(values = c("black", "#08306b", "#2171b5", "#6baed6",
                               "#c6dbef", "#fdd0a2", "#fd8d3c", "grey85")) +
  labs(fill = "Steps from Changsha") +
  theme_void()
```
:::

```{r}
print(MI_corr)
```

Chapter 9 asks what statistical observation the correlogram supports:

- **Decline:** Moran's I falls from 0.30 at lag 1 to 0.03 at lag 4.
- **Sign change:** it turns negative between lags 4 and 5, and is significantly negative at lags 5
  and 6.
- **Reason:** negative values at the higher lags reflect the geometry of the province as well as the
  distribution of GDPPC. At lag 5 the neighbour sets of a county extend across the province.
  Counties in the north east are then compared with counties in the west, and the values at the two
  ends differ.

Lag 4 is the first order at which the null hypothesis of no spatial autocorrelation is not rejected.
Lag 3 is significant at p = 0.04, a marginal result that warrants adjustment, since six tests are
reported in one table. A correction raises the evidence a test must reach, so that a lag counts as
significant only on stronger evidence than 0.05. Holm's correction ranks the six p-values and
applies the strictest threshold to the smallest, then a looser threshold to each in turn.
`print.spcor()` defaults to `p.adj.method = "none"`, so the first table reported unadjusted
p-values.

```{r}
print(MI_corr, p.adj.method = "holm")
```

Under Holm's correction lag 3 is no longer significant (p = 0.081), and the p-value at lag 6
increases to 0.027. At lags 1, 2 and 5 the p-values all increase and all remain below 0.001. Only
lag 4 is unchanged, since it already had the largest p-value in the table. Section 5.3 addresses the
same multiple-comparisons problem on a much larger scale, for the 88 local statistics.

```{r}
#| fig-height: 5
GC_corr <- sp.correlogram(wm_q,
                          hunan$GDPPC,
                          order = 6,
                          method = "C",
                          style = "W")
plot(GC_corr)
```

```{r}
print(GC_corr)
```

Geary's C increases from 0.69 towards 1 and then exceeds it. Its correlogram has the shape of the
Moran correlogram with the vertical axis reversed. Their p-values, however, do not follow the same
pattern.

| Lag | Moran p | Geary p |
|---|---|---|
| 1 | 2.2e-06 | 3.1e-04 |
| 2 | 2.0e-06 | 7.2e-04 |
| 3 | 0.040 | 0.389 |
| 4 | 0.226 | 0.876 |
| 5 | 6.0e-05 | 7.6e-04 |
| 6 | 0.0089 | 0.310 |

Moran's I and Geary's C agree at lags 1, 2, 4 and 5. At lags 3 and 6 Moran's I is significant and
Geary's C is not. This difference follows from the structure of the two statistics. Moran's I is
based on deviations from the provincial mean, and a gradual trend across the province contributes to
it. Geary's C is based on differences between neighbouring values, and a weak trend that leaves
adjacent pairs similar contributes little to it. Section 5.6 finds the opposite for the local forms:
local Geary's c identifies more significant counties than local Moran's I, 21 against 16.

# Part 2: Local measures of spatial autocorrelation

This part follows r4gdsa [Chapter 10](https://r4gdsa.netlify.app/chap10.html). Each statistic in it
is computed for every county, and the results are mapped. Chapter 10 repeats the data import and the
GDPPC maps of chapter 9, which are in sections 2 and 3.1 of this page and are not repeated here.

## Local indicators of spatial association

A global statistic summarises the whole province in a single value. It shows neither where the
clustering is concentrated nor which individual counties differ from their neighbours. A local
indicator of spatial association (LISA) is computed for each county and answers both questions.
Anselin (1995) defines a LISA as any statistic that satisfies two requirements:

1. **Local clustering.** It "gives an indication of the extent of significant spatial clustering of
   similar values around that observation".
2. **Link to the global statistic.** "The sum of LISAs for all observations is proportional to a
   global indicator of spatial association".

Local Moran's I satisfies both. Under the second requirement, its 88 local values sum to a quantity
proportional to the global Moran's I of section 3.3, as slide 22 of the Lesson 5 lecture shows.

### Weights the sfdep way

`sfdep` stores the neighbour list and the weights as columns of the sf object, each holding a list.
Each row holds a county's geometry, its attributes, its neighbours and its weights.

```{r}
wm_q <- hunan |>
  mutate(nb = st_contiguity(geometry),
         wt = st_weights(nb, style = "W"),
         .before = 1)
summary(wm_q$nb)
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses three functions:

```r
mutate(.data, name = value, ..., .before = NULL)          # dplyr
st_contiguity(geometry, queen = TRUE)                     # sfdep
st_weights(nb, style = "W", allow_zero = NULL)            # sfdep
```

- `mutate()` adds columns to a data frame. Each `name = value` pair creates one column, here `nb`
  and `wt`. A later pair can use a column created by an earlier one, so `wt` is computed from the
  `nb` created on the line above.
- `.before` sets where the new columns go, and takes a column position or a column name. `.before =
  1` places them before the column now in position 1, so they become the first two columns. The
  default, `NULL`, adds them after the existing columns. Placing them first makes them visible at
  the start of each printed row.
- `st_contiguity()` finds, for each polygon, the polygons that share a boundary with it, and returns
  a neighbour list: for each county, a vector of the row numbers of its neighbours. With `queen =
  TRUE`, the default, a single shared boundary point is enough. `queen = FALSE` requires a shared
  edge. The function calls `poly2nb()` from section 3.2, so it returns the same list.
- `st_weights()` gives each neighbour a weight. With `style = "W"`, the default, each of a county's
  weights is 1 divided by its number of neighbours, so its weights sum to 1. The function calls
  `nb2listw()` from section 3.2 and keeps only the weights. `allow_zero` concerns counties without
  neighbours, and Hunan has none.

`hunan` before the block, first three counties:

<table class="table table-sm">
<tr><th>NAME_2</th><th>…</th><th>County</th><th>GDPPC</th><th>geometry</th></tr>
<tr><td>Changde</td><td>…</td><td>Anxiang</td><td>23667</td><td>polygon</td></tr>
<tr><td>Changde</td><td>…</td><td>Hanshou</td><td>20981</td><td>polygon</td></tr>
<tr><td>Changde</td><td>…</td><td>Jinshi</td><td>34592</td><td>polygon</td></tr>
<tr><td>…</td><td>…</td><td>…</td><td>…</td><td>…</td></tr>
</table>

`wm_q` after the block, with the two new columns shaded:

<table class="table table-sm">
<tr><th style="background: rgba(255, 193, 7, 0.30)">nb</th><th style="background: rgba(255, 193, 7, 0.30)">wt</th><th>NAME_2</th><th>…</th><th>County</th><th>GDPPC</th><th>geometry</th></tr>
<tr><td style="background: rgba(255, 193, 7, 0.30)">2, 3, 4, 57, 85</td><td style="background: rgba(255, 193, 7, 0.30)">0.2, 0.2, 0.2, 0.2, 0.2</td><td>Changde</td><td>…</td><td>Anxiang</td><td>23667</td><td>polygon</td></tr>
<tr><td style="background: rgba(255, 193, 7, 0.30)">1, 57, 58, 78, 85</td><td style="background: rgba(255, 193, 7, 0.30)">0.2, 0.2, 0.2, 0.2, 0.2</td><td>Changde</td><td>…</td><td>Hanshou</td><td>20981</td><td>polygon</td></tr>
<tr><td style="background: rgba(255, 193, 7, 0.30)">1, 4, 5, 85</td><td style="background: rgba(255, 193, 7, 0.30)">0.25, 0.25, 0.25, 0.25</td><td>Changde</td><td>…</td><td>Jinshi</td><td>34592</td><td>polygon</td></tr>
<tr><td style="background: rgba(255, 193, 7, 0.30)">…</td><td style="background: rgba(255, 193, 7, 0.30)">…</td><td>…</td><td>…</td><td>…</td><td>…</td><td>…</td></tr>
</table>

Each cell of `nb` and `wt` holds a vector, one element per neighbour. Anxiang has five neighbours,
so each receives a weight of 1/5 = 0.2, and Jinshi has four, each weighted 0.25.

The figure follows Anxiang through the two functions, as the figure in section 3.2 does for
`poly2nb()` and `nb2listw()`.

```{r}
#| echo: false
#| fig-height: 4.2
sf_steps <- ggplot() +
  annotate("text", x = 0, y = 0.95, label = "st_contiguity(geometry)", family = "mono",
           size = 3.4, hjust = 0) +
  annotate("label", x = 0, y = 0.78, hjust = 0, family = "mono", size = 3.4,
           label = paste0("wm_q$nb[[1]]\n", paste(wm_q$nb[[1]], collapse = "    "))) +
  annotate("segment", x = 0.25, xend = 0.25, y = 0.64, yend = 0.46,
           arrow = arrow(length = unit(0.25, "cm"))) +
  annotate("text", x = 0.3, y = 0.55, label = 'st_weights(nb, style = "W")',
           family = "mono", size = 3.4, hjust = 0) +
  annotate("label", x = 0, y = 0.3, hjust = 0, family = "mono", size = 3.4,
           label = paste0("wm_q$wt[[1]]\n", paste(wm_q$wt[[1]], collapse = "  "))) +
  coord_cartesian(xlim = c(0, 1), ylim = c(0.1, 1), clip = "off") +
  theme_void()

patchwork::wrap_plots(nb_map, sf_steps, widths = c(1.2, 1))
```
:::

The neighbour summary matches `poly2nb()` exactly: 448 links across 88 counties, average 5.09.

### Local Moran's I

```{r}
set.seed(1234)
lisa <- wm_q |>
  mutate(local_moran = local_moran(GDPPC, nb, wt, nsim = 999),
         .before = 1) |>
  unnest(local_moran)
glimpse(lisa)
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses three functions:

```r
local_moran(x, nb, wt, alternative = "two.sided", nsim = 499)   # sfdep
unnest(data, cols)                                              # tidyr
glimpse(x)                                                      # dplyr
```

- `local_moran()` computes local Moran's I for each county and tests it by conditional permutation.
  It returns a data frame with one row per county and one column per result. `nsim` defaults to
  499, and the call sets 999 to match section 3.3.2. With `alternative = "two.sided"`, the default,
  both unusually high and unusually low values of Ii count as significant.
- Inside `mutate()` (section 4.1.1), that data frame is stored as a single column named
  `local_moran`, and `.before = 1` places it first.
- `unnest()` replaces the column named in `cols` with the columns it contains, so the 12 results
  become 12 columns.
- `glimpse()` prints the name, type and first values of every column, one column per line.

Before `unnest()`, `local_moran` is a single column that holds a 12-column data frame, one row per
county:

<table class="table table-sm">
<tr><th style="background: rgba(255, 193, 7, 0.30)">local_moran</th><th>nb</th><th>wt</th><th>…</th><th>County</th><th>geometry</th></tr>
<tr><td style="background: rgba(255, 193, 7, 0.30)">ii, eii, …, pysal: 12 values</td><td>2, 3, 4, 57, 85</td><td>0.2, …</td><td>…</td><td>Anxiang</td><td>polygon</td></tr>
<tr><td style="background: rgba(255, 193, 7, 0.30)">ii, eii, …, pysal: 12 values</td><td>1, 57, 58, 78, 85</td><td>0.2, …</td><td>…</td><td>Hanshou</td><td>polygon</td></tr>
<tr><td style="background: rgba(255, 193, 7, 0.30)">…</td><td>…</td><td>…</td><td>…</td><td>…</td><td>…</td></tr>
</table>

After `unnest()`, its 12 columns, shaded, replace it:

<table class="table table-sm">
<tr><th style="background: rgba(255, 193, 7, 0.30)">ii</th><th style="background: rgba(255, 193, 7, 0.30)">eii</th><th style="background: rgba(255, 193, 7, 0.30)">…</th><th style="background: rgba(255, 193, 7, 0.30)">p_ii_sim</th><th style="background: rgba(255, 193, 7, 0.30)">…</th><th style="background: rgba(255, 193, 7, 0.30)">mean</th><th style="background: rgba(255, 193, 7, 0.30)">median</th><th style="background: rgba(255, 193, 7, 0.30)">pysal</th><th>nb</th><th>…</th><th>geometry</th></tr>
<tr><td style="background: rgba(255, 193, 7, 0.30)">−0.0015</td><td style="background: rgba(255, 193, 7, 0.30)">…</td><td style="background: rgba(255, 193, 7, 0.30)">…</td><td style="background: rgba(255, 193, 7, 0.30)">0.798</td><td style="background: rgba(255, 193, 7, 0.30)">…</td><td style="background: rgba(255, 193, 7, 0.30)">Low-High</td><td style="background: rgba(255, 193, 7, 0.30)">…</td><td style="background: rgba(255, 193, 7, 0.30)">…</td><td>2, 3, 4, 57, 85</td><td>…</td><td>polygon</td></tr>
<tr><td style="background: rgba(255, 193, 7, 0.30)">0.0259</td><td style="background: rgba(255, 193, 7, 0.30)">…</td><td style="background: rgba(255, 193, 7, 0.30)">…</td><td style="background: rgba(255, 193, 7, 0.30)">0.970</td><td style="background: rgba(255, 193, 7, 0.30)">…</td><td style="background: rgba(255, 193, 7, 0.30)">Low-Low</td><td style="background: rgba(255, 193, 7, 0.30)">…</td><td style="background: rgba(255, 193, 7, 0.30)">…</td><td>1, 57, 58, 78, 85</td><td>…</td><td>polygon</td></tr>
<tr><td style="background: rgba(255, 193, 7, 0.30)">…</td><td style="background: rgba(255, 193, 7, 0.30)">…</td><td style="background: rgba(255, 193, 7, 0.30)">…</td><td style="background: rgba(255, 193, 7, 0.30)">…</td><td style="background: rgba(255, 193, 7, 0.30)">…</td><td style="background: rgba(255, 193, 7, 0.30)">…</td><td style="background: rgba(255, 193, 7, 0.30)">…</td><td style="background: rgba(255, 193, 7, 0.30)">…</td><td>…</td><td>…</td><td>…</td></tr>
</table>

The values shown are those of the first two counties, Anxiang and Hanshou.
:::

Section 5.2 compares two of these p-value columns, `p_ii` and `p_ii_sim`. Each column is described
below.

| Column | What it holds |
|---|---|
| `ii` | the local Moran's I value for that county |
| `eii`, `var_ii`, `z_ii` | the permutation sample mean, variance and standardised deviate. With `nsim` supplied these are computed from the simulations, not from the analytical null |
| `p_ii` | a p-value from `pnorm()` on `z_ii`, so a normal approximation to the permutation distribution |
| `p_ii_sim` | a rank-based p-value from the same permutations, two-sided, on a [0, 1] scale |
| `p_folded_sim` | the same ranking folded onto a [0, 1] range of 0.5, so close to half of `p_ii_sim` |
| `mean`, `median`, `pysal` | three labellings of the county into one of the four scatterplot quadrants |

Under the analytical randomisation null hypothesis with row-standardised weights, every county has
the same expectation, −1/(n−1) = −0.0115. The printed values of `eii` differ between counties, so
they are sample means over the permutations.

Chapter 10 uses `p_ii`. This page uses `p_ii_sim`, for the reasons given in section 5.2.

### The two maps

```{r}
#| fig-height: 10.13
tm_shape(lisa) +
  tm_polygons(fill = "ii",
              fill.scale = tm_scale_intervals(style = "pretty", n = 5,
                                              values = "brewer.rd_bu"),
              fill.legend = tm_legend(title = "Local Moran's I",
                                      position = tm_pos_out())) +
  tm_borders(col_alpha = 0.5) +
  tm_title("Local Moran's I of GDPPC, Queen contiguity")
```

A large positive value indicates a county whose neighbours have similar values. A negative value
indicates a county whose value differs from those of its neighbours. On its own, Ii does not
indicate whether the association is significant. The p-value map below provides that information.

```{r}
#| fig-height: 10.13
tm_shape(lisa) +
  tm_polygons(fill = "p_ii_sim",
              fill.scale = tm_scale_intervals(
                breaks = c(0, 0.01, 0.05, 0.1, 1),
                values = "-brewer.reds"),
              fill.legend = tm_legend(title = "simulated p-value",
                                      position = tm_pos_out())) +
  tm_borders(col_alpha = 0.5) +
  tm_title("Simulated p-values of local Moran's I, 999 permutations")
```

```{r}
min(lisa$p_ii_sim)
```

The legend breaks differ from those of the chapter. `p_ii_sim` is a rank-based p-value, and its
smallest attainable value is `2 / (nsim + 1)`, which is 0.002 at `nsim = 999` and 0.02 at the
chapter's `nsim = 99`. The breaks are set at values attainable with `nsim = 999` (section 5.2).

```{r}
#| fig-height: 4.1
ii.map <- tm_shape(lisa) +
  tm_polygons(fill = "ii",
              fill.scale = tm_scale_intervals(style = "pretty", n = 5,
                                              values = "brewer.rd_bu"),
              fill.legend = tm_legend(title = "Local Moran's I",
                                      position = tm_pos_out("right", "center"))) +
  tm_borders(col_alpha = 0.5) +
  tm_title("Local Moran's I")

p_ii.map <- tm_shape(lisa) +
  tm_polygons(fill = "p_ii_sim",
              fill.scale = tm_scale_intervals(
                breaks = c(0, 0.01, 0.05, 0.1, 1),
                values = "-brewer.reds"),
              fill.legend = tm_legend(title = "p-value",
                                      position = tm_pos_out("right", "center"))) +
  tm_borders(col_alpha = 0.5) +
  tm_title("Simulated p-value")

tmap_arrange(ii.map, p_ii.map, asp = 1, ncol = 2)
```

Neither map is sufficient on its own. Several counties with a large value of `ii` are not
significant, and the set of significant counties is smaller than the set of counties with prominent
values on the first map.

### Moran scatterplot

```{r}
lisa <- lisa |>
  mutate(lag_GDPPC = st_lag(GDPPC, nb, wt), .before = 1)
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses one function:

```r
st_lag(x, nb, wt)                                         # sfdep
```

- `st_lag()` computes the spatial lag of `x`: for each county, the sum of its neighbours' values,
  each multiplied by its weight. With row-standardised weights this is the average GDPPC of the
  neighbours, which the scatterplot below plots against the county's own GDPPC.

The figure works through the lag for Anxiang. Each of its five neighbours has a weight of 0.2, so
its lag is the average GDPPC of the five.

```{r}
#| echo: false
#| fig-height: 4.4
lag_ids <- wm_q$nb[[1]]
lag_labels <- nb_focus[c(1, lag_ids), ] |>
  mutate(txt = paste0(NAME_3, "\n", format(GDPPC, big.mark = ","))) |>
  st_point_on_surface()
lag_map <- nb_map
lag_map$layers[[2]] <- geom_sf_text(data = lag_labels, aes(label = txt), size = 2.8,
                                    lineheight = 0.9)

terms <- paste0(c("  ", rep("+ ", length(lag_ids) - 1)),
                format(wm_q$wt[[1]], nsmall = 1), " x ",
                format(hunan$GDPPC[lag_ids], big.mark = ","),
                "  (", hunan$NAME_3[lag_ids], ")")
lag_lines <- c("st_lag(GDPPC, nb, wt), row 1:", "",
               "lag = sum of weight x neighbour GDPPC", "", terms, "",
               paste0("= ", format(round(lisa$lag_GDPPC[1], 1), big.mark = ",")),
               "", paste0("Anxiang's own GDPPC: ", format(hunan$GDPPC[1], big.mark = ",")))
lag_steps <- ggplot() +
  annotate("segment", x = -0.12, xend = -0.02, y = 0.5, yend = 0.5,
           arrow = arrow(length = unit(0.25, "cm"))) +
  annotate("text", x = 0, y = 0.5, hjust = 0, vjust = 0.5, family = "mono", size = 3.1,
           lineheight = 1.05, label = paste(lag_lines, collapse = "\n")) +
  coord_cartesian(xlim = c(-0.15, 1), ylim = c(0, 1), clip = "off") +
  theme_void()

patchwork::wrap_plots(lag_map, lag_steps, widths = c(1.2, 1))
```
:::

```{r}
#| fig-height: 6
ggplot(data = lisa, aes(x = GDPPC, y = lag_GDPPC)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, colour = "red") +
  labs(x = "GDPPC",
       y = "Spatial lag of GDPPC",
       title = "Moran scatterplot") +
  theme_minimal()
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses two new functions:

```r
geom_point(size)                                          # ggplot2
geom_smooth(method = NULL, se = TRUE, colour)             # ggplot2
```

- `aes(x = GDPPC, y = lag_GDPPC)` places each county's GDPPC on the horizontal axis and its spatial
  lag on the vertical axis.
- `geom_point()` draws one point per county.
- `geom_smooth()` adds a fitted line. `method = "lm"` fits a straight line by least squares, and
  `se = FALSE` leaves out the grey confidence band that is drawn by default.
:::

This is the scatterplot from the end of Hands-on Exercise 4, with a fitted regression line added.
With row-standardised weights, Moran's I is the least-squares slope of the spatial lag on the
variable in deviations from its mean. Anselin (1996) writes I = y'Wy / y'y and treats the plot as
the regression of Wy on y. The slope of the line above is therefore Moran's I.

Anselin (1996) standardises the variable and then takes the spatial lag of the standardised values,
so the vertical axis is Wz. Chapter 10 standardises the lag separately, with `scale(lag_GDPPC)`.
This page follows Anselin (1996):

```{r}
lisa <- lisa |>
  mutate(z_GDPPC = as.numeric(scale(GDPPC)),
         z_lag_GDPPC = st_lag(z_GDPPC, nb, wt),
         .before = 1)
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses two functions:

```r
scale(x, center = TRUE, scale = TRUE)                     # base R
as.numeric(x)                                             # base R
```

- `scale()` standardises a variable: it subtracts the mean and divides by the standard deviation, so
  the result has mean 0 and standard deviation 1.
- `scale()` returns an 88 × 1 matrix, and `as.numeric()` turns it back into a plain vector, so that
  it can be stored as an ordinary column.
- `st_lag()` then takes the spatial lag of the standardised values, which is Anselin's (1996)
  construction.
:::

The two constructions, the lag of the standardised variable and the separately standardised lag,
give different slopes on the same 88 counties:

```{r}
tibble(construction = c("Moran's I (section 3.3.1)",
                        "lag of GDPPC on GDPPC",
                        "Anselin (1996): lag of z on z",
                        "chapter 10: scale(lag) on z"),
       slope = c(moran.test(hunan$GDPPC, rswm_q)$estimate[1],
                 coef(lm(lag_GDPPC ~ GDPPC, data = lisa))[2],
                 coef(lm(z_lag_GDPPC ~ z_GDPPC, data = lisa))[2],
                 coef(lm(as.numeric(scale(lag_GDPPC)) ~ z_GDPPC, data = lisa))[2])) |>
  kable(digits = 4)
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses three new functions:

```r
lm(formula, data)                                         # stats
coef(object)                                              # stats
kable(x, digits)                                          # knitr
```

- `lm()` fits a linear regression. The formula `lag_GDPPC ~ GDPPC` reads "lag_GDPPC explained by
  GDPPC": the variable before `~` is on the vertical axis of the scatterplot and the one after it on
  the horizontal axis.
- `coef()` returns the fitted coefficients, the intercept first and the slope second, so `[2]`
  takes the slope.
- `kable()` prints a data frame as a formatted table, and `digits = 4` rounds the numbers to four
  decimal places.
:::

The chapter's figure is titled a Moran scatterplot, but its regression line does not have Moran's I
as its slope. Under the chapter's construction the slope is 0.485, the Pearson correlation between
GDPPC and its spatial lag. The lag is an average of neighbouring values and varies less than GDPPC
(standard deviations of about 9,300 and 15,000 yuan), so standardising it by its own standard
deviation stretches it. Under Anselin's (1996) construction the slope is 0.301, equal to Moran's I.
The assignment of counties to quadrants is the same under both constructions for this data.


```{r}
#| fig-height: 6
ggplot(data = lisa, aes(x = z_GDPPC, y = z_lag_GDPPC, colour = mean)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE, colour = "black") +
  geom_hline(yintercept = 0, lty = 2) +
  geom_vline(xintercept = 0, lty = 2) +
  scale_colour_manual(values = c("High-High" = "red",
                                 "Low-Low" = "blue",
                                 "Low-High" = "lightblue",
                                 "High-Low" = "pink")) +
  labs(x = "Standardised GDPPC",
       y = "Standardised spatial lag of GDPPC",
       colour = "Quadrant",
       title = "Standardised Moran scatterplot with LISA quadrants") +
  theme_minimal()
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses two new functions:

```r
geom_hline(yintercept, lty)                               # ggplot2
scale_colour_manual(values)                               # ggplot2
```

- `colour = mean` inside `aes()` colours each point by its quadrant label in the `mean` column.
- `geom_hline()` and `geom_vline()` (section 3.3.3) draw the dashed lines at zero, and `lty = 2`
  sets a dashed line.
- `scale_colour_manual()` sets the colour of each quadrant by name, matching the colours of the
  LISA cluster map in section 4.1.5.
:::

The dashed lines mark zero, the mean of a standardised variable.

In the LISA cluster map, counties are classified by the four quadrants:

| Quadrant | Position | Reading |
|---|---|---|
| High-High | top right | a high value surrounded by high values, the core of a cluster |
| Low-Low | bottom left | a low value surrounded by low values |
| High-Low | bottom right | a high value surrounded by low values, a spatial outlier |
| Low-High | top left | a low value surrounded by high values, a spatial outlier |

Counties in the High-High and Low-Low quadrants have a positive local I, and counties in the
High-Low and Low-High quadrants have a negative local I. The sign of Ii alone does not identify the
quadrant: a positive value may indicate a High-High or a Low-Low cluster, and a negative value a
High-Low or a Low-High outlier (Anselin 1995).

### LISA cluster map

```{r}
signif <- 0.05

lisa <- lisa |>
  mutate(LISA_cluster = ifelse(p_ii_sim < signif, as.character(mean), "Insignificant"),
         LISA_cluster = factor(LISA_cluster,
                               levels = c("Insignificant", "Low-Low", "Low-High",
                                          "High-Low", "High-High")))
table(lisa$LISA_cluster)
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses three functions:

```r
ifelse(test, yes, no)                                     # base R
as.character(x)                                           # base R
factor(x, levels)                                         # base R
```

- `signif <- 0.05` stores the significance threshold once. The hot and cold spot classification in
  section 4.2.3 uses it again.
- `ifelse()` works element by element: for each county it returns `yes` where `test` is true and
  `no` where it is false. A significant county takes its quadrant label from the `mean` column, and
  the others are labelled "Insignificant".
- `as.character()` converts the factor `mean` to its text labels. Without it, `ifelse()` would
  return the factor's integer codes.
- The second line of `mutate()` (section 4.1.1) overwrites the column created on the first.
  `factor()` converts the labels to a factor, and `levels` fixes the order of its five classes. The
  paragraph after the map gives the reason. `table()` counts the counties in each class.

A factor stores a set of labels as numbers, together with an ordered list of the labels, called its
levels. The example follows four counties through the second line of `mutate()`:

<div style="display: flex; align-items: center; gap: 1em; flex-wrap: wrap;">
<div>After <code>ifelse()</code>: text<br><table class="table table-sm" style="width: auto; margin: 0;"><tr><th>County</th><th>LISA_cluster</th></tr><tr><td>Anxiang</td><td>Insignificant</td></tr><tr><td>Yuanling</td><td>Low-Low</td></tr><tr><td>Pingjiang</td><td>Low-High</td></tr><tr><td>Liuyang</td><td>High-High</td></tr></table></div>
<div style="font-size: 1.8em;">&rarr;</div>
<div>After <code>factor(levels = ...)</code><br><table class="table table-sm" style="width: auto; margin: 0;"><tr><th>County</th><th style="background: rgba(255, 193, 7, 0.30)">stored as</th><th>shown as</th></tr><tr><td>Anxiang</td><td style="background: rgba(255, 193, 7, 0.30)">1</td><td>Insignificant</td></tr><tr><td>Yuanling</td><td style="background: rgba(255, 193, 7, 0.30)">2</td><td>Low-Low</td></tr><tr><td>Pingjiang</td><td style="background: rgba(255, 193, 7, 0.30)">3</td><td>Low-High</td></tr><tr><td>Liuyang</td><td style="background: rgba(255, 193, 7, 0.30)">5</td><td>High-High</td></tr></table></div>
<div>Its levels, in order<br><table class="table table-sm" style="width: auto; margin: 0;"><tr><th style="background: rgba(255, 193, 7, 0.30)">level</th><th>label</th></tr><tr><td style="background: rgba(255, 193, 7, 0.30)">1</td><td>Insignificant</td></tr><tr><td style="background: rgba(255, 193, 7, 0.30)">2</td><td>Low-Low</td></tr><tr><td style="background: rgba(255, 193, 7, 0.30)">3</td><td>Low-High</td></tr><tr><td style="background: rgba(255, 193, 7, 0.30)">4</td><td>High-Low</td></tr><tr><td style="background: rgba(255, 193, 7, 0.30)">5</td><td>High-High</td></tr></table></div>
</div>

Each county is stored as the number of its level, and the label is shown in its place. The order of
the levels decides the order of the classes in `table()`, in the legend and in the colours of the
map. High-Low is level 4 although no county has it, so it keeps its place in that order.
:::

```{r}
#| fig-height: 10.13
tm_shape(lisa) +
  tm_polygons(fill = "LISA_cluster",
              fill.scale = tm_scale_categorical(
                values = c("grey80", "blue", "lightblue", "pink", "red")),
              fill.legend = tm_legend(title = "LISA cluster",
                                      position = tm_pos_out("right", "center"))) +
  tm_borders() +
  tm_title("Local Moran's I clusters of GDPPC, p < 0.05")
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses one new function:

```r
tm_scale_categorical(values = NA, levels.drop = FALSE)    # tmap
```

- `tm_scale_categorical()` gives each category of a factor its own colour, in the order of the
  factor's levels. The vector of colours has to be written in the same order as the `levels`
  argument of section 4.1.5's `factor()` call, and the levels are set explicitly for that reason.
- `levels.drop = FALSE`, the default, keeps a level in the legend even when no county has it, so the
  colours stay fixed.

The colours are matched to the levels by position. With the levels set in section 4.1.5, each
class gets its intended colour. Without them, R sorts the classes alphabetically and drops High-Low,
which no county has, so the same five colours are matched to four levels in a different order:

<div style="display: flex; align-items: flex-start; gap: 1.5em; flex-wrap: wrap;">
<div>Levels set in section 4.1.5<br><table class="table table-sm" style="width: auto; margin: 0;"><tr><th>Level, in order</th><th>Counties</th><th>Colour given</th></tr><tr><td>Insignificant</td><td>72</td><td><span style="display:inline-block;width:1em;height:1em;vertical-align:middle;border:1px solid #999;background:#CCCCCC"></span> grey80</td></tr><tr><td>Low-Low</td><td>9</td><td><span style="display:inline-block;width:1em;height:1em;vertical-align:middle;border:1px solid #999;background:#0000FF"></span> blue</td></tr><tr><td>Low-High</td><td>1</td><td><span style="display:inline-block;width:1em;height:1em;vertical-align:middle;border:1px solid #999;background:#ADD8E6"></span> lightblue</td></tr><tr><td>High-Low</td><td>0</td><td><span style="display:inline-block;width:1em;height:1em;vertical-align:middle;border:1px solid #999;background:#FFC0CB"></span> pink</td></tr><tr><td>High-High</td><td>6</td><td><span style="display:inline-block;width:1em;height:1em;vertical-align:middle;border:1px solid #999;background:#FF0000"></span> red</td></tr></table></div>
<div>Levels left to R, alphabetical<br><table class="table table-sm" style="width: auto; margin: 0;"><tr><th>Level, in order</th><th>Counties</th><th>Colour given</th></tr><tr><td>High-High</td><td>6</td><td><span style="display:inline-block;width:1em;height:1em;vertical-align:middle;border:1px solid #999;background:#CCCCCC"></span> grey80</td></tr><tr><td>Insignificant</td><td>72</td><td><span style="display:inline-block;width:1em;height:1em;vertical-align:middle;border:1px solid #999;background:#0000FF"></span> blue</td></tr><tr><td>Low-High</td><td>1</td><td><span style="display:inline-block;width:1em;height:1em;vertical-align:middle;border:1px solid #999;background:#ADD8E6"></span> lightblue</td></tr><tr><td>Low-Low</td><td>9</td><td><span style="display:inline-block;width:1em;height:1em;vertical-align:middle;border:1px solid #999;background:#FFC0CB"></span> pink</td></tr></table></div>
</div>

In the second case the six High-High counties would be drawn grey and the 72 insignificant counties
blue.
:::

Chapter 10 asks what statistical observations the LISA map supports:

- **Significant counties:** 16 of the 88, of which six are High-High, nine Low-Low and one
  Low-High. The other 72 are not significant at p < 0.05.
- **High-High cluster:** around Changsha in the north east.
- **Low-Low cluster:** in the west.
- **Spatial outlier:** Pingjiang, the single Low-High county, which lies between the two groups and
  has low GDPPC with high-GDPPC neighbours.

These are uncorrected results. Section 5.3 shows how many remain significant after a correction
for 88 tests.

## Hot and cold spots

Local Moran's I and Getis-Ord Gi\* measure different concepts of spatial association.

| Statistic | What it measures | Significant positive value | Significant negative value |
|---|---|---|---|
| local Moran's I | similarity between a county and its neighbours | the county is similar to its neighbours, High-High or Low-Low | a spatial outlier, High-Low or Low-High |
| Gi\* | whether the sum of values in the neighbourhood is larger or smaller than expected under spatial randomness | clustering of high values, a hot spot | clustering of low values, a cold spot |

Gi\* has no outlier category.

Getis and Ord (1992) state that the variable "has a natural origin and is positive", because G is
computed from the values themselves and not from their deviations from the mean, so a zero that does
not represent absence of the quantity changes the statistic (section 5.4.1). GDPPC is positive in
every county.

### Distance weights

Getis and Ord (1992) define Gi\* on a distance band: it measures the concentration of values within
a given distance of each county. Slide 36 of the Lesson 5 lecture requires the band to be wide
enough that every county has at least one neighbour.

```{r}
ct <- critical_threshold(st_geometry(hunan))
ct
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses two functions:

```r
st_geometry(x)                                            # sf
critical_threshold(geometry, k = 1)                       # sfdep
```

- `st_geometry()` extracts the geometry column of an sf object, without the attributes.
- `critical_threshold()` finds each county's distance to its nearest neighbour (`k = 1`) and
  returns the largest of these distances. A distance band of that width gives every county at least
  one neighbour. The distance is in metres, because `hunan` is in EPSG:32649.

`st_geometry(hunan)` keeps only the shaded column, the polygons:

<div style="display: flex; align-items: center; gap: 1em; flex-wrap: wrap;">
<div><code>hunan</code>, an sf data frame<br><table class="table table-sm" style="width: auto; margin: 0;"><tr><th>NAME_2</th><th>ID_3</th><th>NAME_3</th><th>ENGTYPE_3</th><th>County</th><th>GDPPC</th><th style="background: rgba(255, 193, 7, 0.30)">geometry</th></tr><tr><td>Changde</td><td>21098</td><td>Anxiang</td><td>County</td><td>Anxiang</td><td>23667</td><td style="background: rgba(255, 193, 7, 0.30)">POLYGON ((602726.4 3292136, …))</td></tr><tr><td>Changde</td><td>21100</td><td>Hanshou</td><td>County</td><td>Hanshou</td><td>20981</td><td style="background: rgba(255, 193, 7, 0.30)">POLYGON ((619553.5 3221554, …))</td></tr><tr><td>…</td><td>…</td><td>…</td><td>…</td><td>…</td><td>…</td><td style="background: rgba(255, 193, 7, 0.30)">…</td></tr></table></div>
<div style="font-size: 1.8em;">&rarr;</div>
<div><code>st_geometry(hunan)</code>, an <code>sfc_POLYGON</code><br><table class="table table-sm" style="width: auto; margin: 0;"><tr><th style="background: rgba(255, 193, 7, 0.30)">geometry</th></tr><tr><td style="background: rgba(255, 193, 7, 0.30)">POLYGON ((602726.4 3292136, …))</td></tr><tr><td style="background: rgba(255, 193, 7, 0.30)">POLYGON ((619553.5 3221554, …))</td></tr><tr><td style="background: rgba(255, 193, 7, 0.30)">…</td></tr></table></div>
</div>

The result has no attributes, so `critical_threshold()` receives only the 88 shapes it measures.

The figure draws, for every county, a line to its nearest neighbour. The longest of these lines,
in red, is the value `critical_threshold()` returns.

```{r}
#| echo: false
#| fig-height: 5
ct_pts <- st_point_on_surface(st_geometry(hunan))
ct_nn <- knearneigh(ct_pts, k = 1)$nn[, 1]
ct_xy <- st_coordinates(ct_pts)
ct_links <- tibble(x = ct_xy[, 1], y = ct_xy[, 2],
                   xend = ct_xy[ct_nn, 1], yend = ct_xy[ct_nn, 2]) |>
  mutate(d = sqrt((xend - x)^2 + (yend - y)^2))
far <- which.max(ct_links$d)
ct_mid <- c(mean(unlist(ct_links[far, c("x", "xend")])), mean(unlist(ct_links[far, c("y", "yend")])))
ct_top <- st_bbox(hunan)[["ymax"]] + 25000

ct_map <- ggplot() +
  geom_sf(data = hunan, fill = "grey92", colour = "white", linewidth = 0.3) +
  geom_segment(data = ct_links, aes(x, y, xend = xend, yend = yend),
               colour = "grey45", linewidth = 0.3) +
  geom_point(data = ct_links, aes(x, y), size = 0.8) +
  geom_segment(data = ct_links[far, ], aes(x, y, xend = xend, yend = yend),
               colour = "red", linewidth = 1,
               arrow = arrow(ends = "both", length = unit(0.15, "cm"))) +
  annotate("segment", x = ct_mid[1], y = ct_mid[2], xend = ct_mid[1], yend = ct_top,
           colour = "red", linewidth = 0.3) +
  annotate("label", x = ct_mid[1], y = ct_top, vjust = 0, size = 3.2, colour = "red",
           fill = "white", label.size = 0,
           label = paste0("critical_threshold(): the largest nearest-neighbour distance\n",
                          hunan$NAME_3[far], " to ", hunan$NAME_3[ct_nn[far]], ", ct = ",
                          format(round(ct, 2), big.mark = ","), " m")) +
  coord_sf(datum = NA, ylim = c(st_bbox(hunan)["ymin"], ct_top + 90000)) +
  labs(x = NULL, y = NULL) +
  theme_void()

ct_map
```
:::

`critical_threshold()` returns the smallest distance at which every county has at least one
neighbour, which is the rule on slide 36. Under EPSG:32649 the threshold is 62,955 m. Chapter 10
reports 60,800 m, the value of the same function under EPSG:32650. Section 5.1 examines the
difference.

```{r}
hunan_fdw <- hunan |>
  mutate(nb = include_self(st_dist_band(geometry, upper = ct)),
         wt = st_weights(nb, style = "W"),
         .before = 1)

summary(lengths(hunan_fdw$nb))
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses three functions:

```r
st_dist_band(geometry, lower = 0,
             upper = critical_threshold(geometry))        # sfdep
include_self(nb)                                          # sfdep
lengths(x)                                                # base R
```

- `st_dist_band()` makes a county the neighbour of every county whose distance from it lies between
  `lower` and `upper`, here up to `ct` metres, since `hunan` is in EPSG:32649. For polygons, `sfdep`
  measures the distance between points on the surface of each county, from `st_point_on_surface()`,
  and `critical_threshold()` uses the same points.
- `upper` defaults to `critical_threshold(geometry)`, so passing `ct` gives the same band and shows
  the value used.
- `include_self()` adds each county to its own neighbour list, which turns Gi into Gi\*.
  `st_weights()` then row-standardises the weights as in section 4.1.1.
- `lengths()` returns the length of each element of a list, here the number of neighbours of each
  county, and `summary()` gives their minimum, quartiles, mean and maximum.

The figure applies the band to Taoyuan, the county that sets `ct`. The circle has radius `ct`
around Taoyuan's point, and a county is a neighbour when its point lies inside the circle.

```{r}
#| echo: false
#| fig-height: 5
ty <- which(hunan$NAME_3 == "Taoyuan")
ty_nb <- hunan_fdw$nb[[ty]]
ty_circle <- st_buffer(ct_pts[ty], ct)
band_focus <- hunan |>
  mutate(role = case_when(row_number() == ty ~ "Taoyuan",
                          row_number() %in% ty_nb ~ "neighbour inside the band",
                          TRUE ~ "other"),
         role = factor(role, levels = c("Taoyuan", "neighbour inside the band", "other")))

ty_xy <- st_coordinates(ct_pts[ty])
band_top <- st_bbox(hunan)[["ymax"]] + 25000

ggplot() +
  geom_sf(data = band_focus, aes(fill = role), colour = "white", linewidth = 0.3) +
  geom_sf(data = ty_circle, fill = NA, colour = "red", linewidth = 0.8) +
  geom_sf(data = ct_pts, size = 0.8) +
  annotate("segment", x = ty_xy[1], y = ty_xy[2], xend = ty_xy[1], yend = band_top,
           colour = "red", linewidth = 0.3) +
  annotate("label", x = ty_xy[1], y = band_top, vjust = 0, size = 3.2, colour = "red",
           fill = "white",
           label = paste0("st_dist_band(upper = ct) + include_self()\n",
                          "circle of radius ct around Taoyuan\n",
                          "hunan_fdw$nb[[", ty, "]] = ", paste(ty_nb, collapse = ", "), ": ",
                          paste(hunan$NAME_3[ty_nb], collapse = " and "))) +
  scale_fill_manual(values = c("Taoyuan" = "darkorange",
                               "neighbour inside the band" = "#fdd49e", "other" = "grey92")) +
  coord_sf(datum = NA, ylim = c(st_bbox(hunan)["ymin"], band_top + 110000)) +
  labs(fill = NULL) +
  theme_void() +
  theme(legend.position = "bottom")
```

Taoyuan's only neighbour is Hanshou, whose point lies exactly on the circle, and `include_self()`
adds Taoyuan itself. This is the sparsest neighbourhood in the province.
:::

`include_self()` distinguishes Gi\* from Gi. Gi excludes the county from its own neighbourhood, and
Gi\* includes it. Slide 27 gives both forms, and the chapter uses Gi\* throughout.

```{r}
hunan_adw <- hunan |>
  mutate(nb = include_self(st_knn(geometry, k = 6)),
         wt = st_weights(nb, style = "W"),
         .before = 1)

summary(lengths(hunan_adw$nb))
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses one new function:

```r
st_knn(geometry, k = 1, symmetric = FALSE)                # sfdep
```

- `st_knn()` makes each county's `k` nearest counties its neighbours, measured between points on
  the surface of each county, as for `st_dist_band()` above. `k = 6` gives each county six.
- With `symmetric = FALSE`, the default, the list need not be symmetric: county A can be among
  B's six nearest without B being among A's.
- `include_self()` and `st_weights()` then work as in the fixed-band block above.
:::

An adaptive scheme fixes the number of neighbours instead of the distance. Neighbourhoods are then
smaller where counties are small and larger where they are large. Slide 31 gives the case for
adaptive weights. Slides 30 and 35 give the basis for `k = 6`: both suggest about eight neighbours
per feature for a skewed variable, and slide 35 calls this a rule of thumb. `k = 6` plus the county
itself gives seven.

Chapter 10 creates this object but does not use it. Section 5.7 computes Gi\* on both sets of
weights.

### Local Gi\*

Gi\* is reported as a z-score. Under a standard normal distribution, an absolute value above 1.96
corresponds to p < 0.05 in a two-sided test.

```{r}
set.seed(1234)
HCSA_fdw <- hunan_fdw |>
  mutate(gistar = local_gstar_perm(GDPPC, nb, wt, nsim = 999),
         .before = 1) |>
  unnest(gistar)
glimpse(HCSA_fdw)
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses one function:

```r
local_gstar_perm(x, nb, wt, nsim = 499, alternative = "two.sided")   # sfdep
```

- `local_gstar_perm()` computes Gi\* for each county and tests it by conditional permutation. It
  follows the pattern of `local_moran()` in section 4.1.2: it returns a data frame of results, which
  is stored as the column `gistar` and spread into columns by `unnest()`. `nsim` defaults to 499.
- `unnest()` spreads the 10 results into 10 columns: `gi_star`, `cluster`, `e_gi`, `var_gi`,
  `std_dev`, `p_value`, `p_sim`, `p_folded_sim`, `skewness` and `kurtosis`. Of these, `gi_star` is
  the statistic as a z-score, and `p_sim` the rank-based p-value from the 999 permutations.
:::

```{r}
#| fig-height: 4.1
gi.map <- tm_shape(HCSA_fdw) +
  tm_polygons(fill = "gi_star",
              fill.scale = tm_scale_intervals(style = "pretty", n = 6,
                                              values = "-brewer.rd_bu"),
              fill.legend = tm_legend(title = "Gi*",
                                      position = tm_pos_out("right", "center"))) +
  tm_borders(col_alpha = 0.5) +
  tm_title("Local Gi* of GDPPC")

p.map <- tm_shape(HCSA_fdw) +
  tm_polygons(fill = "p_sim",
              fill.scale = tm_scale_intervals(
                                breaks = c(0, 0.01, 0.05, 0.1, 1),
                values = "-brewer.reds"),
              fill.legend = tm_legend(title = "p-value",
                                      position = tm_pos_out("right", "center"))) +
  tm_borders(col_alpha = 0.5) +
  tm_title("Simulated p-value of Gi*")

tmap_arrange(gi.map, p.map, asp = 1, ncol = 2)
```

::: {.callout-note collapse="false" title="How the code works"}
The block sets the colours with arguments of `tm_scale_intervals()`:

```r
tm_scale_intervals(n = 5, style = "pretty", breaks = NULL, values = NA)   # tmap
```

- `values` names the palette. `"brewer.rd_bu"` is the ColorBrewer "RdBu" palette, which runs from
  red to blue, and `"brewer.reds"` runs from light to dark red.
- A leading minus reverses a palette. In the Gi\* map, `"-brewer.rd_bu"` puts blue at the low end
  and red at the high end, so that negative Gi\* (cold spots) is blue and positive Gi\* (hot spots)
  is red:

<div style="margin: 0.3em 0 0.8em 1.5em;">
<code>&quot;brewer.rd_bu&quot;</code>, low to high: <span style="display:inline-block;width:2.2em;height:1.1em;background:#B2182B;border:1px solid #ccc"></span><span style="display:inline-block;width:2.2em;height:1.1em;background:#EF8A62;border:1px solid #ccc"></span><span style="display:inline-block;width:2.2em;height:1.1em;background:#FDDBC7;border:1px solid #ccc"></span><span style="display:inline-block;width:2.2em;height:1.1em;background:#D1E5F0;border:1px solid #ccc"></span><span style="display:inline-block;width:2.2em;height:1.1em;background:#67A9CF;border:1px solid #ccc"></span><span style="display:inline-block;width:2.2em;height:1.1em;background:#2166AC;border:1px solid #ccc"></span><br>
<code>&quot;-brewer.rd_bu&quot;</code>, low to high: <span style="display:inline-block;width:2.2em;height:1.1em;background:#2166AC;border:1px solid #ccc"></span><span style="display:inline-block;width:2.2em;height:1.1em;background:#67A9CF;border:1px solid #ccc"></span><span style="display:inline-block;width:2.2em;height:1.1em;background:#D1E5F0;border:1px solid #ccc"></span><span style="display:inline-block;width:2.2em;height:1.1em;background:#FDDBC7;border:1px solid #ccc"></span><span style="display:inline-block;width:2.2em;height:1.1em;background:#EF8A62;border:1px solid #ccc"></span><span style="display:inline-block;width:2.2em;height:1.1em;background:#B2182B;border:1px solid #ccc"></span>
</div>

- `style = "pretty"` with `n = 6` chooses about six classes with rounded break points.
- In the p-value map, `breaks = c(0, 0.01, 0.05, 0.1, 1)` fixes the class boundaries at the usual
  significance levels, and `"-brewer.reds"` makes the smallest p-values the darkest red.
:::

### Hot and cold spot map

```{r}
HCSA_fdw <- HCSA_fdw |>
  mutate(HCSA_cluster = case_when(
           p_sim >= signif ~ "Insignificant",
           gi_star > 0 ~ "Hot spot",
           TRUE ~ "Cold spot"),
         HCSA_cluster = factor(HCSA_cluster,
                               levels = c("Insignificant", "Hot spot", "Cold spot")))
table(HCSA_fdw$HCSA_cluster)
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses two functions:

```r
case_when(condition ~ value, ...)                         # dplyr
factor(x, levels)                                         # base R
```

- `case_when()` tests its conditions in order and, for each county, returns the value of the first
  condition that holds. A county with `p_sim >= signif` is labelled "Insignificant" whatever its
  Gi\*, and `TRUE ~ "Cold spot"` labels every county not yet labelled, which is significant with a
  Gi\* of zero or below. With three outcomes, `ifelse()` (section 4.1.5) would need one call inside
  another.
- As in section 4.1.5, `factor()` fixes the order of the classes, and with it the order of the
  colours in the map.
:::

```{r}
#| fig-height: 10.13
tm_shape(HCSA_fdw) +
  tm_polygons(fill = "HCSA_cluster",
              fill.scale = tm_scale_categorical(
                values = c("grey80", "red", "blue")),
              fill.legend = tm_legend(title = "Gi* cluster",
                                      position = tm_pos_out("right", "center"))) +
  tm_borders() +
  tm_title("Hot and cold spots of GDPPC, fixed distance band, p < 0.05")
```

Chapter 10 asks what statistical observations the hot and cold spot map supports:

- **Significant counties:** 12 of the 88, seven hot spots and five cold spots. The other 76 are not
  significant at p < 0.05.
- **Hot spot:** Changsha and the surrounding counties.
- **Cold spot:** in the west.
- **Comparison with the LISA map of section 4.1.5:** both place the division between high and low
  GDPPC in the same location, and differ in how far the clusters on each side of it extend. In the
  west, where the counties are fewer and larger, the fixed distance band on which Gi\* is computed
  reaches further than contiguity.

# Part 3: Review and extension

Parts 1 and 2 reproduced chapters 9 and 10 of r4gdsa. This part examines their method, and adds the
parts of the Lesson 5 lecture and readings that the two chapters do not cover.

| Section | What I do | Source |
|----|--------------------------------------|------------|
| | **Checking chapter 10's method** | |
| 5.1 | Measure distances in the correct UTM zone | UTM zone boundaries |
| 5.2 | Choose the local p-value and the number of simulations | Ord and Getis (1995) |
| 5.3 | Correct the LISA map for 88 simultaneous tests | Ord and Getis (1995), Anselin (1995), Getis (2010) |
| | **Adding what the chapters leave out** | |
| 5.4 | Test whether high or low values drive the clustering | slides 16 and 17 |
| 5.4.1 | Test how G responds to adding a constant to GDPPC | Getis and Ord (1992) |
| 5.5 | Compare the null models for Moran's I | Moran (1950), Geary (1954) |
| 5.6 | Compare local Geary's c with local Moran's I | slide 26 |
| 5.7 | Use the adaptive weights that chapter 10 builds but does not use | slides 30 and 35 |
| 5.7.1 | Vary `k` to find the hot and cold spots that do not depend on it | Mack and Kam (2018) |
| 5.8 | Test Anselin's claim that Gi, Gi\* and local Moran's I share one inference | Anselin (1995) |

## Projected CRS: zone 49N versus zone 50N

Section 2.2.1 projected the Hunan layer to EPSG:32649, and section 4.2.1 computed from it a fixed
bandwidth of 62,955 m, against the 60,800 m reported in chapter 10. Both values are computed by the
same function on the same polygons. The difference arises from the coordinate system in which the
distances are measured.

In the Universal Transverse Mercator (UTM) system:

- the world is divided into sixty zones, each six degrees of longitude wide;
- each zone is projected onto its own plane with its own central meridian;
- coordinates in a zone are metres on that plane, and a distance computed from them is measured on
  that plane and not on the ellipsoid;
- distortion is smallest near the central meridian and increases away from it.

The map below shows the zones over East Asia, with Hunan in orange.

```{r}
#| code-fold: true
#| fig-height: 5.5
s2_was <- sf_use_s2(FALSE)                      # crop in plain longitude and latitude
world <- spData::world |> st_crop(xmin = 80, xmax = 150, ymin = 0, ymax = 60)
sf_use_s2(s2_was)

bounds <- seq(84, 144, by = 6)                  # zone boundaries, 6 degrees apart
zone_lines <- st_sfc(lapply(bounds, function(x) st_linestring(cbind(x, seq(0, 56, 0.5)))),
                     crs = 4326)
zone_labels <- st_sf(label = paste0(45:54, "N"),
                     geometry = st_sfc(lapply(bounds[-length(bounds)] + 3,
                                              function(x) st_point(c(x, 4))), crs = 4326))

ggplot() +
  geom_sf(data = world, fill = "grey90", colour = "grey60", linewidth = 0.2) +
  geom_sf(data = st_transform(hunan, 4326),
          fill = "darkorange", colour = NA) +
  geom_sf(data = zone_lines, colour = "grey20", linewidth = 0.4) +
  geom_sf_text(data = zone_labels, aes(label = label), size = 3.2) +
  coord_sf(crs = "+proj=lcc +lat_1=20 +lat_2=45 +lon_0=114 +datum=WGS84",
           xlim = c(-2600000, 2600000), ylim = c(-500000, 4300000)) +
  labs(x = NULL, y = NULL) +
  theme_minimal()
```

Hence the zone assigned to a province determines the meridian relative to which its distances are
measured. Chapter 10 projects the data to EPSG:32650 (zone 50N), and chapter 9 does not project
them. Neither chapter discusses the choice.

This section addresses three questions:

1. Which zone contains Hunan?
2. How much does the fixed bandwidth differ between the two zones?
3. Does the hot and cold spot map of section 4.2.3 change as a result?

```{r}
hunan_raw <- st_read(dsn = "data/geospatial", layer = "Hunan", quiet = TRUE)
```

The map below marks the UTM zones on the unprojected layer. Dashed lines are zone boundaries, and
solid lines are the central meridians.

```{r}
#| code-fold: true
#| fig-height: 4.5
lines_df <- tibble(lon = c(108, 114, 120, 111, 117),
                   type = c(rep("zone boundary", 3), rep("central meridian", 2)))

ggplot() +
  geom_sf(data = hunan_raw, fill = "grey85", colour = "white") +
  geom_vline(data = lines_df, aes(xintercept = lon, linetype = type)) +
  annotate("label", x = c(111, 117), y = 30.55, label = c("zone 49N", "zone 50N"),
           label.size = 0, fill = "white") +
  scale_linetype_manual(values = c("central meridian" = "solid", "zone boundary" = "dashed")) +
  coord_sf(xlim = c(107.5, 120.5), ylim = c(24.5, 30.8), expand = FALSE) +
  labs(x = NULL, y = NULL, linetype = NULL) +
  theme_minimal()
```

Hunan extends from 108.78°E to 114.25°E. UTM zone 49N covers 108°E to 114°E, and zone 50N covers
114°E to 120°E. All of the province except a narrow strip east of 114°E lies in zone 49N. Its
centre, at 111.5°E, is 2.5° west of the western boundary of zone 50N and 5.5° from the central
meridian of that zone.

Projection to zone 50N thus measures distances on the grid of the wrong zone. These distances are
the input to the fixed distance band.

```{r}
ct49 <- critical_threshold(st_geometry(st_transform(hunan_raw, 32649)))
ct50 <- critical_threshold(st_geometry(st_transform(hunan_raw, 32650)))
tibble(zone = c("49N (EPSG:32649)", "50N (EPSG:32650)"),
       critical_threshold_m = c(ct49, ct50)) |>
  mutate(difference_pct = 100 * (critical_threshold_m - ct49) / ct49) |>
  kable(digits = 2)
```

In zone 50N the computation reproduces the chapter's value of 60,799.91 m exactly, confirming its
source. In the correct zone the bandwidth is 62,954.89 m, 3.4% wider.

**Scale distortion does not explain the difference.** Switching from zone 49N to zone 50N stretches
distances across Hunan by only about 0.3%, because the scale factor of UTM increases with distance
from the central meridian. That would make the zone 50N threshold slightly larger, but it is 3.4%
smaller.

Distances are least distorted along a zone's central meridian and are stretched more the further a
county lies from it. The zone map above shows Hunan around the central meridian of zone 49N, and 3
to 8 degrees west of the central meridian of zone 50N.

The sketch below shows why. A UTM zone is drawn by wrapping a sheet around the globe so that it
touches along the central meridian, and projecting each point outward onto the sheet. Where the
sheet touches the globe, a distance is copied almost exactly. Away from that line the globe curves
away from the sheet, and the same distance lands on a longer piece of the sheet. The sketch is
simplified and its angles are exaggerated. For Hunan in zone 50N the stretch is about 0.3%.

```{r}
#| code-fold: true
#| fig-height: 3.6
deg <- pi / 180
sketch_parts <- function(zone, offset) {
  arc <- seq(offset - 12, offset + 12, by = 1) * deg      # Hunan, 24 degrees wide
  ends <- tan(range(arc))                                   # where Hunan lands on the sheet
  list(globe = tibble(zone, a = seq(-90, 90, 1) * deg, x = sin(a), y = cos(a)),
       hunan = tibble(zone, x = sin(arc), y = cos(arc)),
       image = tibble(zone, x = ends, y = 1),
       rays  = tibble(zone, x = 0, y = 0, xend = ends, yend = 1))
}
zones <- c("Zone 49N: sheet touches inside Hunan", "Zone 50N: sheet touches east of Hunan")
parts <- map2(zones, c(0, -40), sketch_parts) |> transpose() |> map(bind_rows)

ggplot() +
  annotate("segment", x = -1.6, xend = 1.6, y = 1, yend = 1, colour = "grey40") +
  geom_path(data = parts$globe, aes(x, y), colour = "grey60") +
  geom_segment(data = parts$rays, aes(x, y, xend = xend, yend = yend),
               linetype = "dotted", colour = "grey50") +
  geom_path(data = parts$hunan, aes(x, y), colour = "darkorange", linewidth = 2) +
  geom_line(data = parts$image, aes(x, y), colour = "darkorange", linewidth = 2) +
  annotate("point", x = 0, y = 1, size = 2.5) +
  facet_wrap(~ zone) +
  coord_equal(xlim = c(-1.6, 1.6), ylim = c(-0.05, 1.1)) +
  theme_void() +
  theme(strip.text = element_text(size = 11, margin = margin(b = 6)))
```

In each panel the grey arc is the globe, the straight line is the sheet, and the black dot is where
the sheet touches, on the central meridian. The dotted lines project Hunan outward from the centre
of the globe. The orange arc is Hunan on the globe, and the orange bar is Hunan on the map. In zone
49N the bar is almost as long as the arc. In zone 50N it is longer.

**The point inside each county moves.** `critical_threshold()` returns the largest distance from a
county to its nearest neighbour, measured between one point inside each county. In both zones that
distance is between Taoyuan and Hanshou. `sfdep` computes the points with `st_point_on_surface()`
after projection, and the point chosen depends on the projection: Taoyuan's point in zone 50N lies
about 3.3 km from its point in zone 49N. Moving the zone 49N points into zone 50N, without choosing
them again, gives 63,160 m, 0.3% longer, as the scale factor predicts.

Whether the difference affects the results is a separate question from which zone is correct. Below,
the effect on the hot and cold spot classification is tested.

```{r}
h50 <- st_transform(hunan_raw, 32650) |>
  left_join(hunan2012, by = "County") |>
  select(1:4, 7, 15)

fdw50 <- h50 |>
  mutate(nb = include_self(st_dist_band(geometry, upper = ct50)),
         wt = st_weights(nb, style = "W"), .before = 1)

set.seed(1234)
g50 <- local_gstar_perm(fdw50$GDPPC, fdw50$nb, fdw50$wt, nsim = 999)

cls <- function(gi, p) factor(ifelse(p >= 0.05, "Insignificant",
                                     ifelse(gi > 0, "Hot spot", "Cold spot")),
                              levels = c("Insignificant", "Hot spot", "Cold spot"))

table(zone_49N = HCSA_fdw$HCSA_cluster,
      zone_50N = cls(g50$gi_star, g50$p_sim))
```

::: {.callout-note collapse="false" title="How the code works"}
The block repeats earlier steps in zone 50N and defines one function:

```r
cls <- function(gi, p) ...                                # defined here
```

- The first three steps repeat, for zone 50N, the join of section 2.3, the fixed band of section
  4.2.1 and the Gi\* test of section 4.2.2.
- `function(gi, p)` defines a new function, `cls()`, with two inputs: the Gi\* values and their
  p-values. It applies the classification of section 4.2.3 in one line, so that the two zones are
  classified in exactly the same way.
- `table()` then cross-tabulates the two classifications, county by county.
:::

The off-diagonal cells of the table count the counties that the two zones classify differently.
Four of the 88 counties change class: two hot spots in zone 49N are insignificant in zone 50N, and
two insignificant counties in zone 49N become a hot spot and a cold spot. The remaining 84 are
classified the same in both zones.

## Local p-values: normal approximation versus rank

`local_moran()` returns two p-values for each county, and they can disagree:

- `p_ii` standardises the observed statistic by the mean and standard deviation of its permutation
  distribution and passes the result through `pnorm()`. It is a normal approximation to that
  distribution.
- `p_ii_sim` is based on the rank of the observed statistic among the permuted values, and requires
  no assumption about the shape of the distribution.

Chapter 10 classifies counties on `p_ii`. This page uses `p_ii_sim`, from the LISA map of section
4.1.3 onwards, and this section explains the choice.

Both columns come from the same simulation, so the chapter's `p_ii` also uses the permutations it
runs. Chapter 10 calls `local_moran(GDPPC, nb, wt, nsim = 99)`, which runs 99 permutations for each
of the 88 counties, 8,712 in all. Each permutation is conditional: the value of the county is held
fixed and the remaining 87 values are randomly permuted over the other 87 counties.

Two questions decide between the two columns, and this section examines them in turn:

1. **Is the permutation distribution close enough to normal for `p_ii` to be valid?** The
   simulations of Ord and Getis (1995) show when it is not, and `local_moran()` reports the
   skewness of each county's permutation distribution as a direct check.
2. **Can 99 permutations produce a p-value small enough to pass a corrected threshold?** With 88
   counties, the Bonferroni threshold of section 5.3 is 0.05/88. The worked example in
   Anselin (1995) applies the same kind of correction with 9,999 permutations (section 5.3.2).

The section ends with `p_folded_sim`, a third p-value column, and the reason this page does not use
it.

Ord and Getis (1995) is the assigned reading on the conditions under which the normal approximation
fails. They examine the question by simulation, with five thousand random permutations on a 10 by 10
lattice under four distributions.

Their simulations concern Gi\* and Gi and not the local Moran. In their section 4 they show that the
global Moran's I is a weighted average of the local G statistics, and this page takes that relation
as the basis for applying their results to the local Moran.

They report the following patterns:

> i. When the underlying distribution is normal, so is that of the test statistics (an exact
>    result);
> ii. when the underlying distribution is markedly skew, the distribution of the test statistics is
>    non-normal, but approaches normality as the distance is increased.
> iii. the statistics for edge cells approach normality more slowly because they have fewer
>    neighbors; the convergence for corner cells is still slower.

Patterns (ii) and (iii) both apply to these data. GDPPC is right skewed, as in pattern (ii), and the
counties with the fewest neighbours correspond to the edge cells of pattern (iii). `local_moran()`
reports the skewness and kurtosis of the permutation distribution of each county, so the departure
from normality can be measured directly.

```{r}
tibble(measure = c("skewness", "kurtosis"),
       min = c(min(lisa$skewness), min(lisa$kurtosis)),
       median = c(median(lisa$skewness), median(lisa$kurtosis)),
       max = c(max(lisa$skewness), max(lisa$kurtosis))) |>
  kable(digits = 3)
```

A normal distribution has skewness 0 and excess kurtosis 0. Here the permutation distributions have
neither property, and `pnorm()` treats them as normal.

How many counties are significant at p < 0.05 under each p-value? The table has four rows:

- **`p_ii`, `nsim = 99`**: the chapter's choice, a normal approximation.
- **`p_ii_sim`, `nsim = 99`**: the same simulation as the chapter, so only the p-value column
  changes.
- **`p_ii_sim`, `nsim = 999`**: this page's choice, used for the LISA map of section 4.1.3.
- **`p_folded_sim`, `nsim = 999`**: a third column that `local_moran()` returns, examined at the end
  of this section.

```{r}
set.seed(1234)
lm99 <- local_moran(hunan$GDPPC, wm_q$nb, wm_q$wt, nsim = 99)

tibble(p_value = c("p_ii (normal approximation)", "p_ii_sim, nsim = 99",
                   "p_ii_sim, nsim = 999", "p_folded_sim, nsim = 999"),
       significant_at_0.05 = c(sum(lm99$p_ii < 0.05),
                               sum(lm99$p_ii_sim < 0.05),
                               sum(lisa$p_ii_sim < 0.05),
                               sum(lisa$p_folded_sim < 0.05))) |>
  kable()
```

Both the number of significant counties and their identity depend on whether `p_ii` or `p_ii_sim` is
used.

```{r}
table(chapter_p_ii = ifelse(lm99$p_ii < 0.05, as.character(lm99$mean), "Insignificant"),
      simulated = as.character(lisa$LISA_cluster))
```

Of the four groups, Low-Low changes most. With the normal-approximation p-value, three counties in
the west are classified as Low-Low, and with the simulated p-value, nine. The six additional
counties, found by the rank-based p-value from the simulation that the chapter itself runs, have low
GDPPC and neighbours with low GDPPC, the pattern the LISA map is intended to identify.

Chapter 10's `nsim = 99` also limits the smallest attainable p-value, the second condition listed
above. A conditional permutation p-value is `(number of simulations at least as extreme + 1) / (nsim
+ 1)`. Its smallest possible value is `1 / (nsim + 1)`.

For the two-sided `p_ii_sim` the smallest value is twice that. `local_moran()` also returns
`p_folded_sim`, the same ranking folded onto the range [0, 0.5], which is not doubled.

```{r}
tibble(nsim = c(99, 999, 9999),
       floor_p_ii_sim = 2 / (c(99, 999, 9999) + 1),
       floor_p_folded_sim = 1 / (c(99, 999, 9999) + 1)) |>
  kable(digits = 5)
```

Chapter 10 sets breaks for its Gi\* p-value maps at 0.001 and 0.01 with `nsim = 99`, whose minimum
is 0.02, so two of the five classes are empty for any data. This page uses `nsim = 999`, at which
the smallest attainable `p_ii_sim` is 0.002. Section 5.3 shows that 999 is still
insufficient once a correction for multiple comparisons is applied.

`p_folded_sim` is included in the table to show why this page does not use it. In these data it is
close to half of `p_ii_sim`: exactly half in 86 of the 88 rows, with a mean relative difference of
1.4%.

```{r}
all.equal(lisa$p_folded_sim, lisa$p_ii_sim / 2)
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses one function:

```r
all.equal(target, current)                                # base R
```

- `all.equal()` compares two sets of numbers. It returns `TRUE` if they agree up to rounding
  error, and otherwise describes how far they differ, here as a mean relative difference.
:::

The comparison of a quantity on [0, 0.5] against 0.05 is a one-sided test at the 5% level. It
accounts for the 23 significant counties under `p_folded_sim` against 16 under `p_ii_sim`. The
additional counties follow from the scale of the p-value alone and are not evidence that those
counties cluster. This page compares `p_ii_sim` against 0.05 as a two-sided test.

## Multiple comparisons across 88 counties

Section 4.1.5 classified 16 counties as significant at p < 0.05 and mapped them.

A LISA map involves one hypothesis test per county. Each test is evaluated at α = 0.05, the
probability of a false positive for a single test. With no spatial pattern present, 88 tests
evaluated at 0.05 are expected to produce about four significant counties. Section 3.5 met the
problem of multiple comparisons on a smaller scale: with six correlogram lags tested in one table,
lag 3 was no longer significant after Holm's correction.

A correction for multiple comparisons raises the evidence a test must reach, so that a county is
called significant only on stronger evidence than 0.05. Neither chapter applies one. Both assigned
papers on the local statistics state that a correction is needed: Anselin (1995), which introduced
the local Moran, and Ord and Getis (1995), on the distribution of Gi and Gi\*. They disagree over
how many tests the correction should count, and section 5.3.1 sets out the disagreement. Two
corrections are used here, both taken from the readings:

- **Bonferroni** divides α by the number of tests, so every county is evaluated at 0.05/88, the
  same fixed penalty for all. Ord and Getis (1995) recommend it, and Anselin (1995) applies it in
  his worked example.
- **Benjamini-Hochberg** is less severe. It controls the false discovery rate (FDR), the expected
  proportion of false positives among the rejections, and rejects the ith smallest p-value when p(i)
  ≤ i × 0.05 / 88, so its threshold increases with rank. Getis (2010) cites it, from Benjamini and
  Hochberg (1995), among the attempts to address multiple testing in local statistics.

How many of the 16 counties remain significant after each correction, and which conclusions does the
map then support?

```{r}
p <- lisa$p_ii_sim
tibble(correction = c("none", "Benjamini-Hochberg (FDR)", "Bonferroni"),
       significant = c(sum(p < 0.05),
                       sum(p.adjust(p, method = "fdr") < 0.05),
                       sum(p.adjust(p, method = "bonferroni") < 0.05))) |>
  kable()
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses one new function:

```r
p.adjust(p, method = p.adjust.methods)                    # stats
```

- `p.adjust()` adjusts a set of p-values for multiple comparisons, so that each adjusted value can
  be compared directly with 0.05.
- `method = "bonferroni"` multiplies each p-value by the number of tests, 88, with a ceiling of 1.
- `method = "fdr"` applies the Benjamini-Hochberg procedure.
- `sum(... < 0.05)` counts the counties whose adjusted p-value is below 0.05.
:::

Under either correction, none of the 88 counties remains significant, and no county on the LISA map
meets a corrected threshold at this number of simulations.

This follows from the smallest p-value that 999 permutations can produce (section 5.2).

```{r}
c(floor_at_nsim_999 = min(lisa$p_ii_sim), bonferroni_threshold = 0.05 / 88)
```

At `nsim = 999` the smallest attainable `p_ii_sim` is 0.002, and the Bonferroni correction for 88
tests requires p < 0.00057. No rank-based p-value at this number of simulations can meet that bound,
for any data.

Benjamini-Hochberg is not limited in the same way. Its threshold admits a p-value of 0.002 from rank
i = 4 upwards. No county is rejected here: too few counties attain the minimum p-value to meet the
thresholds at those ranks.

```{r}
tibble(rank = 1:6,
       bh_threshold = (1:6) * 0.05 / 88,
       observed_p = sort(lisa$p_ii_sim)[1:6]) |>
  mutate(rejected = observed_p <= bh_threshold) |>
  kable(digits = 5)
```

Ties at the minimum p-value would increase the number of rejections. If four counties had p = 0.002,
all four would meet the i = 4 threshold of 0.00227.

```{r}
set.seed(1234)
lm_big <- local_moran(hunan$GDPPC, wm_q$nb, wm_q$wt, nsim = 9999)
tibble(correction = c("none", "FDR", "Bonferroni"),
       significant_at_nsim_9999 = c(sum(lm_big$p_ii_sim < 0.05),
                                    sum(p.adjust(lm_big$p_ii_sim, "fdr") < 0.05),
                                    sum(p.adjust(lm_big$p_ii_sim, "bonferroni") < 0.05))) |>
  kable()
```

With `nsim` = 9,999 the minimum attainable p-value falls to 0.0002, below the Bonferroni threshold,
and both corrections can then distinguish between counties. At 9,999 permutations 17 counties are
significant without correction, one more than at 999. Three of them remain significant under FDR and
one under Bonferroni.

```{r}
sig_after_fdr <- lisa |>
  st_drop_geometry() |>
  mutate(p_fdr = p.adjust(lm_big$p_ii_sim, "fdr"),
         p_bonf = p.adjust(lm_big$p_ii_sim, "bonferroni")) |>
  filter(p_fdr < 0.05) |>
  select(County, GDPPC, LISA_cluster, p_fdr, p_bonf)
kable(sig_after_fdr, digits = 5)
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses two new functions:

```r
st_drop_geometry(x)                                       # sf
filter(.data, ...)                                        # dplyr
```

- `st_drop_geometry()` removes the geometry column, which turns the sf object into an ordinary data
  frame and keeps the printed table short.
- `filter()` keeps the rows for which the condition is true, here the counties whose FDR-adjusted
  p-value is below 0.05.
- `select()` (section 2.3) then keeps the five columns printed.
:::

### Size of the Bonferroni correction

These counts give the number of counties that remain significant under each correction. They do not
determine whether a correction should be applied to a LISA map at all. Two positions are possible.
If the map is treated as exploratory, each county's p-value describes only that county and no
correction is required. If the map is treated as 88 simultaneous tests, a correction is required and
the remaining question is its size. Neither chapter discusses whether a correction should be
applied. Anselin (1995) and Ord and Getis (1995) both take the second position.

Ord and Getis (1995) recommend the Bonferroni correction:

> we believe that the limits given by the Bonferroni inequality, possibly with the Tukey
> approximation, represent a reasonable basis for making inferences until more precise methods are
> developed.

They also record a referee's objection in the same paragraph: "One of the referees has noted that
the Bonferroni limits may be unduly conservative when n becomes large. This issue needs to be
explored further."

Anselin (1995) agrees that a correction is needed and questions its size:

> Due to this correlation, and the associated problem of multiple comparisons, the usual
> interpretation of significance will be flawed.

> while it is obvious that some correction to the individual significance levels is needed, the
> extent to which it is indeed necessary to take m = n remains to be further investigated.

His reason is that the dependence between local statistics is local and bounded: "the number of
common neighbors does not change with the number of observations". He notes that on a regular
lattice under the Queen rule, first-order neighbours share four elements of their neighbourhood
sets, second-order neighbours three, and higher-order neighbours none. Setting m, the number of
tests, equal to n, the number of observations, therefore corrects for more tests than are correlated
with one another.

He nevertheless applies m = n in his own worked example on 42 African countries: "the individual
significance levels for each observation should be taken as 0.05/42, or 0.0012."

Getis (2010), a chapter on the Lesson 5 reference list, returns to multiple testing in local
statistics fifteen years later. He describes the selection of tests that can withstand "the
simultaneity effects of multiple tests" as "a particularly difficult area of research", since the
observations used for one local test are used again for the tests at neighbouring locations, and
cites Benjamini and Hochberg (1995) and Castro and Singer (2006) among the attempts to resolve it.
The Benjamini-Hochberg counts above are reported alongside Bonferroni because Getis (2010) cites
that procedure as one of these attempts.

### Permutation count and the p-value floor

Anselin's worked example is also relevant to the choice of `nsim`. A corrected bound cannot be met
by a p-value whose minimum attainable value exceeds it, so the number of permutations should be
chosen after the correction. Anselin used **9,999** conditional permutations and gives no reason for
choosing 9,999 rather than some smaller number. His bound is 0.05/42 = 0.0012, and a minimum p-value
of 1/(nsim+1) falls below that bound from about nsim = 840 upwards, so the arithmetic does not
require a count that high. At 9,999 permutations the minimum is well below his bound:

> only Sudan meeting the Bonferroni bound for an overall a = 0.05. For this country, two out of
> the 9,999 statistics computed from the randomized samples exceed the observed one, clearly
> labeling the latter as "extreme."

Chapter 10 sets `nsim = 99` before any correction is considered.

```{r}
tibble(setting = c("chapter 10, nsim = 99", "this page, nsim = 999",
                   "section 5.3, nsim = 9999", "Anselin (1995), nsim = 9999, n = 42"),
       floor = c(2/100, 2/1000, 2/10000, 1/10000),
       bonferroni_bound = c(0.05/88, 0.05/88, 0.05/88, 0.05/42)) |>
  mutate(correction_reachable = floor < bonferroni_bound) |>
  kable(digits = 5)
```

At chapter 10's setting the minimum attainable p-value (0.02) is about thirty-five times the
Bonferroni bound for 88 counties (0.00057), so no county can be significant after the correction,
whatever the data.

### Counties significant after correction

Earlier in this section, no county remained significant under either correction at `nsim = 999`, and
three remained significant under FDR at 9,999. Two decisions follow for the LISA map of section
4.1.5 and the hot and cold spot map of section 4.2.3: whether to redraw them at a corrected
threshold, and which claims the text around them can support.

The map in section 4.1.5 is left uncorrected and treated as exploratory, so that each county's
p-value describes only that county, the first of the two positions in section 5.3.1. The counts
above report how many of its counties a correction would retain.

The two maps below use the same 9,999 permutations, so they differ only in the correction.

```{r}
#| code-fold: true
#| fig-height: 4.1
lvls <- c("Insignificant", "Low-Low", "Low-High", "High-Low", "High-High")
lisa_corr <- lisa |>
  mutate(uncorrected = factor(ifelse(lm_big$p_ii_sim < 0.05,
                                     as.character(mean), "Insignificant"), levels = lvls),
         fdr = factor(ifelse(p.adjust(lm_big$p_ii_sim, "fdr") < 0.05,
                             as.character(mean), "Insignificant"), levels = lvls))

cluster_map <- function(col, title) {
  tm_shape(lisa_corr) +
    tm_polygons(fill = col,
                fill.scale = tm_scale_categorical(
                  values = c("grey80", "blue", "lightblue", "pink", "red")),
                fill.legend = tm_legend(title = "LISA cluster",
                                        position = tm_pos_out("right", "center"))) +
    tm_title(title)
}

tmap_arrange(cluster_map("uncorrected", "No correction, 17 counties"),
             cluster_map("fdr", "Benjamini-Hochberg, 3 counties"),
             asp = 1, ncol = 2)
```

Three counties remain after correction: Longhui and Wugang, Low-Low counties in the west, and
Xiangtan, in the High-High cluster around Changsha. Wugang is the only county that also meets the
Bonferroni bound. The corrected map is therefore weighted towards the Low-Low cluster in the west,
which reverses the emphasis of the uncorrected map. The 16 counties of section 4.1.5 are candidate
clusters. The Gi\* map of section 4.2.3 was not tested against a corrected threshold.

The corrected result also differs from global G in section 5.4, which indicates that high values
dominate the province-wide pattern. The two results are consistent because the statistics measure
different things. Global G weights counties by the magnitude of their values. A corrected LISA
identifies the individual counties whose statistics meet a threshold corrected for 88 simultaneous
tests, and two of the three such counties have low GDPPC.

Anselin (1995) raises a further point that no correction addresses. Local statistics are themselves
spatially autocorrelated, since neighbouring counties share the observations from which their
statistics are computed:

> This spatial clustering (or spatial autocorrelation) of both the Gi\* and the LISA statistics is a
> result of the way they are constructed, and should be kept in mind when visually interpreting a
> map of LISAs.

When significant counties are adjacent, the map cannot show whether they reflect a pattern shared
across the area or one very high value that falls in all of their neighbourhoods.

## Getis-Ord global G

Sections 3.3 and 3.4 found positive spatial autocorrelation in GDPPC under both global measures
(Moran's I = 0.301, Geary's C = 0.691). Neither statistic distinguishes clustering of high values
from clustering of low values.

The Getis-Ord global G, covered in slides 16 and 17 but in neither chapter, measures the overall
concentration of high or low values. Getis and Ord (1992) state the limitation of Moran's I that G
addresses:

> The I(d) statistic has its peculiar weakness in not being able to discriminate between patterns
> that have high values dominant within d or low values dominant.

Getis and Ord (1992) conclude that "any test for spatial association should use both types of
statistics."

Does the positive spatial autocorrelation in GDPPC reflect clustering of high values or of low
values?

```{r}
nb_band <- dnearneigh(st_centroid(st_geometry(hunan)), 0, ct)
lw_band <- nb2listw(nb_band, style = "B", zero.policy = TRUE)
globalG.test(hunan$GDPPC, lw_band, zero.policy = TRUE)
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses three new functions:

```r
st_centroid(x)                                            # sf
dnearneigh(x, d1, d2)                                     # spdep
globalG.test(x, listw, zero.policy = attr(listw, "zero.policy"),
             alternative = "greater")                     # spdep
```

- `st_centroid()` replaces each county by its centroid, and `dnearneigh()` makes every pair of
  centroids between `d1 = 0` and `d2 = ct` metres apart into neighbours.
- `nb2listw(style = "B")` gives each neighbour a weight of 1 without row standardisation, as the
  definition of G requires.
- `globalG.test()` computes G and tests it against its expectation under spatial randomness.
  `alternative = "greater"`, the default, tests whether high values are concentrated.
:::

Getis and Ord (1992) define G on a binary distance-based weights matrix, so `style = "B"` follows
from the definition of the statistic. G is the ratio of the weighted sum of the products of value
pairs to the total sum of the products of value pairs. Only the numerator contains the weights, so
row standardisation would change the meaning of the ratio.

For GDPPC, G is 0.0510, against an expectation of 0.0436 (z = 3.03, p = 0.0012). A significant
positive z-score indicates that high values within d of other high values dominate the pattern.

This appears to conflict with the LISA map of section 4.1.5, which has nine Low-Low counties and six
High-High counties. The two statistics measure different things:

```{r}
mu <- mean(hunan$GDPPC)
lisa |>
  st_drop_geometry() |>
  filter(LISA_cluster %in% c("High-High", "Low-Low")) |>
  group_by(LISA_cluster) |>
  summarise(counties = n(),
            mean_deviation_from_provincial_mean = round(mean(GDPPC - mu))) |>
  kable()
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses three new functions:

```r
group_by(.data, ...)                                      # dplyr
summarise(.data, ...)                                     # dplyr
n()                                                       # dplyr
```

- `filter()` (section 5.3) keeps the High-High and Low-Low counties.
- `group_by()` splits the rows by cluster, and `summarise()` then computes one row per group.
- `n()` counts the counties in each group, and `mean(GDPPC - mu)` averages their deviation from the
  provincial mean, `mu`.
:::

Global G weights each county by the magnitude of its value. On average the six High-High counties
are about 28,900 yuan above the provincial mean, and the nine Low-Low counties about 11,600 yuan
below it. GDPPC is bounded below by zero and unbounded above, so high values deviate further from
the mean than low values, and they dominate a sum-of-products statistic although they are fewer.

The three statistics therefore describe different aspects of the same pattern. Moran's I indicates
positive spatial autocorrelation, global G indicates that high values dominate it, and the LISA map
shows where the clusters are: high values around Changsha and low values in the west.

### Natural origin and location invariance

Section 4.1.4 standardised GDPPC before plotting it against its spatial lag, because the scatterplot
compares each county with the provincial mean. Chapter 10 standardises there but computes Gi\* on
the raw values, and does not say why the two differ. The reason is a condition that
Getis and Ord (1992) place on G, and that neither chapter states.

A variable has a natural origin when zero represents the absence of the quantity measured, as
opposed to an arbitrary reference point. G is computed from the values themselves, not from their
deviations from the mean, so the position of zero on the measurement scale affects the statistic.
GDPPC has a natural origin, since zero yuan per person corresponds to zero output. Regression
residuals do not, since their zero is defined by the fitted line.

Getis and Ord (1992) state two conditions, the first for Gi and the second for G(d):

> Gi is scale-invariant (Yi = bXi yields the same scores as Xi) but not location-invariant
> (Yi = a + Xi gives different results than Xi). The statistic is intended for use only for those
> variables that possess a natural origin.

> it is inappropriate to use G(d) to study residuals from regression.

How far do G and Moran's I change when a constant is added to GDPPC?

```{r}
shift_test <- function(a) {
  x <- hunan$GDPPC + a
  g <- globalG.test(x, lw_band, zero.policy = TRUE)
  m <- moran.test(x, rswm_q, zero.policy = TRUE)
  tibble(shift = a, G = g$estimate[1], z_G = g$statistic, p_G = g$p.value,
         I = m$estimate[1], z_I = m$statistic, p_I = m$p.value)
}
bind_rows(shift_test(0), shift_test(10000), shift_test(100000)) |>
  mutate(across(c(p_G, p_I), ~ format(.x, digits = 3, scientific = TRUE))) |>
  kable(digits = 4)
```

::: {.callout-note collapse="false" title="How the code works"}
The block defines one function and uses two new ones:

```r
shift_test <- function(a) ...                             # defined here
bind_rows(...)                                            # dplyr
across(.cols, .fns)                                       # dplyr
```

- `shift_test()` adds a constant `a` to GDPPC, runs global G and Moran's I on the shifted values,
  and returns one row of results. Defining it once means the three shifts are computed in exactly
  the same way.
- `bind_rows()` stacks the three one-row results into one table.
- `across(c(p_G, p_I), ~ format(...))` applies the same formatting to both p-value columns, which
  prints the very small values in scientific notation.
:::

Moran's I is unchanged: it is computed on deviations from the mean, and adding a constant leaves
every deviation unchanged. Global G is no longer significant. Its z-score decreases from 3.03 to
0.90 and its p-value increases from 0.0012 to 0.19. Multiplying GDPPC by a constant, in place of
adding one, leaves both statistics unchanged. This is the scale invariance that Getis and Ord
describe.

```{r}
g2 <- globalG.test(hunan$GDPPC * 2, lw_band, zero.policy = TRUE)
m2 <- moran.test(hunan$GDPPC * 2, rswm_q, zero.policy = TRUE)
c(G_doubled = g2$estimate[1], z_G_doubled = g2$statistic,
  I_doubled = m2$estimate[1], z_I_doubled = m2$statistic)
```

The natural-origin requirement makes a G result a statement about the variable on the scale on which
it was measured. For a variable centred at zero, or for regression residuals,
a significant G would be an artefact of the position of zero on the scale.

## Null models for Moran's I

Section 3.3.1 ran `moran.test()` on GDPPC and reported a z-score of 4.74. Moran (1950), who
introduced the statistic, states that its variance, and so this z-score, requires normally
distributed values, and GDPPC is right skewed. By default, however, `moran.test()` does not use
Moran's variance. Its `randomisation` argument, which defaults to `TRUE` and is not mentioned in
either chapter, selects a different null model. What does each null model assume?

Moran (1950) derives the expectation of I without any distributional assumption, and states that the
second moment (which gives the variance, how widely I spreads around its expectation) requires an
assumption about the distribution:

> To evaluate the second moment of I we have to assume further that all the xij are distributed
> normally.

`randomisation = FALSE` is Moran's own variance, which assumes the values are drawn from a normal
distribution. Geary (1954) sets out both null models:

> The theory is discussed from the viewpoints of both randomization and classical normal theory.
> With the randomization approach, the observations themselves are the "universe" and no
> assumption need be made as to the character of the frequency distribution.

Having found the distribution of c close to normal for Ireland's 26 counties, he concludes that
"accordingly, one can have confidence in deciding on significance from the standard error."

- **Normality** assumes the 88 values are drawn from a normal distribution, which does not hold for
  a right-skewed variable such as GDPPC.
- **Randomisation**, the default used in section 3.3.1, treats the 88 values as fixed and considers
  every assignment of them to counties. Its z-score still relies on a normal approximation.
- **Permutation**, used in section 3.3.2, samples 999 of those assignments and needs no
  approximation.

Sections 3.3.1 and 3.3.2 reach the same conclusion under the randomisation and permutation null
models: both reject the null hypothesis of no spatial autocorrelation. I is a sum over all 88
counties, so a normal approximation suits it better than a local statistic computed from five
neighbours, which is the case examined in section 5.2.

## Local Geary's c

Section 3.4 computed Geary's C for the whole province. Section 3.5 compared its correlogram with
that of Moran's I, and the two differed in significance at lags 3 and 6.

Neither chapter computes the local form of Geary's C. Slide 26 identifies it as a second LISA,
alongside local Moran's I. Each of the two local statistics misses a pattern that the other
detects, and this section compares them on GDPPC.

Local Moran's I and local Geary's c differ in construction. Up to scaling constants, for county $i$
with neighbours $j$ and weights $w_{ij}$,

$$
I_i = (x_i - \bar{x}) \sum_j w_{ij}(x_j - \bar{x}), \qquad c_i = \sum_j w_{ij}(x_i - x_j)^2 .
$$

- **Local Moran's I compares the county with the provincial mean.** It multiplies the county's
  deviation from the mean by the weighted deviations of its neighbours, so for a county near the
  mean $I_i$ is small, even when the county matches its neighbours closely.
- **Local Geary's c compares the county with its neighbours.** As slide 26 defines it, $c_i$ adds up
  the squared differences between the county and each neighbour, and the provincial mean cancels in
  each difference. $c_i$ is large for a county whose value differs sharply from its neighbours'
  values. It cannot distinguish High-High from Low-Low, since a cluster of high values and a
  cluster of low values both give small differences.

Section 4.1.5 found 16 significant counties with local Moran's I. How many does local Geary's c
find, and which counties do the two statistics classify differently?

```{r}
set.seed(1234)
lgeary <- wm_q |>
  mutate(local_c = local_c_perm(GDPPC, nb, wt, nsim = 999), .before = 1) |>
  unnest(local_c)

tibble(statistic = c("local Moran's I", "local Geary's c"),
       significant_at_0.05 = c(sum(lisa$p_ii_sim < 0.05),
                               sum(lgeary$p_ci_sim < 0.05))) |>
  kable()
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses one new function:

```r
local_c_perm(x, nb, wt, nsim = 499, alternative = "two.sided")   # sfdep
```

- `local_c_perm()` computes local Geary's c for each county and tests it by conditional
  permutation. It follows the pattern of `local_moran()` in section 4.1.2: it returns a data frame
  of results, stored as the column `local_c` and spread into columns by `unnest()`.
- `p_ci_sim` is the rank-based p-value from the 999 permutations. `nsim` defaults to 499.
:::

Local Geary's c identifies 21 significant counties, against 16 for local Moran's I. Eleven counties
are significant under one statistic and not the other.

```{r}
sig_moran <- lisa$p_ii_sim < signif
sig_geary <- lgeary$p_ci_sim < signif
z <- (hunan$GDPPC - mean(hunan$GDPPC)) / sd(hunan$GDPPC)

tibble(County = hunan$County, GDPPC = hunan$GDPPC, z_GDPPC = round(z, 2),
       `local Moran` = sig_moran, `local Geary` = sig_geary) |>
  filter(sig_moran != sig_geary) |>
  arrange(z_GDPPC) |>
  kable()
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses two new functions:

```r
sd(x)                                                     # stats
arrange(.data, ...)                                       # dplyr
```

- `sd()` returns the standard deviation, so `z` is each county's GDPPC in standard deviations from
  the provincial mean.
- `filter(sig_moran != sig_geary)` keeps the counties that are significant under one statistic and
  not the other.
- `arrange()` sorts the rows, here from the lowest `z_GDPPC` to the highest.
:::

Of the eleven:

- **Eight are significant only under local Geary's c.** Six of them have GDPPC within half a
  standard deviation of the provincial mean, where local Moran's I is small whatever the neighbours'
  values are. The exceptions are Jianghua in the south and Shaoshan, whose GDPPC is about two
  standard deviations above the mean.
- **Three are significant only under local Moran's I.** Liuyang and Miluo lie in or beside the
  Changsha hot spot, and Yuanling sits close to the provincial mean.

## Gi\* weights: fixed band versus k nearest neighbours

Section 4.2.1 created `hunan_adw` as chapter 10 does, but chapter 10 never uses it. Section 4.2.2
computed Gi\* on the fixed distance band only.

Fixed and adaptive schemes hold different quantities constant:

- **A fixed distance band fixes the radius.** One radius applies to the whole province, so a county
  surrounded by large counties has few neighbours, and a county in an area of small counties has
  many.
- **An adaptive scheme fixes the number of neighbours.** Every county has the same number, and the
  radius varies.

The maps below show this for two counties: Yuanling, a large county in the west, and Shaodong, a
smaller one in the centre.

```{r}
#| code-fold: true
#| fig-height: 4.1
show_nb <- function(nb, title) {
  yl <- which(hunan$County == "Yuanling")
  sd <- which(hunan$County == "Shaodong")
  hunan |>
    mutate(group = case_when(row_number() %in% nb[[yl]] ~ "Yuanling and its neighbours",
                             row_number() %in% nb[[sd]] ~ "Shaodong and its neighbours",
                             TRUE ~ "Other"),
           group = factor(group, levels = c("Yuanling and its neighbours",
                                            "Shaodong and its neighbours", "Other"))) |>
    tm_shape() +
    tm_polygons(fill = "group",
                fill.scale = tm_scale_categorical(values = c("purple", "orange", "grey90")),
                fill.legend = tm_legend(title = "", position = tm_pos_out("right", "center"))) +
    tm_title(title)
}

tmap_arrange(show_nb(hunan_fdw$nb, "Fixed band, 62,955 m"),
             show_nb(hunan_adw$nb, "Adaptive, k = 6 plus self"),
             asp = 1, ncol = 2)
```

For Gi\*, reported as a z-score, the number of neighbours matters. According to slide 30, a sparse
neighbourhood produces a large variance of the estimate, which reduces the z-score of Gi\* towards
zero, so the county is less likely to reach significance. Slide 35 gives a rule of thumb of about
eight neighbours per feature for a skewed variable.

How many counties do the two schemes classify differently, and are the differences mainly in one
direction?

```{r}
set.seed(1234)
HCSA_adw <- hunan_adw |>
  mutate(gistar = local_gstar_perm(GDPPC, nb, wt, nsim = 999), .before = 1) |>
  unnest(gistar) |>
  mutate(HCSA_cluster = case_when(p_sim >= signif ~ "Insignificant",
                                  gi_star > 0 ~ "Hot spot",
                                  TRUE ~ "Cold spot"),
         HCSA_cluster = factor(HCSA_cluster,
                               levels = c("Insignificant", "Hot spot", "Cold spot")))

table(fixed_band = HCSA_fdw$HCSA_cluster,
      adaptive_k = HCSA_adw$HCSA_cluster)
```

The two schemes classify 15 of the 88 counties differently, and most of the differences are in one
direction:

- **Eleven** counties are significant under the adaptive weights and not under the fixed band.
- **Four** are significant under the fixed band and not under the adaptive weights.

The map below is the adaptive counterpart of the map in section 4.2.3.

```{r}
#| fig-height: 10.13
tm_shape(HCSA_adw) +
  tm_polygons(fill = "HCSA_cluster",
              fill.scale = tm_scale_categorical(
                values = c("grey80", "red", "blue")),
              fill.legend = tm_legend(title = "Gi* cluster",
                                      position = tm_pos_out("right", "center"))) +
  tm_borders() +
  tm_title("Hot and cold spots of GDPPC, adaptive k = 6 plus self, p < 0.05")
```

```{r}
tibble(scheme = c("fixed band", "adaptive k"),
       min_neighbours = c(min(lengths(hunan_fdw$nb)), min(lengths(hunan_adw$nb))),
       median_neighbours = c(median(lengths(hunan_fdw$nb)), median(lengths(hunan_adw$nb))),
       max_neighbours = c(max(lengths(hunan_fdw$nb)), max(lengths(hunan_adw$nb)))) |>
  kable()
```

The neighbour counts explain the direction of the differences:

- **Fixed band: as few as two neighbours.** Some counties have as few as two neighbours, including
  themselves, which is the sparse case described on slide 30. The median county has five.
- **Adaptive: seven for every county.** No county has a small neighbourhood because of its location
  among large counties. Seven is also close to the eight suggested on slide 35.

The adaptive scheme looks preferable, but a larger count of significant counties does not show that
a scheme is more accurate. A larger neighbourhood alone produces more significant counties, with no
change in the data: more neighbours lower the variance of Gi\* (slide 30), which moves its z-score
away from zero. The eight neighbours of slide 35 is a guideline. Section 5.7.1 therefore varies `k`
and checks which classifications change.

### Sensitivity to the number of neighbours

Section 4.2.1 set `k = 6` as chapter 10 does, and the comparison above uses that value alone.

Mack and Kam (2018) take a different approach to the same decision in their study of conflict in
Africa. Before choosing a weights matrix, they tabulate two quantities under several definitions:

- the number of regions left without neighbours;
- the average number of links.

They reject the widest definition because an average of 52 links "may skew the analysis".

The same check can be applied to `k`. A county whose Gi\* classification is the same under several
neighbourhood sizes depends less on any one of them.

How many counties are significant at `k` = 4, 6 and 8, and which keep the same classification under
all three?

```{r}
nb_k4 <- include_self(st_knn(st_geometry(hunan), k = 4))
set.seed(1234)
gi_k4 <- local_gstar_perm(hunan$GDPPC, nb_k4, st_weights(nb_k4), nsim = 999)

nb_k8 <- include_self(st_knn(st_geometry(hunan), k = 8))
set.seed(1234)
gi_k8 <- local_gstar_perm(hunan$GDPPC, nb_k8, st_weights(nb_k8), nsim = 999)

k_compare <- tibble(
  County = hunan$County,
  k4 = case_when(gi_k4$p_sim >= signif ~ "Insignificant",
                 gi_k4$gi_star > 0 ~ "Hot spot", TRUE ~ "Cold spot"),
  k6 = as.character(HCSA_adw$HCSA_cluster),
  k8 = case_when(gi_k8$p_sim >= signif ~ "Insignificant",
                 gi_k8$gi_star > 0 ~ "Hot spot", TRUE ~ "Cold spot"))

tibble(k = c(4, 6, 8),
       significant_counties = c(sum(k_compare$k4 != "Insignificant"),
                                sum(k_compare$k6 != "Insignificant"),
                                sum(k_compare$k8 != "Insignificant"))) |>
  kable()
```

```{r}
k_compare |>
  filter(k4 == k6, k6 == k8, k4 != "Insignificant") |>
  select(County, class = k6) |>
  arrange(class, County) |>
  kable()
```

The number of significant counties rises with `k`, from 17 at `k = 4` to 25 at `k = 8`, as expected
from slide 30: a larger neighbourhood reduces the variance of Gi\* and moves its z-score away from
zero. Twelve counties keep the same classification under all three values:

- **eight hot spots** around Changsha;
- **four cold spots** in the south west: Jingzhou, Suining, Wugang and Xinning. Wugang is also the
  county that meets the Bonferroni bound in section 5.3.3.

The classification of the other significant counties depends on `k`.

## Gi and local Moran's I under permutation

Section 4.1 produced a LISA map from the local Moran and section 4.2 a hot and cold spot map from
Gi\*. Chapter 10 presents the two as separate analyses.

Anselin (1995) argues that the two share one permutation test:

1. Under conditional permutation, only the numerator of each statistic varies.
2. The varying term is the same in both statistics.
3. Therefore:

   > the randomization method applied to [the local Moran] will yield the same empirical reference
   > distribution as when applied to the Getis and Ord Gi and Gi\* statistics. Hence, inference
   > based on this nonparametric approach will be identical for the two statistics.

On the Queen contiguity weights of section 4.1.1, do Gi and Gi\* give the same permutation p-values
as the local Moran? A second run of Gi\* with a different random seed shows how much two runs differ
by chance alone.

```{r}
set.seed(7); li  <- local_moran(hunan$GDPPC, wm_q$nb, wm_q$wt, nsim = 9999)
set.seed(7); gi  <- local_g_perm(hunan$GDPPC, wm_q$nb, wm_q$wt, nsim = 9999)
set.seed(7); gs  <- local_gstar_perm(hunan$GDPPC, wm_q$nb, wm_q$wt, nsim = 9999)
set.seed(8); gs2 <- local_gstar_perm(hunan$GDPPC, wm_q$nb, wm_q$wt, nsim = 9999)

tibble(comparison = c("Gi against local Moran",
                      "Gi* against local Moran",
                      "Gi* against Gi*, different random seed"),
       largest_difference = c(max(abs(gi$p_sim - li$p_ii_sim)),
                              max(abs(gs$p_sim - li$p_ii_sim)),
                              max(abs(gs$p_sim - gs2$p_sim))),
       counties_significant_in_one_only = c(
         sum((gi$p_sim < signif) != (li$p_ii_sim < signif)),
         sum((gs$p_sim < signif) != (li$p_ii_sim < signif)),
         sum((gs$p_sim < signif) != (gs2$p_sim < signif)))) |>
  kable(digits = 4)
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses one new function:

```r
local_g_perm(x, nb, wt, nsim = 499, alternative = "two.sided")   # sfdep
```

- `local_g_perm()` computes Gi, which excludes the county itself, where `local_gstar_perm()`
  (section 4.2.2) computes Gi\*, which includes it. All three statistics use the Queen contiguity
  weights of section 4.1.1 and 9,999 permutations.
- `set.seed()` (section 3.3.2) before each call starts every statistic from the same random numbers,
  except the second Gi\* run, which uses a different seed to measure simulation error.
- `abs()` takes absolute differences, and `max()` keeps the largest across the 88 counties.
:::

- **Gi agrees with the local Moran to within 0.0002**, one step of the p-value grid at 9,999
  permutations.
- **Gi\* differs from the local Moran by up to 0.033**, but two runs of Gi\* differ from each other
  by a similar amount, 0.029.

The difference between Gi\* and the local Moran is therefore simulation error, and Anselin's (1995)
claim holds for both Gi and Gi\* on these data.

Gi\* and the local Moran also reach the same decision at p < 0.05 for every county except Yongshun,
whose p-value is 0.051 under the local Moran and 0.047 under Gi\*. A difference of that size lies
within the simulation error shown in the table.

Where Gi\* and the local Moran give the same p-value, they differ in what a significant result says
about the county:

- **The local Moran identifies similarity and outliers.** A significant positive local Moran
  indicates that the county and its neighbours have similar values, without indicating whether
  those values are high or low. The local Moran also identifies the High-Low and Low-High outliers,
  which Gi\* does not.
- **Gi\* indicates whether the values are high or low.** A positive value marks a cluster of high
  values and a negative value a cluster of low values (Anselin 1995).

This is the sense of the recommendation by Getis and Ord (1992) in section 5.4 that "any
test for spatial association should use both types of statistics".

On the same weights, the permutation tests behind chapter 10's LISA map and hot and cold spot map
would give the same p-values. The two maps differ for two reasons: they use different weights, Queen
contiguity against a fixed distance band, and they classify the significant counties differently.

# Reflection

Chapter 10 frames the analysis as three questions (section 10.2.1). The answers on this page are:

- **Is development evenly distributed across the province?** No. GDPPC ranges from 8,497 yuan in
  Guidong to 88,656 yuan in Changsha.
- **Is there sign of spatial clustering?** Yes. Moran's I is 0.301 and Geary's C is 0.691, both
  significant under the analytical and the permutation tests (sections 3.3 and 3.4).
- **Where are the clusters?** A cluster of high GDPPC around Changsha in the north east and a
  cluster of low GDPPC in the west (sections 4.1.5 and 4.2.3). After correction for 88 tests, three
  counties remain significant, two of them in the western cluster (section 5.3.3).

Chapters 9 and 10 address the question left open by Exercise 4. In the lag scatterplot, the positive
slope reflects significant positive spatial autocorrelation in GDPPC (Moran's I = 0.301, permutation
p = 0.001), and Geary's C, on its inverted scale, gives a consistent result. Global G indicates that
the association is dominated by high values.

The local statistics locate the clusters. Sixteen counties are significant: the High-High group
comprises Changsha and its surroundings, and the Low-Low group is in the west. Gi\* on the same data
identifies a hot spot in the same area. Under every weighting scheme tested here, the division
between high and low GDPPC lies between the north east and the west.

These methods are less independent of one another than their number suggests. Global Moran's I,
global G and the LISA map are all computed from the same 88 values, and in section 5.8 I found that
Gi, Gi\* and the local Moran are the same permutation test when the weights are the same. The
comparisons of weighting schemes in sections 5.7 and 5.7.1 are the only places where the location
of the division between high and low GDPPC was tested against an alternative.

The more specific results depended on choices of method. With the data held the same:

- changing only the p-value column changed the count of significant counties from 13 to 16;
- changing from local Moran to local Geary changed it from 16 to 21;
- 15 counties were classified differently under a fixed and an adaptive Gi\* neighbourhood.

The multiple-comparisons result in section 5.3, that no county remains significant after correction
at 999 permutations, changed how I would report a LISA map, and reading the papers changed my view
of it twice.

In my first draft of that section I attributed to Anselin the argument that a correction does not
apply directly to a LISA map, because the tests are not independent. His paper does not make that
argument. He writes that "it is obvious that some correction to the individual significance levels
is needed" and disputes only whether m should equal n, then applies m = n in his own worked example.
Ord and Getis recommend the Bonferroni bound and record a referee's objection that it may be unduly
conservative when n is large. Both assigned papers treat the correction as necessary.

My second change of view came from the number of permutations in Anselin's worked example, 9,999. He
gives no reason for the number, so the inference that follows is mine. A corrected bound cannot be
met by a p-value whose minimum exceeds it, and the minimum depends on `nsim`. I had treated the
number of simulations as a setting for precision, and I now choose it after deciding on the
correction.

Two of the readings contain claims that I could test directly. Getis and Ord state that G is not
location-invariant. Adding 100,000 to GDPPC changed the p-value of G from 0.0012 to 0.19 and left
Moran's I unchanged. A G result is thus a statement about the variable on the scale on which it was
measured. Anselin states that permutation inference is identical for the local Moran, Gi and Gi\*.
My first test appeared to show the claim failing for Gi\*. The gap turned out to be the same size as
the gap between two runs of Gi\* with different random seeds, so it was simulation error, and the
claim holds for both.

Ord and Getis's simulation results determined my choice of `p_ii_sim` over `p_ii` in section 5.2.
Both of their patterns (ii) and (iii), a skewed variable and few neighbours, apply to these data.
For the counties of most interest on the map, the normal-approximation p-value is least reliable.
Section 5.7 is also affected by these patterns: the adaptive scheme gives every county seven
neighbours and removes the few-neighbour case of pattern (iii).

My main practical difficulty in this exercise was the same as in Exercise 4. Three calls in chapter
10 ran without error and without the effect that the surrounding text described:

- `nsim = 99` ran 8,712 permutations whose rank-based p-value was then not used;
- `hunan_adw` was created and not used;
- `signif` was assigned and not used.

I found them by comparing what each call returned with what the surrounding text said it did.

I first left the choice of `k` as a rule of thumb and then tested it, following the tabulation in
Mack and Kam (2018). The count of significant counties depends on `k`, from 17 at `k = 4` to 25 at
`k = 8`, while twelve counties keep the same classification under every value tested. I would
report those twelve as the hot and cold spots of this map and treat the remainder as dependent on
the neighbourhood size.

A second open question concerns change over time. All results on this page are for a single year,
2012, and the division within the province is treated as fixed. Emerging hot spot analysis, on
slides 37 to 43, would test whether the hot spot around Changsha is intensifying and whether GDPPC
in the west is converging towards that of the north east. It needs a value for every county in every
period, and the Lesson 5 material provides one year.

# References

Course materials:

- Kam, T.S. (2026). *Lesson 5: Global and Local Measures of Spatial Association*. ISSS626
  Geospatial Analytics and Applications, SMU. Slide deck dated 20 September 2026.
- Kam, T.S. (2026). [*R for Geospatial Data Science and Analytics*, Chapter 9: Global Measures of
  Spatial Autocorrelation](https://r4gdsa.netlify.app/chap09.html).
- Kam, T.S. (2026). [*R for Geospatial Data Science and Analytics*, Chapter 10: Local Measures of
  Spatial Autocorrelation](https://r4gdsa.netlify.app/chap10.html).

Assigned self-reading:

- Moran, P.A.P. (1950). "Notes on Continuous Stochastic Phenomena". *Biometrika*, 37(1/2), 17-23.
- Geary, R.C. (1954). "The Contiguity Ratio and Statistical Mapping". *The Incorporated
  Statistician*, 5(3), 115-127 and 129-146.
- Getis, A. and Ord, J.K. (1992). "The Analysis of Spatial Association by Use of Distance
  Statistics". *Geographical Analysis*, 24(3), 189-206.
- Anselin, L. (1995). "Local Indicators of Spatial Association - LISA". *Geographical Analysis*,
  27(2), 93-115.
- Ord, J.K. and Getis, A. (1995). "Local Spatial Autocorrelation Statistics: Distributional Issues
  and an Application". *Geographical Analysis*, 27(4), 286-306.

References listed in the Lesson 5 outline:

- Anselin, L. (1996). "The Moran Scatterplot as an ESDA Tool to Assess Local Instability in Spatial
  Association". In Fischer, M., Scholten, H.J. and Unwin, D. (eds), *Spatial Analytical
  Perspectives on GIS*. Taylor and Francis, 111-125.

- Getis, A. (2010). "B.3 Spatial Autocorrelation". In Fischer, M.M. and Getis, A. (eds), *Handbook
  of Applied Spatial Analysis: Software Tools, Methods and Applications*. Springer, 255-278.

- Mack, Z.W.V. and Kam, T.S. (2018). "Is There Space for Violence?: A Data-driven Approach to the
  Exploration of Spatial-Temporal Dimensions of Conflict". *Proceedings of the 2nd ACM SIGSPATIAL
  Workshop on Geospatial Humanities*, Seattle.

Software:

- Bivand, R. (2022). "R Packages for Analyzing Spatial Data: A Comparative Case Study with
  Areal Data". *Geographical Analysis*, 54(3), 488-518. The `spdep` reference.
- Parry, J. (2023). *sfdep: Spatial Neighbors and Statistics*. R package version 0.2.5.

# About this document

Written and executed in Quarto with R `r getRversion()`, `spdep` `r packageVersion("spdep")`,
`sfdep` `r packageVersion("sfdep")` and `tmap` `r packageVersion("tmap")`. All figures and tables
are generated at render time from the data in `data/`, so nothing on this page is pasted in.

I used AI assistance while working out the approach and the extensions in Part 3. The analysis was
run and checked by me.
