---
title: "In-class Exercise 5"
subtitle: "Emerging Hot Spot Analysis"
author: "An Loc"
date: "26 September 2026"
date-modified: "last-modified"
date-format: "D MMM YYYY"
format:
  html:
    toc: true
    toc-depth: 4
    toc-expand: true
    code-copy: true
    fig-width: 9
    fig-height: 7
execute:
  echo: true
  eval: true
  warning: false
  message: false
  freeze: true
---

# Getting started

```{r}
pacman::p_load(sf, sfdep, spdep, tmap, plotly, tidyverse)
```

# The data

## County boundaries

```{r}
hunan <- st_read(dsn = "data/geospatial", layer = "Hunan", quiet = TRUE) %>%
  st_transform(crs = 32649)

nrow(hunan)
```

## Development indicators for 2012

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")

hunan_GDPPC <- hunan %>%
  left_join(hunan2012, by = "County") %>%
  select(NAME_2, ID_3, NAME_3, ENGTYPE_3,
         Shape_Leng, Shape_Area, County, GDPPC, GIO)

names(hunan_GDPPC)
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses three functions:

``` r
read_csv(...)                   # readr
left_join(..., by = "County")   # dplyr
select(...)                     # dplyr
```

- `read_csv()` reads the 29 development indicators for 2012.
- `left_join()` matches each boundary to its row of indicators on the county name. All 88 names match, so no row is lost.
- `select()` keeps nine of the columns the join produces: the six geography fields, plus `County`, `GDPPC` and `GIO`. The other 26 indicators are dropped because only GDP per capita is analysed here.

`select()` keeps the geometry column automatically, so the result is still an sf object.
:::

# Spatial weights

## Queen contiguity, row-standardised

```{r}
wm_q <- hunan_GDPPC %>%
  mutate(nb = st_contiguity(geometry),
         wt = st_weights(nb, style = "W"),
         .before = 1)
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses three functions:

``` r
st_contiguity(geometry)        # sfdep
st_weights(nb, style = "W")    # sfdep
mutate(..., .before = 1)       # dplyr
```

- `st_contiguity()` builds the neighbour list from the polygons. Queen contiguity is its default, so counties meeting only at a corner count as neighbours. `queen = FALSE` gives Rook contiguity instead, which drops the mean neighbour count from 5.09 to 5.00.
- `st_weights()` turns the neighbour list into weights. `style = "W"` row-standardises them so each county's weights sum to 1 (default setting). The value is case-sensitive: `"w"` raises `Style w invalid`.
- `mutate()` adds `nb` and `wt` as list-columns, and `.before = 1` places them ahead in front of the other attributes.

The result is an sf object rather than a bare weights matrix. `sfdep` functions read `nb` and `wt` directly.
:::

```{r}
wm_q %>%
  st_drop_geometry() %>%
  summarise(counties = n(), mean_neighbours = mean(lengths(nb)))
```

## Reading the neighbour list

::: {.callout-note collapse="false" title="How to read nb and wt"}
`nb` and `wt` are list-columns, so each cell holds a vector rather than a single value. Row 1 is Anxiang:

| Column | Cell 1 | Reads as |
|------------------------|------------------------|------------------------|
| `nb` | `c(2, 3, 4, 57, 85)` | rows 2, 3, 4, 57 and 85 of `wm_q` are Anxiang's neighbours |
| `wt` | `c(0.2, 0.2, 0.2, 0.2, 0.2)` | the weight for each of those five neighbours, in the same order |

- `nb` stores **row positions inside `wm_q`**, not county names and not `ID_3` values. Row 1 is Anxiang, so the value 2 points at row 2.
- `wt` is always the same length as `nb`, one weight per neighbour.
- Each weight is 1 divided by the number of neighbours, because the weights are row-standardised. Anxiang has 5 neighbours so each gets 0.2, Jinshi in row 3 has 4 so each gets 0.25, and the six neighbours in row 9 each get 0.1667. Every county's weights sum to 1.
- RStudio's viewer wraps the values in quotes, as `c("2", "3", "4", ...)`. That is how it displays list-column elements, not the data type.
:::

# Global measures of spatial autocorrelation

## Moran's I

```{r}
moranI <- global_moran(wm_q$GDPPC, wm_q$nb, wm_q$wt)

glimpse(moranI)
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses two functions:

``` r
global_moran(wm_q$GDPPC, wm_q$nb, wm_q$wt)   # sfdep
glimpse(...)                                 # tibble
```

- `global_moran()` takes the variable, then the neighbour list, then the weights. It returns a list of two numbers, and reproduces the value of `spdep::moran()`.
- `glimpse()` prints one line per element.

`I` is **0.301**, against a null expectation of −1/(n−1) = −0.011, so GDP per capita is positively spatially autocorrelated: neighbouring counties tend to hold similar values. The value carries no p-value of its own, because `global_moran()` computes the statistic only. `global_moran_test()` adds an analytic test and `global_moran_perm()` a permutation one.

`K` is the kurtosis of the variable, **7.64**. A normal distribution has a kurtosis of 3, so GDP per capita is strongly heavy-tailed. This matters: the randomisation variance of Moran's I is a function of the kurtosis.
:::

## Moran's I by permutation

```{r}
set.seed(1234)

moranI_perm <- global_moran_perm(wm_q$GDPPC, wm_q$nb, wm_q$wt,
                                 nsim = 99)

moranI_perm
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses two functions:

``` r
set.seed(1234)                                  # base
global_moran_perm(..., nsim = 99)               # sfdep
```

- `set.seed()` fixes the random draw. The permutation test shuffles the values, so without a seed the p-value changes on every run.
- `global_moran_perm()` wraps `spdep::moran.mc()`, which builds a reference distribution by shuffling the values across the counties while the weights stay fixed.

**The number of simulations is always `nsim + 1`.** The reference distribution holds the `nsim` permutations plus the observed statistic, which spdep stores last. So `nsim = 99` produces 100 values, and the output says so:

```         
number of simulations + 1: 100
```

That "+1" sets the smallest p-value the test can report, which is 1/(nsim + 1) = 0.01. A permutation test cannot return anything more extreme than that, however strong the pattern.
:::

⚠️ On this data the two-sided branch returns `p < 2.2e-16`, the stored value being exactly 0. That sits below the 0.01 floor, so it cannot be a real permutation p-value?

# Local measures of spatial autocorrelation

## Local Moran's I

```{r}
set.seed(1234)

lisa <- wm_q %>%
  mutate(local_moran = local_moran(GDPPC, nb, wt, nsim = 99), .before = 1) %>%
  unnest(local_moran)
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses three functions:

``` r
local_moran(GDPPC, nb, wt, nsim = 99)   # sfdep
mutate(..., .before = 1)                # dplyr
unnest(local_moran)                     # tidyr
```

- `local_moran()` computes the statistic for every county in one call and returns a list per county. It takes the same three arguments as the global version, and `nsim` behaves the same way.
- `mutate()` stores that list in a new column, and `.before = 1` puts it first.
- `unnest()` expands the list column, giving each county 11 new columns. The ones that matter here are `ii` (the local statistic), `z_ii`, `p_ii`, `p_ii_sim` and `p_folded_sim`, together with `mean` and `median`, which hold the quadrant labels that the cluster map colours.

The result is an sf object of 88 rows and 24 columns.
:::

## Reading the quadrant columns

::: {.callout-note collapse="false" title="How to read mean, median and pysal"}
These three columns hold quadrant label for every county:

- `mean` compares each county against the **mean** of GDPPC and the mean of its spatial lag. Above both is High-High, below both is Low-Low, and a county above one and below the other is High-Low or Low-High.
- `median` applies the identical rule using the **median** as the cut point.
- `pysal` is PySAL's label for the same classification.

| Cut point | High-High | High-Low | Low-High | Low-Low |
|-----------|-----------|----------|----------|---------|
| `mean`    | 21        | 11       | 13       | 43      |
| `median`  | 31        | 13       | 13       | 31      |
| `pysal`   | 21        | 11       | 13       | 43      |
:::

## LISA cluster map

```{r}
#| fig-width: 9
#| fig-height: 9
lisa_sig <- lisa %>%
  filter(p_ii_sim < 0.05)

tmap_mode("plot")

tm_shape(lisa) +
  tm_polygons() +
  tm_borders(fill_alpha = 0.5) +
  tm_shape(lisa_sig) +
  tm_polygons(fill = "mean",
              fill.legend = tm_legend(text.size = 1.15, width = 6)) +
  tm_borders(fill_alpha = 0.4) +
  tm_layout(legend.title.size = 1.25)
```

::: {.callout-note collapse="false" title="How the code works"}
The block uses four functions:

``` r
filter(p_ii_sim < 0.05)   # dplyr
tmap_mode("plot")         # tmap
tm_shape(...)             # tmap
tm_polygons(...)          # tmap
```

- `filter()` keeps the counties whose permutation p-value is below 0.05. These are the clusters. The other are not distinguishable from spatial randomness.
- `tmap_mode("plot")` switches tmap from interactive to static output.
- The map is built in **two layers**. The first `tm_shape(lisa)` draws all 88 counties as a light backdrop, so the clusters can be read against the whole province. The second `tm_shape(lisa_sig)` draws only the significant counties on top, coloured by quadrant.

Both layers are needed. Without the first, the coloured counties float with nothing to place them against.
:::
