---
title: "In-class Exercise 4"
subtitle: "Spatial Weights and Applications"
author: "An Loc"
date: "19 September 2026"
date-modified: "last-modified"
date-format: "D MMM YYYY"
format:
  html:
    toc: true
    toc-depth: 4
    toc-expand: true
    code-copy: true
    fig-width: 9
    fig-height: 7
execute:
  echo: true
  eval: true
  warning: false
  message: false
  freeze: true
---

# Getting started

```{r}
pacman::p_load(sf, ggstatsplot, spdep, tmap, tidyverse, knitr, GWmodel)
```

# Getting the data into R environment

## Import shapefile into R environment

```{r}
hunan_sf <- st_read(dsn = "data/geospatial", 
                 layer = "Hunan")
```

## Import csv file into R environment

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
class(hunan2012)
match(c("GIO", "Agri", "Service"), names(hunan2012))
```

## Performing relational join

```{r}
# By column name (kept for reference, same result as the version below)
# hunan_sf <- left_join(hunan_sf, hunan2012, by = "County") %>%
#   select(NAME_2, ID_3, NAME_3, ENGTYPE_3, County, GDPPC, GIO, Agri, Service)

# By column position
hunan_sf <- left_join(hunan_sf, hunan2012, by = "County") %>%
  select(1:4, 7, 15:16, 31:32)
```

# Visualising regional development indicator

```{r}
#| fig-width: 16
#| fig-height: 6
#| echo: false
# Left base map

basemap <- tm_shape(hunan_sf) +     # tm_shape is thematic map library
  tm_polygons() +                # Draw polygons, gives the outline
  tm_text("NAME_3", size = 0.5)  # Add labels and use column label NAME_3
                                 #   which is county names, size = 0.5 is small font
                                 #   size to squeeze into small space

# Right map
# Quick thematic mapping function, simplify usage
gdppc <- qtm(hunan_sf, "GDPPC") + tm_layout (legend.position = (c("left", "bottom")))
#, legend.text.size = 0.35,legend.title.size = 0.35)
# Using tm_layout like this to put the legend inside

# To put multiple maps together
tmap_arrange(basemap, gdppc, asp = 1, ncol = 2)
```

# Converting from sf to sp

GWmodel works on sp objects, so convert a copy.

```{r}
hunan_sp <- hunan_sf %>%
  as_Spatial()
```

```{r}
class(hunan_sp)
```

# Determining adaptive bandwidth

## Cross-validation

```{r}
bw_CV <- bw.gwr(GDPPC ~ 1, 
                # ⚬	Usually, regression looks like GDPPC ~ Education + Investment
                # ("predict GDP using education and investment").
                # ⚬	Writing GDPPC ~ 1 strips out all explanatory variables.
                # It tells R: "Don't try to explain GDP using other factors;
                #    just look at GDP itself."
                # What is the optimal neighborhood size (bandwidth) to calculate the
                #   geographically weighted average of GDP per capita across Hunan without
                #   oversmoothing the map?
                data = hunan_sp,
                approach = "CV",
                adaptive = TRUE,
                kernel = "bisquare")
                #longlat = TRUE)

# Adaptive bandwidth: 62 CV score: 15515442343 
# adaptive = TRUE, the bandwidth is measured in a count of spatial units rather
#   than a physical distance (like kilometers)
# CV stands for Cross-Validation
# The algorithm temporarily removes one county at a time,
#   predicts its GDP using the remaining neighbors, and tallies up the difference.
# Rule: You always want the lowest CV score, as a lower score indicates
# higher predictive accuracy.

# e.g. 22 means every computation needs 22 neighbors
# if using FALSE, the great circle distance will be in km, ~72km

# Sometimes it goes down to 52 and the algorihtm came back to higher value and slowly go down again
```

```{r}
bw_CV
```

## AIC

```{r}
bw_AIC <- bw.gwr(GDPPC ~ 1,
                 data = hunan_sp,
                 approach = "AICc",
                 # AIC suitable for big data, less sensitive for small number of observations
                 # Our case area is small, so we use AICc, penalize small number size
                 # Avoid biasness for small locations, ~20 areas
                 adaptive = TRUE,
                 kernel = "bisquare",
                 longlat = TRUE)
```

```{r}
bw_AIC
```

# Computing geographically weighted summary statistics

```{r}
gwstat <- gwss(data = hunan_sp,
               vars = "GDPPC",
               bw = bw_AIC,
               kernel = "bisquare", # need to be consistent
               quantile = TRUE, # calculate both Parametric measures (Mean, Standard Deviation) - assume your data behaves like a normal bell curve, and Non-parametric measures (Median, IQR) - These make no assumptions about distribution shape
               adaptive = TRUE,
               longlat = T)
```

```{r}
names(gwstat$SDF)
```

## Preparing the output data

Extract the summary statistics into a plain data.frame.

```{r}
gwstat_df <- as.data.frame(gwstat$SDF)
```

Append the newly derived data.frame onto `hunan_sf`.

```{r}
hunan_gstat <- cbind(hunan_sf, gwstat_df)
```

`cbind` stands for "column-bind":

- `hunan_sf` is the original map table, containing county names and boundary shapes.

- `gwstat_df` is the new table holding the calculated geographically weighted statistics, such as `GDPPC_LM` and the local standard deviations.

- `cbind(hunan_sf, gwstat_df)` stitches those newly calculated statistical columns directly onto the map polygons.

As we know the sequence is the same, because the sorting was never touched, we simply append directly with `cbind`. If we had done some sorting, the sequence would be incorrect. There is no unique identifier in `gwstat_df` to match on, so row order is the only thing keeping each county with its own statistics.

```{r}
glimpse(hunan_gstat)
```

# Visualising geographically weighted summary statistics

- `fill = "GDPPC_LM"`: Colors each county by its Local Mean GDP per capita.

- `style = "quantile", n = 5`: Divides the 88 counties into 5 equal groups (20% of counties per color bin), ensuring an even spread of color across the map.

- `values = "brewer.blues"`: Shades the bins using a blue color gradient (light blue = poorest neighbor average, dark blue = richest).

- `brew` is professor Dr. Cynthia Brewer, a professor of geography at Pennsylvania State

```{r}
#| fig-width: 9
#| fig-height: 8
tm_shape(hunan_gstat) +
  tm_polygons(fill = "GDPPC_LM",
              fill.scale = tm_scale_intervals(style = "quantile",
                                              n = 5,
                                              values = "brewer.blues"),
              fill.legend = tm_legend(title = "local Mean")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Distribution of geographically weighted mean") +
  tm_layout(legend.position = c("left", "bottom"))
```
